### 2023-07-20 paper/paper.tex
added paper

\documentclass[usenatbib]{mnras}
\usepackage[T1]{fontenc}
\usepackage[pdftex]{graphicx}
\usepackage{bm}
\usepackage{savesym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
\usepackage{relsize}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{afterpage}
\usepackage{txfonts}
\usepackage{pgf}

\newcommand{\logL}{\{ \log \Like_i \}}
\newcommand{\thetab}{\bm{\theta}}
\newcommand{\logLm}{\log \Like_\mathrm{max}}
\newcommand{\set}[1]{\{#1\}}
\newcommand{\nlive}{n_\mathrm{live}}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{D_\mathrm{KL}}

\title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
\author[Z. Hu]{Zixiao Hu, Will Handley}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\begin{abstract}
This project develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration. The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and final prior volume to be calculated. We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods, and by the halfway point for non-Gaussian likelihoods while identifying the correct order of magnitude throughout. 
\end{abstract}

\begin{keywords}
methods: data analysis -- methods: statistical
\end{keywords}

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{split}. The core algorithm is unique for being the only algorithm which estimates high-dimensional volumes using \textit{order statistics}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord} which provide flexible options for nested sampling runs, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to obtain and plot sample statistics. However, currently no such software is able to estimate how long a given run should last. This is of high importance to the end user, who has no indication in the middle of a run whether it will take minutes, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
\par
This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage. The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples. We begin in section \ref{sec:Background} with an overview of Bayesian inference and nested sampling. Section \ref{sec:Endpoint prediction} presents the theory and methodology of making endpoint predictions, before results for both toy and real examples are shown in section \ref{sec:Results}. Finally, conclusions are made in section \ref{sec:Conclusions}.

\section{Background}\label{sec:Background}
\printinunitsof{in}\prntlen{\textwidth}
We present here an overview of Bayesian inference and nested sampling; for a more detailed treatment of both subjects, we refer to \citet{sivia} as well as Skilling's original paper. 
\subsection{Bayesian inference}%
\label{sub:Bayesian inference}

Given some data $D$, Bayes' theorem tells us that the probability of some parameters $\theta$ of a model $M$ is 
\begin{equation}
	\Pr(\theta\mid D, M) = \frac{\Pr(D \mid \theta, M) \Pr(\theta \mid M)}{\Pr(D  \mid  M)}
\end{equation}
Relabelling these terms,
\begin{equation}
    P(\theta) = \frac{L(\theta) \pi(\theta)}{Z}
\end{equation}
where $\pi(\theta)$ is the \textit{prior} - what was known about a model's parameters before knowledge of the data, $P(\theta)$ is the \textit{posterior} - what is known after learning the data, and $L(\theta)$ is the \textit{likelihood} of the data given the parameters.
\par
For parameter estimation, knowledge of $L(\theta)$ and $\pi(\theta)$ is enough, but $Z$, known as the \textit{evidence}, is crucial for model comparison via Bayes factors. By definition, $Z$ is the normalisation factor 
\begin{equation}
    Z = \int L(\theta)\pi(\theta) \ d\theta
\end{equation}
which is a high-dimensional integral over the parameter space. Exponential scaling with dimension means full rasterisation is impractical for most physical applications. This motivates a sampling technique which preferentially selects for regions of high posterior mass. Traditional techniques like Markov-Chain Monte Carlo (MCMC) do not sample from the prior, so it is non-trivial to estimate the evidence from posterior samples. Nested sampling sidesteps these issues by putting the evidence first, while simultaneously yielding the posterior.

\subsection{Nested sampling}
\label{sub:Nested sampling}

We first perform a change of variable to reduce the evidence integral to a single dimension. Define the prior mass 
\begin{equation}\label{X integral}
    X(\lambda) = \int_{L(\theta)>\lambda} \pi(\theta) \ d\theta 
\end{equation}
which allows us to rewrite the evidence as 
\begin{equation}
    Z = \int_0^1 \Like(X) \ dX.
\end{equation}
The integral can be calculated by ordering the prior samples by likelihood then multiplying by the associated prior mass element. Nested sampling translates this into an algorithm which incorporates sorting \textit{de facto}, by taking a set of $\nlive$ \textit{live points} and replacing the point with the lowest likelihood with a resampled point of higher likelihood at every iteration. A choice must be made for this sampling method e.g. MCMC. The number of live points can in principle vary throughout the run, so that it depends on the iteration $k$ i.e. $n_\mathrm{live}=n_k$ \citep{dynamic_ns}. 
\par
The set of live points will therefore occupy a prior volume which tends to zero, shrinking around the maximum of the posterior. Explicit calculation of the volume $X$ within each contour is expensive, so nested sampling makes a statistical estimate using the order statistics of the remaining points: at each iteration, the volume occupied by the live points decreases by the shrinkage ratio $t$, which has probability distribution and statistics
\begin{equation}
	\Pr(t) = \nlive t^{\nlive - 1}, \quad \mathrm{E}[\log t] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t] = \frac{1}{\nlive^2}.
\end{equation}
$X_i$ is the product of the shrinkage ratios
\begin{equation}\label{xi_formula}
    X_i = \prod_{k=0}^{i} t_k,
\end{equation}
so the $X$ themselves are also random variables. It is easier to work with $\log X$, which have the means and correlations
\begin{equation}\label{X_distribution}
	\mathrm{E}[\log X_i] = -\sum_{k=0}^{i} \frac{1}{n_k}, \quad
	\mathrm{Cov}[\log X_i, \log X_j] = \sum_{k=0}^{\mathrm{min}(i, j)} \frac{1}{n_k^2}.
\end{equation}

The evidence integral is then computed using the trapezoid rule at each step, so that
\begin{equation}
	Z = \sum_i L_i \Delta w_i = \frac{1}{2}\sum_i L_i \left( X_{i-1} - X_{i+1} \right).
\end{equation}

The algorithm terminates when the evidence in the live points falls below a user-specified fraction $\epsilon$ of the total accumulated evidence. Skilling's original paper estimates this remaining evidence $\Delta Z$ to have an upper bound of $ \Like_\mathrm{max} X_i$, while some use the average remaining likelihood. Once we have chosen to stop, the remaining evidence is included by killing the live points one by one without replacement, incrementing $Z$ as before. 
\par
The full nested sampling procedure is given in algorithm \ref{alg:ns}. 
\begin{algorithm}
\caption{Nested sampling}\label{alg:ns}
\begin{algorithmic}
\State Start with $N$ points $\theta_1, \ldots, \theta_N$ from prior
\State \hspace{\algorithmicindent} initialise $Z = 0$, $X_0 = 1$
\For{$i = 1, 2, \ldots , j$}
\State $L_i \gets \text{lowest of the current likelihood values}$
\State $X_i \gets \exp(-i/N)$
\State $w_i \gets X_{i-1} - X_i$ or $(X_{i-1} - X_{i+1})/2$
\State $Z \gets Z + L_i w_i$
\State $L_i \gets $ new point $\in \{\pi(\theta): L(\theta) > L_i\}$.
\EndFor
\State Increment $Z$ by $\langle L(\theta) \rangle X_j$.
\end{algorithmic}
\end{algorithm}

\subsection{Uncertainties and the posterior bulk}\label{subsec:Uncertainties and the posterior bulk}
\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/LX.pdf}
\end{center}
\caption{Plot showing the likelihood and the posterior weights $\Like(X)X$. The bulk of posterior mass is contained within a region of size $e^{-\DKL}$, meaning one must iterate through at least $n\DKL$ steps to fully explore the posterior.}
\label{fig:LX}
\end{figure}

\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/X_samples.pdf}
\end{center}
\caption{Plot showing samples of $\log X$ plotted against the likelihoods of the sample points. The uncertainty increases with the number of steps taken.}
\label{fig:X_samples}
\end{figure}

Most of the posterior is contained within a small fraction of the prior of size $X^{*} = e^{-\DKL}$, where
\begin{equation}
\DKL = \int P(\theta) \log  \frac{P(\theta)}{\pi(\theta)} \ d\theta
\end{equation}
is the information gain between the prior and posterior; this is visualised in figure \ref{fig:LX}. To learn anything about the posterior or calculate an accurate evidence, we must compress down to this volume. 
\par
From equation \eqref{X_distribution}, it is easy to see that for a constant number of live points $n$ $\log X$ follows Poisson statistics:
\begin{equation}
	\mathrm{E}[\log X_i] = - \frac{i}{n}, \quad \mathrm{Var}[\log X_i] = \frac{i}{n^2}.
\end{equation}
This is illustrated in figure \ref{fig:X_samples}, where samples are taken from the $X$ distribution and plotted against the likelihood values. To reach the posterior bulk we must therefore take $i^{*} = n\DKL$ steps, which translates to an uncertainty in $\log X^{*} = -\DKL \pm \sqrt{\DKL/n}$. The evidence is $Z = \langle \Like \rangle X^{*}$, or
\begin{equation}
\log Z = \langle \log \Like \rangle + \log X^{*}.
\end{equation}
The first term has low error, since the only difference that errors in $X$ makes is to slightly change the region we average over. Uncertainties in $\log Z$ are therefore dominated by the $\mathcal{O}(1/\sqrt{n})$ errors on $\log X_i$, which are much more significant than the $\mathcal{O}(n^{-2})$ quadrature errors.
\par
The next section will outline the methodology for making endpoint predictions.

\section{Endpoint prediction}\label{sec:Endpoint prediction}
We shall begin by discussing a simple ``eyeball" method for making predictions that is familiar to experienced users of nested sampling, which makes a good sanity check to compare to the more principled approach.
\subsection{Eyeball method}
At an intermediate iteration of a nested sampling run, the output might look something like that shown in the first two columns of table \ref{tab:ns_output}. 
\begin{figure}
\centering
\scalebox{0.9}{\begin{tabular}{|c|c|c|}
\hline
iteration & log Z  & $d\log Z$   \\
\hline
6000 & -34.481 & 1.317 \\ 
6200 & -32.606 & 1.875 \\ 
6400 & -30.700 & 1.906 \\ 
6600 & -29.008 & 1.692 \\ 
6800 & -27.416 & 1.591 \\ 
7000 & -25.865 & 1.551 \\ 
7200 & -24.449 & 1.416 \\ 
7400 & -23.034 & 1.415 \\ 
7600 & -21.617 & 1.417 \\ 
7800 & -20.267 & 1.349 \\ 
8000 & -18.950 & 1.317 \\
\hline
\end{tabular}}
\captionof{table}{Output of accumulated log-evidence during a nested sampling run.}
\label{tab:ns_output}
\end{figure}
\begin{figure}
  \centering
  \includegraphics{Figures/linear_extrapolate.pdf}
  \captionof{figure}{Prediction of endpoint using linear extrapolation. The estimate is much lower than the true endpoint, primarily because of the long tail in $d\log Z$.}
  \label{fig:linear_extrapolate}
\end{figure}
The outputs $\log Z$ appear to converge at a predictable rate, which is shown in terms the increments in column three. One could feasibly extrapolate these to predict when the increments approach zero, which gives an iteration number for convergence. 
\par
This is shown for a linear extrapolation in figure \ref{fig:linear_extrapolate}. The prediction is clearly an underestimate, which is mainly due to the nonlinear profile of $d\log Z$ having a long tail. However, this makes a good benchmark to beat for the more principled method using likelihood approximation.

\subsection{Likelihood approximation method}
We want to make an estimate of the number of iterations until the termination condition is met, which is when the remaining evidence falls below a fraction $\epsilon$ of the accumulated evidence.
Predictions can be made ahead of time because of the live points, which are sampled from higher likelihood regions that we have not yet compressed to. It is important to note that there will always be a small region which the live points do not penetrate because they are too diffuse, which can be seen in figure \ref{fig:dead_live_points}. 
\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/dead_live_points.pdf}
\end{center}
\caption{Plots showing the state of a nested sampling at an intermediate iteration. The live points are samples from higher likelihood regions, which allow us to see the remaining likelihood ahead of time. However, there is always a small high likelihood region where we have no samples, simply because the live points are too diffuse.}
\label{fig:dead_live_points}
\end{figure}
\par
If we could estimate the form of $\Like(X)$ at the higher likelihoods in terms of some \textit{likelihood model} $f(X, \thetab)$, we would be able to express the prior volume $X_f$ which we need to compress to as
\begin{equation}
	\Delta Z = \epsilon Z_\mathrm{tot},
\end{equation}
\begin{equation}\label{model endpoint}
	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $Z_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_f$ can then be identified by solving the above equation either analytically or numerically. Remembering that at each iteration $\log X$ decreases by $1/\nlive$, the total number of iterations $N_f$ will be
\begin{equation}
	N_f = - \nlive \log X_f .
\end{equation}

\subsubsection{Model choice}\label{sec:Model choice}
The choice of the model $f(X, \thetab)$ serves two purposes; to \textit{smooth} over the stochasticity in the live points, since $\Like(X)$ is typically a smooth function, and more importantly to \textit{extrapolate} into the region where we have no samples.
\par
Making this extrapolation necessitates assumptions about the form of the likelihood in the unsampled region. It is sensible to assume it is guided by what came before it, so we fit the model on the known likelihods of the live and dead points. Apart from this, we should make the model $f(X, \thetab)$ as flexible as possible to extrapolate without bias. We chose the model to be a Gaussian with freely varying parameters $\thetab = (\Like_\mathrm{max}, d, \sigma)$:
\begin{equation}
	f(X, \thetab) = \Like_\mathrm{max} \exp\left(-\frac{X^{2/d}}{2\sigma^2}\right).
\end{equation}
The choice was made because it is analytically tractable, and because many likelihoods in practice are near-Gaussian at their peaks. Allowing $d$ to vary freely increases flexibility, and reflects the fact that while the dimensionality of the posterior is fixed, the equivalent dimension for a Gaussian of the same information content can differ \citep{Handley_2019}.

\subsubsection{Endpoint formula}
Now we are equipped with the tools to make endpoint predictions. We will begin by fitting the model over the live points only. At each iteration, estimate the prior volumes of the live points to be those which correspond to killing off each point one by one without replacement and using formula \eqref{X_distribution}. Combined with the live likelihoods $\logL$, the parameters of the model can be found by minimising the cost function
\begin{equation}\label{chi squared}
	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
\end{equation}
with respect to $\thetab$. A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
\begin{equation}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i \log \Like_i X_i^{2/d}}
\end{equation}
\begin{equation}
    \logLm = \frac{1}{N} \sum_i \logL + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d} 
\end{equation}
(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables. The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run. We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
\par
Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \eqref{model endpoint};
\begin{equation}
	\epsilon = \frac{\int_0^{X_f} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX + Z_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX = \frac{d}{2} \cdot 2^{d/2} \cdot \sigma^{d} \cdot \gamma_k
\end{equation}
where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{multline}
	\log X_f = \frac{d}{2}\log 2 + d \log \sigma \\
	+ \log \gamma^{-1} \left( d/2 ,\ \epsilon \left(\gamma_i+ 2^{-d/2} \sigma^{-d} Z_\mathrm{dead}/\Like_\mathrm{max}\right)\right),
\end{multline}
and $N_f$ is just  $-\nlive$ multiplied by this. 
\par
An useful feature is that the formula is robust when we are far away from the posterior bulk. In this phase, the inferred values of $\Like_\mathrm{max}$ turn out to be very uncertain and often divergent, but the dead evidence is also negligible, so $\Like_\mathrm{max}$ has little impact on the prediction. Once $Z_\mathrm{dead}$ becomes comparable to the live evidence $\Like_\mathrm{max}$ does begin to impact the prediction, but at this point we have reached the posterior bulk, so we expect $\Like_\mathrm{max}$ to be well-estimated and thus contribute in a useful way.
\par
Note that the prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.

\subsubsection{Uncertainties}
Our lack of knowledge about the unsampled region means that there is always the chance that there is some huge likelihood spike at small $X$. Since there is no way to anticipate this, all uncertainty estimates are underestimates in the sense that they cannot account for this possibility; there is no estimate which will \textit{always} include the true endpoint. 
\par
We will therefore restrict ourselves to quantifying the uncertainties associated with how we have modelled the \textit{known} data, remembering that because of our lack of knowledge, the true endpoint may fall significantly outside of this range. These uncertainties have two sources: the variance in the prior volume estimates, and the uncertainty in the least squares minimisation. Since the former is the main error in nested sampling, we expect it to dominate. As we saw in section \ref{subsec:Uncertainties and the posterior bulk}, different sets of $\set{X_i}$ drawn from its probability distribution lead to different "maps" of what the remaining likelihood looks like, which lead to different endpoint predictions. The natural way to handle this is to sample from $\Pr(\bm{X})$ then find the best-fit  $\thetab$ and endpoint for each sample, from which we can obtain the mean and variance in the estimates. An visualisation of this approach is shown in figure \ref{fig:gauss_overfit}.

\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/gauss_overfit.pdf}
\end{center}
\caption{Plot showing fits of the likelihood corresponding to different draws of $\log X$. The different realisations of the likelihood lead to a set of endpoint predictions, from which a mean and variance can be obtained.}
\label{fig:gauss_overfit}
\end{figure}

\section{Results}\label{sec:Results}
\subsection{Toy data}
Let us now apply the methodology developed above in practice. We consider the following toy examples:
\begin{enumerate}[leftmargin=0.3in, label=(\alph*)]
	\item A spherically symmetric Gaussian, $\log \Like = -|\bm{r}|^2/2\sigma^2$ in $d$ dimensions
	\item A discontinous "wedding cake" likelihood with Gaussian profile and parameter $\alpha_\mathrm{w}$ controlling the plateau depth, as presented in \cite{plateaus} and shown in figure \ref{fig:wedding_3d}.
	\item A Cauchy likelihood $\log \Like = - \frac{1 + d}{2} \log (1 + |\bm{r}|^2/\gamma^2)$.
\end{enumerate}
\begin{figure}
\begin{center}
	\includegraphics[scale=]{Figures/wedding_3d.pdf}
\end{center}
\caption{An example ``wedding cake" likelihood which has plateaus on top of a Gaussian profile}
\label{fig:wedding_3d}
\end{figure}

Parameters were chosen to have large values of $\DKL$, which are more difficult to model due to the large distance to the posterior bulk.
\par
Nested sampling runs were made for the above likelihoods for $\nlive = 500$ (plots for different numbers of live points are not shown here, but the results are broadly the same). The state at each iteration (i.e. the likelihoods of the dead and live points) was then extracted using \textsc{anesthetic}. At each iteration, the model is fitted over a number of points determined by the bandwidth selection procedure outlined above. For the error bars, we draw an \textit{ensemble} of 25 $\bm{X}$ samples from the distribution \eqref{xi_formula}, and find the optimal $\thetab$ and corresponding  $N_f$ for each sample to get the mean and variance of the prediction. 
\par
Results are shown in figure \ref{fig:toy_predictions}. For all cases, the predictions become more accurate as the iteration increases because the samples shrink around and eventually cross the posterior bulk, leading to an increasingly representative picture of the posterior. The method works very well for the first two likelihoods, recovering the correct endpoint within standard error for the entire run. This is expected, since the model likelihood has the same form as the true likelihoods, so one expects the extrapolation made to be accurate even when we are far away from the posterior bulk.
\par
The Cauchy likelihood shows the result when this is not the case; without samples in the bulk, a Gaussian extrapolation performs poorly. However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance. 
\begin{figure*}
\begin{center}
	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics[scale=]{Figures/gauss_predictions.pdf}}
	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics[scale=]{Figures/wedding_predictions.pdf}}
	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics[scale=]{Figures/cauchy_predictions.pdf}}
\end{center}
\caption{Plot showing endpoint predictions for the toy likelihoods. The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties. For the first two likelihoods (which have Gaussian profile), the true endpoint falls within error for practically all of the run. The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
\label{fig:toy_predictions}
\end{figure*}

\subsection{Real data}
The model was tested on real nested sampling runs for quantification of cosmological parameter tensions. Results are shown in figure \ref{fig:lcdm_predictions}, with the likelihood approximation method compared against the benchmark.
\par
The likelihood model clearly performs better. For all of the chains, the true endpoint was recovered within arund standard error by the halfway point, and falls within the correct order of magnitude throughout all of the runs. In comparison, the benchmark model mostly failed to get the correct order of magnitude on the number of iterations remaining, especially near the end of each run.
\par
Furthermore, the first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian likelihood model. For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations. 
\begin{figure*}
\begin{center}
	\includegraphics[scale=]{Figures/lcdm_predictions.pdf}
\end{center}
\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model. The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains. The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
\label{fig:lcdm_predictions}
\end{figure*}

\section{Conclusions}\label{sec:Conclusions}
In this project we have developed a model to estimate the endpoint of a nested sampling run, which has no precedent in the literature. This was done by predicting the form of the likelihood in the region that the samples have not reached, via fitting of a Gaussian model likelihood to the points already known to us. Assuming that the likelihood continues in a similar trajectory leads to an estimate of the remaining evidence, and hence the end iteration.  
\par
The approach taken was shown to be highly successful for near-Gaussian likelihoods, finding the true endpoint within error from the start of the run; this was expected, because the form of the model matched the likelihood. For a Cauchy likelihood as well as the real nested sampling runs that were tested, the estimates were less accurate, but still able to find the correct order of magnitude which is useful to the end user. Furthermore, the correct endpoint was able to be recovered within standard error at the halfway point of the runs.
\par
Two main refinements are recommended: the uncertainty quantification should be expanded to include the true endpoint from the beginning of the run, and the analysis should be extended to include the iteration prediction when the number of live points is allowed to vary, so that endpoints can be estimated for dynamic nested sampling.

\section*{Data availability}
The data and code for the project can be found in the repository \url{https://github.com/zixiao-h/aeons}.

\bibliographystyle{mnras}
\bibliography{references}

\label{lastpage}
\end{document}

### 2023-07-21 paper/paper.tex
Added sketch of notes on maximisation observations

@@
 \usepackage{pgf}
+\usepackage{cleveref}
 
@@
 
+\section{The anatomy of a Gaussian}
+
+In general, it is well known (to those who know such things) that the anatomy of a Gaussian says that
+\begin{equation}
+    \boxed{\log\mathcal{L}_\mathrm{max} = \langle\log\mathcal{L}\rangle_\mathcal{P} + \frac{d}{2}},  \qquad \boxed{\mathrm{var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}},
+\end{equation}
+namely the posterior average of the log likelihood is $d/2$ less than the maximum, and the variance of the loglikelihood over the posterior is $\frac{d}{2}$.
+Whilst this is exact for gaussians, this provides a good rule of thumb for other distributions.
+
+For example, one may use this to explain why MCMC sampling is a very bad way to maximise, since even in the limit of a very large number of samples from the posterior $n$, the largest of these will be within a few $\sqrt{\frac{d}{2}}$ standard deviations of the mean, and therefore very far away from the maximum $\frac{d}{2}$ away.\footnote{TODO: can we do some analytics on the MCMC eaxmple as well?}
+
+The same is true for nested sampling, although since it is not a posterior sampler, but a likelihood upward scanner, the argument is a little more subtle, which we work through below.
+
+We will work with a Gaussian loglikelihood, lengthscale $\sigma$, defined in terms of prior volumes $0<X<1$
+\begin{equation}
+    \log\mathcal{L}
+    =\log\mathcal{L}_\mathrm{max} - \frac{1}{2\sigma^2}X^{2/d}
+    =\log\mathcal{L}_\mathrm{max} - \frac{1}{2\sigma^2}\exp\left(\frac{2}{d}\log X\right).
+    \label{eq:LX}
+\end{equation}
+We will assume the run has progressed far enough so that prior effects may be neglected, and that $\sigma\ll 1$ is sufficiently small that for the arguments below edge effects may be neglected.
+
+Note that from the Gaussian profile we can see that as $\log X\to-\infty$ at a steady rate of compression, we approach the maximum loglikelihood exponentially slowly, with an decay rate of $\frac{d}{2}$
+
+Taking $X \sim e^{-i/n}$ we see this as
+\begin{equation}
+    \log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \exp\left(-\frac{i-i_0}{nd/2}\right), \qquad i_0 = -\frac{nd}{2}\log 2\sigma^2
+\end{equation}
+in principle this converges very quickly, although never quite reaches the maximum.
+
+\section*{The anatomy of a gaussian}
+\begin{equation}
+    P(\log\mathcal{L}) = \frac{1}{\Gamma(\frac{d}{2})}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
+\end{equation}
+i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})\sim \chi^2_{d} $
+
+\section*{Observation 1: the largest likelihood live point}
+Assume we are at an iteration such that $\mathcal{L}_*$ is the current likelihood contour (i.e.
+the likelihood of the dead point we have just deleted), and that we have $n$ live pointsdistributed uniformly in prior volume beneath this.
+Since the the live points are distributed in prior volume up to $X_*$, $P(X) = [0<X<X_*]$,
+Inverting \cref{eq:LX} we may show that the distribution of loglikelihoods is
+\begin{equation}
+    P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}}[0<\log\mathcal{L}<\log\mathcal{L}_*].
+    \label{eq:PL}
+\end{equation}
+It is helpful at this stage therefore to define a parameter
+\begin{equation}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*} \qquad\Leftrightarrow\qquad \log\mathcal{L} = y\log\mathcal{L}_\mathrm{max} + (1-y)\log\mathcal{L}_*
+    \label{eq:y}
+\end{equation}
+as a normalised likelihood with $y=0$ corresponding to $\mathcal{L}=\mathcal{L}_*$ and $y=1$ to $\mathcal{L}=\mathcal{L}_\mathrm{max}$, which has distribution
+\begin{equation}
+    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1}[0<y<1].
+    \label{eq:Py}
+\end{equation}
+We now seek the distribution of the largest likelihood $\log{\mathcal{L}}_\mathrm{max}^\mathrm{live}$ (or equivalently the largest ${y}_\mathrm{max}^\mathrm{live}$) in the set of $n$ live points.
+We first integrate to find the CDF
+\begin{equation}
+    P(y<Y) = 1-(1-Y)^{\frac{d}{2}}[0<Y<1].
+    \label{eq:Py}
+\end{equation}
+note that to find the largest of $n$ independent variables we simply multiply 
+\begin{equation}
+    P(y_1,y_2,\ldots<Y) = (1-(1-Y)^{\frac{d}{2}})^n[0<Y<1].
+    \label{eq:Py}
+\end{equation}
+and then differentiate with respect to $Y$ to find the PDF for largest likelihood in the live points
+\begin{equation}
+    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
+    \label{eq:Pyhat}
+\end{equation}
+which may be roughly summarised as:
+\begin{equation}
+    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
+    \label{eq:ymax}
+\end{equation}
+or in the large $d$, $n$ limit
+\begin{align}
+    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
+    \label{eq:ymaxd}\\
+    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
+    \label{eq:ymaxdn}
+\end{align}
+where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
+
+This shows that in general the live points are nowhere near the maximum loglikelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
+In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
+Our end result is that:
+\begin{equation}
+    \boxed{
+        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - (1-y_\mathrm{max}^\mathrm{live})(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)
+    },
+\end{equation}
+where you may take any of the $y_\mathrm{max}^\mathrm{live}$ expressions above from \cref{eq:ymax,eq:ymaxd,eq:ymaxdn}.
+
+We may also rewrite the distribution \cref{eq:Pyhat} without $y$ to give
+\begin{equation}
+    \frac{\log\mathcal{L}_\mathrm{max}^\mathrm{live}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
+\end{equation}
+\begin{equation}
+    P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left((\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
+\end{equation}
+
+\subsection*{Observation 2: the termination contour $\mathcal{L}_\mathrm{end}$}
+
+One strategy for terminating nested sampling is to halt when the amount of evidence in the live points is some fraction $f$ of the amount of total evidence (or evidence in the dead points).
+Providing $f\ll 1$, then this is approximately the same as the fraction of evidence left above likelihood contour $\mathcal{L}_\mathrm{end}$:
+\begin{equation}
+    f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L} dX}{\int_0^\infty \mathcal{L} dX}.
+    \label{eq:fint}
+\end{equation}
+Note that we have assumed that prior effects are negligible (so $1=\infty$).
+Computing this for \cref{eq:LX} we find the answer in terms of lower incomplete gamma functions
+\begin{equation}
+    f = 1- \frac{\Gamma_{d/2}(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2})}{\Gamma(\frac{d}{2})}.
+    \label{eq:f}
+\end{equation}
+Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
+\begin{equation}
+    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} f \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d\Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
+\end{equation}
+We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $f$.
+This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a stirling/Stirling approximation:
+\begin{equation}
+    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
+\end{equation}
+In the event that we keep $f$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}f^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
+
+\subsection*{Observation 3: nested sampling as a maximiser}
+
+Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with the large $d,n$ limit of \cref{eq:ymaxdn} we find
+\begin{equation}
+    \boxed{
+        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
+    },
+\end{equation}
+showing that in general the anatomy that nested sampling will finish at a contour $\frac{d}{2e}$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\frac{\log n}{2e}$ closer, with a chance of getting $\sim\frac{\pi}{\sqrt{6}e}=0.472$ closer still by statistical fluctuation.
+
+\begin{figure*}
+\begin{center}
+	\includegraphics{Figures/anatomy.pdf}
+\end{center}
+\caption{}
+\label{fig:anatomy}
+\end{figure*}
+
 \section{Conclusions}\label{sec:Conclusions}

### 2023-07-21 paper/Figures/anatomy.py

@@ -1,0 +1,75 @@
+import matplotlib.pyplot as plt
+import numpy as np
+from matplotlib import rcParams
+rcParams['text.usetex'] = True 
+rcParams['font.family'] = 'serif'
+
+
+fig, ax = plt.subplots()
+
+logLmax = 0
+d = 30
+n = 1000
+avlogL = logLmax-d/2
+
+logL = np.linspace(logLmax-d/2-np.sqrt(d/2)*3.5,0,1000)
+logpdf = logL - logLmax + (d/2-1)*np.log(logLmax-logL)
+logpdf -= np.max(logpdf)
+ax.plot(logL, np.exp(logpdf), label='posterior: $\mathcal{P}(\log\mathcal{L})$')
+
+logLs = logLmax-d/2/np.e
+logL = np.linspace(logLs,0,1000)
+logpdf = (d/2-1) * np.log(logLmax - logL)
+logL = np.concatenate([[logLs], logL])
+logpdf = np.concatenate([[-np.inf], logpdf])
+logpdf -= np.max(logpdf)
+ax.plot(logL, np.exp(logpdf), label='live points: $\mathcal{\pi}(\log\mathcal{L}|\mathcal{L}>\mathcal{L}_*)$')
+
+from scipy.special import logsumexp
+
+logL = np.linspace(logLs,0,1000)
+logpdf = (d/2-1)*np.log(logLmax - logL) + (n-1)* logsumexp([ d/2*np.log(logLmax-logLs)*np.ones_like(logL), d/2*np.log(logLmax-logL)] , b=[np.ones_like(logL),-np.ones_like(logL)], axis=0)
+logpdf -= np.max(logpdf)
+ax.plot(logL, np.exp(logpdf), label='maximum live point: $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$')
+
+
+logLmaxlive = logLmax - d/2/np.e + np.log(n)/np.e
+
+l1 = 1.3
+ax.axvline(avlogL, color='k', linestyle=':')
+ax.axvline(logLmax, color='k', linestyle=':')
+ax.plot([avlogL, logLmax], [l1,l1], 'k-')
+ax.annotate(r'$\frac{d}{2}$', ((avlogL+logLmax)/2,l1), ha='center', va='bottom')
+
+l2 = 0.7
+ax.plot([avlogL-np.sqrt(d/2), avlogL+np.sqrt(d/2)], [l2, l2], 'k-')
[…]

### 2023-07-21 paper/paper.tex
Added additional MCMC content

@@
 
-For example, one may use this to explain why MCMC sampling is a very bad way to maximise, since even in the limit of a very large number of samples from the posterior $n$, the largest of these will be within a few $\sqrt{\frac{d}{2}}$ standard deviations of the mean, and therefore very far away from the maximum $\frac{d}{2}$ away.\footnote{TODO: can we do some analytics on the MCMC eaxmple as well?}
+For example, one may use this to explain why MCMC sampling is a very bad way to maximise, since even in the limit of a very large number of samples from the posterior $n$, the largest of these will be within a few $\sqrt{\frac{d}{2}}$ standard deviations of the mean, and therefore very far away from the maximum $\frac{d}{2}$ away.
+
+We can get someway to the analytics for MCMC. Similar arguments to below show that the distribution of the largest likelihood point found from a posterior is:
+\begin{equation}
+    P(x) = n \frac{e^{-x} x^{\frac{d}{2}-1} }{\Gamma\left(\frac{d}{2}\right)} {\left( \frac{\Gamma_{\frac{d}{2}}(x)}{\Gamma(\frac{d}{2})} \right)}^{n-1},
+\end{equation}
+where $x=2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{max}^\mathrm{posterior})$
+since if we have the CDF of a distribution as $F(x)$, then the lowest of n random variables is $\frac{d}{dx}( 1- (1-F(x))^n)$, and we the upper CDF for a $\chi^2_{d/2}$ distribution is $\Gamma_{\frac{d}{2}}(x)/\Gamma(\frac{d}{2})$.
+I have not yet been able to find an analytic expression/approximation for the center and width of this distribution.
 

### 2023-07-24 aeons/endpoint.py
first implementation of cauchy regression

@@ -28,6 +28,7 @@ class EndModel:
         logL = np.array(points.logL)
         nk = np.array(points.nlive)
         X_mean = X_mu(nk)
+        logZdead = self.logZs.iloc[ndead]
         if live:
             return logL[ndead:], X_mean[ndead:], nk, logZdead
         return logL, X_mean, nk, logZdead
@@ -76,9 +77,12 @@ def theta_bandwidth(logL, X, print_split=False, splits=4):
 def theta_bandwidth_trunc(logL, X, trunc=15, **kwargs):
     theta = theta_bandwidth(logL, X)
     attempts = 1
-    while theta[1]/2 > 180:
-        theta = theta_bandwidth(logL[:-attempts*trunc], X[:-attempts*trunc], **kwargs)
-        attempts += 1
+    while (theta is None) or (theta[1]/2 > 180) or ((trunc * attempts) > len(logL)):
+        try:
+            theta = theta_bandwidth(logL[:-attempts*trunc], X[:-attempts*trunc], **kwargs)
+            attempts += 1
+        except:
+            attempts +=1
     if attempts > 1:
         print(f"{attempts} attempts")
     return theta
@@ -104,9 +108,10 @@ def fit_dlogZ(dlogZ, deg):
 
 class IncrementEndpoint:
     def __init__(self, samples, N_rolling):
-        self.samples = add_logZ(samples)
-        self.logZ = self.samples['logZ']
-        self.dlogZ = get_dlogZ(self.logZ)
+        samples['logZs'] = np.logaddexp.accumulate(samples.logw())
+        self.logZs = samples.logZs
+        self.samples = samples
+        self.dlogZ = get_dlogZ(self.logZs)
         self.dlogZ_rolling = get_dlogZ_rolling(self.dlogZ, N_rolling)
         self.N_rolling = N_rolling
         self.rolling_index = self.dlogZ_rolling.index.get_level_values(0)
@@ -117,7 +122,7 @@ class IncrementEndpoint:
         plt.plot(dlogZ.index.get_level_values(0), dlogZ.values)
 
     def calc_endpoint(self, epsilon=1e-3):
-        logZ = self.logZ
+        logZ = self.logZs
         logZ_tot = logZ.iloc[-1]
         logZ_f = np.log(1 - epsilon) + logZ_tot
         index_f = logZ[logZ > logZ_f].index.get_level_values(0)[0]
[…]

### 2023-07-24 aeons/utils.py

@@ -5,6 +5,9 @@ from scipy.special import gamma, gammainc, logsumexp, gammaincinv
 
 proj_dir = '/home/zixiao/Documents/III/project'
 aeons_dir = '/home/zixiao/Documents/III/project/aeons'
+lcdm_chains = ['BAO', 'lensing', 'lensing_BAO', 'lensing_SH0ES', 'planck', 'planck_BAO', \
+              'planck_lensing', 'planck_SH0ES', 'planck_lensing_BAO', 'planck_lensing_SH0ES', 'SH0ES']
+toy_chains = ["gauss_30_01", "wedding_20_001", "cauchy_10_0001", "planck_gaussian", "gp"]
 
 def pickle_dump(filename, data):
     """Function that pickles data into a file"""
@@ -40,15 +43,10 @@ def read_from_txt(filename):
             data.append(np.fromstring(line.rstrip('\n'), sep=','))
     return data
 
-def get_samples(path):
-    """
-    Example usage:
-    path="lcdm/lensing_BAO" or "toy/planck_gaussian" or "gp"
-    """
-    root = f"{aeons_dir}/samples/"
-    samples = pickle_in(root+path+".pickle")
-    name = path
-    return path, samples
+def get_samples(root, chain):
+    path = f"{aeons_dir}/samples/{root}/{chain}.pickle"
+    samples = pickle_in(path)
+    return chain, samples
 
 def data_split(logLlive, Xlive, splits=1, trunc=None):
     if trunc is not None:
@@ -57,16 +55,19 @@ def data_split(logLlive, Xlive, splits=1, trunc=None):
     X_split, logL_split = Xlive[start:], logLlive[start:]
     return logL_split, X_split
 
-def formatt(theta):
+def formatt(theta, sigfigs=2):
     logLmax, d, sigma = theta
-    return f"[{logLmax:.1e}, {d:.1f}, {sigma:.0e}]"
+    if isinstance(sigfigs, int):
+        return f"[{logLmax:.{sigfigs}e}, {d:.{sigfigs}f}, {sigma:.{sigfigs}e}]"
+    else:
+        s1, s2, s3 = sigfigs
+        return f"[{logLmax:.{s1}e}, {d:.{s2}f}, {sigma:.{s3}e}]"
 
 def reject_outliers(data):
     data = np.array(data)
     dev = np.abs(data - np.median(data))
     median_dev = np.median(dev)
     return data[dev < 2 * median_dev]
-
 
[…]

### 2023-07-24 notebooks/summer_4/cauchy_ns.py

@@ -1,0 +1,63 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+from aeons.utils import *
+from aeons.endpoint import EndModel, theta_basic, theta_bandwidth, theta_bandwidth_trunc
+from aeons.plotting import *
+from aeons.regress import *
+
+def logcauchy(X, theta, torched=False):
+    logLmax, c, gamma = theta
+    if torched:
+        import torch
+        return logLmax - (1 + c)/2 * torch.log( 1 + X**(2/c)/(gamma**2) )
+    return logLmax - (1 + c)/2 * np.log( 1 + X**(2/c)/gamma**2 )
+
+def logLmax_cauchy(logLi, Xi, c, gamma):
+    N = len(logLi)
+    summand = logLi + (1 + c)/2 * np.log( 1 + Xi**(2/c)/gamma**2 )
+    return np.sum(summand)/N
+
+# Load data
+name, samples = get_samples("toy", "cauchy_10_0001")
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+true_logXf = samples.logX().iloc[true_endpoint]
+
+ndead = 5000
+logLall, Xall, nk, logZdead = model.data(ndead, live=False)
+logLall, Xall = np.array(logLall), np.array(Xall)
+window = np.arange(ndead, len(logLall))
+logLd, Xd = logLall[window], Xall[window]
+
+
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import UniformPrior, LogUniformPrior
+
+def likelihood(theta):
+    c, gamma = theta
+    logLmax = logLmax_cauchy(logLd, Xd, c, gamma)
+    Lsq = np.sum((logLd - logcauchy(Xd, [logLmax, c, gamma]))**2)
+    s = np.sqrt(Lsq/len(logLd))
+    return -1/2 * len(logLd) * np.log(2*np.pi*s**2) - Lsq/(2*s**2), []
+
+def prior(hypercube):
+    c = UniformPrior(0, 1e3)(hypercube[0])
+    gamma = LogUniformPrior(1e-20, 1)(hypercube[1])
+    return np.array([c, gamma])
+
+def dumper(live, dead, logweights, logZ, logZerr):
+    print("Last dead point:", dead[-1])
+
+nDims = 2
+nDerived = 0
+settings = PolyChordSettings(nDims, nDerived)
+settings.file_root = 'cauchy_5000'
+settings.nlive = 200
[…]

### 2023-07-24 notebooks/summer_4/cauchy_regress.py

@@ -1,0 +1,26 @@
+import numpy as np
+
+def logcauchy(X, theta, torched=False):
+    logLmax, c, gamma = theta
+    if torched:
+        import torch
+        return logLmax - (1 + c)/2 * torch.log( 1 + X**(2/c)/(gamma**2) )
+    return logLmax - (1 + c)/2 * np.log( 1 + X**(2/c)/gamma**2 )
+
+def logLmax_cauchy(logLi, Xi, c, gamma):
+    N = len(logLi)
+    summand = logLi + (1 + c)/2 * np.log( 1 + Xi**(2/c)/gamma**2 )
+    return np.sum(summand)/N
+
+def cauchy_logXf(theta, logZdead, logX0):
+    # Do likelihood prediction with theta
+    logX_pred = np.linspace(logX0, -200, 2000)
+    X_pred = np.exp(logX_pred)
+    logL_pred = logcauchy(X_pred, theta)
+    dlogX = abs(np.diff(logX_pred)[0])
+    logsummands_pred = logL_pred + logX_pred + np.log(dlogX)
+    logZ_deads_pred = np.logaddexp(logZdead, np.logaddexp.accumulate(logsummands_pred))
+    logZ_tot_pred = logZ_deads_pred[-1]
+    log_differences = logZ_deads_pred - logZ_tot_pred
+    logZ_fracs_pred = np.log(1 - np.exp(log_differences), where=log_differences != 0)
+    return logX_pred[np.argmin(np.abs(logZ_fracs_pred - np.log(1e-3)))]

### 2023-07-24 notebooks/summer_4/theta_samples.py

@@ -1,0 +1,40 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from cauchy_regress import logcauchy, logLmax_cauchy
+
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import UniformPrior, LogUniformPrior
+
+def call_polychord(logL, X, file_root, nlive=50):
+    def likelihood(theta):
+        c, gamma = theta
+        logLmax = logLmax_cauchy(logL, X, c, gamma)
+        Lsq = np.sum((logL - logcauchy(X, [logLmax, c, gamma]))**2)
+        s = np.sqrt(Lsq/len(logL))
+        return -1/2 * len(logL) * np.log(2*np.pi*s**2) - Lsq/(2*s**2), []
+
+    def prior(hypercube):
+        c = UniformPrior(0, 1e3)(hypercube[0])
+        gamma = LogUniformPrior(1e-20, 1)(hypercube[1])
+        return np.array([c, gamma])
+
+    def dumper(live, dead, logweights, logZ, logZerr):
+        print("Last dead point:", dead[-1])
+
+    nDims = 2
+    nDerived = 0
+    settings = PolyChordSettings(nDims, nDerived)
+    settings.file_root = file_root
+    settings.nlive = nlive
+    settings.do_clustering = False
+    settings.read_resume = False
+
+    output = pypolychord.run_polychord(likelihood, nDims, nDerived, settings, prior, dumper)
+    paramnames = [('c', 'c'), ('gamma', r'\gamma')]
+    output.make_paramnames_files(paramnames)
+    
+    from anesthetic import read_chains
+    samples = read_chains(settings.base_dir + '/' + settings.file_root)
+    return samples
+

### 2023-07-24 paper/paper.tex

@@
 since if we have the CDF of a distribution as $F(x)$, then the lowest of n random variables is $\frac{d}{dx}( 1- (1-F(x))^n)$, and we the upper CDF for a $\chi^2_{d/2}$ distribution is $\Gamma_{\frac{d}{2}}(x)/\Gamma(\frac{d}{2})$.
-I have not yet been able to find an analytic expression/approximation for the center and width of this distribution.
+I have not yet been able to find an analytic expression/approximation for the center and width of this distribution, but Jensen's inequality with a Gaussian approximation\footnote{\url{https://math.stackexchange.com/a/89147}} yields
+\begin{equation}
+    \log\mathcal{L}_\mathrm{max}^\mathrm{posterior} \le  \log\mathcal{L}_\mathrm{max} - \frac{d}{2} + \sqrt{d\log n} 
+    %\log\mathcal{L}_\mathrm{max}^\mathrm{posterior} \leq \log\mathcal{L}_\mathrm{max} - \frac{d}{2} + \sqrt{d\log n}.
+\end{equation}
 

### 2023-07-27 aeons/endpoint.py
four parameter model and ignore prior

@@ -106,6 +106,7 @@ def fit_dlogZ(dlogZ, deg):
     return coefs
 
 
+from aeons.endpoint import *
 class IncrementEndpoint:
     def __init__(self, samples, N_rolling):
         samples['logZs'] = np.logaddexp.accumulate(samples.logw())
@@ -122,68 +123,74 @@ class IncrementEndpoint:
         plt.plot(dlogZ.index.get_level_values(0), dlogZ.values)
 
     def calc_endpoint(self, epsilon=1e-3):
-        logZ = self.logZs
-        logZ_tot = logZ.iloc[-1]
+        logZs = self.logZs
+        logZ_tot = logZs.iloc[-1]
         logZ_f = np.log(1 - epsilon) + logZ_tot
-        index_f = logZ[logZ > logZ_f].index.get_level_values(0)[0]
+        index_f = logZs[logZs > logZ_f].index.get_level_values(0)[0]
         return index_f
 
     def index(self, iteration):
         return self.dlogZ.iloc[iteration:].index.get_level_values(0).values
     
-    def dlogZ_fit(self, iteration, N_fit):
-        return self.dlogZ_rolling.iloc[iteration - N_fit - self.N_rolling : iteration - self.N_rolling]
+    def dlogZ_fit(self, iteration, steps, step):
+        index = np.arange(iteration - steps*step, iteration+step, step)
+        logZs = self.logZs.iloc[index]
+        return logZs.diff(1).dropna()
     
-    def pred(self, iteration, N_fit):
-        index = np.arange(iteration - N_fit, len(self.samples))
-        dlogZ_fit = self.dlogZ_fit(iteration, N_fit)
+    def pred(self, iteration, steps, step):
+        index = np.arange(iteration - steps*step, len(self.samples))
+        dlogZ_fit = self.dlogZ_fit(iteration, steps, step)
         coefs = fit_dlogZ(dlogZ_fit, 1)
         dlogZ_pred = poly.polyval(index, coefs)
         index_pred = index[dlogZ_pred > 0]
         dlogZ_pred = dlogZ_pred[dlogZ_pred > 0]
         return index_pred, dlogZ_pred
 
-    def plot_pred(self, iteration, N_fit):
-        index_pred, dlogZ_pred = self.pred(iteration, N_fit)
-        dlogZ_rolling = self.dlogZ_rolling
-        dlogZ_fit = self.dlogZ_fit(iteration, N_fit)
+    def plot_pred(self, iteration, steps, step):
[…]

### 2023-07-27 aeons/plotting.py

@@ -44,7 +44,7 @@ def plot_std(xvals, y_means, y_stds, true, ylim=None, ax=None):
     if isinstance(ylim, float):
         ax.set_ylim(0, true*ylim)
     elif isinstance(ylim, tuple):
-        ax.set_ylim(*ylim)
+        ax.set_ylim(ylim[0]*true, ylim[1]*true)
 
     
 def plot_split(model, ndead, splits=2, trunc=None):

### 2023-07-27 notebooks/summer_4/theta_samples.py

@@ -37,4 +37,3 @@ def call_polychord(logL, X, file_root, nlive=50):
     from anesthetic import read_chains
     samples = read_chains(settings.base_dir + '/' + settings.file_root)
     return samples
-

### 2023-07-27 notebooks/summer_5/four_ns.py

@@ -1,0 +1,66 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+from aeons.utils import *
+from aeons.endpoint import EndModel, theta_basic, theta_bandwidth, theta_bandwidth_trunc
+from aeons.plotting import *
+from aeons.regress import *
+
+def logf4(X, theta):
+    logLmax, Delta, d, gamma = theta
+    return logLmax - Delta * np.log(1 + X**(2/d)/gamma**2)
+
+def logLmax_Delta_estimator(logL, X, d, gamma):
+    N = len(logL)
+    logXsum = np.sum( np.log(1 + X**(2/d)/gamma**2) )
+    logLlogXsum = np.sum( logL * np.log(1 + X**(2/d)/gamma**2) )
+    logXlogXsum = np.sum( (np.log(1 + X**(2/d)/gamma**2))**2 )
+    logLsum = np.sum(logL)
+    denom = (logXsum**2)/N - logXlogXsum
+    logLmax_estimator = (logLlogXsum * logXsum/N - logXlogXsum * logLsum/N)/denom
+    Delta_estimator = (logLlogXsum - logXsum * logLsum/N)/denom
+    return logLmax_estimator, Delta_estimator
+
+# Load data
+name, samples = get_samples("toy", "gauss_30_01")
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+true_logXf = samples.logX().iloc[true_endpoint]
+
+ndead = 5000
+logLall, Xall, nk, logZdead = model.data(ndead, live=False)
+logLall, Xall = np.array(logLall), np.array(Xall)
+window = np.arange(ndead, len(logLall))
+logLd, Xd = logLall[window], Xall[window]
+
+
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import UniformPrior, LogUniformPrior
+
+def likelihood(theta):
+    d, gamma = theta
+    logLmax, Delta = logLmax_Delta_estimator(logLd, Xd, d, gamma)
+    Lsq = np.sum((logLd - logf4(Xd, [logLmax, Delta, d, gamma]))**2)
+    s = np.sqrt(Lsq/len(logLd))
+    return -1/2 * len(logLd) * np.log(2*np.pi*s**2) - Lsq/(2*s**2), []
+
+def prior(hypercube):
+    c = UniformPrior(0, 1e3)(hypercube[0])
+    gamma = LogUniformPrior(1e-10, 1)(hypercube[1])
[…]

### 2023-07-27 paper/paper.tex
added anatomy figure to paper

@@
 \title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
-\author[Z. Hu]{Zixiao Hu, Will Handley}
+\author[Z. Hu et. al]{Zixiao Hu, Artyom Baryshnikov, Will Handley}
 
@@
 \begin{abstract}
-This project develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration.
-The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and final prior volume to be calculated.
-We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods, and by the halfway point for non-Gaussian likelihoods while identifying the correct order of magnitude throughout.
+This paper develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration for any nested sampler.
+The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and prior volume at termination to be calculated.
+We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods.
+For non-Gaussian likelihoods, the estimate is still the correct order of magnitude and converges to the correct answer by the end of the run.
 \end{abstract}
@@
 \par
-An useful feature is that the formula is robust when we are far away from the posterior bulk.
+Noting that the prediction depends on the \textit{ratio} $Z_\mathrm{dead}/\Like_\mathrm{max}$, an useful feature is that the formula is robust when we are far away from the posterior bulk.
 In this phase, the inferred values of $\Like_\mathrm{max}$ turn out to be very uncertain and often divergent, but the dead evidence is also negligible, so $\Like_\mathrm{max}$ has little impact on the prediction.

### 2023-07-28 paper/paper.tex
Fixed figures...

@@
 It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{split}.
-The core algorithm is unique for being the only algorithm which estimates high-dimensional volumes using \textit{order statistics}, which makes high-dimensional integration feasible.
+The core algorithm is unique for being the only algorithm which estimates volumes using \textit{order statistics}, which makes high-dimensional integration feasible.
 It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
@@
 \begin{center}
-	\includegraphics[scale=]{Figures/LX.pdf}
+	\includegraphics{Figures/LX.pdf}
 \end{center}
@@
 \begin{center}
-	\includegraphics[scale=]{Figures/X_samples.pdf}
+	\includegraphics{Figures/X_samples.pdf}
 \end{center}
@@
 \begin{center}
-	\includegraphics[scale=]{Figures/dead_live_points.pdf}
+	\includegraphics{Figures/dead_live_points.pdf}
 \end{center}
@@
 \begin{center}
-	\includegraphics[scale=]{Figures/gauss_overfit.pdf}
+	\includegraphics{Figures/gauss_overfit.pdf}
 \end{center}
@@
 \begin{center}
-	\includegraphics[scale=]{Figures/wedding_3d.pdf}
+	\includegraphics{Figures/wedding_3d.pdf}
 \end{center}
@@
 \begin{center}
-	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics[scale=]{Figures/gauss_predictions.pdf}}
-	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics[scale=]{Figures/wedding_predictions.pdf}}
-	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics[scale=]{Figures/cauchy_predictions.pdf}}
+	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
+	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
+	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
 \end{center}
@@
 \begin{center}
-	\includegraphics[scale=]{Figures/lcdm_predictions.pdf}
+	\includegraphics{Figures/lcdm_predictions.pdf}
 \end{center}

### 2023-07-28 paper/Figures/LX.py

@@ -1,0 +1,34 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import figsettings
+
+figsettings()
+
+
+def full(X, theta):
+    logLmax, d, sigma = theta
+    return logLmax - X**(2/d)/(2*sigma**2)
+
+logXarray = np.linspace(-70, 0, 1000)
+Xarray = np.exp(logXarray)
+theta = [0, 10, 0.01]
+logL = full(Xarray, theta)
+L = np.exp(logL)
+
+Ztrue = np.trapz(L, Xarray)
+LlogL = np.nan_to_num(L/Ztrue * np.log(L/Ztrue))
+H = np.trapz(LlogL, Xarray)
+
+fig, ax1 = plt.subplots(figsize=(3.37, 1.5))
+ax2 = ax1.twinx()
+ax1.plot(logXarray, L, color='black', label=r"$L(X)$")
+ax2.fill(logXarray, L*Xarray, lw=1, color='black', alpha=0.2, label=r"$L(X)X$")
+ax2.axvline(-H, ls='--', lw=0.5, color='black')
+for ax in ax1, ax2:
+    ax.set_yticks([])
+    ax.set_xticks([-H], [f'$-D_\mathrm{{KL}}$'])
+    ax.margins(x=0, y=0)
+ax1.set_xlabel(r'$\log X$');
+fig.legend(loc="upper right", bbox_to_anchor=(1,1), bbox_transform=ax.transAxes)
+fig.tight_layout()
+fig.savefig('LX.pdf')

### 2023-07-28 aeons/utils.py
First implementation of runtime progress bar

@@ -139,4 +139,26 @@ def calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive):
     return endpoints, endpoints_std
 
 
-    
+def figsettings():
+    format = {
+        # Set figure size to (4,2)
+        "figure.figsize": (4, 2),
+        # Use LaTeX to write all text
+        "pgf.texsystem": "pdflatex",
+        "text.usetex": True,
+        "font.family": "serif",
+        # Use 11pt font in plots, to match 11pt font in document
+        "axes.labelsize": 9,
+        "font.size": 9,
+        # Make the legend/label fonts a little smaller
+        "legend.fontsize": 7,
+        "xtick.labelsize": 7,
+        "ytick.labelsize": 7,
+        'axes.linewidth': 0.5,
+        'axes.spines.top': False,
+        'axes.spines.right': False,
+        'patch.linewidth': 0.5,
+        'legend.fancybox': False,
+        'legend.shadow': False
+    }
+    plt.rcParams.update(format)

### 2023-07-28 notebooks/summer_5/runtime/run_pypolychord.py

@@ -1,0 +1,77 @@
+from numpy import pi, log, exp
+import numpy as np
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import UniformPrior
+from aeons.utils import generate_Xs, X_mu, logXf_formula, calc_endpoints, reject_outliers
+from aeons.regress import analytic_lm_params
+from aeons.endpoint import theta_bandwidth_trunc
+
+#| Define a four-dimensional spherical gaussian likelihood,
+#| width sigma=0.1, centered on the 0 with one derived parameter.
+#| The derived parameter is the squared radius
+
+nDims = 10
+nDerived = 0
+sigma = 0.01
+
+def likelihood(theta):
+    """ Simple Gaussian Likelihood"""
+    nDims = len(theta)
+    r2 = sum(theta**2)
+    logL = -log(2*pi*sigma*sigma)*nDims/2.0
+    logL += -r2/2/sigma/sigma
+
+    return logL, [r2]
+
+#| Define a box uniform prior from -1 to 1
+
+def prior(hypercube):
+    """ Uniform prior from [-1,1]^D. """
+    return UniformPrior(-1, 1)(hypercube)
+
+#| Optional dumper function giving run-time read access to
+#| the live points, dead points, weights and evidences
+
+def dumper(live, dead, logweights, logZ, logZerr):
+    ndead = len(dead)
+    nlive = len(live)
+    if nlive == 0:
+        print(f"[{'='*50}>]", end='\n')
+        return
+    logL = live[:, -1]
+    # Sort logL in ascending order
+    logL = np.sort(logL)
+    nk = np.concatenate([nlive * np.ones(ndead), np.arange(nlive, 0, -1)])
+    X_mean = X_mu(nk)[ndead:]
+    # theta = analytic_lm_params(logL, X_mean, d0=2)
+    # theta = theta_bandwidth_trunc(logL, X_mean, trunc=5, splits=[1])
+    Nset = 25
+    logXfs = np.zeros(Nset)
+    for i in range(Nset):
+        X = generate_Xs(nk)[ndead:]
+        theta = analytic_lm_params(logL, X, d0=2)
+        logXfs[i] = logXf_formula(theta, logZ, X_mean[0])
+    logXfs = logXfs[~np.isnan(logXfs)]
+    logXfs = reject_outliers(logXfs)
[…]

### 2023-08-07 paper/first_draft.tex
theory section of paper

--- 2023-07-28 paper/paper.tex
@@
 However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance.
-\begin{figure*}
-\begin{center}
-	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
-	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
-	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
-\end{center}
-\caption{Plot showing endpoint predictions for the toy likelihoods.
+%\begin{figure*}
+%\begin{center}
+%	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
+%	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
+%	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
+%\end{center}
+%\caption{Plot showing endpoint predictions for the toy likelihoods.
 The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties.
@@
 The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
-\label{fig:toy_predictions}
-\end{figure*}
+%\label{fig:toy_predictions}
+%\end{figure*}
 
@@
 For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations.
-\begin{figure*}
-\begin{center}
-	\includegraphics{Figures/lcdm_predictions.pdf}
-\end{center}
-\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model.
+%\begin{figure*}
+%\begin{center}
+%	\includegraphics{Figures/lcdm_predictions.pdf}
+%\end{center}
+%\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model.
 The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains.
 The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
-\label{fig:lcdm_predictions}
-\end{figure*}
+%\label{fig:lcdm_predictions}
+%\end{figure*}
 
@@
 \end{equation}
-\begin{equation}
-    P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left((\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
-\end{equation}
+\begin{multline}
+	P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\Big( (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}\\
+	-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}} \Big)^{n-1} \quad [0<y_\mathrm{max}^\mathrm{live}<1],
+\end{multline}
 

### 2023-08-07 paper/paper.tex
theory section of paper

@@
 \documentclass[usenatbib]{mnras}
-\usepackage[T1]{fontenc}
-\usepackage[pdftex]{graphicx}
+\usepackage{graphicx}
 \usepackage{bm}
-\usepackage{savesym}
 \usepackage{amsmath}
@@
 \usepackage{algpseudocode}
-\usepackage{enumitem}
-\usepackage{relsize}
 \usepackage{caption}
@@
 \usepackage{layouts}
-\usepackage{afterpage}
+\usepackage{cleveref}
 \usepackage{txfonts}
-\usepackage{pgf}
-\usepackage{cleveref}
+\usepackage[T1]{fontenc}
 
@@
 \newcommand{\Like}{\mathcal{L}}
-\newcommand{\DKL}{D_\mathrm{KL}}
+\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
 
@@
 \section{Introduction}
-Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}.
-It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{split}.
-The core algorithm is unique for being the only algorithm which estimates volumes using \textit{order statistics}, which makes high-dimensional integration feasible.
-It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
+Users would like a order of magnitude runtime estimate
+Figure of PolyChord output
+
+\section{Background}
+\subsection{Notation}
+We shall begin with a brief description of nested sampling to establish the necessary notation.
+For a given likelihood $\Like(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
+\begin{equation}\label{eq:evidence}
+	\mathcal{Z} = \int \Like(\theta)\ \pi(\theta)\ \mathrm{d}\theta
+\end{equation}
+while producing samples of the posterior distribution 
+\begin{equation}
+	\mathcal{P}(\theta) = \frac{\Like(\theta) \pi(\theta)}{\mathcal{Z}}.
+\end{equation}
+The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run.
+At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}.
+A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point.
+Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
 \par
-Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord} which provide flexible options for nested sampling runs, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to obtain and plot sample statistics.
-However, currently no such software is able to estimate how long a given run should last.
-This is of high importance to the end user, who has no indication in the middle of a run whether it will take minutes, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
+The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$, defined as
+\begin{equation}
+	\mathcal{Z} = \int_0^1 \Like(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \Like_i (X_{i-1}-X_{i+1}),
+\end{equation}
+where $X(\Like)$ is the fraction of the prior greater than $\Like$.
+The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
+\begin{equation}
+	X_i = \prod_{k=1}^i t_k.
+\end{equation}
+The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
+\begin{equation}\label{eq:t_dist}
+	p(t_i) = \nlive t_i^{\nlive-1}, \quad \mathrm{E}[\log t_i] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t_i] = \frac{1}{\nlive^2}.
+\end{equation}
+The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g.
+$10^{-3}$.
+The remaining live points are then killed off one by one without replacement and added to the evidence.
 \par
-This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage.
-The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples.
-We begin in section \ref{sec:Background} with an overview of Bayesian inference and nested sampling.
-Section \ref{sec:Endpoint prediction} presents the theory and methodology of making endpoint predictions, before results for both toy and real examples are shown in section \ref{sec:Results}.
-Finally, conclusions are made in section \ref{sec:Conclusions}.
+Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.
 
-\section{Background}\label{sec:Background}
-\printinunitsof{in}\prntlen{\textwidth}
-We present here an overview of Bayesian inference and nested sampling; for a more detailed treatment of both subjects, we refer to \citet{sivia} as well as Skilling's original paper. 
-\subsection{Bayesian inference}%
-\label{sub:Bayesian inference}
+\section{Convergence of a nested sampling run}
+Let us now take a top-down view of a nested sampling run as it progresses to observe how it converges.
 
-Given some data $D$, Bayes' theorem tells us that the probability of some parameters $\theta$ of a model $M$ is 
+\subsection{Posterior convergence}
+We will examine how our set of samples evolve over the course of a nested sampling run.
+The conventional perspective for the progress of a run is with $\log X$ along the x-axis; the prior volume is compressed at a steady rate, only picking up the bulk of the integral/posterior at around $\log X = \DKL$.
+\par
+At an intermediate iteration, we have an accurate reconstruction of the posterior only down to the current prior volume $\log X_*$.
+The live points contain information below this volume, but not much more.
+If we kill off all the live points as if we were terminating, then by \cref{eq:t_dist} the prior volume of the final live point $\log X_\mathrm{min}^{\mathrm{live}}$ has mean and variance
 \begin{equation}
-	\Pr(\theta\mid D, M) = \frac{\Pr(D \mid \theta, M) \Pr(\theta \mid M)}{\Pr(D  \mid  M)}
+	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
 \end{equation}
-Relabelling these terms,
 \begin{equation}
-    P(\theta) = \frac{L(\theta) \pi(\theta)}{Z}
+	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6},
 \end{equation}
-where $\pi(\theta)$ is the \textit{prior} - what was known about a model's parameters before knowledge of the data, $P(\theta)$ is the \textit{posterior} - what is known after learning the data, and $L(\theta)$ is the \textit{likelihood} of the data given the parameters.
+where the large $\nlive$ limit is taken for the approximation and $\gamma$ is the Euler-Mascheroni constant. 
 \par
-For parameter estimation, knowledge of $L(\theta)$ and $\pi(\theta)$ is enough, but $Z$, known as the \textit{evidence}, is crucial for model comparison via Bayes factors.
-By definition, $Z$ is the normalisation factor
-\begin{equation}
-    Z = \int L(\theta)\pi(\theta) \ d\theta
-\end{equation}
-which is a high-dimensional integral over the parameter space.
-Exponential scaling with dimension means full rasterisation is impractical for most physical applications.
-This motivates a sampling technique which preferentially selects for regions of high posterior mass.
-Traditional techniques like Markov-Chain Monte Carlo (MCMC) do not sample from the prior, so it is non-trivial to estimate the evidence from posterior samples.
-Nested sampling sidesteps these issues by putting the evidence first, while simultaneously yielding the posterior.
-
-\subsection{Nested sampling}
-\label{sub:Nested sampling}
-
-We first perform a change of variable to reduce the evidence integral to a single dimension. Define the prior mass 
-\begin{equation}\label{X integral}
-    X(\lambda) = \int_{L(\theta)>\lambda} \pi(\theta) \ d\theta 
-\end{equation}
-which allows us to rewrite the evidence as 
-\begin{equation}
-    Z = \int_0^1 \Like(X) \ dX.
-\end{equation}
-The integral can be calculated by ordering the prior samples by likelihood then multiplying by the associated prior mass element.
-Nested sampling translates this into an algorithm which incorporates sorting \textit{de facto}, by taking a set of $\nlive$ \textit{live points} and replacing the point with the lowest likelihood with a resampled point of higher likelihood at every iteration.
-A choice must be made for this sampling method e.g.
-MCMC.
-The number of live points can in principle vary throughout the run, so that it depends on the iteration $k$ i.e.
-$n_\mathrm{live}=n_k$ \citep{dynamic_ns}.
+This shows that the live points only get us a factor of $\log \nlive$ closer to the posterior bulk, or in other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples begin to resemble the posterior.
+Intuitively, if the likelihood is sharply peaked then for most of the run the live points are too diffuse to land there with any significant probability.
 \par
-The set of live points will therefore occupy a prior volume which tends to zero, shrinking around the maximum of the posterior.
-Explicit calculation of the volume $X$ within each contour is expensive, so nested sampling makes a statistical estimate using the order statistics of the remaining points: at each iteration, the volume occupied by the live points decreases by the shrinkage ratio $t$, which has probability distribution and statistics
-\begin{equation}
-	\Pr(t) = \nlive t^{\nlive - 1}, \quad \mathrm{E}[\log t] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t] = \frac{1}{\nlive^2}.
-\end{equation}
-$X_i$ is the product of the shrinkage ratios
-\begin{equation}\label{xi_formula}
-    X_i = \prod_{k=0}^{i} t_k,
-\end{equation}
-so the $X$ themselves are also random variables. It is easier to work with $\log X$, which have the means and correlations
-\begin{equation}\label{X_distribution}
-	\mathrm{E}[\log X_i] = -\sum_{k=0}^{i} \frac{1}{n_k}, \quad
-	\mathrm{Cov}[\log X_i, \log X_j] = \sum_{k=0}^{\mathrm{min}(i, j)} \frac{1}{n_k^2}.
-\end{equation}
-
-The evidence integral is then computed using the trapezoid rule at each step, so that
-\begin{equation}
-	Z = \sum_i L_i \Delta w_i = \frac{1}{2}\sum_i L_i \left( X_{i-1} - X_{i+1} \right).
-\end{equation}
-
-The algorithm terminates when the evidence in the live points falls below a user-specified fraction $\epsilon$ of the total accumulated evidence.
-Skilling's original paper estimates this remaining evidence $\Delta Z$ to have an upper bound of $ \Like_\mathrm{max} X_i$, while some use the average remaining likelihood.
-Once we have chosen to stop, the remaining evidence is included by killing the live points one by one without replacement, incrementing $Z$ as before.
-\par
-The full nested sampling procedure is given in algorithm \ref{alg:ns}. 
-\begin{algorithm}
-\caption{Nested sampling}\label{alg:ns}
-\begin{algorithmic}
-\State Start with $N$ points $\theta_1, \ldots, \theta_N$ from prior
-\State \hspace{\algorithmicindent} initialise $Z = 0$, $X_0 = 1$
-\For{$i = 1, 2, \ldots , j$}
-\State $L_i \gets \text{lowest of the current likelihood values}$
-\State $X_i \gets \exp(-i/N)$
-\State $w_i \gets X_{i-1} - X_i$ or $(X_{i-1} - X_{i+1})/2$
-\State $Z \gets Z + L_i w_i$
-\State $L_i \gets $ new point $\in \{\pi(\theta): L(\theta) > L_i\}$.
-\EndFor
-\State Increment $Z$ by $\langle L(\theta) \rangle X_j$.
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Uncertainties and the posterior bulk}\label{subsec:Uncertainties and the posterior bulk}
 \begin{figure}
 \begin{center}
-	\includegraphics{Figures/LX.pdf}
+	\includegraphics{figures/live_point_distribution.pdf}
 \end{center}
-\caption{Plot showing the likelihood and the posterior weights $\Like(X)X$.
-The bulk of posterior mass is contained within a region of size $e^{-\DKL}$, meaning one must iterate through at least $n\DKL$ steps to fully explore the posterior.}
-\label{fig:LX}
+\caption{}
+\label{fig:}
 \end{figure}
 
-\begin{figure}
-\begin{center}
-	\includegraphics{Figures/X_samples.pdf}
-\end{center}
-\caption{Plot showing samples of $\log X$ plotted against the likelihoods of the sample points. The uncertainty increases with the number of steps taken.}
-\label{fig:X_samples}
-\end{figure}
-
-Most of the posterior is contained within a small fraction of the prior of size $X^{*} = e^{-\DKL}$, where
-\begin{equation}
-\DKL = \int P(\theta) \log  \frac{P(\theta)}{\pi(\theta)} \ d\theta
+\subsection{Loglikelihood convergence}
+We now offer an alternate view of the convergence of a nested sampling run, via the loglikelihood instead of the prior volumes.
+To get an insight into the analytics, work with a Gaussian likelihood of $d$ dimensions and a lengthscale of $\sigma$;
+\begin{equation}\label{eq:logL}
+	\log\Like = \log\Like_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
 \end{equation}
-is the information gain between the prior and posterior; this is visualised in figure \ref{fig:LX}.
-To learn anything about the posterior or calculate an accurate evidence, we must compress down to this volume.
-\par
-From equation \eqref{X_distribution}, it is easy to see that for a constant number of live points $n$ $\log X$ follows Poisson statistics:
-\begin{equation}
-	\mathrm{E}[\log X_i] = - \frac{i}{n}, \quad \mathrm{Var}[\log X_i] = \frac{i}{n^2}.
-\end{equation}
-This is illustrated in figure \ref{fig:X_samples}, where samples are taken from the $X$ distribution and plotted against the likelihood values.
-To reach the posterior bulk we must therefore take $i^{*} = n\DKL$ steps, which translates to an uncertainty in $\log X^{*} = -\DKL \pm \sqrt{\DKL/n}$.
-The evidence is $Z = \langle \Like \rangle X^{*}$, or
-\begin{equation}
-\log Z = \langle \log \Like \rangle + \log X^{*}.
-\end{equation}
-The first term has low error, since the only difference that errors in $X$ makes is to slightly change the region we average over.
-Uncertainties in $\log Z$ are therefore dominated by the $\mathcal{O}(1/\sqrt{n})$ errors on $\log X_i$, which are much more significant than the $\mathcal{O}(n^{-2})$ quadrature errors.
-\par
-The next section will outline the methodology for making endpoint predictions.
-
-\section{Endpoint prediction}\label{sec:Endpoint prediction}
-We shall begin by discussing a simple ``eyeball" method for making predictions that is familiar to experienced users of nested sampling, which makes a good sanity check to compare to the more principled approach.
-\subsection{Eyeball method}
-At an intermediate iteration of a nested sampling run, the output might look something like that shown in the first two columns of table \ref{tab:ns_output}. 
-\begin{figure}
-\centering
-\scalebox{0.9}{\begin{tabular}{|c|c|c|}
-\hline
-iteration & log Z  & $d\log Z$   \\
-\hline
-6000 & -34.481 & 1.317 \\ 
-6200 & -32.606 & 1.875 \\ 
-6400 & -30.700 & 1.906 \\ 
-6600 & -29.008 & 1.692 \\ 
-6800 & -27.416 & 1.591 \\ 
-7000 & -25.865 & 1.551 \\ 
-7200 & -24.449 & 1.416 \\ 
-7400 & -23.034 & 1.415 \\ 
-7600 & -21.617 & 1.417 \\ 
-7800 & -20.267 & 1.349 \\ 
-8000 & -18.950 & 1.317 \\
-\hline
-\end{tabular}}
-\captionof{table}{Output of accumulated log-evidence during a nested sampling run.}
-\label{tab:ns_output}
-\end{figure}
-\begin{figure}
-  \centering
-  \includegraphics{Figures/linear_extrapolate.pdf}
-  \captionof{figure}{Prediction of endpoint using linear extrapolation. The estimate is much lower than the true endpoint, primarily because of the long tail in $d\log Z$.}
-  \label{fig:linear_extrapolate}
-\end{figure}
-The outputs $\log Z$ appear to converge at a predictable rate, which is shown in terms the increments in column three.
-One could feasibly extrapolate these to predict when the increments approach zero, which gives an iteration number for convergence.
-\par
-This is shown for a linear extrapolation in figure \ref{fig:linear_extrapolate}.
-The prediction is clearly an underestimate, which is mainly due to the nonlinear profile of $d\log Z$ having a long tail.
-However, this makes a good benchmark to beat for the more principled method using likelihood approximation.
-
-\subsection{Likelihood approximation method}
-We want to make an estimate of the number of iterations until the termination condition is met, which is when the remaining evidence falls below a fraction $\epsilon$ of the accumulated evidence.
-Predictions can be made ahead of time because of the live points, which are sampled from higher likelihood regions that we have not yet compressed to.
-It is important to note that there will always be a small region which the live points do not penetrate because they are too diffuse, which can be seen in figure \ref{fig:dead_live_points}.
-\begin{figure}
-\begin{center}
-	\includegraphics{Figures/dead_live_points.pdf}
-\end{center}
-\caption{Plots showing the state of a nested sampling at an intermediate iteration.
-The live points are samples from higher likelihood regions, which allow us to see the remaining likelihood ahead of time.
-However, there is always a small high likelihood region where we have no samples, simply because the live points are too diffuse.}
-\label{fig:dead_live_points}
-\end{figure}
-\par
-If we could estimate the form of $\Like(X)$ at the higher likelihoods in terms of some \textit{likelihood model} $f(X, \thetab)$, we would be able to express the prior volume $X_f$ which we need to compress to as
-\begin{equation}
-	\Delta Z = \epsilon Z_\mathrm{tot},
-\end{equation}
-\begin{equation}\label{model endpoint}
-	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
-\end{equation}
-where $X_i$ is the volume of the iteration we have currently compressed to, and $Z_\mathrm{dead}$ is the evidence we have accumulated up to this point.
-$X_f$ can then be identified by solving the above equation either analytically or numerically.
-Remembering that at each iteration $\log X$ decreases by $1/\nlive$, the total number of iterations $N_f$ will be
-\begin{equation}
-	N_f = - \nlive \log X_f .
-\end{equation}
-
-\subsubsection{Model choice}\label{sec:Model choice}
-The choice of the model $f(X, \thetab)$ serves two purposes; to \textit{smooth} over the stochasticity in the live points, since $\Like(X)$ is typically a smooth function, and more importantly to \textit{extrapolate} into the region where we have no samples.
-\par
-Making this extrapolation necessitates assumptions about the form of the likelihood in the unsampled region.
-It is sensible to assume it is guided by what came before it, so we fit the model on the known likelihods of the live and dead points.
-Apart from this, we should make the model $f(X, \thetab)$ as flexible as possible to extrapolate without bias.
-We chose the model to be a Gaussian with freely varying parameters $\thetab = (\Like_\mathrm{max}, d, \sigma)$:
-\begin{equation}
-	f(X, \thetab) = \Like_\mathrm{max} \exp\left(-\frac{X^{2/d}}{2\sigma^2}\right).
-\end{equation}
-The choice was made because it is analytically tractable, and because many likelihoods in practice are near-Gaussian at their peaks.
-Allowing $d$ to vary freely increases flexibility, and reflects the fact that while the dimensionality of the posterior is fixed, the equivalent dimension for a Gaussian of the same information content can differ \citep{Handley_2019}.
-
-\subsubsection{Endpoint formula}
-Now we are equipped with the tools to make endpoint predictions.
-We will begin by fitting the model over the live points only.
-At each iteration, estimate the prior volumes of the live points to be those which correspond to killing off each point one by one without replacement and using formula \eqref{X_distribution}.
-Combined with the live likelihoods $\logL$, the parameters of the model can be found by minimising the cost function
-\begin{equation}\label{chi squared}
-	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
-\end{equation}
-with respect to $\thetab$.
-A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
-\begin{equation}
-    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i \log \Like_i X_i^{2/d}}
-\end{equation}
-\begin{equation}
-    \logLm = \frac{1}{N} \sum_i \logL + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d} 
-\end{equation}
-(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables.
-The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run.
-We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
-\par
-Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \eqref{model endpoint};
-\begin{equation}
-	\epsilon = \frac{\int_0^{X_f} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX + Z_\mathrm{dead}}.
-\end{equation}
-The integrals have the analytic solution
-\begin{equation}
-	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ dX = \frac{d}{2} \cdot 2^{d/2} \cdot \sigma^{d} \cdot \gamma_k
-\end{equation}
-where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
-\begin{multline}
-	\log X_f = \frac{d}{2}\log 2 + d \log \sigma \\
-	+ \log \gamma^{-1} \left( d/2 ,\ \epsilon \left(\gamma_i+ 2^{-d/2} \sigma^{-d} Z_\mathrm{dead}/\Like_\mathrm{max}\right)\right),
-\end{multline}
-and $N_f$ is just  $-\nlive$ multiplied by this. 
-\par
-Noting that the prediction depends on the \textit{ratio} $Z_\mathrm{dead}/\Like_\mathrm{max}$, an useful feature is that the formula is robust when we are far away from the posterior bulk.
-In this phase, the inferred values of $\Like_\mathrm{max}$ turn out to be very uncertain and often divergent, but the dead evidence is also negligible, so $\Like_\mathrm{max}$ has little impact on the prediction.
-Once $Z_\mathrm{dead}$ becomes comparable to the live evidence $\Like_\mathrm{max}$ does begin to impact the prediction, but at this point we have reached the posterior bulk, so we expect $\Like_\mathrm{max}$ to be well-estimated and thus contribute in a useful way.
-\par
-Note that the prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.
-
-\subsubsection{Uncertainties}
-Our lack of knowledge about the unsampled region means that there is always the chance that there is some huge likelihood spike at small $X$.
-Since there is no way to anticipate this, all uncertainty estimates are underestimates in the sense that they cannot account for this possibility; there is no estimate which will \textit{always} include the true endpoint.
-\par
-We will therefore restrict ourselves to quantifying the uncertainties associated with how we have modelled the \textit{known} data, remembering that because of our lack of knowledge, the true endpoint may fall significantly outside of this range.
-These uncertainties have two sources: the variance in the prior volume estimates, and the uncertainty in the least squares minimisation.
-Since the former is the main error in nested sampling, we expect it to dominate.
-As we saw in section \ref{subsec:Uncertainties and the posterior bulk}, different sets of $\set{X_i}$ drawn from its probability distribution lead to different "maps" of what the remaining likelihood looks like, which lead to different endpoint predictions.
-The natural way to handle this is to sample from $\Pr(\bm{X})$ then find the best-fit  $\thetab$ and endpoint for each sample, from which we can obtain the mean and variance in the estimates.
-An visualisation of this approach is shown in figure \ref{fig:gauss_overfit}.
-
-\begin{figure}
-\begin{center}
-	\includegraphics{Figures/gauss_overfit.pdf}
-\end{center}
-\caption{Plot showing fits of the likelihood corresponding to different draws of $\log X$.
-The different realisations of the likelihood lead to a set of endpoint predictions, from which a mean and variance can be obtained.}
-\label{fig:gauss_overfit}
-\end{figure}
-
-\section{Results}\label{sec:Results}
-\subsection{Toy data}
-Let us now apply the methodology developed above in practice. We consider the following toy examples:
-\begin{enumerate}[leftmargin=0.3in, label=(\alph*)]
-	\item A spherically symmetric Gaussian, $\log \Like = -|\bm{r}|^2/2\sigma^2$ in $d$ dimensions
-	\item A discontinous "wedding cake" likelihood with Gaussian profile and parameter $\alpha_\mathrm{w}$ controlling the plateau depth, as presented in \cite{plateaus} and shown in figure \ref{fig:wedding_3d}.
-	\item A Cauchy likelihood $\log \Like = - \frac{1 + d}{2} \log (1 + |\bm{r}|^2/\gamma^2)$.
-\end{enumerate}
-\begin{figure}
-\begin{center}
-	\includegraphics{Figures/wedding_3d.pdf}
-\end{center}
-\caption{An example ``wedding cake" likelihood which has plateaus on top of a Gaussian profile}
-\label{fig:wedding_3d}
-\end{figure}
-
-Parameters were chosen to have large values of $\DKL$, which are more difficult to model due to the large distance to the posterior bulk.
-\par
-Nested sampling runs were made for the above likelihoods for $\nlive = 500$ (plots for different numbers of live points are not shown here, but the results are broadly the same).
-The state at each iteration (i.e.
-the likelihoods of the dead and live points) was then extracted using \textsc{anesthetic}.
-At each iteration, the model is fitted over a number of points determined by the bandwidth selection procedure outlined above.
-For the error bars, we draw an \textit{ensemble} of 25 $\bm{X}$ samples from the distribution \eqref{xi_formula}, and find the optimal $\thetab$ and corresponding  $N_f$ for each sample to get the mean and variance of the prediction.
-\par
-Results are shown in figure \ref{fig:toy_predictions}.
-For all cases, the predictions become more accurate as the iteration increases because the samples shrink around and eventually cross the posterior bulk, leading to an increasingly representative picture of the posterior.
-The method works very well for the first two likelihoods, recovering the correct endpoint within standard error for the entire run.
-This is expected, since the model likelihood has the same form as the true likelihoods, so one expects the extrapolation made to be accurate even when we are far away from the posterior bulk.
-\par
-The Cauchy likelihood shows the result when this is not the case; without samples in the bulk, a Gaussian extrapolation performs poorly.
-However, the correct order of magnitude is still obtained, which is a significant improvement on the status quo of complete ignorance.
-\begin{figure*}
-\begin{center}
-	\subfloat[Gaussian: $d = 30$, $\sigma = 0.01$, $\DKL = 84$]{\includegraphics{Figures/gauss_predictions.pdf}}
-	\subfloat[\centering Wedding cake: $d = 20$, $\sigma = 0.001$, $\alpha_\mathrm{w}=0.5$, $\DKL = 93$]{\includegraphics{Figures/wedding_predictions.pdf}}
-	\subfloat[Cauchy: $d = 10$, $\gamma = 0.0001$, $\DKL = 48$]{\includegraphics{Figures/cauchy_predictions.pdf}}
-\end{center}
-\caption{Plot showing endpoint predictions for the toy likelihoods.
-The mean estimate is dark blue, while the shading shows the $1\sigma$ and $2\sigma$ uncertainties.
-For the first two likelihoods (which have Gaussian profile), the true endpoint falls within error for practically all of the run.
-The Cauchy likelihood is less well-estimated because it is highly non-Gaussian, with the true endpoint outside the reasonable uncertainty range but in the correct order of magnitude.}
-\label{fig:toy_predictions}
-\end{figure*}
-
-\subsection{Real data}
-The model was tested on real nested sampling runs for quantification of cosmological parameter tensions.
-Results are shown in figure \ref{fig:lcdm_predictions}, with the likelihood approximation method compared against the benchmark.
-\par
-The likelihood model clearly performs better.
-For all of the chains, the true endpoint was recovered within arund standard error by the halfway point, and falls within the correct order of magnitude throughout all of the runs.
-In comparison, the benchmark model mostly failed to get the correct order of magnitude on the number of iterations remaining, especially near the end of each run.
-\par
-Furthermore, the first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian likelihood model.
-For all chains, further work needs to be done to place a more accurate upper bound on the prediction uncertainty at early iterations.
-\begin{figure*}
-\begin{center}
-	\includegraphics{Figures/lcdm_predictions.pdf}
-\end{center}
-\caption{Plots showing the endpoint predictions for the benchmark model as well as the likelihood approximation model.
-The likelihood model visibly performs better, managing to recover the true endpoint within around standard error at the halfway point for all the chains.
-The first four likelihoods are highly non-Gaussian, which validates the choice of a Gaussian model.}
-\label{fig:lcdm_predictions}
-\end{figure*}
-
-\section{The anatomy of a Gaussian}
-
-In general, it is well known (to those who know such things) that the anatomy of a Gaussian says that
-\begin{equation}
-    \boxed{\log\mathcal{L}_\mathrm{max} = \langle\log\mathcal{L}\rangle_\mathcal{P} + \frac{d}{2}},  \qquad \boxed{\mathrm{var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}},
-\end{equation}
-namely the posterior average of the log likelihood is $d/2$ less than the maximum, and the variance of the loglikelihood over the posterior is $\frac{d}{2}$.
-Whilst this is exact for gaussians, this provides a good rule of thumb for other distributions.
-
-For example, one may use this to explain why MCMC sampling is a very bad way to maximise, since even in the limit of a very large number of samples from the posterior $n$, the largest of these will be within a few $\sqrt{\frac{d}{2}}$ standard deviations of the mean, and therefore very far away from the maximum $\frac{d}{2}$ away.
-
-We can get someway to the analytics for MCMC. Similar arguments to below show that the distribution of the largest likelihood point found from a posterior is:
-\begin{equation}
-    P(x) = n \frac{e^{-x} x^{\frac{d}{2}-1} }{\Gamma\left(\frac{d}{2}\right)} {\left( \frac{\Gamma_{\frac{d}{2}}(x)}{\Gamma(\frac{d}{2})} \right)}^{n-1},
-\end{equation}
-where $x=2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{max}^\mathrm{posterior})$
-since if we have the CDF of a distribution as $F(x)$, then the lowest of n random variables is $\frac{d}{dx}( 1- (1-F(x))^n)$, and we the upper CDF for a $\chi^2_{d/2}$ distribution is $\Gamma_{\frac{d}{2}}(x)/\Gamma(\frac{d}{2})$.
-I have not yet been able to find an analytic expression/approximation for the center and width of this distribution, but Jensen's inequality with a Gaussian approximation\footnote{\url{https://math.stackexchange.com/a/89147}} yields
-\begin{equation}
-    \log\mathcal{L}_\mathrm{max}^\mathrm{posterior} \le  \log\mathcal{L}_\mathrm{max} - \frac{d}{2} + \sqrt{d\log n} 
-    %\log\mathcal{L}_\mathrm{max}^\mathrm{posterior} \leq \log\mathcal{L}_\mathrm{max} - \frac{d}{2} + \sqrt{d\log n}.
-\end{equation}
-
-The same is true for nested sampling, although since it is not a posterior sampler, but a likelihood upward scanner, the argument is a little more subtle, which we work through below.
-
-We will work with a Gaussian loglikelihood, lengthscale $\sigma$, defined in terms of prior volumes $0<X<1$
-\begin{equation}
-    \log\mathcal{L}
-    =\log\mathcal{L}_\mathrm{max} - \frac{1}{2\sigma^2}X^{2/d}
-    =\log\mathcal{L}_\mathrm{max} - \frac{1}{2\sigma^2}\exp\left(\frac{2}{d}\log X\right).
-    \label{eq:LX}
-\end{equation}
-We will assume the run has progressed far enough so that prior effects may be neglected, and that $\sigma\ll 1$ is sufficiently small that for the arguments below edge effects may be neglected.
-
-Note that from the Gaussian profile we can see that as $\log X\to-\infty$ at a steady rate of compression, we approach the maximum loglikelihood exponentially slowly, with an decay rate of $\frac{d}{2}$
-
-Taking $X \sim e^{-i/n}$ we see this as
-\begin{equation}
-    \log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \exp\left(-\frac{i-i_0}{nd/2}\right), \qquad i_0 = -\frac{nd}{2}\log 2\sigma^2
-\end{equation}
-in principle this converges very quickly, although never quite reaches the maximum.
-
-\section*{The anatomy of a gaussian}
+The distribution of the true posterior in $\log\Like$ is
 \begin{equation}
@@
 \end{equation}
-i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})\sim \chi^2_{d} $
-
-\section*{Observation 1: the largest likelihood live point}
-Assume we are at an iteration such that $\mathcal{L}_*$ is the current likelihood contour (i.e.
-the likelihood of the dead point we have just deleted), and that we have $n$ live pointsdistributed uniformly in prior volume beneath this.
-Since the the live points are distributed in prior volume up to $X_*$, $P(X) = [0<X<X_*]$,
-Inverting \cref{eq:LX} we may show that the distribution of loglikelihoods is
+i.e.
+$2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the same distribution as that of the dead points up to the latest dead likelihood $\log \Like_*$.
+Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
 \begin{equation}
-    P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}}[0<\log\mathcal{L}<\log\mathcal{L}_*].
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\Like_* < \log\Like <\log\Like_\mathrm{max}],
     \label{eq:PL}
 \end{equation}
-It is helpful at this stage therefore to define a parameter
+It is helpful at this stage to define a parameter
 \begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*} \qquad\Leftrightarrow\qquad \log\mathcal{L} = y\log\mathcal{L}_\mathrm{max} + (1-y)\log\mathcal{L}_*
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*} 
     \label{eq:y}
 \end{equation}
-as a normalised likelihood with $y=0$ corresponding to $\mathcal{L}=\mathcal{L}_*$ and $y=1$ to $\mathcal{L}=\mathcal{L}_\mathrm{max}$, which has distribution
+as a normalised measure of how far a point is between the latest dead point and the maximum loglikelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
 \begin{equation}
-    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1}[0<y<1].
+    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
     \label{eq:Py}
 \end{equation}
-We now seek the distribution of the largest likelihood $\log{\mathcal{L}}_\mathrm{max}^\mathrm{live}$ (or equivalently the largest ${y}_\mathrm{max}^\mathrm{live}$) in the set of $n$ live points.
-We first integrate to find the CDF
-\begin{equation}
-    P(y<Y) = 1-(1-Y)^{\frac{d}{2}}[0<Y<1].
-    \label{eq:Py}
-\end{equation}
-note that to find the largest of $n$ independent variables we simply multiply 
-\begin{equation}
-    P(y_1,y_2,\ldots<Y) = (1-(1-Y)^{\frac{d}{2}})^n[0<Y<1].
-    \label{eq:Py}
-\end{equation}
-and then differentiate with respect to $Y$ to find the PDF for largest likelihood in the live points
+We now seek the distribution for the maximum likelihood of the live points, $\log\Like_\mathrm{max}^{\mathrm{live}}$.
+Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
 \begin{equation}
@@
 \end{equation}
-which may be roughly summarised as:
-\begin{equation}
-    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
-    \label{eq:ymax}
+which in the limit of large live points and dimensions can be roughly summarised by
+\begin{equation}\label{eq:ylivemax}
+	    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
 \end{equation}
-or in the large $d$, $n$ limit
-\begin{align}
-    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
-    \label{eq:ymaxd}\\
-    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
-    \label{eq:ymaxdn}
-\end{align}
-where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
-
 This shows that in general the live points are nowhere near the maximum loglikelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
 In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
-Our end result is that:
-\begin{equation}
-    \boxed{
-        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - (1-y_\mathrm{max}^\mathrm{live})(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)
-    },
-\end{equation}
-where you may take any of the $y_\mathrm{max}^\mathrm{live}$ expressions above from \cref{eq:ymax,eq:ymaxd,eq:ymaxdn}.
-
-We may also rewrite the distribution \cref{eq:Pyhat} without $y$ to give
-\begin{equation}
-    \frac{\log\mathcal{L}_\mathrm{max}^\mathrm{live}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
-\end{equation}
-\begin{equation}
-    P(\log\mathcal{L}_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left((\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}-(\log\mathcal{L}_\mathrm{max} - \log\mathcal{L}_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
-\end{equation}
-
-\subsection*{Observation 2: the termination contour $\mathcal{L}_\mathrm{end}$}
-
-One strategy for terminating nested sampling is to halt when the amount of evidence in the live points is some fraction $f$ of the amount of total evidence (or evidence in the dead points).
-Providing $f\ll 1$, then this is approximately the same as the fraction of evidence left above likelihood contour $\mathcal{L}_\mathrm{end}$:
+\par
+This remains true even at the end of the nested sampling run. To see this, we write the halting condition as:
 \begin{equation}
@@
 \end{equation}
-Note that we have assumed that prior effects are negligible (so $1=\infty$).
-Computing this for \cref{eq:LX} we find the answer in terms of lower incomplete gamma functions
+Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $f \ll 1$ so that the denominator is approximately the accumulated evidence.
+Computing this for \cref{eq:logL} we find the answer in terms of lower incomplete gamma functions
 \begin{equation}
-    f = 1- \frac{\Gamma_{d/2}(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2})}{\Gamma(\frac{d}{2})}.
+    f = 1- \frac{\Gamma_{d/2}\left(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2}\right)}{\Gamma\left(\frac{d}{2}\right)}.
     \label{eq:f}
@@
 We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $f$.
-This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a stirling/Stirling approximation:
+This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a Stirling approximation:
 \begin{equation}
@@
 \subsection*{Observation 3: nested sampling as a maximiser}
-
-Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with the large $d,n$ limit of \cref{eq:ymaxdn} we find
+Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
 \begin{equation}
@@
 \end{equation}
-showing that in general the anatomy that nested sampling will finish at a contour $\frac{d}{2e}$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\frac{\log n}{2e}$ closer, with a chance of getting $\sim\frac{\pi}{\sqrt{6}e}=0.472$ closer still by statistical fluctuation.
-
+showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\log (n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
 \begin{figure*}
 \begin{center}
-	\includegraphics{Figures/anatomy.pdf}
+	\includegraphics{figures/anatomy.pdf}
 \end{center}
@@
 \end{figure*}
+\par
+From above analysis we learn that
+\begin{itemize}
+	\item Knowledge of the posterior early on in the run is limited by the maximum live point, which is far from the posterior bulk
+	\item At termination, the maximum live point is far from the true maximum --> nested sampling alone is a poor optimiser (but great for finding starting point of gradient descent)
+\end{itemize}
 
-\section{Conclusions}\label{sec:Conclusions}
-In this project we have developed a model to estimate the endpoint of a nested sampling run, which has no precedent in the literature.
-This was done by predicting the form of the likelihood in the region that the samples have not reached, via fitting of a Gaussian model likelihood to the points already known to us.
-Assuming that the likelihood continues in a similar trajectory leads to an estimate of the remaining evidence, and hence the end iteration.
+\section{Method}
+Let us now tackle the problem of estimating how many steps we still have to take in a nested sampling run before we can terminate.
 \par
-The approach taken was shown to be highly successful for near-Gaussian likelihoods, finding the true endpoint within error from the start of the run; this was expected, because the form of the model matched the likelihood.
-For a Cauchy likelihood as well as the real nested sampling runs that were tested, the estimates were less accurate, but still able to find the correct order of magnitude which is useful to the end user.
-Furthermore, the correct endpoint was able to be recovered within standard error at the halfway point of the runs.
+The above analysis shows that our knowledge of the posterior at an intermediate iteration is limited by the maximum live point, which until just before the end is quite far from the posterior bulk.
+However, we can get an approximate image of the full posterior ahead of time, simply by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points.
 \par
-Two main refinements are recommended: the uncertainty quantification should be expanded to include the true endpoint from the beginning of the run, and the analysis should be extended to include the iteration prediction when the number of live points is allowed to vary, so that endpoints can be estimated for dynamic nested sampling.
+For practical purposes, one would never use this approximate posterior to do inference.
+However, it is more than sufficient for making a prediction for roughly how much evidence remains, and hence the point at which the run should terminate.
+Quantitatively, this proceeds as follows: fitting a model function $f(X, \thetab)$ to the known likelihood profile allows us to express the prior volume we need to compress to as
+\begin{equation}
+	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
+\end{equation}
+\begin{equation}\label{model endpoint}
+	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
+\end{equation}
+where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point.
+$X_f$ can then be identified by solving the above equation either analytically or numerically.
+If $\nlive$ is constant then at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_f$ will be
+\begin{equation}
+	N_f = - \nlive \log X_f .
+\end{equation}
+One has freedom to choose the form of the fitting function $f(X, \thetab)$.
+Following the previous analysis, we use a Gaussian profile, which has the advantage that the regression itself can be sped up analytically, as well as the fact that many real likelihoods are Gaussian or near-Gaussian.
+However, any other function that is sufficiently flexible to capture the shape of the likelihood profile will do.
 
-\section*{Data availability}
-The data and code for the project can be found in the repository \url{https://github.com/zixiao-h/aeons}.
+Fit a Gaussian profile, with advantages that a) Gaussian error common b) analytic optimisation 
+Extraction of termination logX
+Prediction uncertainty dominated by correlations, not least squares i.e. scatter of individual points - get by simulating X
 
-\bibliographystyle{mnras}
-\bibliography{references}
+\section{Results}
+Toy examples - pure Gaussian,  wedding cake, Cauchy
+Real cosmological examples - Planck
 
-\label{lastpage}
 \end{document}

### 2023-08-07 paper/figures/live_point_distribution.py

@@ -1,0 +1,32 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+from aeons.utils import *
+from aeons.likelihoods import full
+
+theta = [0, 10, 0.01]
+logX = np.linspace(-60, 0, 1000)
+X = np.exp(logX)
+logL = full.func(X, theta)
+L = np.exp(logL - logL.max())
+LX = L * X
+
+ndead = 5000
+nlive = 500
+
+def gaussian(x, mu, sig):
+    return 1/(np.sqrt(2*np.pi)*sig)*np.exp(-np.power((x - mu)/sig, 2)/2)
+ElogXlivemin = -ndead/nlive - np.log(nlive) - 0.577
+StdlogXlivemin = np.sqrt(ndead/nlive**2 + (np.pi**2)/6)
+live_logX = logX[(logX > ElogXlivemin) * (logX < -ndead/nlive)]
+
+figsettings()
+plt.figure(figsize=(4,1.5))
+plt.plot(logX, L/L.max(), color='black', label='Likelihood', lw=.5)
+plt.fill(logX, LX/LX.max(), alpha=0.4, color='gray', label='Posterior')
+plt.fill_between(live_logX, np.zeros(len(live_logX)), np.ones_like(live_logX)/(np.log(nlive) + 0.577), alpha=0.5, color='deepskyblue', label='Live points')
+plt.plot(logX, gaussian(logX, ElogXlivemin, StdlogXlivemin), lw=.5, color='red', label='Minimum live point')
+plt.axvline(-ndead/nlive, color='deepskyblue', lw=.5, ls='--')
+plt.xlabel(r'$\log X$')
+plt.legend()
+plt.savefig('live_point_distribution.pdf', bbox_inches='tight')

### 2023-08-25 paper/paper.tex
paper mostly done

@@
+%{{{ Preamble
 \documentclass[usenatbib]{mnras}
 \usepackage{graphicx}
+\usepackage[export]{adjustbox}
 \usepackage{bm}
@@
 \usepackage{algpseudocode}
-\usepackage{caption}
 \usepackage{subcaption}
 \usepackage{layouts}
-\usepackage{cleveref}
+\usepackage{hyperref}
+\usepackage[capitalise]{cleveref}
+\usepackage{fancyvrb}
+\usepackage{xcolor}
 \usepackage{txfonts}
 \usepackage[T1]{fontenc}
+%}}}
 
+%{{{ Commands
 \newcommand{\logL}{\{ \log \Like_i \}}
@@
 \newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
+%}}}
 
@@
 \begin{abstract}
-This paper develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration for any nested sampler.
-The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and prior volume at termination to be calculated.
+This paper develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration which works for any nested sampling implementation.
+The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be calculated.
 We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods.
@@
 \section{Introduction}
+Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}.
+It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}.
+The core algorithm is unique for estimating volumes using \textit{order statistics}, which makes high-dimensional integration feasible.
+It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
+\par
+Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord}, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to plot sample statistics and distributions.
+However, no such software is currently able to estimate how long a given run should last.
+This is of high importance to the end user, who has no indication in the middle of a run whether it will take hours, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
+\par
+This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage, an example of which is shown in \cref{fig:polychord_output}.
+The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples.
+We begin in section \cref{sec:Background} with an overview of nested sampling, before discussing its convergence properties in \cref{sec:convergence}.
+\cref{sec:method} presents the methodology of making endpoint predictions, before results for both toy and real examples are shown in section \cref{sec:results}.
+Finally, conclusions are made in section \cref{sec:conclusions}.
+\par
 Users would like a order of magnitude runtime estimate
-Figure of PolyChord output
-
-\section{Background}
-\subsection{Notation}
-We shall begin with a brief description of nested sampling to establish the necessary notation.
-For a given likelihood $\Like(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
+\begin{figure}
+\begin{Verbatim}[frame=single, commandchars=\\\{\}]
+\textcolor{red}{Predicted endpoint: 25054 +/- 242}
+\textcolor{red}{Progress: [=================>########] 72%}
+___________________
+lives      |   500 |
+phantoms   | 24310 |
+posteriors | 18018 |
+equals     |   245 |
+-------------------
+ncluster   =  1/1
+ndead      =  18018
+nposterior =  18018
+nequals    =  249
+nlike      =  4159049
+<nlike>    =  491.04 (9.82 per slice)
+log(Z)     =  -12.55 +/- 0.27
+\end{Verbatim}
+\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
+\label{fig:polychord_output}
+\end{figure}
+\section{Background}\label{sec:Background}
+Let us begin with a brief description of nested sampling to establish the necessary notation.
+For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
 \begin{equation}\label{eq:evidence}
-	\mathcal{Z} = \int \Like(\theta)\ \pi(\theta)\ \mathrm{d}\theta
+	\mathcal{Z} = \int \mathcal{L}(\theta)\ \pi(\theta)\ \mathrm{d}\theta
 \end{equation}
@@
 \begin{equation}
-	\mathcal{P}(\theta) = \frac{\Like(\theta) \pi(\theta)}{\mathcal{Z}}.
+	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
 \end{equation}
@@
 \begin{equation}
-	\mathcal{Z} = \int_0^1 \Like(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \Like_i (X_{i-1}-X_{i+1}),
+	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
 \end{equation}
-where $X(\Like)$ is the fraction of the prior greater than $\Like$.
+where $X(\mathcal{L})$ is the fraction of the prior greater than $\mathcal{L}$.
 The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
-\begin{equation}
+\begin{equation}\label{eq:X_dist}
 	X_i = \prod_{k=1}^i t_k.
@@
 \begin{equation}\label{eq:t_dist}
-	p(t_i) = \nlive t_i^{\nlive-1}, \quad \mathrm{E}[\log t_i] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t_i] = \frac{1}{\nlive^2}.
+	P(t_i) = \nlive t_i^{\nlive-1}, \quad \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
 \end{equation}
@@
 
-\section{Convergence of a nested sampling run}
-Let us now take a top-down view of a nested sampling run as it progresses to observe how it converges.
-
-\subsection{Posterior convergence}
-We will examine how our set of samples evolve over the course of a nested sampling run.
-The conventional perspective for the progress of a run is with $\log X$ along the x-axis; the prior volume is compressed at a steady rate, only picking up the bulk of the integral/posterior at around $\log X = \DKL$.
+\section{Convergence of a nested sampling run}\label{sec:convergence}
+We will now present several useful ways of measuring the progress of a nested sampling run.
+At a given iteration $i^{*}$, these are the current prior volume $\log X^{*}$ of the latest dead point, the corresponding log-likelihood $\log\Like^{*}$, and the current effective temperature $\beta^{*}$ of the samples.
+\subsection{Current prior volume}\label{sec:volume_convergence}
+The conventional perspective for the progress of a run is with $\log X$ along the x-axis, where the prior volume is compressed at a steady rate until we reach the posterior mass.
+Defining the Kullback-Leibler divergence as
+\begin{equation}\label{eq:DKL}
+	\DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta,
+\end{equation}
+we interpret it as the posterior averaged logarithmic \textit{volume compression} from the prior to the posterior.
+Hence we can define a  \textit{posterior bulk} at below $X = e^{-\DKL}$ containing nearly all of the information, which gives us a measure of the volume we must compress down for the algorithm to converge.
 \par
-At an intermediate iteration, we have an accurate reconstruction of the posterior only down to the current prior volume $\log X_*$.
-The live points contain information below this volume, but not much more.
-If we kill off all the live points as if we were terminating, then by \cref{eq:t_dist} the prior volume of the final live point $\log X_\mathrm{min}^{\mathrm{live}}$ has mean and variance
+At an intermediate iteration $i_*$, we only have samples of the posterior only down to the current prior volume $\log X_*$.
+We do have some information below this volume due to the live points, but we will see now that they do not bring us much further.
+If we kill off all the live points as if we were terminating, then by \cref{eq:t_dist} the prior volume of the final live point $\log X_\mathrm{min}^{\mathrm{live}}$ has mean and variance given by
 \begin{equation}
@@
 \begin{equation}
-	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6},
+	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6},
 \end{equation}
-where the large $\nlive$ limit is taken for the approximation and $\gamma$ is the Euler-Mascheroni constant. 
+where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
 \par
-This shows that the live points only get us a factor of $\log \nlive$ closer to the posterior bulk, or in other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples begin to resemble the posterior.
-Intuitively, if the likelihood is sharply peaked then for most of the run the live points are too diffuse to land there with any significant probability.
+This shows that the live points only get us a factor of $\log \nlive$ closer to the posterior bulk.
+In other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples look anything like the posterior.
+One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities (such as the Planck legacy chains, which have a $\DKL$ of $\sim 40$) and typical live point numbers of  $\sim 1000$, this does not happen until near the end of the run.
 \par
-\begin{figure}
+Intuitively, this is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
+\begin{figure*}
 \begin{center}
@@
 \end{center}
-\caption{}
-\label{fig:}
+\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the prior $\pi(\log X \mid \log X < \log X_*)$ and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i_*$.
+For large values of $\DKL$ i.e.
+informative posteriors and/or large dimensionalities, the maximum live point is very far from the posterior bulk until the very end of the run.
+Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller.
+Plots of the sort from here onwards will be in terms of $-\log X$, where the run will more naturally proceed from rleft to right.
+}
+\label{fig:live_point_distribution}
+\end{figure*}
+
+\subsection{Current log-likelihood}
+We can also view the run as a function of the current log-likelihood value $\log\mathcal{L}^{*}$, which of course monotonically increases throughout the run.
+To illustrate some of the analytics, let us take the example of a Gaussian likelihood with dimension $d$ and lengthscale $\sigma$;
+\begin{equation}\label{eq:logL}
+	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
+\end{equation}
+$\log \mathcal{L}^{*}$ must increase until the posterior bulk is passed, which is located around the posterior averaged log-likelihood 
+\begin{equation}
+    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
+\end{equation}
+Given the value of the latest dead point, we can examine the distribution of the live points in likelihood. Define the normalised likelihood
+\begin{equation}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
+    \label{eq:normalised_likelihood}
+\end{equation}
+as a measure of how far a point is between the current likelihood and the maximum; $y=0$ corresponds to $\log\mathcal{L}_*$ and $y=1$ to $\log\mathcal{L}_\mathrm{max}$.
+It follows that the live points are distributed as
+\begin{equation}
+    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1],
+    \label{eq:Py}
+\end{equation}
+and the maximum live point has mean and variance
+\begin{equation}\label{eq:ylivemax}
+	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}
+\end{equation}
+in the limit of large $d$ and $\nlive$ (see appendix for derivation).
+We therefore see that just as with the prior volumes, in high dimensions the live points only bring us logarithmically further, while $d$ pushes us linearly away.
+
+\subsubsection*{Nested sampling as a maximiser}
+We can use these results to show that nested sampling by itself is a poor maximiser.
+If we set $\log\Like^{*}$ to be the termination point $\log\Like_\mathrm{end}$, we can invert \cref{eq:normalised_likelihood} to find that the maximum live point is located at
+\begin{equation}
+    \boxed{
+        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
+    }.
+\end{equation}
+Hence in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood, and the final set of $n$ live points can get you $(\log n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
+\par
+We find that in higher dimensions (typically greater than ten), the maximum live point is nowhere near the true maximum!
+However, these shortcomings can be mitigated by using the maximum point in the chain as a \textit{starting point} for a gradient-based optimisation algorithm, which would allow us to leverage the advantages of nested sampling to navigate multimodal landscapes while also rapidly approach the maximum point.
+\begin{figure*}
+\begin{center}
+	\includegraphics{figures/anatomy.pdf}
+\end{center}
+\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_*)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$}
+\label{fig:anatomy}
+\end{figure*}
+
+\subsection{Current temperature}\label{sec:current_temperature}
+We now describe a new measure of progress in terms of the \textit{effective temperature} of the current set of samples.
+Nested sampling is invariant to monotonic transformations of the likelihood, so one can transform $\mathcal{L} \rightarrow \mathcal{L}^{\beta}$ (where $\beta$ is an inverse temperature in analogy with statistical mechanics) and obtain an equally valid nested sampling run.
+This has a useful interpretation within the context of Bayesian inference: dialing $\beta$ from zero to one moves us from the prior to the posterior.
+\par
+The nested sampling algorithm does something similar; samples are initially drawn from the prior, which then slowly converge to the posterior as the run progresses.
+It is natural then to expect that any given point during a run can be associated with a $\beta \in [0,1]$, that measures how far the current set of samples lies between prior and posterior.
+We present one such mapping below, which allows us to make useful inferences about the present samples.
+\subsubsection*{The effective temperature}
+\begin{figure}
+\begin{center}
+	\includegraphics{figures/last_live_point.pdf}
+\end{center}
+\caption{The likelihood profile and posterior mass as $\beta$ is lowered.
+The original likelihood is shown in black, which is dominated by the maximum live point.
+As $\beta$ decreases, the posterior mass is shifted back from the maximum point to the other points, until so much of the mass has shifted that the present point $\log X^{*}$ is the termination point.
+The transformed likelihood and posterior at this point is shown in orange.}
+\label{fig:last_live_point}
+\end{figure}
+Midway through the run, it is typical for the posterior mass to be dominated by a single point, the maximum live point, because it is several $\log \mathcal{L}$ and therefore several orders of magnitude $\mathcal{L}$ larger than the next highest point.
+Lowering $\beta$ gradually from 1 has the effect of bringing the likelihoods closer together, which shifts the mass towards higher prior volumes and distributes it more equally (\cref{fig:last_live_point}).
+One can do this until the current point is exactly the termination point, so that all but $\epsilon$ of the posterior mass lies before the current prior volume, which occurs at a value $\beta = \beta^{*}$.
+\par
+$\beta^{*}$ is therefore the temperature at which the nested sampling run would have finished by now, which we can call the \textit{effective temperature} of the current set of samples.
+It increases from zero to one and is monotonic with respect to $\log X^{*}$, and so provides a valid measure of progress from prior to posterior.
+
+\subsubsection*{Intermediate sample statistics}
+The transformed posterior is useful because it allows us to calculate intermediate sample statistics which are otherwise inaccessible because they are dominated by the single maximum live point.
+An example is the \textit{Bayesian model dimensionality}, which is the posterior variance of the information content:
+\begin{equation}
+\frac{\tilde{d}}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
+= \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
+\end{equation}
+$\tilde{d}$ is the number of parameters constrained by the samples, in analogy to the classical heat capacity which is similarly proportional to both variance in the energy and the degrees of freedom in the system.
+If this quantity is computed from the untransformed samples, it will output zero since there is effectively only one sample.
+The transformed posterior, on the other hand, gives a $\tilde{d}$ which is consistent with the distribution of live points, shown in \cref{fig:dG_beta} at various stages for an elongated Gaussian.
+\begin{figure}
+\begin{center}
+	\includegraphics{figures/dG_beta.pdf}
+\end{center}
+\caption{The Bayesian model dimensionality $\tilde{d}$ as a function of $\beta$ for a Gaussian two parameters of variance $10^{-3}$ and $10^{-6}$.
+The number of contrained samples can be seen to increase from zero to the true dimensionality of 2, as the run progresses.}
+\label{fig:dG_beta}
+\end{figure}
+\par
+The gradual increase in $\tilde{d}(\beta)$ and corresponding shape of the live points can be explained in the following way.
+If the likelihood is much narrower for one of the parameters with respect to the prior, then the likelihood gradient is initially much higher in this dimension.
+Since nested sampling gains new samples by climbing the likelihood, the new samples will initially only be taken in this direction, appearing as if there is only one constrained parameter.
+Only once the high variance parameter is constrained to the same level with respect to the prior as the other parameter, compression occurs at the same rate in both dimensions, which yields an effective sample dimensionality of two.
+\par
+Having established several metrics to measure the progress of a nested sampling run, we can use the information they provide to make endpoint predictions.
+
+\section{Method}\label{sec:method}
+The analysis in \cref{sec:volume_convergence} shows that our knowledge of the posterior at an intermediate iteration is limited by the maximum log-likelihood live point, which until just before the end is quite far from the posterior bulk.
+However, we can get an approximate image of the full posterior ahead of time, simply by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\mathcal{L}(X)$ traced out by the live and dead points.
+\par
+One would never use this approximate posterior to do inference, since the extrapolated profile is much less exact than if one were to simply finish the run.
+However, it is more than sufficient for making a prediction for roughly how much evidence remains, and hence the point at which the run should terminate.
+Quantitatively, this proceeds as follows: we fit a model function $f(X, \thetab)$ with some parameters $\thetab$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
+\begin{equation}
+	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
+\end{equation}
+\begin{equation}\label{model endpoint}
+	\int_0^{X_\mathrm{f}} f(X, \thetab)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
+\end{equation}
+where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point.
+$X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically.
+\par
+Once $X_\mathrm{f}$ is known, the corresponding iteration count depend on the live point schedule.
+The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_\mathrm{f}$ will be
+\begin{equation}
+	N_\mathrm{f} = - \nlive \log X_\mathrm{f} .
+\end{equation}
+
+\subsection{Choosing the fitting function}
+One has freedom to choose the form of the fitting function $f(X, \thetab)$.
+Following the previous analysis, we use a Gaussian profile in the form of \cref{eq:logL}.
+It has the advantage that one can speed up its regression analytically, as well as the fact that many real likelihoods are Gaussian or near-Gaussian.
+We allow $d$ to freely vary to increase flexibility, and reflect the fact that while the dimensionality of a likelihood is fixed, the equivalent Gaussian dimensionality may differ \citep{Handley_2019}.
+Regardless of motivation, the proof is in the pudding, so we shall justify the choice on the accuracy of predictions in \cref{sec:results}.
+\par
+We must also choose the set of points over which we fit.
+Using all of the points we have is a poor choice, because unless the profile is actually Gaussian the extrapolation will be insensitive to changes in the profile.
+A natural choice is then to use the live points only, since these represent the most up-to-date likelihood profile.
+For a given set of live points, we have a set of live likelihoods $\logL$, as well as a set of $\set{X_i}$ drawn from the distribution \labelcref{eq:X_dist}.
+The parameters of the fitting function can be found by minimising the cost function
+\begin{equation}\label{chi squared}
+	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
+\end{equation}
+with respect to $\thetab$.
+A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
+\begin{equation}\label{eq:sigma}
+    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
+\end{equation}
+\begin{equation}\label{eq:logLm}
+    \logLm = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d},
+\end{equation}
+(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables.
+The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run.
+We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
+\subsection{The termination prior volume}
+Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \cref{model endpoint};
+\begin{equation}
+	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
+\end{equation}
+The integrals have the analytic solution
+\begin{equation}
+	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
+\end{equation}
+where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
+\begin{equation}
+	\log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \gamma^{-1} \left(\ \frac{d}{2} ,\ \epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),
+\end{equation}
+and $N_\mathrm{f}$ is just  $-\nlive$ multiplied by this.
+\par
+The prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.
+\par
+Something about time complexity?
+
+\subsection{Prediction uncertainties}
+We saw in section X that our knowledge of the posterior is limited by the live point of minimum volume, which is only $\mathcal{O}(\log \nlive)$ less than the log-compression of the dead point.
+Endpoint prediction as we have just done relies on our hope that the likelihood will continue to behave as it has done up to this point, but this may well not be the case; a sharp phase change, for instance, will be our undoing.
+\par
+We will therefore restrict ourselves to quantifying the uncertainties associated with how we have \textit{modelled} the known likelihood profile.
+These have two sources: the variance in the prior volume estimates, and the uncertainty in the least squares regression.
+The former is by far the larger source of error, since the uncertainty in nested sampling is dominated not by the scatter of the individual points but by long-term drift (as shown in \cref{fig:prediction_uncertainty}).
+\par
+To get the uncertainty, we can therefore ignore the least squares uncertainty and simply sample from $\Pr(\bm{X})$, finding the best-fit $\thetab$ and endpoint for each sample, then using this set of estimates to get a mean and variance in the endpoint estimate.
+\begin{figure}
+\begin{center}
+	\includegraphics[max width=\linewidth]{figures/prediction_uncertainty.pdf}
+\end{center}
+\caption{Sources of the prediction uncertainty.
+The navy points are the live points, plotted for three different instances of $\set{X_i}$ drawn from its distribution.
+The light blue lines show the uncertainty in the least squares fits over each instance, which is much less than the uncertainty in the spread of the $X$ samples due to long term drift.}
+\label{fig:prediction_uncertainty}
 \end{figure}
 
-\subsection{Loglikelihood convergence}
-We now offer an alternate view of the convergence of a nested sampling run, via the loglikelihood instead of the prior volumes.
-To get an insight into the analytics, work with a Gaussian likelihood of $d$ dimensions and a lengthscale of $\sigma$;
+\section{Results}\label{sec:results}
+Now that we have a method for predicting the endpoint, we can test it on a range of likelihoods.
+We shall first present results for cosmological chains in \cref{subsec:cosmological_examples}, for which the method works well.
+\cref{subsec:toy_examples} will then show some toy examples that are designed to reveal some of the limitations of the method.
+
+\subsection{Gaussians}\label{subsec:gaussians}
+We will first test our prediction model on truly Gaussian likelihoods to show that it is able to recover the true parameters as well as the correct endpoint.
+Let us begin with a spherically symmetrical Gaussian which matches the form of the fitting function, which we can use to check that the correct parameters are inferred when the model is exact.
+We fix the parameters $\log\mathcal{L}_\mathrm{max} = 0$ and $\sigma = 0.01$ and vary the dimension to see the impact on the predictions.
+\par
+The inferred parameters are shown in \cref{fig:gauss_params}.
+The uncertainty in $d$ and $\sigma$ remain large throughout the run, but the endpoint predictions converge because we have already accumulated most of the evidence.
+\par
+\begin{figure*}
+\begin{center}
+	\includegraphics{figures/gauss_params.pdf}
+\end{center}
+\caption{Parameter inferences for the Gaussian fitting function in the $d=8$ case.
+The true parameters fall within the standard error bars for practically every point, as seen in comparison to the true parameter values shown in the dotted red.
+The uncertainties remain relatively constant for $d$ and $\sigma$, but the endpoint predictions converge because we have already accumulated most of the evidence.}
+\label{fig:gauss_params}
+\end{figure*}
+Predictions are shown in \cref{fig:gauss_predictions}, where the x-axis plots the negative log-compression as a measure of the progress, so that the run proceeds in the rightward direction.
+The y-axis shows the predicted log-compression $\log\hat{X}_\mathrm{f}$ at which the termination condition is met.
+\par
+The true endpoint is recovered to within $2\sigma$ for all runs, but the uncertainty is greater for the higher dimension and $\DKL$ likelihoods.
+This reflects the fact that the live points typically probe only $\sim \log n_\mathrm{live}$ away from the current dead point.
+For larger $\DKL$ therefore, the set of samples at any given iteration before the end are quite far from the posterior bulk, and hence poorly resemble the shape of the true posterior.
+
+\subsubsection*{Comparison to extrapolating evidence increments}
+Seasoned users of nested sampling might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges.
+We do this for a spherical Gaussian as a comparison to our method.
+At an intermediate stage of the run, the most recent outputs might look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate}.
+Extrapolating those data to a linear and exponential profile yields endpoint estimates plotted in the graph to the right.
+\par
+The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile.
+The increments are also not quite exponential, since the exponential fit leads to a large overprediction.
+The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result.
+One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
+\par
+More importantly, we find that for real likelihoods which have an element of noise the extrapolation often diverges, for instance when the increments do not monotonically decrease.
+Directly extrapolating the evidence increments is therefore far less stable than the previous method, and generally not a reliable method for prediction.
+\begin{figure}
+\begin{adjustbox}{valign=t, scale=0.72}
+\subfloat{\begin{tabular}{|c|c|c|}
+\hline
+iteration & log Z  & $d\log Z$   \\
+\hline
+5000 & -1435.8 & 190.8 \\
+5500 & -1264.6 & 171.2 \\
+6000 & -1123.7 & 140.9 \\
+6500 & -991.5 & 132.2 \\
+7000 & -885.0 & 106.6 \\
+7500 & -790.3 & 94.7 \\
+8000 & -702.6 & 87.7 \\
+8500 & -619.7 & 82.9 \\
+9000 & -551.8 & 67.9 \\
+9500 & -492.7 & 59.1 \\
+\hline
+\end{tabular}}
+\end{adjustbox}
+\quad
+\begin{adjustbox}{valign=t}
+\subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
+\end{adjustbox}
+\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation}
+\label{fig:inc_extrapolate}
+\end{figure}
+
+\begin{figure}
+\includegraphics{figures/inc_predictions.pdf}
+\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation}
+\label{fig:inc_predictions}
+\end{figure}
+
+\subsection{Cosmological examples}\label{subsec:cosmological_examples}
+Having established that the prediction framework works for Gaussian likelihoods, we now turn to some cosmological examples, namely chains based on the \textsc{Planck 2018} dataset.
+Predictions are shown in \cref{fig:lcdm_predictions}, with the correct endpoint inferred to within error before the halfway point for all likelihoods.
+\par
+Most predictions fail to converge at early stages of the run.
+We find that this is not a consequence of the prediction model, but due to the effective dimensionality of the samples being lower than the true value at the beginning of the run, as was seen in \cref{sec:current_temperature}.
+
+\subsubsection*{Compressive phase transitions}
+\begin{figure}
+\begin{center}
+\includegraphics{figures/phase_transition.pdf}
+\end{center}
+\caption{Endpoint predictions and effective dimensionality as a function of compression for a two-dimensional Gaussian in a box prior $\in [0, 1]$.
+The parameter variances are $10^{-2}$ and $10^{-5}$ respectively, leading to compression only along one direction initially, which is reflected in $\tilde{d}$.
+The endpoint only converges when the dimensionality does; before that point the samples genuinely suggest that the likelihood is one-dimensional.}
+\label{fig:phase_transition}
+\end{figure}
+The initial dimensionality of the $\mathcal{L}(X)$ curve will be $1$, so extrapolating from the curve would fail to account for the true dimensionality of the likelihood, giving inaccurate predictions.
+\par
+.
+The extrapolation now accounts for the true dimensionality, so the predictions become accurate.
+The trouble is that before this has happened, the samples really do behave as if they are only constrained in a single dimension!
+It could really be the case that there is no second parameter, and without our extra knowledge there is no way to detect whether there are two, or even three or more unseen parameters hiding at higher compressions.tem
+\par
+The different rates of compression for different parameters therefore induce an effective phase transition in the number of degrees of freedom.
+One can put this on a more quantitative footing by considering the \textit{effective dimensionality} of the posterior.
+Define the Bayesian model dimensionality $\tilde{d}$ as the posterior variance of the information content, which is given by
+\par
+\cref{fig:phase_transition} plots the effective dimensionality alongside the endpoint predictions, where it is evident that the predictions are directly dependent on $\tilde{d}$.
+Introducing correlations between the parameters (not shown) causes the transition to be less sharp, since the more constrained parameter is able to `drag' the other parameter with it, allowing the predictions can be converge shortly before the dimensionality.
+
+\subsection{Edge cases}\label{subsec:toy_examples}
+In the cosmological cases, underestimated predictions at early stages was due to the difference in parameter compression rates.
+However, there are also cases where the form of the fitting function causes problems.
+\par
+One such instance is a spherical Cauchy likelihood.
+In fact, at early stages direct least squares fitting fails completely by predicting a remaining evidence that overflows.
+This can be explained by the fact that a long tail of a Cauchy resembles a Gaussian of much higher dimension, and inferring the dimensionality is unstable.
+Since we have a stable measure of dimension from the Bayesian model dimensionality, a partial fix is to set $d = \tilde{d}$ instead of inferring it, then using \cref{eq:sigma,,eq:logLm} to get $\log\mathcal{L}_\mathrm{max}$ and $\sigma$.
+\par
+Results are shown in \cref{fig:cauchy_predictions}, where the likelihood has the form
+\begin{equation}
+	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
+\end{equation}
+choosing $d = 10$ and allowing $\gamma$ to vary.
+However, the extrapolations still fall outside fall outside the standard error range despite isotropic compression.
+The issue in this case is that the Gaussian fitting function is unable to capture the true shape of the likelihood.
+A more comprehensive fix would be to use a more flexible fitting function that is able to capture a broader family of likelihoods, which is outside the scope of this paper.
+Nevertheless, the order of magnitude is still accurate, which remains highly useful to the end user.
+
+\begin{figure*}
+\begin{center}
+	\subfloat[Predictions for a spherically symmetric Gaussian likelihood.
+The correct endpoint is ascertained to within standard error from the beginning for all runs, but the uncertainty is larger for the higher dimension and $\DKL$ likelihoods, reflecting the fact that the live points are further away from the posterior bulk in those cases.]{\includegraphics{figures/gauss_predictions.pdf}\label{fig:gauss_predictions}}
+	\\[5ex]
+	\subfloat[Predictions for the $\Lambda$CDM model.
+The chains without \textsc{Planck} have low $\DKL$ of below 10, so the live points are relatively close to the posterior bulk at all times and the predictions are quite accurate.
+]{\includegraphics{figures/lcdm_predictions.pdf}\label{fig:lcdm_predictions}}
+	\\[5ex]
+	\subfloat[Predictions for a Cauchy likelihood.
+The parameters of the fitting function are obtained by fixing $d = \tilde{d}$, then least squares optimising $\log\mathcal{L}_\mathrm{max}$ and $\sigma$.]{\includegraphics{figures/cauchy_predictions.pdf}\label{fig:cauchy_predictions}}
+\end{center}
+\caption{Endpoint predictions for different likelihoods.
+The x-axis shows the negative log-compression, and the y-axis the predicted endpoint $\log\hat{X}_\mathrm{f}$.
+The shaded regions show the 1 and 2$\sigma$ uncertainties, while the red dashed line shows the true value.}
+\end{figure*}
+
+\section{Conclusions}\label{sec:conclusions}
+We have presented a method for predicting the endpoint of a nested sampling run, based on extrapolating the known likelihood profile into higher likelihood regions using a Gaussian fitting function.
+\par
+The method in general converges to the correct endpoint by about the halfway point, and gets the correct order of magnitude throughout.
+Predictions are poor when the effective dimension of the samples is lower than the true posterior, but this is a limitation of the information content in the samples rather than our method.
+\par
+Further work can be done to incorporate more flexible fitting functions to extrapolate non-Gaussian likelihoods, as well as developing a more rigorous framework to choose between the inferred $d$ and $\tilde{d}$.
+
+\bibliographystyle{mnras}
+\bibliography{references}
+
+\appendix
+\section{Derivation of the termination log-likelihood }\label{app:fitting_function}
+Assume a Gaussian likelihood
 \begin{equation}\label{eq:logL}
-	\log\Like = \log\Like_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
+	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
 \end{equation}
-The distribution of the true posterior in $\log\Like$ is
+The distribution of the true posterior in $\log\mathcal{L}$ is
 \begin{equation}
@@
 \end{equation}
-i.e.
-$2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the same distribution as that of the dead points up to the latest dead likelihood $\log \Like_*$.
+i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the distribution of the weighted samples. The posterior average and variance of $\log \mathcal{L}$ are given by
+\begin{equation}
+    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
+\end{equation}
 Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
 \begin{equation}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\Like_* < \log\Like <\log\Like_\mathrm{max}],
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\mathcal{L}_* < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
     \label{eq:PL}
@@
 \end{equation}
-as a normalised measure of how far a point is between the latest dead point and the maximum loglikelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
+as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
 \begin{equation}
@@
 \end{equation}
-We now seek the distribution for the maximum likelihood of the live points, $\log\Like_\mathrm{max}^{\mathrm{live}}$.
+We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$.
 Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
@@
 \end{equation}
-This shows that in general the live points are nowhere near the maximum loglikelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
+This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
 In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
@@
 \begin{equation}
-    f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L} dX}{\int_0^\infty \mathcal{L} dX}.
+	f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
     \label{eq:fint}
@@
 In the event that we keep $f$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}f^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
-
-\subsection*{Observation 3: nested sampling as a maximiser}
+\par
 Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
@@
 \end{equation}
-showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\log (n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
-\begin{figure*}
-\begin{center}
-	\includegraphics{figures/anatomy.pdf}
-\end{center}
-\caption{}
-\label{fig:anatomy}
-\end{figure*}
-\par
-From above analysis we learn that
-\begin{itemize}
-	\item Knowledge of the posterior early on in the run is limited by the maximum live point, which is far from the posterior bulk
-	\item At termination, the maximum live point is far from the true maximum --> nested sampling alone is a poor optimiser (but great for finding starting point of gradient descent)
-\end{itemize}
+showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood, and the final set of $n$ live points can get you $\log (n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
+This general anatomy is shown in \cref{fig:anatomy}.
 
-\section{Method}
-Let us now tackle the problem of estimating how many steps we still have to take in a nested sampling run before we can terminate.
-\par
-The above analysis shows that our knowledge of the posterior at an intermediate iteration is limited by the maximum live point, which until just before the end is quite far from the posterior bulk.
-However, we can get an approximate image of the full posterior ahead of time, simply by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points.
-\par
-For practical purposes, one would never use this approximate posterior to do inference.
-However, it is more than sufficient for making a prediction for roughly how much evidence remains, and hence the point at which the run should terminate.
-Quantitatively, this proceeds as follows: fitting a model function $f(X, \thetab)$ to the known likelihood profile allows us to express the prior volume we need to compress to as
-\begin{equation}
-	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
-\end{equation}
-\begin{equation}\label{model endpoint}
-	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
-\end{equation}
-where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point.
-$X_f$ can then be identified by solving the above equation either analytically or numerically.
-If $\nlive$ is constant then at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_f$ will be
-\begin{equation}
-	N_f = - \nlive \log X_f .
-\end{equation}
-One has freedom to choose the form of the fitting function $f(X, \thetab)$.
-Following the previous analysis, we use a Gaussian profile, which has the advantage that the regression itself can be sped up analytically, as well as the fact that many real likelihoods are Gaussian or near-Gaussian.
-However, any other function that is sufficiently flexible to capture the shape of the likelihood profile will do.
-
-Fit a Gaussian profile, with advantages that a) Gaussian error common b) analytic optimisation 
-Extraction of termination logX
-Prediction uncertainty dominated by correlations, not least squares i.e. scatter of individual points - get by simulating X
-
-\section{Results}
-Toy examples - pure Gaussian,  wedding cake, Cauchy
-Real cosmological examples - Planck
-
+\label{lastpage}
 \end{document}

### 2023-08-25 paper/figures/LX.py

@@ -1,34 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from aeons.utils import figsettings
-
-figsettings()
-
-
-def full(X, theta):
-    logLmax, d, sigma = theta
-    return logLmax - X**(2/d)/(2*sigma**2)
-
-logXarray = np.linspace(-70, 0, 1000)
-Xarray = np.exp(logXarray)
-theta = [0, 10, 0.01]
-logL = full(Xarray, theta)
-L = np.exp(logL)
-
-Ztrue = np.trapz(L, Xarray)
-LlogL = np.nan_to_num(L/Ztrue * np.log(L/Ztrue))
-H = np.trapz(LlogL, Xarray)
-
-fig, ax1 = plt.subplots(figsize=(3.37, 1.5))
-ax2 = ax1.twinx()
-ax1.plot(logXarray, L, color='black', label=r"$L(X)$")
-ax2.fill(logXarray, L*Xarray, lw=1, color='black', alpha=0.2, label=r"$L(X)X$")
-ax2.axvline(-H, ls='--', lw=0.5, color='black')
-for ax in ax1, ax2:
-    ax.set_yticks([])
-    ax.set_xticks([-H], [f'$-D_\mathrm{{KL}}$'])
-    ax.margins(x=0, y=0)
-ax1.set_xlabel(r'$\log X$');
-fig.legend(loc="upper right", bbox_to_anchor=(1,1), bbox_transform=ax.transAxes)
-fig.tight_layout()
-fig.savefig('LX.pdf')

### 2023-08-25 paper/figures/anatomy.py

@@ -1,10 +1,9 @@
 import matplotlib.pyplot as plt
 import numpy as np
 from matplotlib import rcParams
-rcParams['text.usetex'] = True 
-rcParams['font.family'] = 'serif'
+from aeons.utils import figsettings
 
-
+figsettings()
 fig, ax = plt.subplots()
 
 logLmax = 0
@@ -15,7 +14,7 @@ avlogL = logLmax-d/2
 logL = np.linspace(logLmax-d/2-np.sqrt(d/2)*3.5,0,1000)
 logpdf = logL - logLmax + (d/2-1)*np.log(logLmax-logL)
 logpdf -= np.max(logpdf)
-ax.plot(logL, np.exp(logpdf), label='posterior: $\mathcal{P}(\log\mathcal{L})$')
+ax.fill(logL, np.exp(logpdf), color='gray', alpha=.4, label='Posterior')
 
 logLs = logLmax-d/2/np.e
 logL = np.linspace(logLs,0,1000)
@@ -23,53 +22,58 @@ logpdf = (d/2-1) * np.log(logLmax - logL)
 logL = np.concatenate([[logLs], logL])
 logpdf = np.concatenate([[-np.inf], logpdf])
 logpdf -= np.max(logpdf)
-ax.plot(logL, np.exp(logpdf), label='live points: $\mathcal{\pi}(\log\mathcal{L}|\mathcal{L}>\mathcal{L}_*)$')
+ax.fill(logL, .7*np.exp(logpdf), color='deepskyblue', alpha=0.5, label='Live points')
 
 from scipy.special import logsumexp
 
 logL = np.linspace(logLs,0,1000)
 logpdf = (d/2-1)*np.log(logLmax - logL) + (n-1)* logsumexp([ d/2*np.log(logLmax-logLs)*np.ones_like(logL), d/2*np.log(logLmax-logL)] , b=[np.ones_like(logL),-np.ones_like(logL)], axis=0)
 logpdf -= np.max(logpdf)
-ax.plot(logL, np.exp(logpdf), label='maximum live point: $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$')
+ax.plot(logL, .7*np.exp(logpdf), lw=.5, color='red', label='Max live point')
 
 
 logLmaxlive = logLmax - d/2/np.e + np.log(n)/np.e
 
-l1 = 1.3
[…]

### 2023-08-25 paper/figures/beta_move.py

@@ -1,0 +1,37 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+figsettings()
+
+def weights(logL, logX, beta):
+    logw = beta*logL + logX
+    return np.exp(logw - np.max(logw))
+
+name, samples = get_samples('toy', 'planck_gaussian')
+samples = samples.iloc[:int(len(samples)*0.7)]
+logL = samples.logL
+logX = samples.logX()
+betastar = 1e-3
+DKL = samples.D_KL()
+DKLstar = samples.set_beta(betastar).D_KL()
+
+
+fig, axl = plt.subplots(figsize=(3.5, 1.4))
+axw = axl.twinx()
+axw.fill(-logX, weights(logL, logX, beta=1), label='$\\beta=1$', alpha=.5, color='gray')
+axw.fill(-logX, weights(logL, logX, beta=betastar), label='$\\beta=10^{-2}$', alpha=.5, color='orange')
+axl.plot(-logX, np.exp(logL - np.max(logL)), lw=1, color='k')
+axl.plot(-logX, np.exp(betastar*(logL - np.max(logL))), lw=1, color='darkorange')
+axl.axvline(x=DKL, ls='--', color='k', lw=.5)
+axl.axvline(x=DKLstar, ls='--', color='darkorange', lw=.5)
+
+for ax in [axw, axl]:
+    ax.set_yticks([])
+    ax.margins(x=0)
+    ax.set_ylim(0, 1.1)
+axl.set_xticks([DKL, DKLstar], [r'$\mathcal{D}_{\mathrm{KL}}$', r'$-\log X^*$'])
+axl.set_xlabel(r'$-\log X$')
+axw.legend(loc='upper left', fontsize=6)
+
+fig.tight_layout()
+fig.savefig('beta_move.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/cauchy_predictions.py

@@ -1,0 +1,18 @@
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import *
+figsettings()
+
+fig, axs = plt.subplots(1, 3, figsize=(8, 1.4))
+for i, exp in enumerate([2, 3, 4]):
+    ax = axs[i]
+    name, samples = get_samples('toy', f'cauchy_8{exp}')
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/cauchy/cauchy_8{exp}.txt')
+    logXs = samples.logX().iloc[iterations]
+    logXf_true = samples.logX().iloc[int(true_endpoint[0])]
+    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.6), color='deepskyblue', ax=ax)
+    ax.set_title(f'$\\gamma=10^{{{-exp}}}$, $\\mathcal{{D}}_\\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
+
+fig.supxlabel('$-\\log X$', y=-0.15)
+fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.06)
+fig.savefig('cauchy_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/dG_beta.py

@@ -1,0 +1,54 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import plot_std
+from matplotlib.gridspec import GridSpec
+from anesthetic import make_2d_axes
+figsettings()
+
+name, samples = get_samples('toy', 'gauss_36')
+iterations, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_36_dG.txt')
+
+fig = plt.figure(figsize=(3.5, 4))
+gs = GridSpec(3, 3, wspace=0.1, hspace=.7)
+_, ax1 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 0])
+_, ax2 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 1])
+_, ax3 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 2])
+ax4 = fig.add_subplot(gs[1, :])
+ax5 = fig.add_subplot(gs[2, :])
+
+samples.set_beta(1e-6).plot_2d(ax1)
+samples.set_beta(1e-4).plot_2d(ax2)
+samples.set_beta(1).plot_2d(ax3)
+
+betas = np.logspace(-6, 0, 100)
+# samples.d_G(beta=betas).plot(ax=ax4, logx=True, lw=1, color='k')
+d_Gs_beta = samples.d_G(nsamples=10, beta=betas).unstack()
+plot_std(np.log10(betas), d_Gs_beta.mean(axis=1), d_Gs_beta.std(axis=1), ax=ax4)
+plot_std(-samples.logX().iloc[iterations], d_Gs, d_Gs_std, ax=ax5)
+
+ax4.axvline(x=-4.3, lw=.5, ls='--', color='gray')
+ax4.axvline(x=-2, lw=.5, ls='--', color='gray')
+ax5.axvline(x=.5, lw=.5, ls='--', color='gray')
+ax5.axvline(x=2.8, lw=.5, ls='--', color='gray')
+
+for ax in [ax1, ax2, ax3]:
+    ax = ax.iloc[0,0]
+    ax.set_xlim(0, 1)
+    ax.set_ylim(0, 1)
+    ax.set_xticks([])
[…]

### 2023-08-25 paper/figures/gauss_params.py

@@ -1,0 +1,23 @@
+import matplotlib.pyplot as plt
+
+from aeons.utils import *
+from aeons.plotting import *
+figsettings()
+
+samples = get_samples('toy', 'gauss_8')[1]
+iterations, logLmaxs, logLmaxs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_8_logLmaxs.txt')
+_, ds, ds_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_8_ds.txt')
+_, sigmas, sigmas_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_8_sigmas.txt')
+logXs = samples.logX().iloc[iterations]
+
+fig, axs = plt.subplots(1, 3, figsize=(7, 1.6)) 
+plot_std(-logXs, logLmaxs, logLmaxs_std, true=0, ax=axs[0], color='salmon')
+plot_std(-logXs, ds, ds_std, true=8, ylim=(0, 2), ax=axs[1], color='salmon')
+plot_std(-logXs, sigmas, sigmas_std, true=0.01, ylim=(0, 3), ax=axs[2], color='salmon')
+axs[0].set_title("$\\log \\mathcal{L}_\\mathrm{max}$", fontsize=8)
+axs[1].set_title("$d$", fontsize=8)
+axs[2].set_title("$\\sigma$", fontsize=8)
+# fig.suptitle('Inferred parameters', y=.9, fontsize=8)
+fig.supxlabel(r"$-\log X$", y=.05, fontsize=8)
+fig.tight_layout()
+fig.savefig('gauss_params.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/gauss_predictions.py

@@ -1,0 +1,21 @@
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import *
+figsettings()
+
+d = [8, 16, 32]
+fig, axs = plt.subplots(1, 3, figsize=(7, 1.8))
+axs = axs.flatten()
+for i, nDims in enumerate(d):
+    samples = pickle_in(f'{aeons_dir}/data/predictions/gauss/gauss_{nDims}.pkl')
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_{nDims}.txt')
+    logXf_true = samples.logX().iloc[int(true_endpoint[0])]
+    logXs = samples.logX().iloc[iterations]
+    # endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=500)
+    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.8), ax=axs[i])
+    axs[i].set_title(rf'$d$ = {nDims}, $\mathcal{{D}}_\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
+    axs[i].set_yticks([-logXf_true], [f"{-logXf_true:.1f}"])
+fig.supxlabel('$-\\log X$', y=0.1, fontsize=8)
+fig.supylabel('$-\\log \hat{X}_\mathrm{f}$',  fontsize=8)
+fig.tight_layout()
+fig.savefig('gauss_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/inc_extrapolate.py

@@ -1,0 +1,40 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.increment import IncrementModel
+figsettings()
+
+
+name, samples = get_samples('toy', 'gauss_16')
+inc = IncrementModel(samples, nlive=500)
+true_endpoint = inc.true_endpoint()
+true_logXf = samples.logX().iloc[true_endpoint]
+
+ndead = 10000
+endpoint_linear = inc.get_endpoint(ndead, method='linear', steps=2)
+endpoint_exp = inc.get_endpoint(ndead, method='exp', steps=10)
+logXf_linear = endpoint_linear/-500
+logXf_exp = endpoint_exp/-500
+
+fig, ax = plt.subplots(figsize=(2, 1.5))
+inc.delta_logZ(ndead).plot(marker='x', ms=3, ls='none', ax=ax, label='Known $\\Delta\\log\\mathcal{Z}$')
+inc.true_delta_logZ(ndead).plot(color='black', marker='x', ms=3, ls='none', ax=ax, label='True $\\Delta \\log \\mathcal{Z}$')
+inc.exponential_extrapolate(ndead, steps=10).plot(lw=1, ax=ax, color='deeppink')
+inc.linear_extrapolate(ndead, steps=2).plot(lw=1, ax=ax, color='deeppink')
+
+ax.axvline(true_endpoint, lw=1, linestyle='--', color='black')
+ax.axvline(x=endpoint_exp, lw=1, linestyle='--', color='deeppink')
+ax.axvline(x=endpoint_linear, lw=1, linestyle='--', color='deeppink')
+ax.set_xticks([endpoint_linear, true_endpoint, endpoint_exp], [-logXf_linear, f"{-true_logXf:.1f}", -logXf_exp])
+height = 210
+ax.text(endpoint_linear, height, "Linear", ha='center', va='top', fontsize=6)
+ax.text(true_endpoint, height, "True", ha='center', va='top', fontsize=6)
+ax.text(endpoint_exp, height, "Exponential", ha='center', va='top', fontsize=6)
+ax.margins(y=0.01)
+
+ax.set_xlim(ndead - 5000, true_endpoint * 2.5)
+ax.set_xlabel("$-\\log X$")
+ax.set_ylabel('$\\Delta \\log \\mathcal{Z}$')
+ax.legend(loc='upper right', fontsize=6, handletextpad=.1)
+
[…]

### 2023-08-25 paper/figures/inc_predictions.py

@@ -1,0 +1,37 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import *
+from aeons.increment import IncrementModel
+figsettings()
+
+name, samples = get_samples('toy', 'gauss_16')
+inc = IncrementModel(samples, nlive=500)
+true_endpoint = inc.true_endpoint()
+
+steps = 10
+start = steps * 500/true_endpoint
+
+iterations = make_iterations(true_endpoint, 50, start=start)
+endpoints_exp = np.zeros(len(iterations))
+endpoints_linear = np.zeros(len(iterations))
+for i, ndead in enumerate(iterations):
+    endpoints_exp[i] = inc.get_endpoint(ndead, method='exp', steps=steps)
+    endpoints_linear[i] = inc.get_endpoint(ndead, method='linear', steps=2)
+
+ndeads, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_16.txt')
+logXs = samples.logX().iloc[ndeads]
+endpoints, endpoints_std = calc_endpoints(ndeads, logXs, logXfs, logXfs_std, nlive=500)
+true_logXf = samples.logX().iloc[true_endpoint].values
+
+fig, ax = plt.subplots(figsize=(3.5, 1.5))
+ax.plot(-logXs, endpoints_exp/500, lw=.8, color='green', label='Exponential')
+ax.plot(-logXs, endpoints_linear/500, lw=.8, color='orange', label='Linear')
+plot_std(-logXs, -logXfs, logXfs_std, -true_logXf, ax=ax, label='Fit $\\mathcal{L}(X)$')
+
+ax.set_ylim(0, 160)
+ax.set_ylabel(r'$-\log \hat{X}_f$')
+ax.set_xlabel(r'$-\log X$')
+ax.legend(loc='upper right', fontsize=6)
+
+fig.savefig('inc_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/last_live_point.py

@@ -1,0 +1,49 @@
+import matplotlib.pyplot as plt
+from aeons.utils import *
+import matplotlib.patches as patches
+figsettings()
+
+def get_beta(points, ndead):
+    logX = points.logX().iloc[ndead]
+    if logX < -points.D_KL():
+        return 1
+    def func(beta):
+        return logX + points.set_beta(beta).D_KL()
+    from scipy import optimize
+    res = optimize.root_scalar(func, bracket=[0, 1])
+    return res.root
+
+name, samples = get_samples('toy', 'planck_gaussian')
+ndead = 8000
+points = points_at_iteration(samples, ndead).recompute()
+betas = [1, get_beta(points, ndead) * 1.5, get_beta(points, ndead) * 0.5, get_beta(points, ndead) * 0.05]
+labels = ['$\\beta=1$', '$\\beta^* < \\beta < 1$', None, '$\\beta = \\beta^*$']
+colors = ['black', 'gray', 'darkgray', 'orange']
+
+fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
+for i, beta in enumerate(betas):
+    axs[0].plot(-points.logX(), np.exp(beta*(points.logL - points.logL.max())), 'x', ms=1, color=colors[i], label=labels[i])
+    axs[0].plot(-points.logX(), np.exp(beta*(points.logL - points.logL.max())), lw=.5, alpha=.2, color=colors[i])
+    axs[1].plot(-points.logX(), weights(points.logL, points.logX(), beta=beta), lw=.8, color=colors[i])
+axs[0].plot(-points.logX().iloc[-1], np.exp(points.logL.iloc[-1] - points.logL.max()), 'x', ms=2, color='k')
+
+l = .2
+arrow_start = (6, l)
+arrow_end = (2, l)
+arrow = patches.FancyArrowPatch(arrow_start, arrow_end, arrowstyle='->', mutation_scale=5)
+axs[0].add_patch(arrow)
+axs[0].annotate(r'Lower $\beta$', xy=(4, l), xytext=(4, l+.1), ha='center', va='bottom', fontsize=5)
+
+axs[1].axvline(x=-points.logX().iloc[ndead], ls='--', color='k', lw=.5)
+for ax in axs:
+    ax.margins(x=0.02)
+    ax.set_ylim(-0.01, 1.1)
+    ax.set_xticks([])
+    ax.set_yticks([])
[…]

### 2023-08-25 paper/figures/lcdm_predictions.py

@@ -1,0 +1,33 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+from aeons.utils import *
+from aeons.plotting import *
+
+figsettings()
+fig, axs = plt.subplots(3, 4, figsize=(7, 4))
+axs = axs.flatten()
+fig.delaxes(axs[-1])
+for i, chain in enumerate(lcdm_chains):
+    ax = axs[i]
+    axd = ax.twinx()
+    name, samples = get_samples('lcdm', chain)
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f"{aeons_dir}/data/predictions/lcdm/nos/{chain}_bt_25_nos.txt")
+    iterations_dG, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/lcdm/dG/{chain}_dG.txt')
+    logXf_true = samples.logX().iloc[int(true_endpoint[0])]
+    logXs = samples.logX().iloc[iterations]
+    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.3), ax=ax)
+    plot_std(-logXs, d_Gs, d_Gs_std, d_Gs[-1], ylim=(0, 2), ax=axd, color='lightsalmon')
+    axd.set_yticks([d_Gs[-1]], [f'{d_Gs[-1]:.1f}'])
+    axs[i].set_title(name, fontsize=8)
+# Make legend with custom patches
+from matplotlib.patches import Patch
+lines = [Patch(color='deepskyblue', label='Endpoints'), Patch(color='lightsalmon', label='Effective dimensionality')]
+fig.legend(handles=lines, ncol=1, fontsize=6, bbox_to_anchor=(0.97, 0.2))
+fig.tight_layout()
+fig.supxlabel('$-\\log X$')
+fig.supylabel('$-\\log \hat{X}_\mathrm{f}$', x=0.02)
+# Another ylabel on the right
+fig.text(1, 0.5, '$\\tilde{d}$', va='center')
+fig.tight_layout()
+fig.savefig('lcdm_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/live_point_distribution.py

@@ -1,5 +1,6 @@
 import numpy as np
 import matplotlib.pyplot as plt
+import matplotlib.patches as patches
 
 from aeons.utils import *
 from aeons.likelihoods import full
@@ -10,9 +11,12 @@ X = np.exp(logX)
 logL = full.func(X, theta)
 L = np.exp(logL - logL.max())
 LX = L * X
+Ztrue = np.trapz(L, X)
+LlogL = np.nan_to_num(L/Ztrue * np.log(L/Ztrue))
+DKL = np.trapz(LlogL, X)
 
-ndead = 5000
-nlive = 500
+ndead = 7000
+nlive = 1000
 
 def gaussian(x, mu, sig):
     return 1/(np.sqrt(2*np.pi)*sig)*np.exp(-np.power((x - mu)/sig, 2)/2)
@@ -21,12 +25,35 @@ StdlogXlivemin = np.sqrt(ndead/nlive**2 + (np.pi**2)/6)
 live_logX = logX[(logX > ElogXlivemin) * (logX < -ndead/nlive)]
 
 figsettings()
-plt.figure(figsize=(4,1.5))
-plt.plot(logX, L/L.max(), color='black', label='Likelihood', lw=.5)
-plt.fill(logX, LX/LX.max(), alpha=0.4, color='gray', label='Posterior')
-plt.fill_between(live_logX, np.zeros(len(live_logX)), np.ones_like(live_logX)/(np.log(nlive) + 0.577), alpha=0.5, color='deepskyblue', label='Live points')
-plt.plot(logX, gaussian(logX, ElogXlivemin, StdlogXlivemin), lw=.5, color='red', label='Minimum live point')
-plt.axvline(-ndead/nlive, color='deepskyblue', lw=.5, ls='--')
-plt.xlabel(r'$\log X$')
-plt.legend()
-plt.savefig('live_point_distribution.pdf', bbox_inches='tight')
+fig, ax = plt.subplots(figsize=(8, 3))
+ax.plot(logX, L/L.max(), color='black', label='Likelihood', lw=.5)
+ax.fill(logX, LX/LX.max(), alpha=0.4, color='gray', label='Posterior mass')
+ax.fill_between(live_logX, np.zeros_like(live_logX), 0.5*np.exp(live_logX + ndead/nlive), alpha=0.5, color='deepskyblue', label='Live points') 
[…]

### 2023-08-25 paper/figures/phase_transition.py

@@ -1,0 +1,20 @@
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import *
+figsettings()
+
+samples = pickle_in(f'{aeons_dir}/data/predictions/gauss/two.pkl')
+iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/gauss/two_logXfs.txt')
+iterations, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/two_d_Gs.txt')
+
+logXf_true = samples.logX().iloc[int(true_endpoint[0])]
+logXs = samples.logX().iloc[iterations]
+endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=705)
+
+fig, axs = plt.subplots(1, 2, figsize=(4, 1.2))
+plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0.6, 1.1), color='deepskyblue', ax=axs[0])
+plot_std(-logXs, d_Gs, d_Gs_std, d_Gs[-1], (0, 1.3), color='orange', ax=axs[1])
+axs[0].set_title('$-\\log \hat{X_\mathrm{f}}$')
+axs[1].set_title(f'$\\tilde{{d}}$')
+fig.supxlabel('$-\\log X$', y=-0.2)
+fig.savefig('phase_transition.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 paper/figures/prediction_uncertainty.py

@@ -1,0 +1,44 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+from aeons.utils import *
+from aeons.endpoint import EndModel
+from aeons.toy_samples import gaussian_samples
+from aeons.regress import analytic_lm_params, GaussianRegress
+from aeons.likelihoods import full
+
+samples = gaussian_samples(nlive=200, ndims=4, sigma=0.1, seed=20)
+model = EndModel(samples)
+ndead = 1200
+logL, X_mean, nk, logZdead = model.data(ndead)
+logX_mean = logX_mu(nk)
+
+logXarray = np.linspace(0, model.logX_mean[-1], 1000)
+Xarray = np.exp(logXarray)
+
+figsettings()
+fig, ax1 = plt.subplots(figsize=(4, 1.6))
+Nset = 3
+colors = ['navy', 'cornflowerblue', 'lightsteelblue']
+for i in range(Nset):
+    X = generate_Xs(nk)[ndead:]
+    regress = GaussianRegress(logL[ndead:], X)
+    logLmax, d, sigma = regress.theta
+    covtheta = regress.covtheta()
+    ax1.plot(-np.log(X), np.exp(logL[ndead:]), '+', ms=2, color=colors[i]) # samples to fit over
+    for _ in range(10):
+        t = np.random.multivariate_normal(regress.theta, covtheta)
+        ax1.plot(-logXarray, np.exp(full.func(Xarray, t)), lw=.1, color=colors[i]) # reconstructed likelihood
+
+ax1.plot([], [], '+', ms=1, color='navy', label='Live points')
+ax1.plot([], [], lw=.5, color='deepskyblue', label='Least squares fits')
+ax1.plot(-logXarray, np.exp(full.func(Xarray, [0, 4, 0.1])), lw=2, color='black', label='True likelihood') # true likelihood
+ax1.set_xticks([]);
+ax1.set_yticks([]);
+# ax1.set_xlim(logX_mean[-1], logX_mean[ndead])
+# ax1.set_ylim(np.exp(logL[ndead]), np.exp(logL[-1]))
+ax1.set_xlabel('$-\log X$')
+ax1.set_ylabel('$\mathcal{L}(X)$')
+ax1.margins(x=0, y=0)
+ax1.legend(fontsize=6)
+plt.savefig('prediction_uncertainty.pdf', pad_inches=0, bbox_inches='tight')

### 2023-08-25 aeons/endpoint.py
week 7-9 work

@@ -23,6 +23,9 @@ class EndModel:
         true_endpoint = self.true_endpoint(epsilon)
         return self.logX_mean[true_endpoint]
 
+    def points(self, ndead):
+        return points_at_iteration(self.samples, ndead)
+
     def data(self, ndead, live=False):
         points = points_at_iteration(self.samples, ndead)
         logL = np.array(points.logL)
@@ -48,7 +51,7 @@ class EndModel:
             logXf_i = reject_outliers(logXf_i)
             logXfs[i] = np.mean(logXf_i)
             logXfs_std[i] = np.std(logXf_i)
-            print(f"Iteration {ndead} complete, {len(logXf_i)} samples")
+            print('\r', f"Iteration {ndead} of {iterations[-1]}, {len(logXf_i)} samples", end='')
         return logXfs, logXfs_std
 
 def theta_basic(logL, X):
@@ -84,7 +87,7 @@ def theta_bandwidth_trunc(logL, X, trunc=15, **kwargs):
         except:
             attempts +=1
     if attempts > 1:
-        print(f"{attempts} attempts")
+        print(f"{attempts} attempts", end='\r')
     return theta
 
 def get_dlogZ(logZ):
@@ -106,7 +109,6 @@ def fit_dlogZ(dlogZ, deg):
     return coefs
 
 
-from aeons.endpoint import *
 class IncrementEndpoint:
     def __init__(self, samples, N_rolling):
         samples['logZs'] = np.logaddexp.accumulate(samples.logw())
@@ -157,7 +159,7 @@ class IncrementEndpoint:
         ax1.plot(index_pred, dlogZ_pred, color='orange', lw=1)
         ax1.axvline(x = iteration, lw=.5, ls='--', color='deepskyblue')
         ax1.set_ylim(0, dlogZ_rolling.iloc[0])
-        ax1.set_ylabel('$d\log Z$')
+        ax1.set_ylabel('$d\\log Z$')
 
         ax2.plot(dlogZ_rolling.index.get_level_values(0), dlogZ_rolling)
         ax2.plot(dlogZ_fit.index.get_level_values(0), dlogZ_fit, color='deepskyblue')

### 2023-08-25 aeons/increment.py

@@ -1,0 +1,65 @@
+import numpy as np
+import pandas as pd
+from aeons.endpoint import EndModel
+from aeons.regress import analytic_lm_params
+from aeons.likelihoods import full
+
+
+import numpy.polynomial.polynomial as poly
+def linear_regress(x, y):
+    coefs = poly.polyfit(x, y, deg=1)
+    return coefs
+
+def stopping_index(logZ_start, dlogZ_predict, epsilon=1e-3):
+    logZ_predicted = logZ_start + dlogZ_predict.cumsum()
+    logZ_tot = logZ_predicted[-1]
+    logZ_f = np.log(1 - epsilon) + logZ_tot
+    index_f = np.argmax([logZ_predicted > logZ_f])
+    return index_f
+
+class IncrementModel(EndModel):
+    # Inherit from EndModel
+    def __init__(self, samples, nlive=500):
+        super().__init__(samples)
+        self.nlive = nlive
+    
+    def true_delta_logZ(self, ndead):
+        """True change in logZ every <nlive> ndeads from <ndead> onward"""
+        logZs = self.logZs.iloc[ndead - self.nlive: len(self.samples): self.nlive]
+        return logZs.diff(1).dropna()
+    
+    def delta_logZ(self, ndead, steps=10):
+        if ndead < steps * self.nlive:
+            raise ValueError("ndead must be greater than steps*nlive")
+        logZs = self.logZs.iloc[:ndead:self.nlive]
+        return logZs.diff(1).dropna().iloc[-steps:]
+    
+    def linear_extrapolate(self, ndead, steps=10):
+        deltas = self.delta_logZ(ndead, steps)
+        coefs = linear_regress(deltas.index.get_level_values(0), deltas)
+        index_predict = np.arange(ndead - self.nlive, len(self.samples), self.nlive)
+        dlogZ_predict = poly.polyval(index_predict, coefs)
+        return pd.Series(index=index_predict[dlogZ_predict > 0], data=dlogZ_predict[dlogZ_predict > 0])
+    
+    def exponential_extrapolate(self, ndead, steps=10):
+        nlive = self.nlive
+        deltas = self.delta_logZ(ndead, steps)
+        deltas_index = deltas.index.get_level_values(0).values
+        x = np.linspace(0, 1, len(deltas))
+        theta = analytic_lm_params(np.log(deltas.values), x, d0=1)
+        index_predict = np.arange(ndead, 5*len(self.samples))
[…]

### 2023-08-25 aeons/plotting.py

@@ -14,7 +14,7 @@ def plot_lx(samples):
     L_norm = np.exp(logL_norm)
     ax1.plot(logX_mean, L_norm, lw=1, color='black')
     ax2.plot(logX_mean, L_norm*np.exp(logX_mean), lw=1, color='navy')
-    ax1.set_title(f'logL_max = {logL.max():.2f}')
+    ax1.set_title(f'$\\log\\mathcal{{L}}_\\mathrm{{max}}$ = {logL.max():.2f}')
 
 def plot_logLx(samples, ndead=0, lives=True, ax=None, N_points=50, ms=2):
     points = points_at_iteration(samples, ndead)
@@ -34,29 +34,30 @@ def plot_logLx(samples, ndead=0, lives=True, ax=None, N_points=50, ms=2):
         ax.plot(X_mean[index], logL[index], '+', ms=ms, color='navy')
     ax.ticklabel_format(style='sci')
 
-def plot_std(xvals, y_means, y_stds, true, ylim=None, ax=None):
+def plot_std(xvals, y_means, y_stds, true=None, ylim=None, ax=None, color='deepskyblue', label=None):
     # plt.plot(xvals, y_means, lw=1, color='navy')
     if ax is None:
         fig, ax = plt.subplots(figsize=(4,2))
-    ax.fill_between(xvals, y_means - y_stds, y_means + y_stds, alpha=1, color='deepskyblue')
-    ax.fill_between(xvals, y_means - 2*y_stds, y_means + 2*y_stds, alpha=.2, color='deepskyblue')
-    ax.axhline(y=true, lw=1, color='navy')
-    if isinstance(ylim, float):
-        ax.set_ylim(0, true*ylim)
-    elif isinstance(ylim, tuple):
-        ax.set_ylim(ylim[0]*true, ylim[1]*true)
+    ax.fill_between(xvals, y_means - y_stds, y_means + y_stds, alpha=1, color=color, label=label)
+    ax.fill_between(xvals, y_means - 2*y_stds, y_means + 2*y_stds, alpha=.2, color=color)
+    if true is not None:
+        ax.axhline(y=true, lw=.5, color='red', ls='--')
+        if isinstance(ylim, float):
+            ax.set_ylim(0, true*ylim)
+        elif isinstance(ylim, tuple):
+            ax.set_ylim(ylim[0]*true, ylim[1]*true)
 
     
[…]

### 2023-08-25 aeons/utils.py

@@ -5,8 +5,8 @@ from scipy.special import gamma, gammainc, logsumexp, gammaincinv
 
 proj_dir = '/home/zixiao/Documents/III/project'
 aeons_dir = '/home/zixiao/Documents/III/project/aeons'
-lcdm_chains = ['BAO', 'lensing', 'lensing_BAO', 'lensing_SH0ES', 'planck', 'planck_BAO', \
-              'planck_lensing', 'planck_SH0ES', 'planck_lensing_BAO', 'planck_lensing_SH0ES', 'SH0ES']
+lcdm_chains = ['BAO', 'lensing', 'lensing_BAO', 'lensing_SH0ES', 'SH0ES', 'planck', 'planck_BAO', \
+              'planck_lensing', 'planck_SH0ES', 'planck_lensing_BAO', 'planck_lensing_SH0ES']
 toy_chains = ["gauss_30_01", "wedding_20_001", "cauchy_10_0001", "planck_gaussian", "gp"]
 
 def pickle_dump(filename, data):
@@ -25,10 +25,12 @@ def pickle_in(filename):
 
 
 def write_to_txt(filename, data):
-    try:
-        open(filename, 'w').close()
-    except:
-        print(f'creating {filename}.txt')
+    import os
+    # Desired file path
+    # Create directories if they do not exist
+    directory = os.path.dirname(filename)
+    if not os.path.exists(directory):
+        os.makedirs(directory)
     with open(filename, 'a') as f:
         for item in data:
             if not np.shape(item):
@@ -44,7 +46,7 @@ def read_from_txt(filename):
     return data
 
 def get_samples(root, chain):
-    path = f"{aeons_dir}/samples/{root}/{chain}.pickle"
+    path = f"{aeons_dir}/data/samples/{root}/{chain}.pkl"
     samples = pickle_in(path)
     return chain, samples
 
@@ -63,26 +65,20 @@ def formatt(theta, sigfigs=2):
         s1, s2, s3 = sigfigs
         return f"[{logLmax:.{s1}e}, {d:.{s2}f}, {sigma:.{s3}e}]"
 
-def reject_outliers(data):
+def reject_outliers(data, degree=2):
     data = np.array(data)
     dev = np.abs(data - np.median(data))
     median_dev = np.median(dev)
-    return data[dev < 2 * median_dev]
+    return data[dev < degree * median_dev]
 
 def points_at_iteration(samples, ndead):
[…]

### 2023-08-25 notebooks/summer_5/runtime/run_pypolychord.py

@@ -12,11 +12,12 @@ from aeons.endpoint import theta_bandwidth_trunc
 #| The derived parameter is the squared radius
 
 nDims = 10
-nDerived = 0
+nDerived = 1
 sigma = 0.01
 
 def likelihood(theta):
     """ Simple Gaussian Likelihood"""
+
     nDims = len(theta)
     r2 = sum(theta**2)
     logL = -log(2*pi*sigma*sigma)*nDims/2.0
@@ -34,18 +35,18 @@ def prior(hypercube):
 #| the live points, dead points, weights and evidences
 
 def dumper(live, dead, logweights, logZ, logZerr):
+    # print(dead[-1])
     ndead = len(dead)
     nlive = len(live)
     if nlive == 0:
-        print(f"[{'='*50}>]", end='\n')
+        print(f"Endpoint: {ndead}")
+        print(f"[{'='*50}>] 100%", end='\n')
         return
     logL = live[:, -1]
     # Sort logL in ascending order
     logL = np.sort(logL)
     nk = np.concatenate([nlive * np.ones(ndead), np.arange(nlive, 0, -1)])
     X_mean = X_mu(nk)[ndead:]
-    # theta = analytic_lm_params(logL, X_mean, d0=2)
-    # theta = theta_bandwidth_trunc(logL, X_mean, trunc=5, splits=[1])
     Nset = 25
     logXfs = np.zeros(Nset)
     for i in range(Nset):
@@ -57,8 +58,8 @@ def dumper(live, dead, logweights, logZ, logZerr):
     endpoint, endpoint_std = calc_endpoints(len(dead), np.log(X_mean[0]), logXfs.mean(), logXfs.std(), nlive)
     end_predict = int(endpoint) + nlive
     # Print graphic showing progress of the run
-    print(f"Predicted endpoint: {end_predict} +/- {int(endpoint_std)}, progress {len(dead)/end_predict*100:.0f}%")
-    print(f"[{'='*int(len(dead)/end_predict*50)}>{'#'*(50-int(len(dead)/end_predict*50))}]", end='\n')
+    print(f"Predicted endpoint: {end_predict} +/- {int(endpoint_std)}, progress ")
+    print(f"[{'='*int(len(dead)/end_predict*50)}>{'#'*(50-int(len(dead)/end_predict*50))}] {len(dead)/end_predict*100:.0f}%", end='\n')
 
 #| Initialise the settings
 
@@ -69,7 +70,6 @@ settings.nlive = 500
[…]

### 2023-08-25 notebooks/summer_5/runtime/test.py

@@ -1,0 +1,79 @@
+from numpy import pi, log
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import UniformPrior
+try:
+    from mpi4py import MPI
+except ImportError:
+    pass
+
+
+#| Define a four-dimensional spherical gaussian likelihood,
+#| width sigma=0.1, centered on the 0 with one derived parameter.
+#| The derived parameter is the squared radius
+
+nDims = 4
+nDerived = 1
+sigma = 0.1
+
+def likelihood(theta):
+    """ Simple Gaussian Likelihood"""
+
+    nDims = len(theta)
+    r2 = sum(theta**2)
+    logL = -log(2*pi*sigma*sigma)*nDims/2.0
+    logL += -r2/2/sigma/sigma
+
+    return logL, [r2]
+
+#| Define a box uniform prior from -1 to 1
+
+def prior(hypercube):
+    """ Uniform prior from [-1,1]^D. """
+    return UniformPrior(-1, 1)(hypercube)
+
+#| Optional dumper function giving run-time read access to
+#| the live points, dead points, weights and evidences
+
+def dumper(live, dead, logweights, logZ, logZerr):
+    print(logZ)
+    print("Last dead point:", dead[-1])
+
+#| Initialise the settings
+
+settings = PolyChordSettings(nDims, nDerived)
+settings.file_root = 'gaussian'
+settings.nlive = 200
+settings.do_clustering = True
+settings.read_resume = False
+
+#| Run PolyChord
+
+output = pypolychord.run_polychord(likelihood, nDims, nDerived, settings, prior, dumper)
+
+#| Create a paramnames file
+
+paramnames = [('p%i' % i, r'\theta_%i' % i) for i in range(nDims)]
+paramnames += [('r*', 'r')]
+output.make_paramnames_files(paramnames)
+
+#| Make an anesthetic plot (could also use getdist)
+try:
+    from matplotlib import pyplot as plt
+    from anesthetic import read_chains
+    samples = read_chains(settings.base_dir + '/' + settings.file_root)
+    samples.plot_2d(['p0','p1','p2','p3','r'])
+    plt.savefig('posterior.pdf')
+
+except ImportError:
+    try:
+        import getdist.plots
+        posterior = output.posterior
+        g = getdist.plots.getSubplotPlotter()
+        g.triangle_plot(posterior, filled=True)
+        g.export('posterior.pdf')
+    except ImportError:
[…]

### 2023-08-25 notebooks/summer_7/d_prior.py

@@ -1,0 +1,66 @@
+from numpy import pi, log
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import GaussianPrior
+from aeons.utils import *
+from aeons.endpoint import EndModel
+from aeons.regress import params_from_d
+from aeons.likelihoods import full
+
+try:
+    from mpi4py import MPI
+except ImportError:
+    pass
+
+
+#| Define a four-dimensional spherical gaussian likelihood,
+#| width sigma=0.1, centered on the 0 with one derived parameter.
+#| The derived parameter is the squared radius
+
+nDims = 1
+nDerived = 0
+
+name, samples = get_samples('toy', 'gauss_16')
+model = EndModel(samples)
+ndead = 10000
+points = points_at_iteration(samples, ndead)
+logL, X_mean, nk, logZdead = model.data(ndead)
+logLd, Xd = logL[ndead:], X_mean[ndead:]
+N = len(logLd)
+beta = get_beta(points, ndead)
+dG = points.set_beta(beta).d_G(25).values
+dG_mean = dG.mean()
+dG_std = dG.std()
+
+
+def likelihood(d):
+    """ Simple Gaussian Likelihood"""
+    theta = params_from_d(logLd, Xd, d)
+    loss = logLd - full.func(Xd, theta)
+    L_sq = np.sum(loss**2)
+    s = np.sqrt(L_sq/N)
+    logPr = -1/2 * N * log(2*pi*s**2) - L_sq/(2*s**2)
+    return logPr, []
+
+def prior(hypercube):
+    """ Uniform prior from [-1,1]^D. """
+    return GaussianPrior(dG_mean, dG_std)(hypercube)
+
+#| Optional dumper function giving run-time read access to
+#| the live points, dead points, weights and evidences
+
+def dumper(live, dead, logweights, logZ, logZerr):
+    print(logZ)
+    print("Last dead point:", dead[-1])
+
+#| Initialise the settings
+
+settings = PolyChordSettings(nDims, nDerived)
+settings.file_root = 'gaussian'
+settings.nlive = 200
+settings.do_clustering = True
+settings.read_resume = False
+
+#| Run PolyChord
+
+output = pypolychord.run_polychord(likelihood, nDims, nDerived, settings, prior, dumper)

### 2023-08-25 notebooks/summer_7/spherical_gaussian.py

@@ -1,0 +1,44 @@
+#%%
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.endpoint import EndModel, theta_bandwidth_trunc
+from aeons.plotting import *
+figsettings()
+
+# %%
+from aeons.toy_samples import gaussian_samples
+samples = gaussian_samples(500, 16, 0.01, seed=10)
+
+# %%
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+
+# %%
+samples = pickle_in('gauss/gauss_8.pkl')
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+iterations = make_iterations(true_endpoint, 50)
+logXfs, logXfs_std = model.logXfs(theta_bandwidth_trunc, iterations, Nset=25, splits=1)
+
+# %%
+logXs = samples.logX().iloc[iterations]
+endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=500)
+plot_std(iterations, endpoints, endpoints_std, true_endpoint, ylim=(0, 1.6))
+plt.title(f"DKL = {samples.D_KL():.2f}")
+
+#%%
+pickle_dump('gauss_16.pkl', samples)
+write_to_txt('gauss_16.txt', [iterations, logXfs, logXfs_std, true_endpoint])
+
+# %%
+ds = [8, 16, 32]
+fig, axs = plt.subplots(1, 3, figsize=(8, 1.5))
+for i, d in enumerate(ds):
+    samples = pickle_in(f'gauss/gauss_{d}.pkl')
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'gauss/gauss_{d}.txt')
+    logXs = samples.logX().iloc[iterations]
+    endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=500)
+    plot_std(iterations, endpoints, endpoints_std, true_endpoint, ylim=(0, 1.5), ax=axs[i])
+    axs[i].set_title(f'$d={d}$')
+fig.tight_layout()

### 2023-08-25 notebooks/summer_7/d_prior.py
removed old notebooks

@@ -1,66 +1,0 @@
-from numpy import pi, log
-import pypolychord
-from pypolychord.settings import PolyChordSettings
-from pypolychord.priors import GaussianPrior
-from aeons.utils import *
-from aeons.endpoint import EndModel
-from aeons.regress import params_from_d
-from aeons.likelihoods import full
-
-try:
-    from mpi4py import MPI
-except ImportError:
-    pass
-
-
-#| Define a four-dimensional spherical gaussian likelihood,
-#| width sigma=0.1, centered on the 0 with one derived parameter.
-#| The derived parameter is the squared radius
-
-nDims = 1
-nDerived = 0
-
-name, samples = get_samples('toy', 'gauss_16')
-model = EndModel(samples)
-ndead = 10000
-points = points_at_iteration(samples, ndead)
-logL, X_mean, nk, logZdead = model.data(ndead)
-logLd, Xd = logL[ndead:], X_mean[ndead:]
-N = len(logLd)
-beta = get_beta(points, ndead)
-dG = points.set_beta(beta).d_G(25).values
-dG_mean = dG.mean()
-dG_std = dG.std()
-
-
-def likelihood(d):
-    """ Simple Gaussian Likelihood"""
-    theta = params_from_d(logLd, Xd, d)
-    loss = logLd - full.func(Xd, theta)
-    L_sq = np.sum(loss**2)
-    s = np.sqrt(L_sq/N)
-    logPr = -1/2 * N * log(2*pi*s**2) - L_sq/(2*s**2)
-    return logPr, []
-
-def prior(hypercube):
-    """ Uniform prior from [-1,1]^D. """
-    return GaussianPrior(dG_mean, dG_std)(hypercube)
-
-#| Optional dumper function giving run-time read access to
-#| the live points, dead points, weights and evidences
-
-def dumper(live, dead, logweights, logZ, logZerr):
-    print(logZ)
-    print("Last dead point:", dead[-1])
-
-#| Initialise the settings
-
-settings = PolyChordSettings(nDims, nDerived)
-settings.file_root = 'gaussian'
-settings.nlive = 200
-settings.do_clustering = True
-settings.read_resume = False
-
-#| Run PolyChord
-
-output = pypolychord.run_polychord(likelihood, nDims, nDerived, settings, prior, dumper)

### 2023-09-27 paper/paper.tex
third draft created

--- 2023-08-07 paper/paper.tex
@@
 \usepackage{algpseudocode}
-\usepackage{caption}
 \usepackage{subcaption}
 \usepackage{layouts}
-\usepackage{cleveref}
+\usepackage{hyperref}
+\usepackage[capitalise]{cleveref}
+\usepackage{fancyvrb}
+\usepackage{xcolor}
 \usepackage{txfonts}
 \usepackage[T1]{fontenc}
+%}}}
 
-\newcommand{\logL}{\{ \log \Like_i \}}
-\newcommand{\thetab}{\bm{\theta}}
-\newcommand{\logLm}{\log \Like_\mathrm{max}}
-\newcommand{\set}[1]{\{#1\}}
 \newcommand{\nlive}{n_\mathrm{live}}
@@
 \begin{abstract}
-This paper develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration for any nested sampler.
-The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model likelihood, which allows the remaining evidence and prior volume at termination to be calculated.
-We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods.
-For non-Gaussian likelihoods, the estimate is still the correct order of magnitude and converges to the correct answer by the end of the run.
 \end{abstract}
@@
 \section{Introduction}
-Users would like a order of magnitude runtime estimate
-Figure of PolyChord output
+Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}.
+It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}.
+The core algorithm is unique for estimating volumes by \textit{counting}, which makes high-dimensional integration feasible.
+It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
+\par
+The order of magnitude runtime of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user.
+Currently, existing implementations of nested sampling either do not give any indication of the runtime, or only provide crude estimates.
+In the latter case, these are always underpredictions which usually fail to recover the correct order of magnitude.
+\par
+This paper sets out a more principled manner of endpoint estimation for nested sampling at each intermediate stage, the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from.
+Outline of paper.
+\begin{figure}
+\begin{Verbatim}[frame=single, commandchars=\\\{\}]
+\textcolor{red}{Predicted endpoint: 25054 +/- 242}
+\textcolor{red}{Progress: [=================>########] 72%}
+___________________
+lives      |   500 |
+phantoms   | 24310 |
+posteriors | 18018 |
+equals     |   245 |
+-------------------
+ncluster   =  1/1
+ndead      =  18018
+nposterior =  18018
+nequals    =  249
+nlike      =  4159049
+<nlike>    =  491.04 (9.82 per slice)
+log(Z)     =  -12.55 +/- 0.27
+\end{Verbatim}
+\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
+\label{fig:polychord_output}
+\end{figure}
 
 \section{Background}
-\subsection{Notation}
-We shall begin with a brief description of nested sampling to establish the necessary notation.
-For a given likelihood $\Like(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
-\begin{equation}\label{eq:evidence}
-	\mathcal{Z} = \int \Like(\theta)\ \pi(\theta)\ \mathrm{d}\theta
-\end{equation}
-while producing samples of the posterior distribution 
-\begin{equation}
-	\mathcal{P}(\theta) = \frac{\Like(\theta) \pi(\theta)}{\mathcal{Z}}.
-\end{equation}
-The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run.
-At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}.
-A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point.
-Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
+
+\subsection{The progress of a nested sampling run}
+Present some theoretical results on anatomy of nested sampling run.
+
+\subsection{Current prior volume}
+
+\subsection{Current log-likelihood}
+
+\subsubsection{Dimensionality}
+Define Bayesian model dimensionality
+Goal to observe how the number of constrained parameters changes throughout the run
+Only works at the end of the run, because BMD is a property of the posterior - heat capacity analogy
+Need to transform intermediate set of samples into a posterior. Useful tool is the transformation using beta. 
+Which beta to use? Several existing temperatures in the literature - microcanonical temperature, canonical temperature used by Habeck.
+Propose a temperature from Bayesian inference.
+
+Ridge geometries contain phase transitions in nested sampling, because the dimensionality of the samples increases throughout the run. 
 \par
-The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$, defined as
-\begin{equation}
-	\mathcal{Z} = \int_0^1 \Like(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \Like_i (X_{i-1}-X_{i+1}),
-\end{equation}
-where $X(\Like)$ is the fraction of the prior greater than $\Like$.
-The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
-\begin{equation}
-	X_i = \prod_{k=1}^i t_k.
-\end{equation}
-The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
-\begin{equation}\label{eq:t_dist}
-	p(t_i) = \nlive t_i^{\nlive-1}, \quad \mathrm{E}[\log t_i] = -\frac{1}{\nlive}, \quad \mathrm{Var}[\log t_i] = \frac{1}{\nlive^2}.
-\end{equation}
-The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g.
-$10^{-3}$.
-The remaining live points are then killed off one by one without replacement and added to the evidence.
+As an example, consider an elongated Gaussian in a unit hypercube.
+
+\section{Endpoint prediction}
+Discussion of how to get endpoint and the necessity of extrapolation. Compare to UltraNest/dynesty slab methods and why they underpredict.
 \par
-Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.
+The key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian.
+Fitting a Gaussian of this dimension to the likelihood profile should be a reasonable approximation to the true distribution.
+\par
+Show fits
+Find BMD
+Fit data to find sigma
+Extrapolate using a Gaussian likelihood with this parameterisation
 
-\section{Convergence of a nested sampling run}
-Let us now take a top-down view of a nested sampling run as it progresses to observe how it converges.
+\bibliographystyle{mnras}
+\bibliography{references}
 
-\subsection{Posterior convergence}
-We will examine how our set of samples evolve over the course of a nested sampling run.
-The conventional perspective for the progress of a run is with $\log X$ along the x-axis; the prior volume is compressed at a steady rate, only picking up the bulk of the integral/posterior at around $\log X = \DKL$.
-\par
-At an intermediate iteration, we have an accurate reconstruction of the posterior only down to the current prior volume $\log X_*$.
-The live points contain information below this volume, but not much more.
-If we kill off all the live points as if we were terminating, then by \cref{eq:t_dist} the prior volume of the final live point $\log X_\mathrm{min}^{\mathrm{live}}$ has mean and variance
-\begin{equation}
-	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
-\end{equation}
-\begin{equation}
-	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6},
-\end{equation}
-where the large $\nlive$ limit is taken for the approximation and $\gamma$ is the Euler-Mascheroni constant. 
-\par
-This shows that the live points only get us a factor of $\log \nlive$ closer to the posterior bulk, or in other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples begin to resemble the posterior.
-Intuitively, if the likelihood is sharply peaked then for most of the run the live points are too diffuse to land there with any significant probability.
-\par
-\begin{figure}
-\begin{center}
-	\includegraphics{figures/live_point_distribution.pdf}
-\end{center}
-\caption{}
-\label{fig:}
-\end{figure}
-
-\subsection{Loglikelihood convergence}
-We now offer an alternate view of the convergence of a nested sampling run, via the loglikelihood instead of the prior volumes.
-To get an insight into the analytics, work with a Gaussian likelihood of $d$ dimensions and a lengthscale of $\sigma$;
-\begin{equation}\label{eq:logL}
-	\log\Like = \log\Like_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
-\end{equation}
-The distribution of the true posterior in $\log\Like$ is
-\begin{equation}
-    P(\log\mathcal{L}) = \frac{1}{\Gamma(\frac{d}{2})}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
-\end{equation}
-i.e.
-$2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the same distribution as that of the dead points up to the latest dead likelihood $\log \Like_*$.
-Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
-\begin{equation}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\Like_* < \log\Like <\log\Like_\mathrm{max}],
-    \label{eq:PL}
-\end{equation}
-It is helpful at this stage to define a parameter
-\begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*} 
-    \label{eq:y}
-\end{equation}
-as a normalised measure of how far a point is between the latest dead point and the maximum loglikelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
-\begin{equation}
-    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
-    \label{eq:Py}
-\end{equation}
-We now seek the distribution for the maximum likelihood of the live points, $\log\Like_\mathrm{max}^{\mathrm{live}}$.
-Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
-\begin{equation}
-    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
-    \label{eq:Pyhat}
-\end{equation}
-which in the limit of large live points and dimensions can be roughly summarised by
-\begin{equation}\label{eq:ylivemax}
-	    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
-\end{equation}
-This shows that in general the live points are nowhere near the maximum loglikelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
-In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
-\par
-This remains true even at the end of the nested sampling run. To see this, we write the halting condition as:
-\begin{equation}
-    f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L} dX}{\int_0^\infty \mathcal{L} dX}.
-    \label{eq:fint}
-\end{equation}
-Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $f \ll 1$ so that the denominator is approximately the accumulated evidence.
-Computing this for \cref{eq:logL} we find the answer in terms of lower incomplete gamma functions
-\begin{equation}
-    f = 1- \frac{\Gamma_{d/2}\left(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2}\right)}{\Gamma\left(\frac{d}{2}\right)}.
-    \label{eq:f}
-\end{equation}
-Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
-\begin{equation}
-    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} f \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d\Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
-\end{equation}
-We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $f$.
-This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a Stirling approximation:
-\begin{equation}
-    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
-\end{equation}
-In the event that we keep $f$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}f^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
-
-\subsection*{Observation 3: nested sampling as a maximiser}
-Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
-\begin{equation}
-    \boxed{
-        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
-    },
-\end{equation}
-showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum loglikelihood, and the final set of $n$ live points can get you $\log (n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
-\begin{figure*}
-\begin{center}
-	\includegraphics{figures/anatomy.pdf}
-\end{center}
-\caption{}
-\label{fig:anatomy}
-\end{figure*}
-\par
-From above analysis we learn that
-\begin{itemize}
-	\item Knowledge of the posterior early on in the run is limited by the maximum live point, which is far from the posterior bulk
-	\item At termination, the maximum live point is far from the true maximum --> nested sampling alone is a poor optimiser (but great for finding starting point of gradient descent)
-\end{itemize}
-
-\section{Method}
-Let us now tackle the problem of estimating how many steps we still have to take in a nested sampling run before we can terminate.
-\par
-The above analysis shows that our knowledge of the posterior at an intermediate iteration is limited by the maximum live point, which until just before the end is quite far from the posterior bulk.
-However, we can get an approximate image of the full posterior ahead of time, simply by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points.
-\par
-For practical purposes, one would never use this approximate posterior to do inference.
-However, it is more than sufficient for making a prediction for roughly how much evidence remains, and hence the point at which the run should terminate.
-Quantitatively, this proceeds as follows: fitting a model function $f(X, \thetab)$ to the known likelihood profile allows us to express the prior volume we need to compress to as
-\begin{equation}
-	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
-\end{equation}
-\begin{equation}\label{model endpoint}
-	\int_0^{X_f} f(X, \thetab)\ dX = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ dX + Z_\mathrm{dead} \right),
-\end{equation}
-where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point.
-$X_f$ can then be identified by solving the above equation either analytically or numerically.
-If $\nlive$ is constant then at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_f$ will be
-\begin{equation}
-	N_f = - \nlive \log X_f .
-\end{equation}
-One has freedom to choose the form of the fitting function $f(X, \thetab)$.
-Following the previous analysis, we use a Gaussian profile, which has the advantage that the regression itself can be sped up analytically, as well as the fact that many real likelihoods are Gaussian or near-Gaussian.
-However, any other function that is sufficiently flexible to capture the shape of the likelihood profile will do.
-
-Fit a Gaussian profile, with advantages that a) Gaussian error common b) analytic optimisation 
-Extraction of termination logX
-Prediction uncertainty dominated by correlations, not least squares i.e. scatter of individual points - get by simulating X
-
-\section{Results}
-Toy examples - pure Gaussian,  wedding cake, Cauchy
-Real cosmological examples - Planck
-
+\label{lastpage}
 \end{document}

### 2023-09-27 paper/second.tex
third draft created

%{{{ Preamble
\documentclass[usenatbib]{mnras}
\usepackage{graphicx}
\usepackage[export]{adjustbox}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{subcaption}
\usepackage{layouts}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{fancyvrb}
\usepackage{xcolor}
\usepackage{txfonts}
\usepackage[T1]{fontenc}
%}}}

%{{{ Commands
\newcommand{\logL}{\{ \log \Like_i \}}
\newcommand{\thetab}{\bm{\theta}}
\newcommand{\logLm}{\log \Like_\mathrm{max}}
\newcommand{\set}[1]{\{#1\}}
\newcommand{\nlive}{n_\mathrm{live}}
\newcommand{\Like}{\mathcal{L}}
\newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
%}}}

\title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
\author[Z. Hu et. al]{Zixiao Hu, Artyom Baryshnikov, Will Handley}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

\begin{abstract}
This paper develops a technique to estimate the runtime of a nested sampling run at an intermediate iteration which works for any nested sampling implementation. The known likelihood at an iteration is fitted, then extrapolated using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be calculated. We find that the method succesfully recovers the true endpoint within standard error from the beginning for near-Gaussian likelihoods. For non-Gaussian likelihoods, the estimate is still the correct order of magnitude and converges to the correct answer by the end of the run.
\end{abstract}

\begin{keywords}
methods: data analysis -- methods: statistical
\end{keywords}

\section{Introduction}
Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously accomplishes the two tasks of Bayesian inference, model comparison and parameter estimation \citep{skilling}. It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}. The core algorithm is unique for estimating volumes using \textit{order statistics}, which makes high-dimensional integration feasible. It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
\par
Many instances of software for nested sampling exist, including popular implementations such as \textsc{MultiNest} \citep{multinest} and \textsc{PolyChord} \citep{polychord}, as well as post-processing packages such as \textsc{anesthetic} \citep{anesthetic} to plot sample statistics and distributions. However, no such software is currently able to estimate how long a given run should last. This is of high importance to the end user, who has no indication in the middle of a run whether it will take hours, or weeks and months to finish; even an order of magnitude estimate is useful compared to the status quo of complete ignorance.
\par
This project sets out a novel approach for estimating the endpoint of a nested sampling run at an intermediate stage, an example of which is shown in \cref{fig:polychord_output}. The idea is to predict the form of the likelihood function in the region we have yet to sample from, using the information we have already gained from our previous samples. We begin in section \cref{sec:Background} with an overview of nested sampling, before discussing its convergence properties in \cref{sec:convergence}. \cref{sec:method} presents the methodology of making endpoint predictions, before results for both toy and real examples are shown in section \cref{sec:results}. Finally, conclusions are made in section \cref{sec:conclusions}.
\par
Users would like a order of magnitude runtime estimate
\begin{figure}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
\textcolor{red}{Predicted endpoint: 25054 +/- 242}
\textcolor{red}{Progress: [=================>########] 72%}
___________________
lives      |   500 |
phantoms   | 24310 |
posteriors | 18018 |
equals     |   245 |
-------------------
ncluster   =  1/1
ndead      =  18018
nposterior =  18018
nequals    =  249
nlike      =  4159049
<nlike>    =  491.04 (9.82 per slice)
log(Z)     =  -12.55 +/- 0.27
\end{Verbatim}
\caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
\label{fig:polychord_output}
\end{figure}
\section{Background}\label{sec:Background}
Let us begin with a brief description of the nested sampling algorithm and several adjacent concepts to establish the necessary notation. 
\subsection{Nested sampling}
For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
\begin{equation}\label{eq:evidence}
	\mathcal{Z} = \int \mathcal{L}(\theta)\ \pi(\theta)\ \mathrm{d}\theta
\end{equation}
while producing samples of the posterior distribution
\begin{equation}
	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
\end{equation}
The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run. At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}. A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point. Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
\par
The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$, defined as
\begin{equation}
	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
\end{equation}
where $X(\mathcal{L})$ is the fraction of the prior greater than $\mathcal{L}$. The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
\begin{equation}\label{eq:X_dist}
	X_i = \prod_{k=1}^i t_k.
\end{equation}
The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
\begin{equation}\label{eq:t_dist}
	P(t_i) = \nlive t_i^{\nlive-1}, \quad \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
\end{equation}
The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g. $10^{-3}$. The remaining live points are then killed off one by one without replacement and added to the evidence.
\par
Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.
\par
The time complexity of nested sampling, as analysed in \citet{supernest}, is
\begin{equation}
    T \propto \nlive \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle \mathcal{T}\{ \mathrm{Impl.}\} \rangle \times \mathcal{D}_\pi \{ \mathcal{P} \}.
\end{equation}
For a given resolution of $\nlive$, the runtime is therefore proportional to the average time for a likelihood evaluation, the implementation-dependent average time to sample a higher likelihood live point, and the information gain from prior to posterior. The first two of these are typically constant over the course of a run, so the runtime is primarily determined by the compression factor $\mathcal{D}$.

\subsection{Thermodynamic connection}\label{sec:thermodynamic}
The thermodynamic generalisation of Bayes' theorem is given by
\begin{equation}\label{eq:thermo_bayes}
    \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta) \pi(\theta)}{\mathcal{Z}(\beta)},
    \quad
    \mathcal{Z}(\beta) = \int \mathcal{L}(\theta)^\beta \pi(\theta)\ \mathrm{d}\theta,
\end{equation}
where $\beta$ is the inverse temperature $1/T$. If each $\theta$ is a microstate, then $\log \mathcal{L}$ becomes the negative energy, $\pi$ becomes the density of states, and $\mathcal{Z}$ naturally the partition function. The advantage of nested sampling for computing partition functions is that it compresses at a constant thermodynamic speed by taking samples equally spaced in volume entropy, which sidesteps the problem of designing an annealing schedule \citep{statmech}.

\subsection{Bayesian model dimensionality}\label{sec:dimensionality}
Following the thermodynamic connection, an important construct is the analogue of the heat capacity, which is the \textit{Bayesian model dimensionality}, defined as the posterior variance of the Shannon information: 
\begin{equation}
\frac{\tilde{d}}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
= \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
\end{equation}
Just as the heat capacity is a measure of the number of degrees of freedom in a system, the Bayesian model dimensionality is a measure of the number of constrained parameters in a set of samples.

\section{The progress of a nested sampling run}\label{sec:convergence}
We will now present several useful ways of measuring the progress of a nested sampling run, together with an assortment of results we shall use to make endpoint predictions. At a given iteration $i^{*}$, these are the current prior volume $\log X^{*}$ of the latest dead point, the corresponding log-likelihood $\log\Like^{*}$, and the current effective temperature $\beta^{*}$ of the samples. 
\subsection{Current prior volume}\label{sec:volume_convergence}
The conventional perspective for the progress of a run is with $\log X$ along the x-axis, where the prior volume is compressed at a steady rate until we reach the posterior mass. Defining the Kullback-Leibler divergence as
\begin{equation}\label{eq:DKL}
	\DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta,
\end{equation}
we interpret it as the posterior averaged logarithmic \textit{volume compression} from the prior to the posterior. Hence we can define a  \textit{posterior bulk} at below $X = e^{-\DKL}$ containing nearly all of the information, which gives us a measure of the volume we must compress down for the algorithm to converge.
\par
At an intermediate iteration $i_*$, we only have samples of the posterior only down to the current prior volume $\log X_*$. We do have some information below this volume due to the live points, but we will see now that they do not bring us much further. If we kill off all the live points as if we were terminating, then by \cref{eq:t_dist} the prior volume of the final live point $\log X_\mathrm{min}^{\mathrm{live}}$ has mean and variance given by
\begin{equation}
	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
\end{equation}
\begin{equation}
	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6},
\end{equation}
where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
\par
This shows that the live points only get us a factor of $\log \nlive$ closer to the posterior bulk. In other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples look anything like the posterior. One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities (such as the Planck legacy chains, which have a $\DKL$ of $\sim 40$) and typical live point numbers of  $\sim 1000$, this does not happen until near the end of the run.
\par
Intuitively, this is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
\begin{figure*}
\begin{center}
	\includegraphics{figures/live_point_distribution.pdf}
\end{center}
\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the prior $\pi(\log X \mid \log X < \log X_*)$ and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i_*$. For large values of $\DKL$ i.e. informative posteriors and/or large dimensionalities, the maximum live point is very far from the posterior bulk until the very end of the run. Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller. Plots of the sort from here onwards will be in terms of $-\log X$, where the run will more naturally proceed from rleft to right. }
\label{fig:live_point_distribution}
\end{figure*}

\subsection{Current log-likelihood}
We can also view the run as a function of the current log-likelihood value $\log\mathcal{L}^{*}$, which of course monotonically increases throughout the run. To illustrate some of the analytics, let us take the example of a Gaussian likelihood with dimension $d$ and lengthscale $\sigma$;
\begin{equation}\label{eq:logL}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
\end{equation}
$\log \mathcal{L}^{*}$ must increase until the posterior bulk is passed, which is located around the posterior averaged log-likelihood 
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Given the value of the latest dead point, we can examine the distribution of the live points in likelihood. Define the normalised likelihood
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
    \label{eq:normalised_likelihood}
\end{equation}
as a measure of how far a point is between the current likelihood and the maximum; $y=0$ corresponds to $\log\mathcal{L}_*$ and $y=1$ to $\log\mathcal{L}_\mathrm{max}$. It follows that the live points are distributed as 
\begin{equation}
    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1],
    \label{eq:Py}
\end{equation}
and the maximum live point has mean and variance
\begin{equation}\label{eq:ylivemax}
	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}
\end{equation}
in the limit of large $d$ and $\nlive$ (see appendix for derivation). We therefore see that just as with the prior volumes, in high dimensions the live points only bring us logarithmically further, while $d$ pushes us linearly away.

\subsubsection*{Nested sampling as a maximiser}
Previous literature has explored the potential of nested sampling algorithms as maximisers \citep{Akrami_2010, Feroz_2011}. The latter authors emphasised the terminatioin criterion controls the global maximum obtained. 
\par
We can use these results to show that nested sampling by itself is a poor maximiser. If we set $\log\Like^{*}$ to be the termination point $\log\Like_\mathrm{end}$, we can invert \cref{eq:normalised_likelihood} to find that the maximum live point is located at
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
    }.
\end{equation}
Hence in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood, and the final set of $n$ live points can get you $(\log n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation. 
\par
We find that in higher dimensions (typically greater than ten), the maximum live point is nowhere near the true maximum! However, these shortcomings can be mitigated by using the maximum point in the chain as a \textit{starting point} for a gradient-based optimisation algorithm, which would allow us to leverage the advantages of nested sampling to navigate multimodal landscapes while also rapidly approach the maximum point.
\begin{figure*}
\begin{center}
	\includegraphics{figures/anatomy.pdf}
\end{center}
\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_*)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$}
\label{fig:anatomy}
\end{figure*}

\subsection{Current temperature}\label{sec:current_temperature}
As previously discussed, nested sampling chooses an annealing schedule that moves at constant thermodynamic speed, which constructs a gradual transition from prior ($\beta = 0$) to posterior ($\beta = 1$). One can expect then that any given point during a run can be associated with a $\beta \in [0,1]$ that measures how far the current set of samples lies between prior and posterior. 
\par
We present here the idea that one can assign each iteration of a nested sampling run $i^{*}$ with a \textit{current temperature} $\beta^*$.
\par
As noted by Skilling, the posterior with respect to $\log X$ $\Like^{\beta} X$ has a peak at $\log X^*$ when  
\begin{equation}
    \beta^* = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like}
\end{equation}

\subsubsection*{Choice of temperature mapping}
The typical behaviour at an intermediate stage of a nested sampling run is that the posterior mass is dominated by a single point. This is because the highest live point is typically several log-likelihoods, and therefore orders of magnitude higher in weight than the next point. Decreasing $\beta$, equivalent to increasing the temperature, causes the mass to redistribute more equally across the samples. A natural choice for the present temperature $\beta^{*}$ is therefore one where the posterior mass is concentrated around the current dead point. What that constitutes is subjective, and several plausible choices are presented below:
\begin{itemize}
    \item Setting the posterior averaged log-likelihood $\langle \log\Like \rangle_{\mathcal{P}(\beta)}$ to be equal to the current $\log\Like^{*}$.
    \item Setting the Kullback-Leibler divergence of current set of samples $\mathcal{D}(\beta)$ to be equal to the current $-\log X^{*}$.
    \item Setting $\beta$ so that the set of samples meets some criterion for the remaining posterior mass, for example so that a specified fraction $\epsilon$ of the total mass lies beyond the current point.
\end{itemize}
Note that the second condition is a special case of the third, since $\DKL = \langle \log\Like \rangle_\mathcal{P} - \log \mathcal{Z} = -\log X^{*}$ corresponds to $\mathcal{Z} \sim \mathcal{L}X$ so that $\epsilon \sim 0.5$. One could account for the uncertainty in choice by sampling uniformly between the $\beta^{*}$ that correspond to fractions $\epsilon$ and $1 - \epsilon$ of the posterior mass remaining, since all such choices have a posterior mass 
\par

\subsubsection*{Inferring the current temperature}
Transforming $\mathcal{L} \to \Like^{\beta}$ leads to the posterior distribution for $\log X$ to instead follow
\begin{equation}
    \mathcal{P}\left(\log X \mid \beta\right) = \frac{\Like^{\beta} X}{\int \Like^{\beta} X \ \mathrm{d}\log X}.
\end{equation}
Using Bayes' rule one can get reverse, which is the probability of the temperature given a set of $\log X$;
\begin{equation}
    P\left(\beta \mid \log X\right) \propto \mathcal{P}\left(\log X \mid \beta\right)P\left(\beta\right).
\end{equation}
Evaluating the left hand side for $\log X = \log X^{*}$ then gives the distribution of $\beta$ at the current iteration. We note that we would like $P\left(\beta = 0 \mid \log X^{*}\right) = 0$ since the likelihood must always increase, but $\Like^{\beta} X^{*} = X^{}$

\subsubsection*{Intermediate sample dimensionalities}
One problem with the Bayesian model dimensionality is that it is practically zero for most of the run, only converging at the very end when the posterior mass is no longer concentrated at a single point with zero variance. From the thermodynamic perspective, this is expected; the present number of degrees of freedom only corresponds to the heat capacity at the present, not the final temperature. The number of parameters constrained at an intermediate stage should really be calculated from the \textit{transformed} posterior, not the original set of samples.
\par
Plotted below are the temperature adjusted sample dimensionalities for a spherical Gaussian, for which the ground truth is known.

\begin{figure}
\begin{center}
	\includegraphics{figures/last_live_point.pdf}
\end{center}
\caption{The likelihood profile and posterior mass as $\beta$ is lowered. The original likelihood is shown in black, which is dominated by the maximum live point. As $\beta$ decreases, the posterior mass is shifted back from the maximum point to the other points, until so much of the mass has shifted that the present point $\log X^{*}$ is the termination point. The transformed likelihood and posterior at this point is shown in orange.}
\label{fig:last_live_point}
\end{figure}

If this quantity is computed from the untransformed samples, it will output zero since there is effectively only one sample. The transformed posterior, on the other hand, gives a $\tilde{d}$ which is consistent with the distribution of live points, shown in \cref{fig:dG_beta} at various stages for an elongated Gaussian. 
\begin{figure}
\begin{center}
	\includegraphics{figures/dG_beta.pdf}
\end{center}
\caption{The Bayesian model dimensionality $\tilde{d}$ as a function of $\beta$ for a Gaussian two parameters of variance $10^{-3}$ and $10^{-6}$. The number of contrained samples can be seen to increase from zero to the true dimensionality of 2, as the run progresses.}
\label{fig:dG_beta}
\end{figure}

Use beta to calculate the current Bayesian model dimensionality, the number of parameters constrained by the current set of samples. 
\par
Plot of the model dimensionality throughout the run for an elongated Gaussian, comparing to that obtained with beta = 1. Find that it matches up well with the apparent dimension of the current set of samples. 
\par
Find that a feature of nested sampling is that parameters with high variance are initially `hidden'. Compression occurs in the direction which is most likely to have a sample of higher likelihood. Much easier to find a better point along the direction of a parameter that is initially poorly constrained. So compressions do not occur in parameters that are already well-constrained; appears as though they are uniformly distributed.

\section{Method}\label{sec:method}
The analysis in \cref{sec:volume_convergence} shows that our knowledge of the posterior at an intermediate iteration is limited by the maximum log-likelihood live point, which until just before the end is quite far from the posterior bulk. However, we can get an approximate image of the full posterior ahead of time, simply by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\mathcal{L}(X)$ traced out by the live and dead points.
\par
One would never use this approximate posterior to do inference, since the extrapolated profile is much less exact than if one were to simply finish the run. However, it is more than sufficient for making a prediction for roughly how much evidence remains, and hence the point at which the run should terminate. Quantitatively, this proceeds as follows: we fit a model function $f(X, \thetab)$ with some parameters $\thetab$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
\begin{equation}
	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
\end{equation}
\begin{equation}\label{endpoint}
	\int_0^{X_\mathrm{f}} f(X, \thetab)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \thetab)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
\end{equation}
where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point. $X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically. 
\par
Once $X_\mathrm{f}$ is known, the corresponding iteration count depend on the live point schedule. The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_\mathrm{f}$ will be
\begin{equation}
	N_\mathrm{f} = - \nlive \log X_\mathrm{f} .
\end{equation}

\subsection{Choosing the extrapolation}
One has freedom to choose the form of the fitting function $f(X, \thetab)$. Following the previous analysis, we use a Gaussian profile in the form of \cref{eq:logL}. It has the advantage that one can speed up its regression analytically, as well as the fact that many real likelihoods are Gaussian or near-Gaussian. We allow $d$ to freely vary to increase flexibility, and reflect the fact that while the dimensionality of a likelihood is fixed, the equivalent Gaussian dimensionality may differ \citep{Handley_2019}. Regardless of motivation, the proof is in the pudding, so we shall justify the choice on the accuracy of predictions in \cref{sec:results}.
\par
We must also choose the set of points over which we fit. Using all of the points we have is a poor choice, because unless the profile is actually Gaussian the extrapolation will be insensitive to changes in the profile. A natural choice is then to use the live points only, since these represent the most up-to-date likelihood profile. For a given set of live points, we have a set of live likelihoods $\logL$, as well as a set of $\set{X_i}$ drawn from the distribution \labelcref{eq:X_dist}. The parameters of the fitting function can be found by minimising the cost function
\begin{equation}\label{chi squared}
	C^2(\thetab) = \sum_i \left| \log \Like_i - \log f(X_i, \thetab) \right| ^2
\end{equation}
with respect to $\thetab$. A key feature is that we can analytically differentiate  $C^2$ for the Gaussian model, which allows us to write the optimal values of $\logLm$ and $\sigma$ in terms of the optimal values of $d$:
\begin{equation}\label{eq:sigma}
    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
\end{equation}
\begin{equation}\label{eq:logLm}
    \logLm = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d},
\end{equation}
(where $N$ here is the number of data points in we fit over), and simply minimise with respect to $d$ instead of all three variables. The result is that we are able to consistently find \textit{global} rather than local minima in a single optimisation run. We do not need to concern ourselves with the usual difficulties of global optimisation like comparing an ensemble of initial conditions, which is orders of magnitude slower.
\subsection{The termination prior volume}
Once we have obtained the best-fit $\thetab$, the endpoint can be calculated in the manner described in \cref{endpoint};
\begin{equation}
	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
\end{equation}
The integrals have the analytic solution
\begin{equation}
	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
\end{equation}
where $\gamma_k = \gamma\left(d/2, X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
\begin{equation}
	\log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \gamma^{-1} \left(\ \frac{d}{2} ,\ \epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),
\end{equation}
and $N_\mathrm{f}$ is just  $-\nlive$ multiplied by this.
\par
The prediction procedure should not contribute any significant time complexity to the nested sampling run, since it involves no likelihood evaluations.

\subsection{Prediction uncertainties}
We saw in section X that our knowledge of the posterior is limited by the live point of minimum volume, which is only $\mathcal{O}(\log \nlive)$ less than the log-compression of the dead point. Endpoint prediction as we have just done relies on our hope that the likelihood will continue to behave as it has done up to this point, but this may well not be the case; a sharp phase change, for instance, will be our undoing.
\par
We will therefore restrict ourselves to quantifying the uncertainties associated with how we have \textit{modelled} the known likelihood profile. These have two sources: the variance in the prior volume estimates, and the uncertainty in the least squares regression. The former is by far the larger source of error, since the uncertainty in nested sampling is dominated not by the scatter of the individual points but by long-term drift (as shown in \cref{fig:prediction_uncertainty}).
\par
To get the uncertainty, we can therefore ignore the least squares uncertainty and simply sample from $\Pr(\bm{X})$, finding the best-fit $\thetab$ and endpoint for each sample, then using this set of estimates to get a mean and variance in the endpoint estimate.
\begin{figure}
\begin{center}
	\includegraphics[max width=\linewidth]{figures/prediction_uncertainty.pdf}
\end{center}
\caption{Sources of the prediction uncertainty. The navy points are the live points, plotted for three different instances of $\set{X_i}$ drawn from its distribution. The light blue lines show the uncertainty in the least squares fits over each instance, which is much less than the uncertainty in the spread of the $X$ samples due to long term drift.}
\label{fig:prediction_uncertainty}
\end{figure}

\section{Results}\label{sec:results}
Now that we have a method for predicting the endpoint, we can test it on a range of likelihoods. We shall first present results for cosmological chains in \cref{subsec:cosmological_examples}, for which the method works well. \cref{subsec:toy_examples} will then show some toy examples that are designed to reveal some of the limitations of the method.

\subsection{Gaussians}\label{subsec:gaussians}
We will first test our prediction model on truly Gaussian likelihoods to show that it is able to recover the true parameters as well as the correct endpoint. Let us begin with a spherically symmetrical Gaussian which matches the form of the fitting function, which we can use to check that the correct parameters are inferred when the model is exact. We fix the parameters $\log\mathcal{L}_\mathrm{max} = 0$ and $\sigma = 0.01$ and vary the dimension to see the impact on the predictions.
\par
The inferred parameters are shown in \cref{fig:gauss_params}. The uncertainty in $d$ and $\sigma$ remain large throughout the run, but the endpoint predictions converge because we have already accumulated most of the evidence.
\par
\begin{figure*}
\begin{center}
	\includegraphics{figures/gauss_params.pdf}
\end{center}
\caption{Parameter inferences for the Gaussian fitting function in the $d=8$ case. The true parameters fall within the standard error bars for practically every point, as seen in comparison to the true parameter values shown in the dotted red. The uncertainties remain relatively constant for $d$ and $\sigma$, but the endpoint predictions converge because we have already accumulated most of the evidence.}
\label{fig:gauss_params}
\end{figure*}
Predictions are shown in \cref{fig:gauss_predictions}, where the x-axis plots the negative log-compression as a measure of the progress, so that the run proceeds in the rightward direction. The y-axis shows the predicted log-compression $\log\hat{X}_\mathrm{f}$ at which the termination condition is met.
\par
The true endpoint is recovered to within $2\sigma$ for all runs, but the uncertainty is greater for the higher dimension and $\DKL$ likelihoods. This reflects the fact that the live points typically probe only $\sim \log n_\mathrm{live}$ away from the current dead point. For larger $\DKL$ therefore, the set of samples at any given iteration before the end are quite far from the posterior bulk, and hence poorly resemble the shape of the true posterior.

\subsubsection{Elongated Gaussian}
Consider same example as that presented in \cref{sec:constrained_parameters}. The predictions are initially awry because samples from three of the parameters are uniform. Cannot determine whether these parameters are unconstrained or simply high variance at the midway point. So the predictions are fundamentally limited by the information contained in the samples, not by the prediction method.

\subsubsection*{Comparison to extrapolating evidence increments}
Seasoned users of nested sampling might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges. We do this for a spherical Gaussian as a comparison to our method. At an intermediate stage of the run, the most recent outputs might look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate}. Extrapolating those data to a linear and exponential profile yields endpoint estimates plotted in the graph to the right. 
\par
The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile. The increments are also not quite exponential, since the exponential fit leads to a large overprediction. The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result. One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
\par
More importantly, we find that for real likelihoods which have an element of noise the extrapolation often diverges, for instance when the increments do not monotonically decrease. Directly extrapolating the evidence increments is therefore far less stable than the previous method, and generally not a reliable method for prediction.
\begin{figure}
\begin{adjustbox}{valign=t, scale=0.72}
\subfloat{\begin{tabular}{|c|c|c|}
\hline
iteration & log Z  & $d\log Z$   \\
\hline
5000 & -1435.8 & 190.8 \\
5500 & -1264.6 & 171.2 \\
6000 & -1123.7 & 140.9 \\
6500 & -991.5 & 132.2 \\
7000 & -885.0 & 106.6 \\
7500 & -790.3 & 94.7 \\
8000 & -702.6 & 87.7 \\
8500 & -619.7 & 82.9 \\
9000 & -551.8 & 67.9 \\
9500 & -492.7 & 59.1 \\
\hline
\end{tabular}}
\end{adjustbox}
\quad
\begin{adjustbox}{valign=t}
\subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
\end{adjustbox}
\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation}
\label{fig:inc_extrapolate}
\end{figure}

\begin{figure}
\includegraphics{figures/inc_predictions.pdf}
\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation}
\label{fig:inc_predictions}
\end{figure}

\subsection{Cosmological examples}\label{subsec:cosmological_examples}
Having established that the prediction framework works for Gaussian likelihoods, we now turn to some cosmological examples, namely chains based on the \textsc{Planck 2018} dataset as described in \citet{curvature_tension}. Predictions are shown in \cref{fig:lcdm_predictions}, with the correct endpoint inferred to within error before the halfway point for all likelihoods.
\par
Most predictions fail to converge at early stages of the run. We find that this is not a consequence of the prediction model, but due to the effective dimensionality of the samples being lower than the true value at the beginning of the run, as was seen in \cref{sec:current_temperature}.

\subsubsection*{Compressive phase transitions}
\begin{figure}
\begin{center}
\includegraphics{figures/phase_transition.pdf}
\end{center}
\caption{Endpoint predictions and effective dimensionality as a function of compression for a two-dimensional Gaussian in a box prior $\in [0, 1]$. The parameter variances are $10^{-2}$ and $10^{-5}$ respectively, leading to compression only along one direction initially, which is reflected in $\tilde{d}$. The endpoint only converges when the dimensionality does; before that point the samples genuinely suggest that the likelihood is one-dimensional.}
\label{fig:phase_transition}
\end{figure}
The initial dimensionality of the $\mathcal{L}(X)$ curve is so extrapolating from the curve would fail to account for the true dimensionality of the likelihood, giving inaccurate predictions.
\par
The extrapolation now accounts for the true dimensionality, so the predictions become accurate. The trouble is that before this has happened, the samples really do behave as if they are only constrained in a single dimension! It could really be the case that there is no second parameter, and without our extra knowledge there is no way to detect whether there are two, or even three or more unseen parameters hiding at higher compressions
\par
The different rates of compression for different parameters therefore induce an effective phase transition in the number of degrees of freedom. One can put this on a more quantitative footing by considering the \textit{effective dimensionality} of the posterior. Define the Bayesian model dimensionality $\tilde{d}$ as the posterior variance of the information content, which is given by
\par
\cref{fig:phase_transition} plots the effective dimensionality alongside the endpoint predictions, where it is evident that the predictions are directly dependent on $\tilde{d}$. Introducing correlations between the parameters (not shown) causes the transition to be less sharp, since the more constrained parameter is able to `drag' the other parameter with it, allowing the predictions can be converge shortly before the dimensionality.

\subsection{Edge cases}\label{subsec:toy_examples}
In the cosmological cases, underestimated predictions at early stages was due to the difference in parameter compression rates. However, there are also cases where the form of the fitting function causes problems.
\par
One such instance is a spherical Cauchy likelihood. In fact, at early stages direct least squares fitting fails completely by predicting a remaining evidence that overflows. This can be explained by the fact that a long tail of a Cauchy resembles a Gaussian of much higher dimension, and inferring the dimensionality is unstable. Since we have a stable measure of dimension from the Bayesian model dimensionality, a partial fix is to set $d = \tilde{d}$ instead of inferring it, then using \cref{eq:sigma,,eq:logLm} to get $\log\mathcal{L}_\mathrm{max}$ and $\sigma$.
\par
Results are shown in \cref{fig:cauchy_predictions}, where the likelihood has the form
\begin{equation}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
\end{equation}
choosing $d = 10$ and allowing $\gamma$ to vary. However, the extrapolations still fall outside fall outside the standard error range despite isotropic compression. The issue in this case is that the Gaussian fitting function is unable to capture the true shape of the likelihood. A more comprehensive fix would be to use a more flexible fitting function that is able to capture a broader family of likelihoods, which is outside the scope of this paper. Nevertheless, the order of magnitude is still accurate, which remains highly useful to the end user.

\begin{figure*}
\begin{center}
	\subfloat[Predictions for a spherically symmetric Gaussian likelihood. The correct endpoint is ascertained to within standard error from the beginning for all runs, but the uncertainty is larger for the higher dimension and $\DKL$ likelihoods, reflecting the fact that the live points are further away from the posterior bulk in those cases.]{\includegraphics{figures/gauss_predictions.pdf}\label{fig:gauss_predictions}}
	\\[5ex]
	\subfloat[Predictions for the $\Lambda$CDM model. The chains without \textsc{Planck} have low $\DKL$ of below 10, so the live points are relatively close to the posterior bulk at all times and the predictions are quite accurate. ]{\includegraphics{figures/lcdm_predictions.pdf}\label{fig:lcdm_predictions}}
	\\[5ex]
	\subfloat[Predictions for a Cauchy likelihood. The parameters of the fitting function are obtained by fixing $d = \tilde{d}$, then least squares optimising $\log\mathcal{L}_\mathrm{max}$ and $\sigma$.]{\includegraphics{figures/cauchy_predictions.pdf}\label{fig:cauchy_predictions}}
\end{center}
\caption{Endpoint predictions for different likelihoods. The x-axis shows the negative log-compression, and the y-axis the predicted endpoint $\log\hat{X}_\mathrm{f}$. The shaded regions show the 1 and 2$\sigma$ uncertainties, while the red dashed line shows the true value.} 
\end{figure*}

\section{Conclusions}\label{sec:conclusions}
We have presented a method for predicting the endpoint of a nested sampling run, based on extrapolating the known likelihood profile into higher likelihood regions using a Gaussian fitting function.
\par
The method in general converges to the correct endpoint by about the halfway point, and gets the correct order of magnitude throughout. Predictions are poor when the effective dimension of the samples is lower than the true posterior, but this is a limitation of the information content in the samples rather than our method.
\par
Further work can be done to incorporate more flexible fitting functions to extrapolate non-Gaussian likelihoods, as well as developing a more rigorous framework to choose between the inferred $d$ and $\tilde{d}$.

\bibliographystyle{mnras}
\bibliography{references}

\appendix
\section{Derivation of the termination log-likelihood }
Assume a Gaussian likelihood
\begin{equation}\label{eq:logL}
	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{X^{2/d}}{2\sigma^2}.
\end{equation}
The distribution of the true posterior in $\log\mathcal{L}$ is
\begin{equation}
    P(\log\mathcal{L}) = \frac{1}{\Gamma(\frac{d}{2})}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
\end{equation}
i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the distribution of the weighted samples. The posterior average and variance of $\log \mathcal{L}$ are given by
\begin{equation}
    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
\end{equation}
Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
\begin{equation}
	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\mathcal{L}_* < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
    \label{eq:PL}
\end{equation}
It is helpful at this stage to define a parameter
\begin{equation}
    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
    \label{eq:y}
\end{equation}
as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
\begin{equation}
    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
    \label{eq:Py}
\end{equation}
We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$. Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
\begin{equation}
    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}[0<y_\mathrm{max}^\mathrm{live}<1],
    \label{eq:Pyhat}
\end{equation}
which in the limit of large live points and dimensions can be roughly summarised by
\begin{equation}\label{eq:ylivemax}
	    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
\end{equation}
This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$. In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
\par
This remains true even at the end of the nested sampling run. To see this, we write the halting condition as:
\begin{equation}
	f = \frac{\int_0^{X_\mathrm{end}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
    \label{eq:fint}
\end{equation}
Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $f \ll 1$ so that the denominator is approximately the accumulated evidence. Computing this for \cref{eq:logL} we find the answer in terms of lower incomplete gamma functions
\begin{equation}
    f = 1- \frac{\Gamma_{d/2}\left(\frac{X_\mathrm{end}^{2/d}}{2\sigma^2}\right)}{\Gamma\left(\frac{d}{2}\right)}.
    \label{eq:f}
\end{equation}
Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
\begin{equation}
    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} f \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d \ \Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
\end{equation}
We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $f$. This becomes yet more pleasing in the large $d$ limit, since $f^{2/d}\to 1$, we find via a Stirling approximation:
\begin{equation}
    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
\end{equation}
In the event that we keep $f$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}f^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
\par
Putting this together, taking $\mathcal{L}_*$ in \cref{eq:y} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
\begin{equation}
    \boxed{
        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
    },
\end{equation}
showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood, and the final set of $n$ live points can get you $\log (n)/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation. This general anatomy is shown in \cref{fig:anatomy}.

\label{lastpage}
\end{document}

### 2023-09-27 paper/figures/cauchy_predictions.py

@@ -3,16 +3,18 @@ from aeons.utils import *
 from aeons.plotting import *
 figsettings()
 
-fig, axs = plt.subplots(1, 3, figsize=(8, 1.4))
-for i, exp in enumerate([2, 3, 4]):
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
+for i, exp in enumerate([2, 3, 4, 5]):
     ax = axs[i]
-    name, samples = get_samples('toy', f'cauchy_8{exp}')
-    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/cauchy/cauchy_8{exp}.txt')
+    name, samples = get_samples(f'cauchy_8{exp}')
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/dG_range/02_cauchy_8{exp}.txt')
     logXs = samples.logX().iloc[iterations]
     logXf_true = samples.logX().iloc[int(true_endpoint[0])]
-    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.6), color='deepskyblue', ax=ax)
+    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 2), color='deepskyblue', ax=ax)
     ax.set_title(f'$\\gamma=10^{{{-exp}}}$, $\\mathcal{{D}}_\\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
+    ax.set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
 
-fig.supxlabel('$-\\log X$', y=-0.15)
-fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.06)
+fig.supxlabel('$-\\log X$', y=0.02)
+fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.01)
+fig.tight_layout()
 fig.savefig('cauchy_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-09-27 paper/figures/gauss_predictions.py

@@ -3,18 +3,18 @@ from aeons.utils import *
 from aeons.plotting import *
 figsettings()
 
-d = [8, 16, 32]
-fig, axs = plt.subplots(1, 3, figsize=(7, 1.8))
+d = [4, 8, 16, 32]
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.8))
 axs = axs.flatten()
 for i, nDims in enumerate(d):
-    samples = pickle_in(f'{aeons_dir}/data/predictions/gauss/gauss_{nDims}.pkl')
-    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_{nDims}.txt')
+    name, samples = get_samples(f'gauss_{nDims}')
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/dG_range/02_gauss_{nDims}.txt')
     logXf_true = samples.logX().iloc[int(true_endpoint[0])]
     logXs = samples.logX().iloc[iterations]
     # endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=500)
-    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.8), ax=axs[i])
+    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.5), ax=axs[i])
     axs[i].set_title(rf'$d$ = {nDims}, $\mathcal{{D}}_\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
-    axs[i].set_yticks([-logXf_true], [f"{-logXf_true:.1f}"])
+    axs[i].set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
 fig.supxlabel('$-\\log X$', y=0.1, fontsize=8)
 fig.supylabel('$-\\log \hat{X}_\mathrm{f}$',  fontsize=8)
 fig.tight_layout()

### 2023-09-27 paper/figures/lcdm_predictions.py

@@ -11,8 +11,8 @@ fig.delaxes(axs[-1])
 for i, chain in enumerate(lcdm_chains):
     ax = axs[i]
     axd = ax.twinx()
-    name, samples = get_samples('lcdm', chain)
-    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f"{aeons_dir}/data/predictions/lcdm/nos/{chain}_bt_25_nos.txt")
+    name, samples = get_samples(chain)
+    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f"{aeons_dir}/data/predictions/dG_range/02_{chain}.txt")
     iterations_dG, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/lcdm/dG/{chain}_dG.txt')
     logXf_true = samples.logX().iloc[int(true_endpoint[0])]
     logXs = samples.logX().iloc[iterations]
@@ -22,7 +22,7 @@ for i, chain in enumerate(lcdm_chains):
     axs[i].set_title(name, fontsize=8)
 # Make legend with custom patches
 from matplotlib.patches import Patch
-lines = [Patch(color='deepskyblue', label='Endpoints'), Patch(color='lightsalmon', label='Effective dimensionality')]
+lines = [Patch(color='deepskyblue', label='Endpoints'), Patch(color='lightsalmon', label='Sample dimensionality')]
 fig.legend(handles=lines, ncol=1, fontsize=6, bbox_to_anchor=(0.97, 0.2))
 fig.tight_layout()
 fig.supxlabel('$-\\log X$')

### 2023-10-04 paper/paper.tex
started temperature section

@@
 \usepackage{fancyvrb}
+\usepackage{enumitem}
 \usepackage{xcolor}
@@
 
+\renewcommand\thesubsubsection{\Alph{subsubsection}.}
 \newcommand{\nlive}{n_\mathrm{live}}
@@
 \newcommand{\DKL}{\mathcal{D}_\mathrm{KL}}
+\newcommand{\logLmax}{\log \Like_\mathrm{max}}
 
@@
 The order of magnitude runtime of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user.
-Currently, existing implementations of nested sampling either do not give any indication of the runtime, or only provide crude estimates.
-In the latter case, these are always underpredictions which usually fail to recover the correct order of magnitude.
+Currently, existing implementations of nested sampling \citep{multinest, polychord} either do not give an indication of runtime, or only provide crude measures of progress that do not directly correspond to the runtime.
 \par
@@
 \section{Background}
+Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation.
+For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
+\begin{equation}\label{eq:evidence}
+	\mathcal{Z} = \int \mathcal{L}(\theta)\ \pi(\theta)\ \mathrm{d}\theta
+\end{equation}
+while producing samples of the posterior distribution
+\begin{equation}
+	\mathcal{P}(\theta) = \frac{\mathcal{L}(\theta) \pi(\theta)}{\mathcal{Z}}.
+\end{equation}
+The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}.
+At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}.
+A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point.
+Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
+\par
+The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
+\begin{equation}
+	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
+\end{equation}
+where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$.
+The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
+\begin{equation}\label{eq:X_dist}
+	X_i = \prod_{k=1}^i t_k.
+\end{equation}
+The $t_i$ are the maximum of $\nlive$ points drawn from $[0,1]$, so follow the distribution
+\begin{equation}\label{eq:t_dist}
+	P(t_i) = \nlive t_i^{\nlive-1}, \quad \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
+\end{equation}
+The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g.
+$10^{-3}$.
+The remaining live points are then killed off one by one without replacement and added to the evidence.
+\par
+Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.
+For any given problem, the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{\nlive}$, so $\nlive$ sets the resolution.
 
-\subsection{The progress of a nested sampling run}
-Present some theoretical results on anatomy of nested sampling run.
+\section{The anatomy of a nested sampling run}
+The following section aims to make an inventory of the information available to us at an intermediate iteration $i^{*}$, which we shall eventually use to make endpoint predictions.
+We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature schedule, and the dimensionality of the samples.
 
-\subsection{Current prior volume}
+\subsection{Prior volume}
+The key feature of nested sampling is that the sampling is controlled by prior volume compression.
+The task is to find the posterior typically lying in a tiny fraction of the prior volume, a total compression which is quantified by the average information gain, or \textit{Kullback-Leibler divergence}:
+\begin{equation}\label{eq:DKL}
+   \DKL = \int \mathcal{P}(\theta) \log \frac{\mathcal{P}(\theta)}{\pi(\theta)}\ \mathrm{d}\theta. 
+\end{equation}
+The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression. One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that 
+\begin{equation}
+    \mathrm{E}(\log X_i) = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
+\end{equation}
+A constant step size in $\log X$ corresponds to a geometrically decreasing measure for the dead points (as shown in fig X), which is exactly needed to overcome the curse of dimensionality.
+\par
+Dead measure plot
+\par
+The same is not true for the live points, which are uniformly distributed in prior volume. As a result, the maximum live point is found at 
+\begin{equation}\label{eq:Xmin}
+	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
+\end{equation}
+with variance
+\begin{equation}
+	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6}, 
+\end{equation}
+where the large $\nlive$ limit is taken for the approximation to the harmonic series, $\gamma$ being the Euler-Mascheroni constant.
+\par
+Hence the live points only get us a factor of $\log \nlive$ closer to the posterior bulk.
+In other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples look anything like the posterior.
+One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
+\par
+Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
 
-\subsection{Current log-likelihood}
+\subsection{Log-likelihood}
+It is also useful to observe the distribution of the live and dead points in log-likelihood. We choose to examine a representative case of the $d$-dimensional multivariate Gaussian:
+\begin{equation}\label{eq:gaussian_logL}
+    \log \Like = \logLmax - X^{2/d}/2\sigma^2
+\end{equation}
+with maximum point $\logLmax$ and lengthscale $\sigma$ to get an insight into the analytics.
+The likelihood monotonically increases towards $\logLmax$, with the posterior samples eventually concentrated around the bulk at
+\begin{equation}
+    \langle\log\Like\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
+\end{equation}
+We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Define a likelihood normalised by the distance to the maximum 
+\begin{equation}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
+    \label{eq:normalised_likelihood}
+\end{equation}
+as a measure of how far a point is between the current likelihood and the maximum; $y=0$ corresponds to $\log\mathcal{L}_*$ and $y=1$ to $\log\mathcal{L}_\mathrm{max}$.
+At each iteration, $y$ increases by roughly
+\begin{equation}
+    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}  
+\end{equation}
+in the large $\nlive$ limit.
+We can again sum the harmonic series to see that the maximum live point is expected to be at $y = 2\log \nlive/d$: a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
+\par
+The above also implies that the normalised distance between the highest and second highest live point is roughly $2/d$.
+Before reaching the posterior bulk, $\logLmax - \log\Like^{*} > d/2$, so the log-likelihood distance between the two points is larger than one, hence at least an order of magnitude apart in likelihood.
+It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.
 
-\subsubsection{Dimensionality}
-Define Bayesian model dimensionality
-Goal to observe how the number of constrained parameters changes throughout the run
-Only works at the end of the run, because BMD is a property of the posterior - heat capacity analogy
-Need to transform intermediate set of samples into a posterior. Useful tool is the transformation using beta. 
-Which beta to use? Several existing temperatures in the literature - microcanonical temperature, canonical temperature used by Habeck.
-Propose a temperature from Bayesian inference.
+\subsubsection*{Aside: nested sampling as a maximiser}
+Previous literature \citep{Akrami_2010, Feroz_2011} has explored the potential for nested sampling to be used as a global maximiser, given its ability to handle multi-modalities.
+In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion.
+We now use the machinery we have developed to put this statement in more quantitative terms.
+\par
+A more rigorous derivation (see Appendix) shows that the maximum live point has mean and variance
+\begin{equation}\label{eq:ylivemax}
+	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
+\end{equation}
+Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{end}$ and prior volume $X_\mathrm{end}$, so that
+\begin{equation}
+	\epsilon = \frac{\int_0^{X_\mathrm{end}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
+\end{equation}
+Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence.
+Computing this for \cref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
+\begin{equation}
+\epsilon = 1- \frac{\Gamma_{d/2}\left(X_\mathrm{end}^{2/d}/2\sigma^2\right)}{\Gamma(d/2)}.
+\end{equation}
+Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
+\begin{equation}
+    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} \epsilon \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d \ \Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
+\end{equation}
+We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $\epsilon$.
+This becomes yet more pleasing in the large $d$ limit, since $\epsilon^{2/d}\to 1$, we find via a Stirling approximation:
+\begin{equation}
+    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
+\end{equation}
+In the event that we keep $\epsilon$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}\epsilon^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
+\par
+Putting this together, taking $\mathcal{L}_*$ in \cref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
+\begin{equation}
+    \boxed{
+        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
+    },
+\end{equation}
+showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood.
+The final set of $n$ live points gets you $\log n/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
 
-Ridge geometries contain phase transitions in nested sampling, because the dimensionality of the samples increases throughout the run. 
+\subsection{Temperature}
+\subsubsection*{Motivations}
+As shown in the previous section, midway through the run nearly all of the posterior mass is concentrated at a single point.
+However, this does not capture the \textit{structure} of the posterior that has been explored and all of the information it provides.
 \par
-As an example, consider an elongated Gaussian in a unit hypercube.
+We have the potential to fix this because nested sampling is invariant to monotonic transformations; softening the likelihood via $L \to L^{\beta}$ brings the likelihoods closer together.
+Large $\beta$ and small $\beta$ both discard information, but there is an inbetween range of $\beta$ which is significant, because it causes the posterior mass to be concentrated at the current point; see plot.
+\par
+At this point relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived.
+If one takes $- \log \Like$ to be the energy $E_i$ etc, then $\beta$ is the inverse temperature $1/T$.
+Softening or hardening is equivalent to setting the temperature of the samples, which one can easily do in NS because it is athermal; get the posterior for $\Like^{\beta}$ trivially by reweighting.
+\par
+Bayesian inference has a correspondence to statistical mechanics in that one can equate the parameters to microstates $i$, the negative log-lihood to the microstate energy $E_i$, and the prior to the density of states $g_i$.
+One can generalise Bayes' theorem to
+\begin{equation}
+p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{Z(\beta)}
+\end{equation}
+Clear similarity to thermodynamic integration, which uses temperature as a control parameter to construct intermediate distributions between the prior at $\beta = 0$ and posterior at $\beta = 1$.
+NS uses compression instead of temp, sidestepping difficulty of designing the annealing schedule.
+However, it still does something similar, starting with unweighted prior samples and ending with likelihood-weighted prior samples i.e.
+the posterior, during which $\beta$ presumably increases from 0 to 1.
+Inferring the temperature from above, one would hope to get another quantity that determines the progress of a run.
+\par
+Important difference to annealing is that here we are working backwards; instead of choosing a temperature to sample at, we obtain the temperature from the samples after the fact. 
+An useful discussion can be found in \citet{demons}, which we summarise here as a starting point.
+In contrast to annealing methods which attempt to sample from a series of canonical ensembles $\pi(\theta) \exp{-\beta \mathrm{E}(\theta)}$, nested sampling evolves a series of \textit{microcanonical} ensembles and computing the density of states by tracking the compression.
+\par
+We now present several methods of obtaining the temperature that one can plausibly consider to be the current temperature of a nested sampling run.
+
+\subsubsection{Microcanonical temperature}
+As discussed in Skilling's original paper, the density of states, given by the negative slope of the log-likelihood curve, directly corresponds to a temperature;
+\begin{equation}
+    \beta^* = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like^{*}}.
+\end{equation}
+This is the microcanonical temperature $\partial S/\partial E$, where the volume entropy $S=\log X$ and the energy $E=-\log \Like$.
+Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing (in practice, we find that above a threshold of around 10 iterations the result is fairly insensitive to this choice).
+
+\subsubsection{Canonical temperature}
+Briefly considered by Habeck, it is the temperature of an ensemble whose average energy is the current energy, which one can obtain by inverting 
+\begin{equation}
+    \langle \log \Like \rangle_{\mathcal{P}_\beta} = \log \Like^{*}
+\end{equation}
+
+\subsubsection{Bayesian temperature}
+We furthermore propose a temperature that is obtained via Bayesian inference, which returns a distribution rather than a point estimate for $\beta$.
+Since each value of $\beta$ leads to a different likelihood $\Like^{\beta}$, one can consider the posterior distribution as a function of $\log X$ to be \textit{conditioned} on $\beta$.
+We can therefore write
+\begin{equation}
+   \mathcal{P}(\log X \mid \beta) = \frac{\Like^{\beta}(X) X}{\mathcal{Z(\beta)}}.
+\end{equation}
+What we would really like is the distribution of $\beta$ at the present iteration, so the natural step is to invert this via Bayes' rule;
+\begin{equation}
+    P\left(\beta \mid \log X^{*}\right) = \frac{\mathcal{P}\left(\log X^{*} \mid \beta\right) P\left(\beta\right)}{P\left(\log X\right)}.
+\end{equation}
+As with all Bayesian analyses, the distribution of $\beta$ is fixed up to a prior, which we choose to be uniform in $\beta$.
+The obtained temperatures are consistent with the previous two choices, which may seem oddly coincidental.
+However, closer inspection reveals that large values of $P(\beta \mid \log X^{*})$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence.
+Thus the peak is just the temperature that causes the posterior to be sharply peaked at the current contour i.e.
+the location of the posterior bulk.
+Heuristically, it is doing a similar thing as the previous methods.
+
+\subsubsection*{Comparisons}
+Unlike annealing, the temperature is not monotonic with progress, which is precisely what allows nested sampling to handle phase transitions. 
+\par
+Free parameter; choose definition based on what you would like to use $\beta$ to do.
+
+\subsection{Dimensionality}
+We can immediately use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible.
+\citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality (BMD), defined as the posterior variance of the information content:
+\begin{equation}
+\frac{\tilde{d}}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
+= \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
+\end{equation}
+\par
+Calculating the quantity using intermediate set of weighted samples (which is concentrated at a single point) leads to vanishing variance, hence also dimensionality.
+However, we can recover the structure of the posterior together with the true dimensionality by adjusting the temperature.
+Dimensionality estimates are plotted in fig X for a spherical Gaussian of various dimensions.
+The Bayesian temperature appears to make the best estimates of the dimension, so that is the one we choose.
+\par
+Plot of results for spherical Gaussian, for which we know the correct answer.
+
+\subsubsection*{Parameter compression rate}
+Plots of samples dimensionality against compression also draw attention to the \textit{speed} at which different parameters are constrained throughout the run.
+As a concrete example, consider an elongated Gaussian in a unit hypercube prior with $\mu = 0$ and $\Sigma = \mathrm{diag}\left(1, 1, 0.01, 0.01\right) \times 10^{-2}$.
+\par
+Plot of dimensionality vs compression factor, showing step changes 
+\par
+It is important to appreciate that at lower compressions the samples truly lie in a lower-dimensional space, because the algorithm has not yet begun to constrain the other parameters.
+Anticipating the full dimensionality of the space is therefore just as impossible as that associated with a slab-spike geometry, so in this sense such geometries contain a phase transition.
 
 \section{Endpoint prediction}
-Discussion of how to get endpoint and the necessity of extrapolation. Compare to UltraNest/dynesty slab methods and why they underpredict.
+The time complexity of nested sampling \citep{supernest} is
+\begin{equation}
+    T \propto \nlive \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle \mathcal{T}\{ \mathrm{Impl.}\} \rangle \times \DKL.
+\end{equation}
+The second term is the time per likelihood evaluation, which is constant.
+The third is the cost to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude.
+The primary unknown during a run, and therefore our primary interest, is the final term: the compression required to get from prior to posterior.
+
+\subsection{The termination prior volume}
+To be exact, one wishes to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is slightly larger in magnitude than $\DKL$ (Fig X).
+The problem is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is quite far from the posterior bulk.
 \par
-The key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian.
-Fitting a Gaussian of this dimension to the likelihood profile should be a reasonable approximation to the true distribution.
+In order to get an idea of where the true posterior bulk sits, we need to predict what the posterior looks like past the highest live point.
+We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points.
 \par
-Show fits
-Find BMD
-Fit data to find sigma
-Extrapolate using a Gaussian likelihood with this parameterisation
+One would never use this predicted posterior to do inference, since more accuracy can always be achieved by simply finishing the run.
+However, it is more than sufficient for making a prediction for $\log X_\mathrm{f}$.
+Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
+\begin{equation}
+	\Delta \mathcal{Z} = \epsilon \mathcal{Z}_\mathrm{tot},
+\end{equation}
+\begin{equation}\label{endpoint}
+	\int_0^{X_\mathrm{f}} f(X, \phi)\ \mathrm{d}X = \epsilon \left( \int_0^{X_i} f(X, \phi)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead} \right),
+\end{equation}
+where $X_i$ is the volume of the iteration we have currently compressed to, and $\mathcal{Z}_\mathrm{dead}$ is the evidence we have accumulated up to this point.
+$X_\mathrm{f}$ can then be identified by solving the above equation either analytically or numerically.
+\par
+Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule.
+The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_\mathrm{f}$ will be
+\begin{equation}
+	N_\mathrm{f} = - \nlive \log X_\mathrm{f} .
+\end{equation}
+
+\subsection{Regression procedure}
+A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian.
+Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function (see plot).
+The explicit form of the Gaussian that we fit is the same as that given in section X, which we shall repeat here for clarity;
+\begin{equation}\label{gaussian}
+    f(X; \phi) = \logLmax - X^{2/d}/2\sigma^2
+\end{equation}
+The extrapolation then proceeds thus:
+\begin{enumerate}[leftmargin=*]
+    \item Find the current dimensionality $\tilde{d}^{*}$ of the posterior at the Bayesian temperature
+    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = \tilde{d}$ to infer $\logLmax$ and $\sigma$ 
+    \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
+\end{enumerate}
+The advantage of fitting a Gaussian is that the procedure can be sped up analytically. Firstly, the least squares regression is trivial because analytic estimators exist; the cost function 
+\begin{equation}\label{chi squared}
+	C^2(\logLmax, \sigma) = \sum_i \left| \log \Like_i - f(X_i; \logLmax, \sigma) \right| ^2
+\end{equation}
+is minimised with respect to $(\logLmax, \sigma)$ when
+\begin{equation}\label{eq:sigma}
+    \sigma^2 = \frac{N \sum_i X_i^{4/d} - \left(\sum_i X_i^{2/d}\right)^2}{2 \sum_i \log \Like_i \sum_i X_i^{2/d} - 2N \sum_i X_i^{2/d}\log \Like_i },
+\end{equation}
+and
+\begin{equation}\label{eq:logLmax}
+    \logLmax = \frac{1}{N} \sum_i \log \mathcal{L}_i + \frac{1}{2N\sigma^2} \sum_i X_i^{2/d}.
+\end{equation}
+Secondly, the termination prior volume can also be obtained analytically. Rewriting \cref{endpoint} in terms of the Gaussian parameters gives
+\begin{equation}
+	\epsilon = \frac{\int_0^{X_\mathrm{f}} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X}{\int_0^{X_i} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X + \mathcal{Z}_\mathrm{dead}}.
+\end{equation}
+The integrals have the analytic solution
+\begin{equation}
+	\int_0^{X_k} \Like_\mathrm{max} \exp\left(-X^{2/d}/2\sigma^2\right)\ \mathrm{d}X = \frac{d}{2} \cdot \left(\sqrt{2}\sigma\right)^d \cdot \gamma_k
+\end{equation}
+where $\gamma_k = \Gamma_{d/2}\left(X_k^{2/d}/2\sigma^2\right)$ is the lower incomplete gamma function. After taking the inverse of  $\gamma$ and a few more steps of algebra, we arrive at
+\begin{equation}
+    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),
+\end{equation}
+and $N_\mathrm{f}$ is of course just $-\nlive$ multiplied by this.
+Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained.
+\par
+Plot of fits to Gaussian for different distributions
+\par
+One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown it to be far less stable.
+
+\subsection{Alternative approaches}
+Ultranest \citep{ultranest} tracks progress based on the remaining integral, approximated as $\Like_\mathrm{max} X_i$.
+does not correspond directly to runtime, since it will be zero for most of the run before the bulk is crossed.
+\par
+Extrapolating evidence increments is considerably less stable
+
+\section{Results}
+Now that we have a method for predicting the endpoint, it can be tested on a range of distributions.
+We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
+\subsection{Toy examples}
+\subsubsection{Gaussians}
+Spherical Gaussian, elongated Gaussian
+\subsubsection{Cauchy}
+Cauchy\\
+\subsection{Cosmological examples}
+
+\section{Conclusion}
+
+\appendix
+\section{The maximum live log-likelihood}
+Assume a Gaussian likelihood
+\begin{equation}\label{eq:logL}
+	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - X^{2/d}/2\sigma^2.
+\end{equation}
+The distribution of the true posterior in $\log\mathcal{L}$ is
+\begin{equation}
+    P(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
+\end{equation}
+i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the distribution of the weighted samples. The posterior average and variance of $\log \mathcal{L}$ are given by
+\begin{equation}
+    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
+\end{equation}
+Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
+\begin{equation}
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\mathcal{L}_* < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
+    \label{eq:PL}
+\end{equation}
+It is helpful at this stage to define a parameter
+\begin{equation}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
+    \label{eq:y}
+\end{equation}
+as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
+\begin{equation}
+    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
+    \label{eq:Py}
+\end{equation}
+We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$.
+Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
+\begin{equation}
+    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1} \quad [0<y_\mathrm{max}^\mathrm{live}<1],
+    \label{eq:Pyhat}
+\end{equation}
+which may be roughly summarised as
+\begin{align}
+    y_\mathrm{max}^\mathrm{live} \sim &1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
+    & \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
+    \label{eq:ymax}
+\end{align}
+or in the large $d$, $n$ limit
+\begin{align}
+    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
+    \label{eq:ymaxd}\\
+    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
+    \label{eq:ymaxdn}
+\end{align}
+where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
+\par
+This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
+In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
 

### 2023-10-04 paper/figures/dead_measure.py

@@ -1,0 +1,110 @@
+# import lecture_style
+from anesthetic.examples.perfect_ns import planck_gaussian
+import matplotlib.pyplot as plt
+from matplotlib.patches import Rectangle
+from matplotlib.lines import Line2D
+import numpy as np
+from matplotlib import patheffects, rcParams
+rcParams['path.effects'] = [patheffects.withStroke(linewidth=1, foreground='white')]
+
+nlive = 100
+samples = planck_gaussian(nlive)
+
+x = 'omegabh2'
+y = 'omegach2'
+
+fig, axes = plt.subplots(1,4, figsize=(7*4*0.5,7*0.5*1.5))
+
+for ax in axes:
+    ax.set_xticks([])
+    ax.set_yticks([])
+    ax.set_box_aspect(1)
+    ax.plot(samples[x], samples[y], 'C0.', ms=0.5)
+
+axes[0].get_children()[-1].get_zorder()
+axes[0].set_zorder(10)
+axes[0].get_children()[-1].get_zorder()
+
+def get_box(i):
+    live = samples.live_points(i)
+    xmin = live[x].min()
+    xmax = live[x].max()
+    ymin = live[y].min()
+    ymax = live[y].max()
+    return xmin, xmax, ymin, ymax
+
+def inter_axis_line(fig, ax0, ax1, x, y, *args, **kwargs):
+    coord0 = fig.transFigure.inverted().transform(ax0.transData.transform([x, y]))
+    coord1 = fig.transFigure.inverted().transform(ax1.transData.transform([x, y]))
+    return Line2D([coord0[0], coord1[0]], [coord0[1], coord1[1]], transform=fig.transFigure,*args, **kwargs)
+
+def draw_zoom_lines(ax0, ax1, xmin, xmax, ymin, ymax):
+    ax1.set_xlim(xmin, xmax)
+    ax1.set_ylim(ymin, ymax)
+    zorder = max([child.get_zorder() for child in ax0.get_children()] +
+                 [ax0.get_zorder()])
+    ax1.set_zorder(zorder+2)
+    lines = list(fig.lines)
+    for x_ in [xmin, xmax]:
+        for y_ in [ymin, ymax]:
+            lines.append(inter_axis_line(fig, ax0, ax1, x_, y_, color='k',
+                                         zorder=zorder+1,
[…]

### 2023-10-25 paper/paper.tex
revising paper

@@
 \usepackage{xcolor}
-\usepackage{txfonts}
 \usepackage[T1]{fontenc}
@@
 \renewcommand\thesubsubsection{\Alph{subsubsection}.}
-\newcommand{\nlive}{n_\mathrm{live}}
+\newcommand{\nlive}{n_i}
 \newcommand{\Like}{\mathcal{L}}
@@
 \begin{abstract}
+This paper develops a technique to estimate the runtime of nested sampling which works for any nested sampling implementation.
+At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be calculated.
+The method finds the right order of magnitude at all times, obtaining the true endpoint within standard error generally around the halfway point, and converges to the true value by the end of the run.
 \end{abstract}
@@
 It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}.
-The core algorithm is unique for estimating volumes by \textit{counting}, which makes high-dimensional integration feasible.
+The core algorithm is unique in its estimation of volumes by \textit{counting}, which makes high-dimensional integration feasible.
 It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
@@
 The order of magnitude runtime of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user.
-Currently, existing implementations of nested sampling \citep{multinest, polychord} either do not give an indication of runtime, or only provide crude measures of progress that do not directly correspond to the runtime.
+Currently, existing implementations of nested sampling (e.g.
+\citealt{multinest, polychord, ultranest}) either do not give an indication of runtime, or only provide crude measures of progress that do not directly correspond to the runtime.
 \par
-This paper sets out a more principled manner of endpoint estimation for nested sampling at each intermediate stage, the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from.
-Outline of paper.
+This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from.
+We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction.
+\cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts.
+Finally, \cref{sec:results} presents the results and discussions for toy and cosmological chains, before we conclude.
 \begin{figure}
@@
 
-\section{Background}
-Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation.
+\section{Background}\label{sec:background}
+Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation; for a more comprehensive treatment, we recommend Skilling's paper and the textbook \citet{sivia}.
 For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
 \begin{equation}\label{eq:evidence}
-	\mathcal{Z} = \int \mathcal{L}(\theta)\ \pi(\theta)\ \mathrm{d}\theta
+	\mathcal{Z} = \int \mathcal{L}(\theta)\pi(\theta)\ \mathrm{d}\theta
 \end{equation}
@@
 \par
-The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
+The integral in \eqref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
 \begin{equation}
@@
 \begin{equation}\label{eq:t_dist}
-	P(t_i) = \nlive t_i^{\nlive-1}, \quad \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
+	P(t_i) = \nlive t_i^{\nlive-1}
+\end{equation}
+\begin{equation}\label{eq:t_moments}
+    \langle\log t_i\rangle = -\frac{1}{\nlive}, \quad \mathrm{Var}(\log t_i) = \frac{1}{\nlive^2}.
 \end{equation}
@@
 Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.
-For any given problem, the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{\nlive}$, so $\nlive$ sets the resolution.
+For any given problem, the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{\nlive}$, so $\nlive$ sets the resolution of the algorithm.
 
@@
 The following section aims to make an inventory of the information available to us at an intermediate iteration $i^{*}$, which we shall eventually use to make endpoint predictions.
-We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature schedule, and the dimensionality of the samples.
+We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature, and the dimensionality of the samples.
 
@@
 \end{equation}
-The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression. One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that 
+The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression.
+One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
 \begin{equation}
@@
 \end{equation}
-A constant step size in $\log X$ corresponds to a geometrically decreasing measure for the dead points (as shown in fig X), which is exactly needed to overcome the curse of dimensionality.
-\par
-Dead measure plot
+A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points (as shown in \cref{fig:dead_measure}), which is exactly needed to overcome the curse of dimensionality.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/dead_measure_live.pdf}
+\end{center}
+\caption{The dead points of a nested sampling run, recursively zoomed in.
+Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} V/V$, and hence scale invariant.
+The live points (larger dots) are uniform across the prior, plotted for comparison.}
+\label{fig:dead_measure}
+\end{figure*}
 \par
@@
 \begin{equation}\label{eq:Xmin}
-	\mathrm{E}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{E}[\log X_*] - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
+	\mathrm{E}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{E}(\log X_*) - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
 \end{equation}
@@
 \begin{equation}
-	\mathrm{Var}[\log X_\mathrm{min}^{\mathrm{live}}] = \mathrm{Var}[\log X_*] + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6}, 
+	\mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_*) + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6}, 
 \end{equation}
@@
 \par
-Hence the live points only get us a factor of $\log \nlive$ closer to the posterior bulk.
-In other words, it is not until we are around $\log \nlive$ away from $\log X = \DKL$ that the samples look anything like the posterior.
-One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
+The live points therefore only get us a factor of $\log \nlive$ closer to the posterior bulk.
+In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples look anything like the posterior.
+One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
+Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
 \par
-Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
+The above results are summarised in \cref{fig:logX_distribution}
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/logX_distribution.pdf}
+\end{center}
+\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the constrained prior and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i_*$.
+For large values of $\DKL$ i.e.
+informative posteriors and/or large dimensionalities, the maximum live point is very far from the posterior bulk until the very end of the run.
+Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller.
+Plots of the sort from here onwards will be in terms of $-\log X$, where the run will more naturally proceed from left to right.
+}
+\label{fig:logX_distribution}
+\end{figure*}
 
-\subsection{Log-likelihood}
+\subsection{Log-likelihood}\label{sec:logL}
 It is also useful to observe the distribution of the live and dead points in log-likelihood. We choose to examine a representative case of the $d$-dimensional multivariate Gaussian:
@@
 \end{equation}
-We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Define a likelihood normalised by the distance to the maximum 
+We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Let us define a likelihood normalised by the distance to the maximum:
 \begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}^*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}^*};
     \label{eq:normalised_likelihood}
 \end{equation}
-as a measure of how far a point is between the current likelihood and the maximum; $y=0$ corresponds to $\log\mathcal{L}_*$ and $y=1$ to $\log\mathcal{L}_\mathrm{max}$.
-At each iteration, $y$ increases by roughly
+$y=0$ corresponds to the current point $\log\mathcal{L}^*$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
 \begin{equation}
@@
 \end{equation}
-in the large $\nlive$ limit.
-We can again sum the harmonic series to see that the maximum live point is expected to be at $y = 2\log \nlive/d$: a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
+in the large $\nlive$ limit. We can again sum the harmonic series to see that the maximum live point is expected to be at 
+\begin{equation}
+    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d},
+\end{equation}
+where we have used the notation of order statistics to denote the $x_{(n)}$ as the maximum of $n$ points.
+This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
 \par
-The above also implies that the normalised distance between the highest and second highest live point is roughly $2/d$.
-Before reaching the posterior bulk, $\logLmax - \log\Like^{*} > d/2$, so the log-likelihood distance between the two points is larger than one, hence at least an order of magnitude apart in likelihood.
+The above also implies that the normalised distance between the highest and second highest live point is roughly 
+\begin{equation}
+    y_{(\nlive)}^{\mathrm{live}} - y_{(\nlive - 1)}^{\mathrm{live}} \approx \frac{2}{d}.
+\end{equation}
+Before reaching the posterior bulk, $\logLmax - \log\Like^{*} > d/2$, so we must have
+\begin{equation}
+    \log \Like^{\mathrm{live}}_{(\nlive)} - \log \Like^\mathrm{live}_{(\nlive - 1)} > 1.
+\end{equation}
+In other words, the highest likelihood point is always at least an order of magnitude greater than the second highest.
 It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.
@@
 \par
-A more rigorous derivation (see Appendix) shows that the maximum live point has mean and variance
+A more rigorous derivation (\cref{sec:logLmax}) shows that the maximum live point has mean and variance
 \begin{equation}\label{eq:ylivemax}
@@
 Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence.
-Computing this for \cref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
+Computing this for \eqref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
 \begin{equation}
@@
 \par
-Putting this together, taking $\mathcal{L}_*$ in \cref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{end}$, and combining this with \cref{eq:ylivemax} we find
+Putting this together, taking $\mathcal{L}_*$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{end}$, and combining this with \eqref{eq:ylivemax} we find
 \begin{equation}
@@
 The final set of $n$ live points gets you $\log n/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
+\par
+A plot of the log-likelihood distribution at the end of a run is shown in \cref{fig:logL_distribution}
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/logL_distribution.pdf}
+\end{center}
+\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_*)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$.
+The distances are shown between these locations at the end of the run, the key takeaway being that in high dimensions the highest log-likelihood point of a nested sampling run is nowhere near the maximum in high dimensions.}
+\label{fig:logL_distribution}
+\end{figure*}
 
@@
 \par
-We have the potential to fix this because nested sampling is invariant to monotonic transformations; softening the likelihood via $L \to L^{\beta}$ brings the likelihoods closer together.
-Large $\beta$ and small $\beta$ both discard information, but there is an inbetween range of $\beta$ which is significant, because it causes the posterior mass to be concentrated at the current point; see plot.
+We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially reweighting the samples.
+Increasing $\beta$ worsens the situation, while $\beta \to 0$ simply gives back the prior.
+There is, however, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples.
+A schematic of the procedure is shown in \cref{fig:last_live_point}.
 \par
-At this point relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived.
-If one takes $- \log \Like$ to be the energy $E_i$ etc, then $\beta$ is the inverse temperature $1/T$.
-Softening or hardening is equivalent to setting the temperature of the samples, which one can easily do in NS because it is athermal; get the posterior for $\Like^{\beta}$ trivially by reweighting.
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/last_live_point.pdf}
+\end{center}
+\caption{The likelihood and posterior as a function of $\log X$ in the middle of a nested sampling run.
+Almost all of the posterior mass is concentrated at a single point with the highest compression, because it is orders of magnitude higher in likelihood.
+Reducing $\beta$ reweights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta^*$ which is significant because it centres the posterior at the current contour.}
+\label{fig:last_live_point}
+\end{figure}
 \par
-Bayesian inference has a correspondence to statistical mechanics in that one can equate the parameters to microstates $i$, the negative log-lihood to the microstate energy $E_i$, and the prior to the density of states $g_i$.
-One can generalise Bayes' theorem to
+At this point it is relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived.
+If one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble
 \begin{equation}
-p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{Z(\beta)}
+    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}
 \end{equation}
-Clear similarity to thermodynamic integration, which uses temperature as a control parameter to construct intermediate distributions between the prior at $\beta = 0$ and posterior at $\beta = 1$.
-NS uses compression instead of temp, sidestepping difficulty of designing the annealing schedule.
-However, it still does something similar, starting with unweighted prior samples and ending with likelihood-weighted prior samples i.e.
-the posterior, during which $\beta$ presumably increases from 0 to 1.
-Inferring the temperature from above, one would hope to get another quantity that determines the progress of a run.
+at the inverse temperature $\beta = 1/T$.
+As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, but nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours.
+Each step moves at constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{statmech}.
 \par
-Important difference to annealing is that here we are working backwards; instead of choosing a temperature to sample at, we obtain the temperature from the samples after the fact. 
-An useful discussion can be found in \citet{demons}, which we summarise here as a starting point.
-In contrast to annealing methods which attempt to sample from a series of canonical ensembles $\pi(\theta) \exp{-\beta \mathrm{E}(\theta)}$, nested sampling evolves a series of \textit{microcanonical} ensembles and computing the density of states by tracking the compression.
-\par
-We now present several methods of obtaining the temperature that one can plausibly consider to be the current temperature of a nested sampling run.
+Because the temperature of a microcanonical ensemble is a derived property rather than a parameter, there is some freedom in its definition.
+Returning to our original motivation, we make the connection that the temperature is the reweighting $\Like \to \Like^{\beta}$ which centers the ensemble around the current energy.
+We now present several temperatures that achieve this aim, each of which one can plausibly consider to be the current temperature of a nested sampling run.
 
 \subsubsection{Microcanonical temperature}
-As discussed in Skilling's original paper, the density of states, given by the negative slope of the log-likelihood curve, directly corresponds to a temperature;
+The obvious candidate is the microcanonical temperature $\partial S/\partial E$, where the volume entropy is $\log X$ and the energy is as usual $-\log \Like$.
+This gives the density of states; as discussed in Skilling's original paper,
 \begin{equation}
-    \beta^* = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like^{*}}.
+    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like^{*}}.
 \end{equation}
-This is the microcanonical temperature $\partial S/\partial E$, where the volume entropy $S=\log X$ and the energy $E=-\log \Like$.
+is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X^{*}$, if we assume differentiability, which is exactly the intuition we were aiming for to put the ensemble bulk at the current contour.
 Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing (in practice, we find that above a threshold of around 10 iterations the result is fairly insensitive to this choice).
+\par
+The ensemble bulk can be placed at the current point in several other plausible ways, which are discussed below.
 
 \subsubsection{Canonical temperature}
-Briefly considered by Habeck, it is the temperature of an ensemble whose average energy is the current energy, which one can obtain by inverting 
+Another temperature considered by Habeck is that at which the current energy (i.e. $-\log \Like^{*}$) is the average energy of the entire ensemble. One can obtain it by inverting 
 \begin{equation}
@@
 \end{equation}
+to get the `canonical' temperature $\beta_\mathrm{C}$.
+While $\beta_\mathrm{M}$ is derived from (the gradient of) a single contour, this temperature uses the entire ensemble.
+It has the desirable property that it rises monotonically with compression, in analogy to a monotonic annealing schedule.
 
 \subsubsection{Bayesian temperature}
-We furthermore propose a temperature that is obtained via Bayesian inference, which returns a distribution rather than a point estimate for $\beta$.
+We furthermore propose a temperature $\beta_\mathrm{B}$ that is obtained via Bayesian inference, which returns a distribution rather than a point estimate.
 Since each value of $\beta$ leads to a different likelihood $\Like^{\beta}$, one can consider the posterior distribution as a function of $\log X$ to be \textit{conditioned} on $\beta$.
@@
 However, closer inspection reveals that large values of $P(\beta \mid \log X^{*})$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence.
-Thus the peak is just the temperature that causes the posterior to be sharply peaked at the current contour i.e.
-the location of the posterior bulk.
-Heuristically, it is doing a similar thing as the previous methods.
+Thus the Bayesian temperature uses the same idea as the microcanonical one, except it accounts for the spread in the result.
 
 \subsubsection*{Comparisons}
-Unlike annealing, the temperature is not monotonic with progress, which is precisely what allows nested sampling to handle phase transitions. 
+\cref{fig:beta_comparison} shows the three temperatures as a function of compression for two cases, one containing a phase transition and one without.
+They are consistent in both cases when there is a single dominant phase, but differ during a phase transition.
+The canonical temperature is the only one that rises monotonically with compression.
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/beta_comparison.pdf}
+\end{center}
+\caption{Inferred temperatures using the microcanonical, canonical and Bayesian definitions. All are consistent for a single phase, but differ during a phase transition.}
+\label{fig:beta_comparison}
+\end{figure}
 \par
-Free parameter; choose definition based on what you would like to use $\beta$ to do.
+One should keep in mind that despite the above theoretical reasoning, our introduction of the likelihood transformation was ultimately motivated by our wish to utilise the extra degree of freedom it provides.
+As we will see below, we recommend choosing the exact definition depending on what is useful for the transformation to do.
 
@@
 \begin{equation}
-\frac{\tilde{d}}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
+\frac{d_G}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
 = \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
@@
 However, we can recover the structure of the posterior together with the true dimensionality by adjusting the temperature.
-Dimensionality estimates are plotted in fig X for a spherical Gaussian of various dimensions.
-The Bayesian temperature appears to make the best estimates of the dimension, so that is the one we choose.
+Dimensionality estimates are plotted in \cref{fig:d_G_spherical} for a spherical 32-d Gaussian, for which the true dimensionality is known.
 \par
-Plot of results for spherical Gaussian, for which we know the correct answer.
+The different choices of temperature are again consistent, but for the rest of this paper we choose the Bayesian $\beta$, because it provides a better reflection of the uncertainty in the estimate; the others, while fluctuating around the true value, are often many standard errors away from the true value at each single point.
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/d_G_spherical.pdf}
+\end{center}
+\caption{Dimensionality estimates using the different temperatures for a spherical 32-d Gaussian.
+Again, all are consistent, but the Bayesian definition has a uncertainty which includes the true value far more consistently than for the other definitions.}
+\label{fig:d_G_spherical}
+\end{figure}
 
-\subsubsection*{Parameter compression rate}
-Plots of samples dimensionality against compression also draw attention to the \textit{speed} at which different parameters are constrained throughout the run.
-As a concrete example, consider an elongated Gaussian in a unit hypercube prior with $\mu = 0$ and $\Sigma = \mathrm{diag}\left(1, 1, 0.01, 0.01\right) \times 10^{-2}$.
+\subsubsection*{Anisotropic compression}\label{sec:elongated}
+Plots of samples dimensionality against compression also draw attention to the \textit{directions} in which the samples are constrained throughout the run.
+As a concrete example, consider an elongated Gaussian in a unit hypercube prior with $\mu = \bm{0}$ and $\Sigma = \mathrm{diag}\left(10^{-3}, 10^{-3}, 10^{-3}, 10^{-6}, 10^{-6}, 10^{-6}\right)$, for which the dimensionality estimates are plotted in \cref{fig:d_G_elongated}.
+Alongside is a view of the distribution of live points across the prior for two directions with different scales, which shows the level to which those parameters have been constrained at different times.
 \par
-Plot of dimensionality vs compression factor, showing step changes 
+A feature of nested sampling made apparent here is that parameters with high variance are initially `hidden'.
+Compression occurs in the direction which is most likely to have a sample of higher likelihood, and initially it is much easier to find a better point along the direction of a parameter that is poorly constrained.
+Lower variance parameters are constrained much later, and before that happens it appears as though those parameters have a uniform distribution.
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/d_G_elongated.pdf}
+\end{center}
+\caption{Dimensionality estimates for a Gaussian that is elongated in half of its dimensions.
+The locations of the live points in the prior are shown at three stages, indicated by the vertical lines.
+As can be seen from the live point distribution, the prior does not compress in the higher variance direction until much later in the run, and early on it appears as if those directions are completely unconstrained.}
+\label{fig:d_G_elongated}
+\end{figure}
 \par
-It is important to appreciate that at lower compressions the samples truly lie in a lower-dimensional space, because the algorithm has not yet begun to constrain the other parameters.
-Anticipating the full dimensionality of the space is therefore just as impossible as that associated with a slab-spike geometry, so in this sense such geometries contain a phase transition.
+It is important to appreciate that at lower compressions the samples truly lie in a lower-dimensional space, rather than some artefact of the way we view them.
+Anticipating the full dimensionality of the space is therefore just as impossible as that associated with a slab-spike geometry, so in this sense such geometries contain a `compressive phase transition'.
 
-\section{Endpoint prediction}
+\section{Endpoint prediction}\label{sec:endpoint}
 The time complexity of nested sampling \citep{supernest} is
 \begin{equation}
-    T \propto \nlive \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle \mathcal{T}\{ \mathrm{Impl.}\} \rangle \times \DKL.
+    T \propto \nlive \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
 \end{equation}
-The second term is the time per likelihood evaluation, which is constant.
-The third is the cost to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude.
+The second term is the relatively constant time per likelihood evaluation.
+The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude.
 The primary unknown during a run, and therefore our primary interest, is the final term: the compression required to get from prior to posterior.
+\par
+Further exploration can be found in the talk [\citenum{kcl_talk}] and the upcoming paper \citet{scaling_frontier}.
 
 \subsection{The termination prior volume}
-To be exact, one wishes to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is slightly larger in magnitude than $\DKL$ (Fig X).
+To be exact, one wishes to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}).
 The problem is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is quite far from the posterior bulk.
@@
 Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule.
-The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/\nlive$, so the total number of iterations $N_\mathrm{f}$ will be
+The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ will be
 \begin{equation}
-	N_\mathrm{f} = - \nlive \log X_\mathrm{f} .
+	N_\mathrm{f} = - n\log X_\mathrm{f} .
 \end{equation}
 
-\subsection{Regression procedure}
+\subsection{How to extrapolate?}
 A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian.
-Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function (see plot).
-The explicit form of the Gaussian that we fit is the same as that given in section X, which we shall repeat here for clarity;
+Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function.
+The parameterisation of the Gaussian that we fit is the same as that given in \cref{sec:logL}, which we shall repeat here for clarity;
 \begin{equation}\label{gaussian}
@@
 \begin{enumerate}[leftmargin=*]
-    \item Find the current dimensionality $\tilde{d}^{*}$ of the posterior at the Bayesian temperature
-    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = \tilde{d}$ to infer $\logLmax$ and $\sigma$ 
+    \item Find the current dimensionality ${d}_G^{*}$ of the posterior at the Bayesian temperature
+    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = {d}_G^{*}$ to infer $\logLmax$ and $\sigma$ 
     \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
@@
 \par
-Plot of fits to Gaussian for different distributions
-\par
 One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown it to be far less stable.
@@
 \subsection{Alternative approaches}
-Ultranest \citep{ultranest} tracks progress based on the remaining integral, approximated as $\Like_\mathrm{max} X_i$.
-does not correspond directly to runtime, since it will be zero for most of the run before the bulk is crossed.
+We consider here several alternative approaches to endpoint estimation to compare against our method. 
+\subsubsection{Integral progress}
+One alternative approach, used in \textsc{Ultranest} \citep{ultranest}, derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
+\begin{equation}\label{eq:integral_progress}
+    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X^{*}.
+\end{equation}
+There are a few problems here.
+Firstly, runtime is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width.
+Secondly, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is simply be zero for most of the run.
+Furthermore, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
 \par
-Extrapolating evidence increments is considerably less stable
+This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.
 
-\section{Results}
+\subsubsection{Extrapolating evidence increments}
+Seasoned users of nested sampling might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges.
+We do this for a spherical Gaussian and compare it to our method.
+At an intermediate stage of the run, the most recent outputs might look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate}.
+Extrapolating those data to a linear and exponential profile yields endpoint estimates plotted in the graph to the right.
+\par
+The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile.
+The increments are also not quite exponential, since the exponential fit leads to a large overprediction.
+The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result.
+One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
+\par
+More importantly, we find that for real likelihoods which have an element of noise the extrapolation often diverges, for instance when the increments do not monotonically decrease.
+Directly extrapolating the evidence increments is therefore far less stable than the previous method, and generally not a reliable method for prediction.
+\begin{figure}
+\begin{adjustbox}{valign=t, scale=0.72}
+\subfloat{\begin{tabular}{|c|c|c|}
+\hline
+iteration & log Z  & $\Delta\log Z$   \\
+\hline
+5000 & -1435.8 & 190.8 \\
+5500 & -1264.6 & 171.2 \\
+6000 & -1123.7 & 140.9 \\
+6500 & -991.5 & 132.2 \\
+7000 & -885.0 & 106.6 \\
+7500 & -790.3 & 94.7 \\
+8000 & -702.6 & 87.7 \\
+8500 & -619.7 & 82.9 \\
+9000 & -551.8 & 67.9 \\
+9500 & -492.7 & 59.1 \\
+\hline
+\end{tabular}}
+\end{adjustbox}
+\quad
+\begin{adjustbox}{valign=t}
+\subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
+\end{adjustbox}
+\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation.}
+\label{fig:inc_extrapolate}
+\end{figure}
+\begin{figure}
+\includegraphics{figures/inc_predictions.pdf}
+\caption{Endpoint predictions for a spherical Gaussian.
+Extrapolating the evidence increments in a linear/exponential manner under/overpredicts the endpoint, both of which do considerably worse than the method of extrapolating the likelihood.}
+\label{fig:inc_predictions}
+\end{figure}
+
+\section{Results}\label{sec:results}
 Now that we have a method for predicting the endpoint, it can be tested on a range of distributions.
@@
 \subsubsection{Gaussians}
-Spherical Gaussian, elongated Gaussian
+Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact.
+All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$.
+The correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained.
+\par
+We note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, which can be seen from the middle two plots.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/gauss_predictions.pdf}
+\end{center}
+\caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left).
+The correct endpoint is obtained for all but the earliest iterations, and the uncertainty is controlled by the number of live points, which can be seen from the two $d = 16$ plots.}
+\label{fig:gauss_predictions}
+\end{figure*}
+We also observe the effect of elongating the Gaussian, using the same example as \cref{sec:elongated}.
+\cref{fig:elongated_logXfs} shows a step-like trend similar to the inferred dimensionalities, reflecting the fact that the full dimensionality is undetectable at lower compression factors.
+The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/elongated_logXfs}
+\end{center}
+\caption{Endpoint prediction for an elongated Gaussian.
+At early stages, the full dimensionality is undetectable, and the endpoint is predicted to be the same as for a likelihood with three unconstrained directions.
+Only once the prior has been compressed enough to constrain the other three directions does the prediction converge to the true value.}
+\label{fig:elongated_logXfs}
+\end{figure*}
+
 \subsubsection{Cauchy}
-Cauchy\\
+One case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian.
+\cref{fig:cauchy_predictions} shows the predictions for a likelihood of the form
+\begin{equation}
+	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
+\end{equation}
+choosing $d = 10$ and allowing $\gamma$ to vary.
+The correct estimate is obtained to within standard error by about halfway, but before that is quite off.
+The key limitation is that the estimate is wrong early on not because the compression is anisotropic, or that there is a phase transition; rather, it is a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
+\par
+However, the right order of magnitude is obtained at all times, so this is not a major concern.
+The Cauchy is also a pathological case, and the same problem does not rear its head in more realistic cases, as we shall see next.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/cauchy_predictions.pdf}
+\end{center}
+\caption{Predictions for the Cauchy distribution for various widths $\gamma$ and $\nlive = 500$.
+The endpoint is underestimated for the first half of the run, but this is a limitation of the Gaussian approximation rather than a lack of information mid-run.}
+
+\label{fig:cauchy_predictions}
+\end{figure*}
+
 \subsection{Cosmological examples}
+Finally, we evaluate the method on real cosmological chains.
+\cref{fig:lcdm_logXfs} presents the endpoints (calculated after the fact) for nested sampling runs for curvature quantification on several common cosmological datasets (details in \citet{curvature_tension}).
+\par
+The \textsc{SH0ES}, \textsc{BAO} and \textsc{lensing} chains are `easy' low $\DKL$ inferences, so it is expected that the correct endpoint is inferred practically from the start.
+The \textsc{Planck} endpoints, on the other hand, are not correct until at least midway through.
+However, this is expected from the covariance of the \textsc{Planck} likelihood, which consists of principal components of many scales and therefore elongated in many dimensions.
+It is therefore of the same class as the elongated Gaussian presented in \cref{sec:elongated}; the samples exist in a lower dimensional subspace mid-run, which slowly increases to the full dimensionality only at the end of the run.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/lcdm_logXfs.pdf}
+\end{center}
+\caption{Endpoint predictions for cosmological likelihoods.
+The first three low $\DKL$ inferences get the correct endpoint from the start, while the \textsc{Planck} chain takes longer to converge because the likelihood is highly covariant and thus subject to anisotropic compression, which makes the samples lie in a lower dimensional subspace mid-run.}
+\label{fig:lcdm_logXfs}
+\end{figure*}
 
 \section{Conclusion}
+We have presented a method for predicting the endpoint of a nested sampling run, by inferring the Bayesian model dimensionality of a set of samples mid-run and using it to extrapolate the known likelihood profile.
+\par
+The method in general converges to the correct endpoint by about the halfway point, and gets the correct order of magnitude throughout.
+Consistent predictions are obtained for both toy and cosmological examples.
+The accuracy is typically limited by the information available mid-run, either because of a phase transition or because the anisotropy of the nested sampling compression.
+Pathological distributions, such as a Cauchy, lead to less stable inferences of the dimensionality and expose the limitations of a Gaussian approximation, though the order of magnitude is still correct.
+\par
+Further work can be done to experiment with more flexible basis functions for regression of the likelihood profile, so that it is less dependent on the Gaussian approximation.
 
 \appendix
-\section{The maximum live log-likelihood}
+\section{The maximum live log-likelihood}\label{sec:logLmax}
 Assume a Gaussian likelihood
@@
 \end{equation}
-i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, which is the distribution of the weighted samples. The posterior average and variance of $\log \mathcal{L}$ are given by
+i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$. The posterior average and variance of $\log \mathcal{L}$ are given by
 \begin{equation}
@@
 Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
-\begin{equation}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \quad [\log\mathcal{L}_* < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
+\begin{multline}
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \\
+    [\log\mathcal{L}_* < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
     \label{eq:PL}
-\end{equation}
+\end{multline}
 It is helpful at this stage to define a parameter
@@
 Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
-\begin{equation}
-    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1} \quad [0<y_\mathrm{max}^\mathrm{live}<1],
+\begin{multline}
+    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}\\ 
+    [0<y_\mathrm{max}^\mathrm{live}<1],
     \label{eq:Pyhat}
-\end{equation}
+\end{multline}
 which may be roughly summarised as
-\begin{align}
-    y_\mathrm{max}^\mathrm{live} \sim &1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
-    & \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
+\begin{multline}
+    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
+     \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
     \label{eq:ymax}
-\end{align}
+\end{multline}
 or in the large $d$, $n$ limit

### 2023-10-25 post_prop.tex
revising paper

\documentclass[10pt, a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=.5in, footskip=0.5in]{geometry}
\usepackage{graphicx}

\begin{document}
\begin{figure}
\begin{center}
    \includegraphics{notebooks/temperature/figures/dimensionalities_post.pdf}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
    \includegraphics{notebooks/temperature/figures/endpoints_post.pdf}
\end{center}
\end{figure}

\newpage
\begin{figure}
\begin{center}
    \includegraphics{notebooks/temperature/figures/dimensionalities_post_log.pdf}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
    \includegraphics{notebooks/temperature/figures/endpoints_post_log.pdf}
\end{center}
\end{figure}

\newpage
\begin{figure}
\begin{center}
    \includegraphics{notebooks/temperature/figures/dimensionalities_prop.pdf}
\end{center}
\end{figure}
\begin{figure}
\begin{center}
    \includegraphics{notebooks/temperature/figures/endpoints_prop.pdf}
\end{center}
\end{figure}

\end{document}

### 2023-10-25 aeons/beta.py

@@ -1,0 +1,112 @@
+import numpy as np
+from scipy.special import logsumexp
+from aeons.utils import *
+from aeons.regress import *
+
+## Different betas
+
+def get_logbeta_analytic(logXs, d, sigma):
+    return np.log(d*sigma**2) - 2/d * logXs
+
+def get_logbeta_grad(points, ndead, interval=100):
+    logX = points.logX()
+    logL = points.logL
+    delta_logX = logX.iloc[ndead+interval] - logX.iloc[ndead-interval]
+    delta_logL = logL.iloc[ndead+interval] - logL.iloc[ndead-interval]
+    return np.log(-delta_logX/delta_logL)
+
+def get_logbeta_post(points, ndead):
+    betas = np.logspace(-5, 1, 1000)
+    logX = points.logX()
+    logL = points.logL
+    logXs = logX.iloc[ndead]
+    logLs = logL.iloc[ndead]
+    logLbetasX = betas * logLs + logXs - points.logZ(beta=betas) + np.log(betas)
+    logprob = logLbetasX - logsumexp(logLbetasX)
+    mean = np.sum(np.exp(logprob)*np.log(betas))
+    var = np.sum(np.exp(logprob)*(np.log(betas)-mean)**2)
+    return mean, np.sqrt(var)
+
+def get_betas_DKL(samples):
+    from scipy.interpolate import interp1d
+    beta = np.logspace(-14, 14, 1000)
+    logX = -samples.D_KL(beta=beta)
+    beta = np.concatenate([[0], beta])
+    logX = np.concatenate([[0], logX])
+    f = interp1d(logX, beta)
+    return f(samples.logX())
+
+def get_betas_logL(samples):
+    from scipy.interpolate import interp1d
+    beta = np.logspace(-14, 14, 1000)
+    logL = samples.logL_P(beta=beta)/beta
+    i = np.argmin(logL < samples.logL.max())
+    beta = np.concatenate([[0],beta[:i+1]])
+    logL = np.concatenate([[samples.logL.min()],logL.iloc[:i+1]])
+    f = interp1d(logL, beta)
+    return f(samples.logL)
+
+### d_G
+def get_d_G_grad(points, ndead, Nset=25):
+    logbeta = get_logbeta_grad(points, ndead)
+    beta = np.exp(logbeta)
+    d_G = points.d_G(Nset, beta=beta)
+    return d_G.values, ""
+
[…]

### 2023-10-25 aeons/endpoint.py

@@ -58,6 +58,41 @@ class EndModel:
                 print(f"Iteration {ndead} of {iterations[-1]}, NaN")
         return logXfs, logXfs_std
 
+    def d_Gs(self, method, iterations, **kwargs):
+        N = len(iterations)
+        d_Gs, d_Gs_std = np.zeros(N), np.zeros(N)
+        for i, ndead in enumerate(iterations):
+            points = self.points(ndead)
+            try:
+                d_Gs_set, message = method(points, ndead, **kwargs)
+                d_Gs[i] = np.mean(d_Gs_set)
+                d_Gs_std[i] = np.std(d_Gs_set)
+                print('\r', f"Iteration {ndead} of {iterations[-1]}, {message}", end='')
+            except:
+                d_Gs[i] = np.nan
+                d_Gs_std[i] = np.nan
+                print(f"Iteration {ndead} of {iterations[-1]}, NaN")
+        return d_Gs, d_Gs_std
+    
+    def inferences(self, d_G_method, iterations, Nset=25):
+        logXfs, d_Gs = np.zeros((len(iterations), Nset)), np.zeros((len(iterations), Nset))
+        for i, ndead in enumerate(iterations):
+            points = self.points(ndead)
+            d_G = d_G_method(points, ndead, Nset)[0]
+            from scipy.stats import gaussian_kde
+            kde = gaussian_kde(d_G)
+            logL, X_mean, nk, logZdead = data(points)
+            logXf_set = np.zeros(Nset)
+            ds = kde.resample()
+            ds = ds[ds > 0]
+            for j in range(Nset):
+                X = generate_Xs(nk)
+                d = np.random.choice(ds)
+                theta = params_from_d(logL[ndead:], X[ndead:], d)
+                logXf_set[j] = logXf_formula(theta, logZdead, X_mean[ndead])
+            logXfs[i], d_Gs[i] = logXf_set, d_G
+            print('\r', f"Iteration {ndead} of {iterations[-1]}, d={d_G.mean():.1f}", end='')
+        return logXfs, d_Gs
 
 def logXf_basic(points, ndead, Nset=25):
     logL, X_mean, nk, logZdead = data(points)
@@ -88,3 +123,80 @@ def logXf_dG_DKL(points, ndead, Nset=25):
     logXf_set = logXf_set[~np.isnan(logXf_set)]
[…]

### 2023-10-25 aeons/likelihoods.py

@@ -1,5 +1,5 @@
 import numpy as np
-import torch
+# import torch
 
 
 class likelihood:

### 2023-10-25 aeons/plotting.py

@@ -1,15 +1,34 @@
 import numpy as np
 import matplotlib.pyplot as plt
 from aeons.utils import *
-from aeons.regress import GaussianRegress
+from aeons.regress import GaussianRegress, analytic_lm_params, params_from_d
 from aeons.likelihoods import full
 
+from anesthetic.utils import quantile
+from anesthetic.plot import quantile_plot_interval
+def plot_quantiles(x, y, true=None, ylim=None, label=None, ax=None, color='deepskyblue'):
+    q1 = quantile_plot_interval(1)
+    q2 = quantile_plot_interval(2)
+    y1 = np.apply_along_axis(quantile, 1, y, q1).T
+    y2 = np.apply_along_axis(quantile, 1, y, q2).T
+    if ax is None:
+        fig, ax = plt.subplots()
+    ax.fill_between(x, *y1, alpha=0.8, color=color, label=label)
+    ax.fill_between(x, *y2, alpha=0.2, color=color)
+    if true is not None:
+        ax.axhline(y=true, lw=.5, color='red', ls='--')
+        if isinstance(ylim, float):
+            ax.set_ylim(0, true*ylim)
+        elif isinstance(ylim, tuple):
+            ax.set_ylim(ylim[0]*true, ylim[1]*true)
 
-def plot_lx(samples):
+
+def plot_lx(samples, beta=1):
     logL = samples.logL
     logX_mean = samples.logX()
     fig, ax1 = plt.subplots(figsize=(6.7,2))
     ax2 = plt.twinx(ax1)
+    logL *= beta
     logL_norm = logL - logL.max()
     L_norm = np.exp(logL_norm)
     ax1.plot(logX_mean, L_norm, lw=1, color='black')
@@ -71,4 +90,26 @@ def plot_split(model, ndead, nlive, splits=2, trunc=None):
     axs[1].legend()
     fig.suptitle(f"ndead = {ndead}, true = {model.true_endpoint():.0f}, {model.true_logXf():.1f}", y=0.95)
     fig.tight_layout()
+    
+def plot_fit(model, ndead, d=None, ax=None, **kwargs):
+    points = model.points(ndead)
+    logL, X_mean, nk, logZdead = data(points)
+    logLd, Xd = logL[ndead:], X_mean[ndead:]
+    if ax is None:
+        fig, ax = plt.subplots()
[…]

### 2023-10-25 aeons/utils.py

@@ -5,10 +5,40 @@ from scipy.special import gamma, gammainc, logsumexp, gammaincinv, loggamma
 
 proj_dir = '/home/zixiao/Documents/III/project'
 aeons_dir = '/home/zixiao/Documents/III/project/aeons'
+data_dir = f'{aeons_dir}/data'
 lcdm_chains = ['BAO', 'lensing', 'lensing_BAO', 'lensing_SH0ES', 'SH0ES', 'planck', 'planck_BAO', \
               'planck_lensing', 'planck_SH0ES', 'planck_lensing_BAO', 'planck_lensing_SH0ES']
-toy_chains = ["gauss_30_01", "wedding_20_001", "cauchy_10_0001", "planck_gaussian", "gp"]
 test_chains = ['gauss_32', 'cauchy_83', 'BAO', 'planck']
+gauss_chains = ['gauss_4', 'gauss_8', 'gauss_16', 'gauss_32', 'gauss_64']
+cauchy_chains = ['cauchy_82', 'cauchy_83', 'cauchy_84', 'cauchy_85']
+all_chains = gauss_chains + cauchy_chains + lcdm_chains
+endpoints = {'BAO': 18129,
+ 'lensing': 19034,
+ 'lensing_BAO': 22108,
+ 'lensing_SH0ES': 20672,
+ 'SH0ES': 15096,
+ 'planck': 57372,
+ 'planck_BAO': 58216,
+ 'planck_lensing': 58100,
+ 'planck_SH0ES': 57424,
+ 'planck_lensing_BAO': 57852,
+ 'planck_lensing_SH0ES': 58032,
+ 'gauss_4': 11672,
+ 'gauss_8': 18665,
+ 'gauss_16': 31329,
+ 'gauss_32': 53346,
+ 'gauss_64': 90737,
+ 'cauchy_82': 20961,
+ 'cauchy_83': 29889,
+ 'cauchy_84': 39236,
+ 'cauchy_85': 48405,
+ 'slab_spike': 6593,
+ 'correlated_3d': 26107,
+ 'correlated_6d': 30234,
+ 'gp': 91721,
+ 'gauss_32_2000': 213249,
+ 'gauss_16_2000': 125308,
+ 'gauss_16_1000': 62973}
 
 def pickle_dump(filename, data):
     """Function that pickles data into a file"""
@@ -97,10 +127,10 @@ def reject_outliers(data, degree=2):
     return data[dev < degree * median_dev]
 
[…]

### 2023-10-25 notebooks/temperature/beta.py

@@ -1,0 +1,101 @@
+import numpy as np
+from scipy.special import logsumexp
+from aeons.utils import *
+from aeons.regress import *
+
+## Different betas
+
+def get_beta_analytic(logXs, d, sigma):
+    return np.exp(np.log(d*sigma**2) - 2/d * logXs)
+
+def get_betas_logL(samples):
+    from scipy.interpolate import interp1d
+    beta = np.logspace(-14, 14, 1000)
+    logL = samples.logL_P(beta=beta)/beta
+    i = np.argmin(logL < samples.logL.max())
+    beta = np.concatenate([[0],beta[:i+1]])
+    logL = np.concatenate([[samples.logL.min()],logL.iloc[:i+1]])
+    f = interp1d(logL, beta)
+    return f(samples.logL)
+    
+def get_betas_DKL(samples):
+    from scipy.interpolate import interp1d
+    beta = np.logspace(-14, 14, 1000)
+    logX = -samples.D_KL(beta=beta)
+    beta = np.concatenate([[0], beta])
+    logX = np.concatenate([[0], logX])
+    f = interp1d(logX, beta)
+    return f(samples.logX())
+
+def get_beta_post(points, ndead):
+    betas = np.logspace(-5, 1, 1000)
+    logX = points.logX()
+    logL = points.logL
+    logXs = logX.iloc[ndead]
+    logLs = logL.iloc[ndead]
+    logLbetasX = betas * logLs + logXs
+    logLbetasX -= points.logZ(beta=betas)
+    logprob = logLbetasX - logsumexp(logLbetasX)
+    mean = np.sum(np.exp(logprob)*np.log(betas))
+    var = np.sum(np.exp(logprob)*(np.log(betas)-mean)**2)
+    return mean, np.sqrt(var)
+
+### d_G
+
+def get_d_G_logL(points, ndead):
+    beta = points.beta_logL[ndead]
+    d_G = points.d_G(100, beta=beta)
+    return d_G.values, ""
+
+def get_d_G_post(points, ndead):
+    beta_mean, beta_std = get_beta_post(points, ndead)
+    betas_post = np.exp(np.random.normal(beta_mean, beta_std, 100))
+    d_G_post = points.d_G(beta=betas_post)
+    return d_G_post.values, f"{len(betas_post)} samples"
+
+def get_d_G_prop(points, ndead):
+    betas = np.logspace(-10, 0, 1000)
[…]

### 2023-10-25 notebooks/temperature/calc_betas.py

@@ -1,0 +1,26 @@
+# Calculates microcanonical, canonical and Bayesian temperatures and writes them to file
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.regress import *
+from aeons.endpoint import *
+from aeons.plotting import *
+from aeons.beta import *
+figsettings()
+
+name, samples = get_samples('gauss_32')
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+true_logXf = model.true_logXf()
+
+iterations = make_iterations(true_endpoint, 100)
+logXs = samples.logX().iloc[iterations]
+logbetas_mean, logbetas_std = np.ones(len(iterations)), np.ones(len(iterations))
+logbetas_grad = np.ones(len(iterations))
+for i, ndead in enumerate(iterations):
+    points = points_at_iteration(samples, ndead)
+    logbetas_mean[i], logbetas_std[i] = get_logbeta_post(points, ndead)
+    logbetas_grad[i] = get_logbeta_grad(points, ndead)
+    print('\r', i, end='')
+logbetas_logL = np.log(get_betas_logL(samples)[iterations])
+# write_to_txt(f'{data_dir}/logbetas/{name}.txt', [iterations, logbetas_grad, logbetas_logL, logbetas_mean, logbetas_std])

### 2023-10-25 notebooks/temperature/inferences.py

@@ -1,0 +1,58 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.regress import *
+from aeons.endpoint import *
+from aeons.plotting import *
+from aeons.beta import *
+figsettings()
+
+def inferences(d_G_method, points, ndead, Nset=25):
+    d_G = d_G_method(points, ndead, Nset)[0]
+    from scipy.stats import gaussian_kde
+    kde = gaussian_kde(d_G)
+    logL, X_mean, nk, logZdead = data(points)
+    logXf_set = np.zeros(Nset)
+    ds = kde.resample()
+    ds = ds[ds > 0]
+    for i in range(Nset):
+        X = generate_Xs(nk)
+        d = np.random.choice(ds)
+        theta = params_from_d(logL[ndead:], X[ndead:], d)
+        logXf_set[i] = logXf_formula(theta, logZdead, X_mean[ndead])
+    # logXf_set = logXf_set[~np.isnan(logXf_set)]
+    # logXf_set = reject_outliers(logXf_set)
+    return [logXf_set, d_G], f"{len(logXf_set)} samples, {d_G.mean():.1f}"
+
+# Take in as command line argument name of chain
+import sys
+chain = sys.argv[1]
+name, samples = get_samples(chain)
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+samples['beta_logL'] = get_betas_logL(samples)
+print(f'Loaded chain {name}')
+
+Nset = 25
+iterations = make_iterations(true_endpoint, 50)
+logXs = samples.logX().iloc[iterations]
+
+logXfs_grad, d_Gs_grad = np.zeros((len(iterations), Nset)), np.zeros((len(iterations), Nset))
+logXfs_logL, d_Gs_logL = np.zeros((len(iterations), Nset)), np.zeros((len(iterations), Nset))
+logXfs_post, d_Gs_post = np.zeros((len(iterations), Nset)), np.zeros((len(iterations), Nset))
+for i, ndead in enumerate(iterations):
+    points = points_at_iteration(samples, ndead)
+    logXfs_grad[i], d_Gs_grad[i] = inferences(get_d_G_grad, points, ndead, Nset)[0]
[…]

### 2023-10-25 notebooks/temperature/run_pypolychord.py

@@ -1,0 +1,54 @@
+from numpy import pi, log
+import pypolychord
+from pypolychord.settings import PolyChordSettings
+from pypolychord.priors import UniformPrior
+from scipy.special import logsumexp
+
+#| Define a four-dimensional spherical gaussian likelihood,
+#| width sigma=0.1, centered on the 0 with one derived parameter.
+#| The derived parameter is the squared radius
+
+nDims = 8
+nDerived = 0
+sigma1 = 0.1
+sigma2 = 0.01
+
+def likelihood(theta):
+    """ Simple Gaussian Likelihood"""
+
+    nDims = len(theta)
+    r2 = sum(theta**2)
+    logL1 = -log(2*pi*sigma1*sigma1)*nDims/2.0
+    logL1 += -r2/2/sigma1/sigma1
+
+    logL2 = -log(2*pi*sigma2*sigma2)*nDims/2.0
+    logL2 += -r2/2/sigma2/sigma2
+    logL = logsumexp([logL1, logL2])
+
+    return logL, []
+
+#| Define a box uniform prior from -1 to 1
+
+def prior(hypercube):
+    """ Uniform prior from [-1,1]^D. """
+    return UniformPrior(-1, 1)(hypercube)
+
+#| Optional dumper function giving run-time read access to
+#| the live points, dead points, weights and evidences
+
+def dumper(live, dead, logweights, logZ, logZerr):
+    print(logZ)
+    print("Last dead point:", dead[-1])
+
+#| Initialise the settings
+
+settings = PolyChordSettings(nDims, nDerived)
+settings.file_root = 'slab_spike'
+settings.nlive = 200
+settings.do_clustering = True
+settings.read_resume = False
+settings.precision_criterion = 1e-4
+
+#| Run PolyChord
+
+output = pypolychord.run_polychord(likelihood, nDims, nDerived, settings, prior, dumper)

### 2023-10-25 paper/figures/anatomy.py

@@ -1,79 +1,0 @@
-import matplotlib.pyplot as plt
-import numpy as np
-from matplotlib import rcParams
-from aeons.utils import figsettings
-
-figsettings()
-fig, ax = plt.subplots()
-
-logLmax = 0
-d = 30
-n = 1000
-avlogL = logLmax-d/2
-
-logL = np.linspace(logLmax-d/2-np.sqrt(d/2)*3.5,0,1000)
-logpdf = logL - logLmax + (d/2-1)*np.log(logLmax-logL)
-logpdf -= np.max(logpdf)
-ax.fill(logL, np.exp(logpdf), color='gray', alpha=.4, label='Posterior')
-
-logLs = logLmax-d/2/np.e
-logL = np.linspace(logLs,0,1000)
-logpdf = (d/2-1) * np.log(logLmax - logL)
-logL = np.concatenate([[logLs], logL])
-logpdf = np.concatenate([[-np.inf], logpdf])
-logpdf -= np.max(logpdf)
-ax.fill(logL, .7*np.exp(logpdf), color='deepskyblue', alpha=0.5, label='Live points')
-
-from scipy.special import logsumexp
-
-logL = np.linspace(logLs,0,1000)
-logpdf = (d/2-1)*np.log(logLmax - logL) + (n-1)* logsumexp([ d/2*np.log(logLmax-logLs)*np.ones_like(logL), d/2*np.log(logLmax-logL)] , b=[np.ones_like(logL),-np.ones_like(logL)], axis=0)
-logpdf -= np.max(logpdf)
-ax.plot(logL, .7*np.exp(logpdf), lw=.5, color='red', label='Max live point')
-
-
-logLmaxlive = logLmax - d/2/np.e + np.log(n)/np.e
-
-pad_text = 0.02
-l1 = 1.2
-ax.axvline(avlogL, color='k', linestyle=':', lw=1)
-ax.axvline(logLmax, color='k', linestyle=':', lw=1)
-ax.plot([avlogL, logLmax], [l1,l1], 'k-', lw=.5)
-ax.annotate(r'$\frac{d}{2}$', ((avlogL+logLmax)/2,l1+pad_text), ha='center', va='bottom', fontsize=8)
-
-l2 = 0.7
-ax.plot([avlogL-np.sqrt(d/2), avlogL+np.sqrt(d/2)], [l2, l2], 'k-', lw=.5)
[…]

### 2023-10-25 paper/figures/beta_comparison.py

@@ -1,0 +1,37 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.regress import *
+from aeons.endpoint import *
+from aeons.plotting import *
+from aeons.beta import *
+figsettings()
+
+fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
+
+name, samples = get_samples('gauss_32')
+iterations, logbetas_grad, logbetas_logL, logbetas_mean, logbetas_std = read_from_txt(f'{data_dir}/logbetas/{name}.txt')
+logXs = samples.logX().iloc[iterations]
+plot_std(-logXs, logbetas_mean, logbetas_std, label=f'Bayesian', ax=axs[0], color='red')
+axs[0].plot(-logXs, logbetas_logL, color='orange', label='Canonical')
+axs[0].plot(-logXs, logbetas_grad, color='black', lw=2, label='Microcanonical')
+
+name, samples = get_samples('slab_spike')
+iterations, logbetas_grad, logbetas_logL, logbetas_mean, logbetas_std = read_from_txt(f'{data_dir}/logbetas/{name}.txt')
+logXs = samples.logX().iloc[iterations]
+plot_std(-logXs, logbetas_mean, logbetas_std, label=f'Bayesian', ax=axs[1], color='red')
+axs[1].plot(-logXs, logbetas_logL, color='orange', label='Canonical')
+axs[1].plot(-logXs, logbetas_grad, color='black', lw=2, label='Microcanonical')
+
+for ax in axs:
+    ax.set_xlabel(r"$-\log X$")
+
+axs[0].set_ylabel(r'$\log \beta$')    
+axs[1].set_ylim(-8, 8)
+
+handles, labels = axs[0].get_legend_handles_labels()
+axs[0].legend(handles[::-1], labels[::-1], ncol=3, bbox_to_anchor=(2.2, -0.4), fontsize=7)
+axs[0].set_title('Spherical Gaussian', fontsize=7)
+axs[1].set_title('Slab-spike Gaussian', fontsize=7)
+fig.suptitle('Temperature vs compression for different distributions', y=1.15, fontsize=8)
+fig.savefig('beta_comparison.pdf', bbox_inches='tight', pad_inches=0)

### 2023-10-25 paper/figures/beta_move.py

@@ -7,7 +7,7 @@ def weights(logL, logX, beta):
     logw = beta*logL + logX
     return np.exp(logw - np.max(logw))
 
-name, samples = get_samples('toy', 'planck_gaussian')
+name, samples = get_samples('planck_gaussian')
 samples = samples.iloc[:int(len(samples)*0.7)]
 logL = samples.logL
 logX = samples.logX()

### 2023-10-25 paper/figures/cauchy_predictions.py

@@ -7,12 +7,15 @@ fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
 for i, exp in enumerate([2, 3, 4, 5]):
     ax = axs[i]
     name, samples = get_samples(f'cauchy_8{exp}')
-    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/dG_range/02_cauchy_8{exp}.txt')
+    true_endpoint = endpoints[name]
+    iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
+    logXfs = np.array(logXfs)
     logXs = samples.logX().iloc[iterations]
-    logXf_true = samples.logX().iloc[int(true_endpoint[0])]
-    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 2), color='deepskyblue', ax=ax)
+    logXf_true = samples.logX().iloc[true_endpoint]
+    plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 2), color='deepskyblue', ax=ax)
     ax.set_title(f'$\\gamma=10^{{{-exp}}}$, $\\mathcal{{D}}_\\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
     ax.set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
+    ax.margins(x=0)
 
 fig.supxlabel('$-\\log X$', y=0.02)
 fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.01)

### 2023-10-25 paper/figures/dG_beta.py

@@ -1,54 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from aeons.utils import *
-from aeons.plotting import plot_std
-from matplotlib.gridspec import GridSpec
-from anesthetic import make_2d_axes
-figsettings()
-
-name, samples = get_samples('toy', 'gauss_36')
-iterations, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_36_dG.txt')
-
-fig = plt.figure(figsize=(3.5, 4))
-gs = GridSpec(3, 3, wspace=0.1, hspace=.7)
-_, ax1 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 0])
-_, ax2 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 1])
-_, ax3 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 2])
-ax4 = fig.add_subplot(gs[1, :])
-ax5 = fig.add_subplot(gs[2, :])
-
-samples.set_beta(1e-6).plot_2d(ax1)
-samples.set_beta(1e-4).plot_2d(ax2)
-samples.set_beta(1).plot_2d(ax3)
-
-betas = np.logspace(-6, 0, 100)
-# samples.d_G(beta=betas).plot(ax=ax4, logx=True, lw=1, color='k')
-d_Gs_beta = samples.d_G(nsamples=10, beta=betas).unstack()
-plot_std(np.log10(betas), d_Gs_beta.mean(axis=1), d_Gs_beta.std(axis=1), ax=ax4)
-plot_std(-samples.logX().iloc[iterations], d_Gs, d_Gs_std, ax=ax5)
-
-ax4.axvline(x=-4.3, lw=.5, ls='--', color='gray')
-ax4.axvline(x=-2, lw=.5, ls='--', color='gray')
-ax5.axvline(x=.5, lw=.5, ls='--', color='gray')
-ax5.axvline(x=2.8, lw=.5, ls='--', color='gray')
-
-for ax in [ax1, ax2, ax3]:
-    ax = ax.iloc[0,0]
-    ax.set_xlim(0, 1)
-    ax.set_ylim(0, 1)
-    ax.set_xticks([])
[…]

### 2023-10-25 paper/figures/dG_elongated.py

@@ -1,0 +1,54 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import plot_std
+from matplotlib.gridspec import GridSpec
+from anesthetic import make_2d_axes
+figsettings()
+
+name, samples = get_samples('correlated_6d')
+iterations, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_36_dG.txt')
+
+fig = plt.figure(figsize=(3.5, 4))
+gs = GridSpec(3, 3, wspace=0.1, hspace=.7)
+_, ax1 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 0])
+_, ax2 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 1])
+_, ax3 = make_2d_axes([0, 1], upper=False, diagonal=False, fig=fig, subplot_spec=gs[0, 2])
+ax4 = fig.add_subplot(gs[1, :])
+ax5 = fig.add_subplot(gs[2, :])
+
+samples.set_beta(1e-6).plot_2d(ax1)
+samples.set_beta(1e-4).plot_2d(ax2)
+samples.set_beta(1).plot_2d(ax3)
+
+betas = np.logspace(-6, 0, 100)
+# samples.d_G(beta=betas).plot(ax=ax4, logx=True, lw=1, color='k')
+d_Gs_beta = samples.d_G(nsamples=10, beta=betas).unstack()
+plot_std(np.log10(betas), d_Gs_beta.mean(axis=1), d_Gs_beta.std(axis=1), ax=ax4)
+plot_std(-samples.logX().iloc[iterations], d_Gs, d_Gs_std, ax=ax5)
+
+ax4.axvline(x=-4.3, lw=.5, ls='--', color='gray')
+ax4.axvline(x=-2, lw=.5, ls='--', color='gray')
+ax5.axvline(x=.5, lw=.5, ls='--', color='gray')
+ax5.axvline(x=2.8, lw=.5, ls='--', color='gray')
+
+for ax in [ax1, ax2, ax3]:
+    ax = ax.iloc[0,0]
+    ax.set_xlim(0, 1)
+    ax.set_ylim(0, 1)
+    ax.set_xticks([])
+    ax.set_yticks([])
[…]

### 2023-10-25 paper/figures/d_G_elongated.py

@@ -1,0 +1,46 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import *
+from matplotlib.gridspec import GridSpec
+figsettings()
+
+name, samples = get_samples('correlated_6d', reduced=False)
+iterations, *d_Gs = read_from_txt(f'{data_dir}/d_Gs/post/{name}.txt')
+logXs = samples.logX().iloc[iterations]
+d_Gs = np.array(d_Gs)
+
+fig = plt.figure(figsize=(3.5, 3))
+gs = GridSpec(2, 3, wspace=0.1, hspace=0.35)
+ax1 = fig.add_subplot(gs[0, 0])
+ax2 = fig.add_subplot(gs[0, 1])
+ax3 = fig.add_subplot(gs[0, 2])
+ax4 = fig.add_subplot(gs[1, :])
+axs = [ax1, ax2, ax3, ax4]
+
+logXstars = [-1, -8, -30]
+logX = samples.logX()
+# Find index of logX closest to logXstar
+ndeads = [np.argmin(np.abs(logX - logXstar)) for logXstar in logXstars]
+
+for i, ndead in enumerate(ndeads):
+    ax = axs[i]
+    live_points = points_at_iteration(samples, ndead).iloc[ndead:]
+    live_points.drop_weights().plot.scatter(0, 4, ax=ax, s=5)
+    ax.set_aspect('equal')
+    ax.set_xlim(0, 1)
+    ax.set_ylim(0, 1)
+    ax.set_xticks([])
+    ax.set_yticks([])
+    ax.set_xlabel('')
+    ax.set_ylabel('')
+    ax4.axvline(-logXstars[i], lw=1)
+ax2.set_xlabel(r"$\sigma = 10^{-3}$", fontsize=8)
+ax1.set_ylabel(r"$\sigma = 10^{-6}$", fontsize=8)
+plot_quantiles(-logXs, d_Gs, 6, (0, 1.2), color='lightseagreen', ax=ax4)
+ax4.set_title('Inferred dimensionality as run proceeds', fontsize=8)
+ax4.set_xlabel(r'$-\log X$', fontsize=8)
+ax4.set_ylabel(r'$\hat{d}_G$', fontsize=8, rotation=0, labelpad=10)
+ax4.margins(x=0)
+fig.suptitle('Set of live points as run proceeds', fontsize=8, y=0.92)
+fig.savefig('d_G_elongated.pdf', pad_inches=0.02, bbox_inches='tight')

### 2023-10-25 paper/figures/d_G_spherical.py

@@ -1,0 +1,28 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.regress import *
+from aeons.endpoint import *
+from aeons.plotting import *
+from aeons.beta import *
+figsettings()
+
+name, samples = get_samples('gauss_32')
+iterations_grad, *d_Gs_grad = read_from_txt(f'{data_dir}/d_Gs/grad/{name}.txt')
+iterations_logL, *d_Gs_logL = read_from_txt(f'{data_dir}/d_Gs/logL/{name}.txt')
+iterations_post, *d_Gs_post = read_from_txt(f'{data_dir}/d_Gs/post/{name}.txt')
+d_Gs_grad, d_Gs_logL, d_Gs_post = np.array(d_Gs_grad), np.array(d_Gs_logL), np.array(d_Gs_post)
+logXs_grad = samples.logX().iloc[iterations_grad]
+logXs_logL = samples.logX().iloc[iterations_logL]
+logXs_post = samples.logX().iloc[iterations_post]
+
+fig, ax = plt.subplots(figsize=(3.5, 1.3))
+plot_quantiles(-logXs_post, d_Gs_post, samples.d_G(), ylim=(0, 1.5), ax=ax, color='red', label='Bayesian')
+plot_quantiles(-logXs_logL, d_Gs_logL, samples.d_G(), ylim=(0, 1.5), ax=ax, color='orange', label='Canonical')
+plot_quantiles(-logXs_grad, d_Gs_grad, samples.d_G(), ylim=(0, 1.5), ax=ax, color='black', label='Microcanonical')
+handles, labels = ax.get_legend_handles_labels()
+ax.legend(handles[::-1], labels[::-1], ncol=3, bbox_to_anchor=(1.05, -0.4))
+ax.set_xlabel(r'$-\log X$')
+ax.set_ylabel(r'$d_G$', rotation=0, labelpad=10)
+ax.set_title('Intermediate dimension for spherical 32-d Gaussian')
+fig.savefig('d_G_spherical.pdf', bbox_inches='tight', pad_inches=0)

### 2023-10-25 paper/figures/dead_measure.py

@@ -6,20 +6,21 @@ from matplotlib.lines import Line2D
 import numpy as np
 from matplotlib import patheffects, rcParams
 rcParams['path.effects'] = [patheffects.withStroke(linewidth=1, foreground='white')]
+from aeons.utils import *
 
 nlive = 100
-samples = planck_gaussian(nlive)
+samples = get_samples('planck_gaussian_100', reduced=False)[1]
 
 x = 'omegabh2'
 y = 'omegach2'
 
-fig, axes = plt.subplots(1,4, figsize=(7*4*0.5,7*0.5*1.5))
+fig, axes = plt.subplots(1,4, figsize=(7,2))
 
 for ax in axes:
     ax.set_xticks([])
     ax.set_yticks([])
     ax.set_box_aspect(1)
-    ax.plot(samples[x], samples[y], 'C0.', ms=0.5)
+    ax.scatter(samples[x], samples[y], s=0.1)
 
 axes[0].get_children()[-1].get_zorder()
 axes[0].set_zorder(10)
@@ -84,18 +85,18 @@ ax = fig.add_subplot(1,4,4)
 ax.set_box_aspect(1)
 ax.set_xticks([])
 ax.set_yticks([])
-ax.plot(samples[x], samples[y], 'C0.', ms=0.5)
+ax.scatter(samples[x], samples[y], s=0.2)
 axes = np.concatenate([axes, [ax]])
 
 fig.canvas.draw()
 fig.canvas.flush_events()
-plot_live_points = False
+plot_live_points = True
 
 for k, (i, ax0, ax1) in enumerate(zip(range(nlive*20, nlive*100, nlive*20), axes[:-1], axes[1:])):
     if plot_live_points:
         live = samples.live_points(i)
-        ax0.plot(live[x], live[y], f'C{k+1}o')
-        ax1.plot(live[x], live[y], f'C{k+1}o')
+        ax0.scatter(live[x], live[y], s=2, color=f'C{k+1}')
+        ax1.scatter(live[x], live[y], s=2, color=f'C{k+1}')
     draw_zoom_lines(ax0, ax1, *get_box(i))
 
 fig.canvas.draw()

### 2023-10-25 paper/figures/elongated_logXfs.py

@@ -1,0 +1,26 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.plotting import *
+figsettings()
+
+
+name, samples = get_samples('correlated_6d')
+iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
+true_endpoint = endpoints[name]
+logXfs = np.array(logXfs)
+logXs = samples.logX().iloc[iterations]
+logXf_true = samples.logX().iloc[true_endpoint]
+logXf_3d = get_samples('correlated_3d')[1].logX().iloc[endpoints['correlated_3d']]
+
+fig, ax = plt.subplots(figsize=(3, 1.6))
+plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 1.2), color='deepskyblue', ax=ax)
+
+ax.axhline(-logXf_3d, color='orange', linestyle='--', linewidth=.5)
+ax.set_title('Elongated Gaussian', fontsize=8)
+ax.set_yticks([0, -logXf_3d, -logXf_true], [0, r"$-\log X_\mathrm{f}^\mathrm{3d}$", r"$-\log X_\mathrm{f}^\mathrm{true}$"], rotation=0)
+ax.set_xlabel('$-\\log X$')
+ax.margins(x=0)
+
+fig.savefig('elongated_logXfs.pdf', pad_inches=0, bbox_inches='tight')
+

### 2023-10-25 paper/figures/gauss_params.py

@@ -1,23 +1,0 @@
-import matplotlib.pyplot as plt
-
-from aeons.utils import *
-from aeons.plotting import *
-figsettings()
-
-samples = get_samples('toy', 'gauss_8')[1]
-iterations, logLmaxs, logLmaxs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_8_logLmaxs.txt')
-_, ds, ds_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_8_ds.txt')
-_, sigmas, sigmas_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_8_sigmas.txt')
-logXs = samples.logX().iloc[iterations]
-
-fig, axs = plt.subplots(1, 3, figsize=(7, 1.6)) 
-plot_std(-logXs, logLmaxs, logLmaxs_std, true=0, ax=axs[0], color='salmon')
-plot_std(-logXs, ds, ds_std, true=8, ylim=(0, 2), ax=axs[1], color='salmon')
-plot_std(-logXs, sigmas, sigmas_std, true=0.01, ylim=(0, 3), ax=axs[2], color='salmon')
-axs[0].set_title("$\\log \\mathcal{L}_\\mathrm{max}$", fontsize=8)
-axs[1].set_title("$d$", fontsize=8)
-axs[2].set_title("$\\sigma$", fontsize=8)
-# fig.suptitle('Inferred parameters', y=.9, fontsize=8)
-fig.supxlabel(r"$-\log X$", y=.05, fontsize=8)
-fig.tight_layout()
-fig.savefig('gauss_params.pdf', pad_inches=0, bbox_inches='tight')

### 2023-10-25 paper/figures/gauss_predictions.py

@@ -3,19 +3,23 @@ from aeons.utils import *
 from aeons.plotting import *
 figsettings()
 
-d = [4, 8, 16, 32]
-fig, axs = plt.subplots(1, 4, figsize=(7, 1.8))
+d = [8, 16, '16_2000', 32]
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
 axs = axs.flatten()
 for i, nDims in enumerate(d):
     name, samples = get_samples(f'gauss_{nDims}')
-    iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/dG_range/02_gauss_{nDims}.txt')
-    logXf_true = samples.logX().iloc[int(true_endpoint[0])]
+    true_endpoint = endpoints[name]
+    logXf_true = samples.logX().iloc[true_endpoint]
+    iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
+    logXfs = np.array(logXfs)
     logXs = samples.logX().iloc[iterations]
-    # endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=500)
-    plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0, 1.5), ax=axs[i])
+    plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 1.5), color='deepskyblue', ax=axs[i])
     axs[i].set_title(rf'$d$ = {nDims}, $\mathcal{{D}}_\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
     axs[i].set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
-fig.supxlabel('$-\\log X$', y=0.1, fontsize=8)
-fig.supylabel('$-\\log \hat{X}_\mathrm{f}$',  fontsize=8)
+    axs[i].margins(x=0)
+
+axs[2].set_title(r"$d$ = 16, $n_\mathrm{live} = 2000$", fontsize=8)
+fig.supxlabel('$-\\log X$', y=0.02, fontsize=8)
+fig.supylabel('$-\\log \hat{X}_\mathrm{f}$', fontsize=8, x=0.01)
 fig.tight_layout()
 fig.savefig('gauss_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-10-25 paper/figures/inc_predictions.py

@@ -5,7 +5,7 @@ from aeons.plotting import *
 from aeons.increment import IncrementModel
 figsettings()
 
-name, samples = get_samples('toy', 'gauss_16')
+name, samples = get_samples('gauss_16')
 inc = IncrementModel(samples, nlive=500)
 true_endpoint = inc.true_endpoint()
 
@@ -19,19 +19,22 @@ for i, ndead in enumerate(iterations):
     endpoints_exp[i] = inc.get_endpoint(ndead, method='exp', steps=steps)
     endpoints_linear[i] = inc.get_endpoint(ndead, method='linear', steps=2)
 
-ndeads, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/gauss/gauss_16.txt')
+true_endpoint = endpoints[name]
+ndeads, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/gauss_16.txt')
+logXfs = np.array(logXfs)
 logXs = samples.logX().iloc[ndeads]
-endpoints, endpoints_std = calc_endpoints(ndeads, logXs, logXfs, logXfs_std, nlive=500)
-true_logXf = samples.logX().iloc[true_endpoint].values
+
+true_logXf = samples.logX().iloc[true_endpoint]
 
 fig, ax = plt.subplots(figsize=(3.5, 1.5))
 ax.plot(-logXs, endpoints_exp/500, lw=.8, color='green', label='Exponential')
 ax.plot(-logXs, endpoints_linear/500, lw=.8, color='orange', label='Linear')
-plot_std(-logXs, -logXfs, logXfs_std, -true_logXf, ax=ax, label='Fit $\\mathcal{L}(X)$')
+plot_quantiles(-logXs, -logXfs, -true_logXf, ax=ax, label='Fit $\\mathcal{L}(X)$')
 
 ax.set_ylim(0, 160)
 ax.set_ylabel(r'$-\log \hat{X}_f$')
 ax.set_xlabel(r'$-\log X$')
 ax.legend(loc='upper right', fontsize=6)
+ax.margins(x=0)
 
 fig.savefig('inc_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2023-10-25 paper/figures/last_live_point.py

@@ -1,30 +1,24 @@
 import matplotlib.pyplot as plt
 from aeons.utils import *
+from aeons.beta import *
 import matplotlib.patches as patches
 figsettings()
 
-def get_beta(points, ndead):
-    logX = points.logX().iloc[ndead]
-    if logX < -points.D_KL():
-        return 1
-    def func(beta):
-        return logX + points.set_beta(beta).D_KL()
-    from scipy import optimize
-    res = optimize.root_scalar(func, bracket=[0, 1])
-    return res.root
-
-name, samples = get_samples('toy', 'planck_gaussian')
+name, samples = get_samples('planck_gaussian')
 ndead = 8000
-points = points_at_iteration(samples, ndead).recompute()
-betas = [1, get_beta(points, ndead) * 1.5, get_beta(points, ndead) * 0.5, get_beta(points, ndead) * 0.05]
-labels = ['$\\beta=1$', '$\\beta^* < \\beta < 1$', None, '$\\beta = \\beta^*$']
-colors = ['black', 'gray', 'darkgray', 'orange']
+points = points_at_iteration(samples, ndead)
+beta_grad = np.exp(get_logbeta_grad(points, ndead))
+betas = [0, beta_grad, 1]
+labels = ['$\\beta=1$', '$\\beta = \\beta^*$', '$\\beta = 0$']
+colors = ['darkgray', 'orange', 'black']
 
 fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
 for i, beta in enumerate(betas):
     axs[0].plot(-points.logX(), np.exp(beta*(points.logL - points.logL.max())), 'x', ms=1, color=colors[i], label=labels[i])
     axs[0].plot(-points.logX(), np.exp(beta*(points.logL - points.logL.max())), lw=.5, alpha=.2, color=colors[i])
-    axs[1].plot(-points.logX(), weights(points.logL, points.logX(), beta=beta), lw=.8, color=colors[i])
+    logw = beta * points.logL + points.logX()
+    logw -= logw.max()
+    axs[1].fill_between(-points.logX(), 0, np.exp(logw), lw=.8, color=colors[i])
 axs[0].plot(-points.logX().iloc[-1], np.exp(points.logL.iloc[-1] - points.logL.max()), 'x', ms=2, color='k')
 
[…]

### 2023-10-25 paper/figures/lcdm_logXfs.py

@@ -1,0 +1,28 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.regress import *
+from aeons.endpoint import *
+from aeons.plotting import *
+from aeons.beta import *
+figsettings()
+
+chains = ['SH0ES', 'BAO', 'lensing', 'planck']
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
+for i, chain in enumerate(chains):
+    ax = axs[i]
+    name, samples = get_samples(chain)
+    true_endpoint = endpoints[name]
+    iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
+    logXfs = np.array(logXfs)
+    logXs = samples.logX().iloc[iterations]
+    logXf_true = samples.logX().iloc[true_endpoint]
+    plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 1.5), color='deepskyblue', ax=ax)
+    ax.set_title(f'{name.upper()}, $\\mathcal{{D}}_\\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
+    ax.set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
+    ax.margins(x=0)
+
+fig.supxlabel('$-\\log X$', y=0.02)
+fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.01)
+fig.tight_layout()
+fig.savefig('lcdm_logXfs.pdf', pad_inches=0, bbox_inches='tight')

### 2023-10-25 paper/figures/lcdm_predictions.py

@@ -1,10 +1,9 @@
 import numpy as np
 import matplotlib.pyplot as plt
-
 from aeons.utils import *
 from aeons.plotting import *
-
 figsettings()
+
 fig, axs = plt.subplots(3, 4, figsize=(7, 4))
 axs = axs.flatten()
 fig.delaxes(axs[-1])

### 2023-10-25 paper/figures/live_point_distribution.py

@@ -1,59 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-import matplotlib.patches as patches
-
-from aeons.utils import *
-from aeons.likelihoods import full
-
-theta = [0, 10, 0.01]
-logX = np.linspace(-60, 0, 1000)
-X = np.exp(logX)
-logL = full.func(X, theta)
-L = np.exp(logL - logL.max())
-LX = L * X
-Ztrue = np.trapz(L, X)
-LlogL = np.nan_to_num(L/Ztrue * np.log(L/Ztrue))
-DKL = np.trapz(LlogL, X)
-
-ndead = 7000
-nlive = 1000
-
-def gaussian(x, mu, sig):
-    return 1/(np.sqrt(2*np.pi)*sig)*np.exp(-np.power((x - mu)/sig, 2)/2)
-ElogXlivemin = -ndead/nlive - np.log(nlive) - 0.577
-StdlogXlivemin = np.sqrt(ndead/nlive**2 + (np.pi**2)/6)
-live_logX = logX[(logX > ElogXlivemin) * (logX < -ndead/nlive)]
-
-figsettings()
-fig, ax = plt.subplots(figsize=(8, 3))
-ax.plot(logX, L/L.max(), color='black', label='Likelihood', lw=.5)
-ax.fill(logX, LX/LX.max(), alpha=0.4, color='gray', label='Posterior mass')
-ax.fill_between(live_logX, np.zeros_like(live_logX), 0.5*np.exp(live_logX + ndead/nlive), alpha=0.5, color='deepskyblue', label='Live points') 
-ax.plot(logX, gaussian(logX, ElogXlivemin, StdlogXlivemin), lw=.5, color='red', label='Min live point')
-ax.axvline(-ndead/nlive, color='deepskyblue', lw=.5, ls='--')
-
-# Arrows showing log n + gamma
-l = 0.32
-pad_text = 0.05
-ax.annotate(r'$\log n+\gamma$', ((-ndead/nlive+ ElogXlivemin)/2 - 3,l+pad_text), ha='center', va='bottom', fontsize=8)
-arrow_start = (-ndead/nlive, l)
-arrow_end = (ElogXlivemin, l)
-arrow = patches.FancyArrowPatch(arrow_start, arrow_end, arrowstyle='->', mutation_scale=5)
-ax.add_patch(arrow)
-
-# Arrow showing direction of iterations
[…]

### 2023-10-25 paper/figures/logL_distribution.py

@@ -1,0 +1,84 @@
+import matplotlib.pyplot as plt
+import numpy as np
+from matplotlib import rcParams
+from aeons.utils import figsettings
+
+figsettings()
+fig, ax = plt.subplots()
+
+logLmax = 0
+d = 30
+n = 1000
+avlogL = logLmax-d/2
+
+logLs = logLmax-d/2/np.e
+logLdead = np.linspace(logLmax-d/2-np.sqrt(d/2)*4, logLs, 1000)
+logLlive = np.linspace(logLs,0,1000)
+logLall = np.concatenate([logLdead, logLlive])
+
+logpdf = logLall - logLmax + (d/2-1)*np.log(logLmax-logLall)
+logpdf -= np.max(logpdf)
+ax.fill(logLall, np.exp(logpdf), color='gray', alpha=.4, label='Posterior')
+
+ldist = .5
+pdf_dead = d/2 / (-logLdead) 
+pdf_dead /= np.max(pdf_dead)
+ax.fill_between(logLdead, 0, ldist*pdf_dead, color='deepskyblue', alpha=0.5, label='Dead/live distribution')
+
+logpdf_live = (d/2-1) * np.log(logLmax - logLlive)
+logpdf_live -= np.max(logpdf_live)
+ax.fill_between(logLlive, 0, ldist*np.exp(logpdf_live), color='deepskyblue', alpha=0.5)
+
+
+from scipy.special import logsumexp
+logpdf_maxlive = (d/2-1)*np.log(logLmax - logLlive) + (n-1)* logsumexp([ d/2*np.log(logLmax-logLs)*np.ones_like(logLlive), d/2*np.log(logLmax-logLlive)] , b=[np.ones_like(logLlive),-np.ones_like(logLlive)], axis=0)
+logpdf_maxlive -= np.max(logpdf_maxlive)
+ax.plot(logLlive, .7*np.exp(logpdf_maxlive), lw=.5, color='red', label='Max live point')
+
+
+logLmaxlive = logLmax - d/2/np.e + np.log(n)/np.e
+
+pad_text = 0.02
+l1 = 1.2
+ax.axvline(avlogL, color='k', linestyle=':', lw=1)
+ax.axvline(logLmax, color='k', linestyle=':', lw=1)
+ax.plot([avlogL, logLmax], [l1,l1], 'k-', lw=.5)
[…]

### 2023-10-25 paper/figures/logX_distribution.py

@@ -1,0 +1,73 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib.patches as patches
+from aeons.utils import *
+from aeons.likelihoods import full
+figsettings()
+
+theta = [0, 10, 0.01]
+logX = np.linspace(-60, 0, 1000)
+X = np.exp(logX)
+logL = full.func(X, theta)
+L = np.exp(logL - logL.max())
+LX = L * X
+Ztrue = np.trapz(L, X)
+LlogL = np.nan_to_num(L/Ztrue * np.log(L/Ztrue))
+DKL = np.trapz(LlogL, X)
+
+logXf = -44.4
+
+ndead = 7000
+nlive = 1000
+
+def gaussian(x, mu, sig):
+    return 1/(np.sqrt(2*np.pi)*sig)*np.exp(-np.power((x - mu)/sig, 2)/2)
+ElogXlivemin = -ndead/nlive - np.log(nlive) - 0.577
+StdlogXlivemin = np.sqrt(ndead/nlive**2 + (np.pi**2)/6)
+live_logX = logX[(logX > ElogXlivemin) * (logX < -ndead/nlive)]
+dead_logX = logX[(logX > -ndead/nlive)]
+
+fig, ax = plt.subplots(figsize=(8, 3))
+ax.plot(logX, L/L.max(), color='black', label='Likelihood', lw=.5)
+ax.fill(logX, LX/LX.max(), alpha=0.4, color='gray', label='Posterior mass')
+lpoints = 0.4
+# Live distribution
+ax.fill_between(live_logX, np.zeros_like(live_logX), lpoints*np.exp(live_logX + ndead/nlive), alpha=0.5, color='deepskyblue', label='Dead/live distribution') 
+# Dead distribution
+ax.fill_between(dead_logX, 0, lpoints, alpha=0.5, color='deepskyblue')
+ax.plot(logX, gaussian(logX, ElogXlivemin, StdlogXlivemin), lw=.5, color='red', label='Min live point')
+ax.axvline(-ndead/nlive, color='deepskyblue', lw=.5, ls='--')
+
+# Arrows showing log n + gamma
+l = 0.32
+pad_text = 0.05
+ax.annotate(r'$\log n+\gamma$', ((-ndead/nlive+ ElogXlivemin)/2 - 3,l+pad_text), ha='center', va='bottom', fontsize=8)
[…]

### 2023-10-25 paper/figures/run_aeons.py

@@ -1,0 +1,28 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.utils import *
+from aeons.regress import *
+from aeons.endpoint import *
+from aeons.plotting import *
+from aeons.beta import *
+figsettings()
+
+import sys
+chain = sys.argv[1]
+name, samples = get_samples(chain)
+model = EndModel(samples)
+true_endpoint = model.true_endpoint()
+true_logXf = model.true_logXf()
+
+iterations = make_iterations(true_endpoint, 50)
+logXs = samples.logX().iloc[iterations]
+logXfs, d_Gs = model.inferences(get_d_G_post, iterations, Nset=50)
+
+write_to_txt(f'{data_dir}/logXfs/post/{name}.txt', [iterations, *logXfs])
+write_to_txt(f'{data_dir}/d_Gs/post/{name}.txt', [iterations, *d_Gs])
+print('Written to data/logXfs and data/d_Gs with 50 iterations')
+
+fig, axs = plt.subplots(1, 2, figsize=(8, 2))
+plot_quantiles(-logXs, d_Gs, samples.d_G(), (0, 1.5), ax=axs[0], color='lightseagreen')
+plot_quantiles(-logXs, -logXfs, -true_logXf, (0, 1.5), ax=axs[1])
+fig.suptitle(f'{name}')

### 2023-11-09 paper/paper.tex
added missing figures and latest changes to paper

@@
 \title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
-\author[Z. Hu et. al]{Zixiao Hu, Artyom Baryshnikov, Will Handley}
+\author[Z. Hu et al.]{Zixiao Hu, Artyom Baryshnikov, Will Handley}
 
@@
 \begin{abstract}
-This paper develops a technique to estimate the runtime of nested sampling which works for any nested sampling implementation.
+This paper presents new analytic results on the anatomy of nested sampling, from which a technique is developed to estimate the run-time of the algorithm that works for any implementation.
 At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be calculated.
@@
 Nested sampling is a multi-purpose algorithm invented by John Skilling which simultaneously functions as a probabilistic sampler, integrator and optimiser \citep{skilling}.
-It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics, materials science \citep{physical_scientists} and machine learning \citep{sparse_reconstruction}.
+It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}.
 The core algorithm is unique in its estimation of volumes by \textit{counting}, which makes high-dimensional integration feasible.
@@
 \par
-The order of magnitude runtime of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user.
+The order of magnitude run-time of an algorithm, that is, whether termination is hours or weeks and months away, is of high importance to the end user.
 Currently, existing implementations of nested sampling (e.g.
-\citealt{multinest, polychord, ultranest}) either do not give an indication of runtime, or only provide crude measures of progress that do not directly correspond to the runtime.
+\citealt{multinest, polychord, ultranest}) either do not give an indication of run-time, or only provide crude measures of progress that do not directly correspond to the run-time.
 \par
 This paper sets out a principled manner of endpoint estimation for nested sampling at each intermediate stage (as shown in \cref{fig:polychord_output}), the key idea being to use the existing samples to predict the likelihood in the region we have yet to sample from.
-We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction.
+We begin with an overview of nested sampling in \cref{sec:background}, followed by an examination of the anatomy of a nested sampling run to establish key concepts for endpoint prediction in \cref{sec:anatomy}.
 \cref{sec:endpoint} then outlines the methodology we use, including discussion and comparisons to previous attempts.
@@
 \section{Background}\label{sec:background}
-Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation; for a more comprehensive treatment, we recommend Skilling's paper and the textbook \citet{sivia}.
+Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation.
+For a more comprehensive treatment, we recommend Skilling's paper and textbook \citet{sivia}, as well as the excellent reviews \citet{Buchner_2023} and \citet{physical_scientists}.
+\par
 For a given likelihood $\mathcal{L}(\theta)$ and prior $\pi(\theta)$, nested sampling simultaneously calculates the Bayesian evidence
 \begin{equation}\label{eq:evidence}
-	\mathcal{Z} = \int \mathcal{L}(\theta)\pi(\theta)\ \mathrm{d}\theta
+	\mathcal{Z} = \int \mathcal{L}\left(\theta\right)\pi(\theta)\ \mathrm{d}\theta
 \end{equation}
@@
 The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}.
-At each iteration, the point with the lowest likelihood is removed and added to a list of \textit{dead points}.
+At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points}.
 A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point.
@@
 \begin{equation}\label{eq:t_dist}
-	P(t_i) = \nlive t_i^{\nlive-1}
+	P(t_i) = \nlive t_i^{\nlive-1},
 \end{equation}
@@
 \end{equation}
-The algorithm terminates when an user-specified condition is met; a popular choice is to terminate when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g.
-$10^{-3}$.
-The remaining live points are then killed off one by one without replacement and added to the evidence.
+The algorithm terminates when an user-specified condition is met; a popular choice is when the evidence in the live points falls below some fraction $\epsilon$ of the accumulated evidence e.g.
+$10^{-3}$, which is proven to be a valid convergence criterion \citep{evans, Chopin_2010}.
+Much of the existing literature treats this remaining evidence separately, for instance by estimating it as the termination $X$ multiplied by the average likelihood amongst the remaining live points.
+It is, however, quantitatively equivalent but qualitatively neater to consider termination as killing the remaining live points off one-by-one, incrementing the evidence exactly as during the run with decreasing $\nlive$ \citep{dynesty}.
 \par
 Uncertainties in the evidence are dominated by the spread in the prior volume distribution, and the simplest way to estimate them is by Monte Carlo sampling over sets of $\bm{t}$.
-For any given problem, the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{\nlive}$, so $\nlive$ sets the resolution of the algorithm.
+As Skilling and others \citep{Chopin_2010, Keeton_2011} have shown, for any given problem the uncertainty in $\log \mathcal{Z}$ is proportional to $1/\sqrt{n_\mathrm{live}}$, so $n_\mathrm{live}$ sets the resolution of the algorithm.
 
-\section{The anatomy of a nested sampling run}
-The following section aims to make an inventory of the information available to us at an intermediate iteration $i^{*}$, which we shall eventually use to make endpoint predictions.
+\section{The anatomy of a nested sampling run}\label{sec:anatomy}
+The following section aims to make an inventory of the information available to us at an intermediate iteration $i$, which we shall eventually use to make endpoint predictions.
 We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature, and the dimensionality of the samples.
@@
 \begin{equation}
-    \mathrm{E}(\log X_i) = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
+    \langle\log X_i\rangle = -\sum_{k=1}^i \frac{1}{n_k}, \quad \mathrm{Var}(\log X_i) = \sum_{k=1}^i \frac{1}{n_k^2}.
 \end{equation}
@@
 \begin{equation}\label{eq:Xmin}
-	\mathrm{E}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{E}(\log X_*) - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i_*}{\nlive} - \log \nlive - \gamma,
+	\langle\log X_\mathrm{min}^{\mathrm{live}}\rangle = \langle\log X_i\rangle - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i}{\nlive} - \log \nlive - \gamma,
 \end{equation}
@@
 \begin{equation}
-	\mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_*) + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i_*}{\nlive^2} + \frac{\pi^2}{6}, 
+	\mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_i) + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6}, 
 \end{equation}
@@
 One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
+\par
+The result is consistent with that in \citet{statmech}, which states that a spike at a volume smaller than $X_i/n_i$ will go undetected.
 Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
 \par
-The above results are summarised in \cref{fig:logX_distribution}
+These results are summarised in \cref{fig:logX_distribution}
 \begin{figure*}
@@
 \end{center}
-\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the constrained prior and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i_*$.
+\caption{The distribution of the posterior mass in terms of $\log X$, the live points over the constrained prior and the smallest live point prior volume $\log X_\mathrm{min}^{\mathrm{live}}$ at an intermediate iteration $i$.
 For large values of $\DKL$ i.e.
@@
 Note that the x-axis in this plot is $\log X$, so that the run proceeds from right to left to emphasise that the enclosed prior volume iteratively gets smaller.
-Plots of the sort from here onwards will be in terms of $-\log X$, where the run will more naturally proceed from left to right.
+Plots of the sort from here onward will be in terms of $-\log X$, where the run will more naturally proceed from left to right.
 }
@@
 \end{equation}
-with maximum point $\logLmax$ and lengthscale $\sigma$ to get an insight into the analytics.
+with maximum point $\logLmax$ and length scale $\sigma$ to get an insight into the analytics.
 The likelihood monotonically increases towards $\logLmax$, with the posterior samples eventually concentrated around the bulk at
@@
 \begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}^*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}^*};
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i};
     \label{eq:normalised_likelihood}
 \end{equation}
-$y=0$ corresponds to the current point $\log\mathcal{L}^*$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
+$y=0$ corresponds to the current point $\log\mathcal{L}_i$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
 \begin{equation}
@@
 \end{equation}
-where we have used the notation of order statistics to denote the $x_{(n)}$ as the maximum of $n$ points.
+where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points.
 This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
@@
 \end{equation}
-Before reaching the posterior bulk, $\logLmax - \log\Like^{*} > d/2$, so we must have
+Before reaching the posterior bulk, $\logLmax - \log\Like_i > d/2$, so we must have
 \begin{equation}
@@
 In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion.
-We now use the machinery we have developed to put this statement in more quantitative terms.
+We now use the machinery we have developed to put this statement on a more quantitative footing.
 \par
@@
 \end{equation}
-Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{end}$ and prior volume $X_\mathrm{end}$, so that
+Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{f}$ and prior volume $X_\mathrm{f}$, so that
 \begin{equation}
-	\epsilon = \frac{\int_0^{X_\mathrm{end}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
+	\epsilon = \frac{\int_0^{X_\mathrm{f}} \mathcal{L}\ \mathrm{d}X}{\int_0^\infty \mathcal{L}\ \mathrm{d}X}.
 \end{equation}
-Note that we have assumed that prior effects are negligible (so $1=\infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence.
+Note that we have assumed that prior effects are negligible (so $1\approx \infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence.
 Computing this for \eqref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
 \begin{equation}
-\epsilon = 1- \frac{\Gamma_{d/2}\left(X_\mathrm{end}^{2/d}/2\sigma^2\right)}{\Gamma(d/2)}.
+\epsilon = 1- \frac{\Gamma_{d/2}\left(X_\mathrm{f}^{2/d}/2\sigma^2\right)}{\Gamma(d/2)}.
 \end{equation}
-Taking the $X_\mathrm{end}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
+Taking the $X_\mathrm{f}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
 \begin{equation}
-    \lim_{X_\mathrm{end}\ll (\sqrt{2}\sigma)^d} \epsilon \approx \frac{X_\mathrm{end}}{(\sqrt{2}\sigma)^d \ \Gamma(1+\frac{d}{2})} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{end})^{\frac{d}{2}}}{\Gamma(1+\frac{d}{2})}.
+    \lim_{X_\mathrm{f}\ll (\sqrt{2}\sigma)^d} \epsilon \approx \frac{X_\mathrm{f}}{(\sqrt{2}\sigma)^d \ \Gamma\left(1+\frac{d}{2}\right)} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{f})^{\frac{d}{2}}}{\Gamma\left(1+\frac{d}{2}\right)}.
 \end{equation}
-We thus have an expression relating $\mathcal{L}_\mathrm{end}$ at termination to the termination fraction $\epsilon$.
+We thus have an expression relating $\mathcal{L}_\mathrm{f}$ at termination to the termination fraction $\epsilon$.
 This becomes yet more pleasing in the large $d$ limit, since $\epsilon^{2/d}\to 1$, we find via a Stirling approximation:
 \begin{equation}
-    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{end} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
+    \lim_{d\to\infty} \log\mathcal{L}_\mathrm{f} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e}.
 \end{equation}
@@
 \par
-Putting this together, taking $\mathcal{L}_*$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{end}$, and combining this with \eqref{eq:ylivemax} we find
+Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$, and combining this with \eqref{eq:ylivemax} we find
 \begin{equation}
@@
 showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood.
-The final set of $n$ live points gets you $\log n/2e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
+The final set of $n$ live points gets you $\log n/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
 \par
@@
 \end{center}
-\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_*)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$.
+\caption{Distribution of samples as a function of $\log\mathcal{L}$, showing the posterior $\mathcal{P}(\log\mathcal{L})$, the distribution of the live points $\mathcal{\pi}(\log\mathcal{L} \mid \mathcal{L}>\mathcal{L}_i)$, and the distribution of the maximum likelihood live point $P(\log\mathcal{L}_\mathrm{max}^\mathrm{live})$.
 The distances are shown between these locations at the end of the run, the key takeaway being that in high dimensions the highest log-likelihood point of a nested sampling run is nowhere near the maximum in high dimensions.}
@@
 \par
-We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially reweighting the samples.
+We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially re-weighting the samples.
 Increasing $\beta$ worsens the situation, while $\beta \to 0$ simply gives back the prior.
-There is, however, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples.
+There is, on the other hand, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples.
 A schematic of the procedure is shown in \cref{fig:last_live_point}.
@@
 Almost all of the posterior mass is concentrated at a single point with the highest compression, because it is orders of magnitude higher in likelihood.
-Reducing $\beta$ reweights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta^*$ which is significant because it centres the posterior at the current contour.}
+Reducing $\beta$ re-weights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta_i$ which is significant because it centres the posterior at the current contour.}
 \label{fig:last_live_point}
@@
 As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, but nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours.
-Each step moves at constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{statmech}.
+Instead of using temperature \citep{simulated_annealing, parallel_tempering} or energy \citep{wang_landau} as a control parameter, nested sampling chooses a series of ensembles with constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{baldock}.
 \par
 Because the temperature of a microcanonical ensemble is a derived property rather than a parameter, there is some freedom in its definition.
-Returning to our original motivation, we make the connection that the temperature is the reweighting $\Like \to \Like^{\beta}$ which centers the ensemble around the current energy.
+Returning to our original motivation, we make the connection that the temperature is the re-weighting $\Like \to \Like^{\beta}$ which centres the ensemble around the current energy.
 We now present several temperatures that achieve this aim, each of which one can plausibly consider to be the current temperature of a nested sampling run.
@@
 \begin{equation}
-    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like^{*}}.
+    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like_i}.
 \end{equation}
-is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X^{*}$, if we assume differentiability, which is exactly the intuition we were aiming for to put the ensemble bulk at the current contour.
-Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing (in practice, we find that above a threshold of around 10 iterations the result is fairly insensitive to this choice).
+is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X_i$, if we assume differentiability, which is exactly the intuition we were aiming for to put the ensemble bulk at the current contour. 
+\par
+Its value can be easily obtained via finite difference of the $\log \Like$ and $\log X$ intervals, albeit subject to an arbitrary window size for the differencing.
+Indeed, material science applications \citep{Baldock_2017} use this estimator to monitor the `cooling' progress of nested sampling, with a window size of 1000 iterations.
 \par
@@
 \subsubsection{Canonical temperature}
-Another temperature considered by Habeck is that at which the current energy (i.e. $-\log \Like^{*}$) is the average energy of the entire ensemble. One can obtain it by inverting 
+Another temperature considered by Habeck is that at which the current energy (i.e. $-\log \Like_i$) is the average energy of the entire ensemble. One can obtain it by inverting 
 \begin{equation}
-    \langle \log \Like \rangle_{\mathcal{P}_\beta} = \log \Like^{*}
+    \langle \log \Like \rangle_{\mathcal{P}_\beta} = \log \Like_i
 \end{equation}
@@
 \begin{equation}
-    P\left(\beta \mid \log X^{*}\right) = \frac{\mathcal{P}\left(\log X^{*} \mid \beta\right) P\left(\beta\right)}{P\left(\log X\right)}.
+    P\left(\beta \mid \log X_i\right) = \frac{\mathcal{P}\left(\log X_i \mid \beta\right) P\left(\beta\right)}{P\left(\log X\right)}.
 \end{equation}
@@
 The obtained temperatures are consistent with the previous two choices, which may seem oddly coincidental.
-However, closer inspection reveals that large values of $P(\beta \mid \log X^{*})$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence.
+In fact, closer inspection reveals that large values of $P(\beta \mid \log X_i)$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence.
 Thus the Bayesian temperature uses the same idea as the microcanonical one, except it accounts for the spread in the result.
@@
 \end{center}
-\caption{Inferred temperatures using the microcanonical, canonical and Bayesian definitions. All are consistent for a single phase, but differ during a phase transition.}
+\caption{Inferred temperatures using the microcanonical, canonical and Bayesian definitions.
+The shaded regions show the $1-2\sigma$ uncertainties.
+All are consistent for a single phase, but differ during a phase transition.}
 \label{fig:beta_comparison}
@@
 \end{center}
-\caption{Dimensionality estimates for a Gaussian that is elongated in half of its dimensions.
-The locations of the live points in the prior are shown at three stages, indicated by the vertical lines.
+\caption{Dimensionality estimates for a Gaussian that is elongated in half of its dimensions, with $1-2\sigma$ uncertainties shaded.
+The locations of the live points in the prior are shown at three stages, indicated by the connecting lines.
 As can be seen from the live point distribution, the prior does not compress in the higher variance direction until much later in the run, and early on it appears as if those directions are completely unconstrained.}
@@
 \par
-It is important to appreciate that at lower compressions the samples truly lie in a lower-dimensional space, rather than some artefact of the way we view them.
+It is important to appreciate that at lower compression the samples truly lie in a lower-dimensional space, rather than some artefact of the way we view them.
 Anticipating the full dimensionality of the space is therefore just as impossible as that associated with a slab-spike geometry, so in this sense such geometries contain a `compressive phase transition'.
@@
 \section{Endpoint prediction}\label{sec:endpoint}
-The time complexity of nested sampling \citep{supernest} is
+As described in \citet{supernest} and further explored in the talk and upcoming paper \citet{kcl_talk, scaling_frontier}, the time complexity of nested sampling is
 \begin{equation}
-    T \propto \nlive \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
+    T \propto \langle \nlive \rangle \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
 \end{equation}
@@
 The primary unknown during a run, and therefore our primary interest, is the final term: the compression required to get from prior to posterior.
-\par
-Further exploration can be found in the talk [\citenum{kcl_talk}] and the upcoming paper \citet{scaling_frontier}.
 
@@
 \begin{enumerate}[leftmargin=*]
-    \item Find the current dimensionality ${d}_G^{*}$ of the posterior at the Bayesian temperature
-    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = {d}_G^{*}$ to infer $\logLmax$ and $\sigma$ 
+    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
+    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
     \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
@@
 \end{equation}
-and $N_\mathrm{f}$ is of course just $-\nlive$ multiplied by this.
+and $N_\mathrm{f}$ is of course just $-n$ multiplied by this.
 Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained.
@@
 \subsection{Alternative approaches}
-We consider here several alternative approaches to endpoint estimation to compare against our method. 
+More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could certainly do better than what we have just presented.
+However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more practical approach we have adopted.
+Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity.
 \subsubsection{Integral progress}
@@
 \begin{equation}\label{eq:integral_progress}
-    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X^{*}.
+    \mathcal{Z}_\mathrm{rem} \approx \Like_\mathrm{max}^{\mathrm{live}} X_i.
 \end{equation}
 There are a few problems here.
-Firstly, runtime is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width.
+Firstly, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width.
 Secondly, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is simply be zero for most of the run.
@@
 The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile.
-The increments are also not quite exponential, since the exponential fit leads to a large overprediction.
+The increments are also not quite exponential, since the exponential fit leads to a large over-prediction.
 The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result.
@@
 \includegraphics{figures/inc_predictions.pdf}
-\caption{Endpoint predictions for a spherical Gaussian.
-Extrapolating the evidence increments in a linear/exponential manner under/overpredicts the endpoint, both of which do considerably worse than the method of extrapolating the likelihood.}
+\caption{Endpoint predictions for a spherical Gaussian, with $1-2\sigma$ uncertainties shaded.
+Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which do considerably worse than the method of extrapolating the likelihood.}
 \label{fig:inc_predictions}
@@
 Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact.
+In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties.
 All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$.
@@
 \par
-However, the right order of magnitude is obtained at all times, so this is not a major concern.
+Nonetheless, the right order of magnitude is obtained at all times, so this is not a major concern.
 The Cauchy is also a pathological case, and the same problem does not rear its head in more realistic cases, as we shall see next.
@@
 Finally, we evaluate the method on real cosmological chains.
-\cref{fig:lcdm_logXfs} presents the endpoints (calculated after the fact) for nested sampling runs for curvature quantification on several common cosmological datasets (details in \citet{curvature_tension}).
+\cref{fig:lcdm_logXfs} presents the endpoints (calculated after the fact) for nested sampling runs for curvature quantification on several common cosmological data sets (details in \citet{curvature_tension}).
 \par
@@
 \section{Conclusion}
-We have presented a method for predicting the endpoint of a nested sampling run, by inferring the Bayesian model dimensionality of a set of samples mid-run and using it to extrapolate the known likelihood profile.
+We have derived new analytic results to make an anatomy of nested sampling, understanding the progression of a run via the compression of prior volume, the increase in log-likelihood, the inferred temperature schedule, and the convergence of the sample dimensionality.
+From these analyses, we developed a method for predicting the endpoint of a nested sampling run, by using the inferred Bayesian model dimensionality mid-run to extrapolate the known likelihood profile.
 \par
@@
 Further work can be done to experiment with more flexible basis functions for regression of the likelihood profile, so that it is less dependent on the Gaussian approximation.
+
+\section{Code availability}
+A package implementing the endpoint prediction mechanism is in development for easy plug-in to existing implementations of nested sampling.
+The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.
 
@@
 \begin{multline}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*)^{\frac{d}{2}}} \\
-    [\log\mathcal{L}_* < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i)^{\frac{d}{2}}} \\
+    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
     \label{eq:PL}
@@
 \begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_*}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_*}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
     \label{eq:y}
 \end{equation}
-as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_*$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
+as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_i$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
 \begin{equation}
@@
 \par
-This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_*,\log\mathcal{L}_\mathrm{max}]$.
+This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_i,\log\mathcal{L}_\mathrm{max}]$.
 In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.

### 2023-11-09 paper/figures/d_G_elongated.py

@@ -2,8 +2,8 @@ import numpy as np
 import matplotlib.pyplot as plt
 from aeons.utils import *
 from aeons.plotting import *
-from matplotlib.gridspec import GridSpec
 figsettings()
+from matplotlib.gridspec import GridSpec
 
 name, samples = get_samples('correlated_6d', reduced=False)
 iterations, *d_Gs = read_from_txt(f'{data_dir}/d_Gs/post/{name}.txt')
@@ -11,14 +11,14 @@ logXs = samples.logX().iloc[iterations]
 d_Gs = np.array(d_Gs)
 
 fig = plt.figure(figsize=(3.5, 3))
-gs = GridSpec(2, 3, wspace=0.1, hspace=0.35)
-ax1 = fig.add_subplot(gs[0, 0])
-ax2 = fig.add_subplot(gs[0, 1])
-ax3 = fig.add_subplot(gs[0, 2])
-ax4 = fig.add_subplot(gs[1, :])
+gs = GridSpec(2, 3, wspace=0.1, hspace=0.1)
+ax1 = fig.add_subplot(gs[0, 0], zorder=1)
+ax2 = fig.add_subplot(gs[0, 1], zorder=1)
+ax3 = fig.add_subplot(gs[0, 2], zorder=1)
+ax4 = fig.add_subplot(gs[1, :], zorder=0)
 axs = [ax1, ax2, ax3, ax4]
 
-logXstars = [-1, -8, -30]
+logXstars = [-0.5, -8, -30]
 logX = samples.logX()
 # Find index of logX closest to logXstar
 ndeads = [np.argmin(np.abs(logX - logXstar)) for logXstar in logXstars]
@@ -26,7 +26,7 @@ ndeads = [np.argmin(np.abs(logX - logXstar)) for logXstar in logXstars]
 for i, ndead in enumerate(ndeads):
     ax = axs[i]
     live_points = points_at_iteration(samples, ndead).iloc[ndead:]
-    live_points.drop_weights().plot.scatter(0, 4, ax=ax, s=5)
+    live_points.drop_weights().plot.scatter(0, 4, ax=ax, s=5, color='lightseagreen')
     ax.set_aspect('equal')
     ax.set_xlim(0, 1)
     ax.set_ylim(0, 1)
@@ -34,13 +34,23 @@ for i, ndead in enumerate(ndeads):
     ax.set_yticks([])
     ax.set_xlabel('')
     ax.set_ylabel('')
[…]

### 2023-11-09 paper/figures/inc_predictions.py

@@ -32,7 +32,7 @@ ax.plot(-logXs, endpoints_linear/500, lw=.8, color='orange', label='Linear')
 plot_quantiles(-logXs, -logXfs, -true_logXf, ax=ax, label='Fit $\\mathcal{L}(X)$')
 
 ax.set_ylim(0, 160)
-ax.set_ylabel(r'$-\log \hat{X}_f$')
+ax.set_ylabel(r'$-\log \hat{X}_\mathrm{f}$')
 ax.set_xlabel(r'$-\log X$')
 ax.legend(loc='upper right', fontsize=6)
 ax.margins(x=0)

### 2023-11-09 paper/figures/last_live_point.py

@@ -9,7 +9,7 @@ ndead = 8000
 points = points_at_iteration(samples, ndead)
 beta_grad = np.exp(get_logbeta_grad(points, ndead))
 betas = [0, beta_grad, 1]
-labels = ['$\\beta=1$', '$\\beta = \\beta^*$', '$\\beta = 0$']
+labels = ['$\\beta=1$', '$\\beta = \\beta_i$', '$\\beta = 0$']
 colors = ['darkgray', 'orange', 'black']
 
 fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
@@ -33,7 +33,7 @@ for ax in axs:
     ax.margins(x=0.02)
     ax.set_ylim(-0.01, 1.1)
     ax.set_yticks([])
-    ax.set_xticks([0, -points.logX().iloc[ndead]], [0, r'$-\log X^*$'])
+    ax.set_xticks([0, -points.logX().iloc[ndead]], [0, r'$-\log X_i$'])
 axs[0].set_title(r'$\mathcal{L}^\beta$', fontsize=8)
 axs[1].set_title('$\\mathcal{P}(\\log X) = \\mathcal{L}^\\beta X$', fontsize=8)
 axs[0].legend(loc='upper left', fontsize=5)

### 2023-11-09 paper/figures/logL_distribution.py

@@ -68,7 +68,7 @@ ax.set_xticks([avlogL,
                logLmaxlive,
                logLmax])
 ax.set_xticklabels([r'$\langle\log\mathcal{L}\rangle_\mathcal{P}$',
-                    r'$\log\mathcal{L}_\mathrm{end}$',
+                    r'$\log\mathcal{L}_i$',
                     r'$\log\mathcal{L}_\mathrm{max}^\mathrm{live}$',
                     r'$\log\mathcal{L}_\mathrm{max}$'],
                     fontsize=8)

### 2023-11-09 paper/figures/logX_distribution.py

@@ -49,7 +49,7 @@ ax.add_patch(arrow)
 
 # Arrows showing \Delta\log X
 pad_text = 0.05
-ax.annotate(r'$\Delta\log X = -\frac{1}{n_\mathrm{live}}$', (-ndead/nlive - 1, lpoints+pad_text), ha='center', va='bottom', fontsize=8)
+ax.annotate(r'$\Delta\log X = -\frac{1}{n_{i}}$', (-ndead/nlive - 1, lpoints+pad_text), ha='center', va='bottom', fontsize=8)
 arrow_start = (-ndead/nlive, lpoints)
 arrow_end = (-ndead/nlive - 1.5, lpoints)
 arrow = patches.FancyArrowPatch(arrow_start, arrow_end, arrowstyle='->', mutation_scale=5)
@@ -63,7 +63,7 @@ arrow = patches.FancyArrowPatch(arrow_start, arrow_end, arrowstyle='->', mutatio
 ax.add_patch(arrow)
 ax.annotate("Iterations", (-25, l1+pad_text), ha='center', va='bottom', fontsize=8)
 
-ax.set_xticks([logXf, -DKL, ElogXlivemin, -ndead/nlive, 0], [r"$\log X_\mathrm{f}$", r'$-\mathcal{D}_\mathrm{KL}$', r'$\langle\log X_\mathrm{min}^\mathrm{live}\rangle$', r'$\langle\log X^*\rangle$', '$0$'], fontsize=8)
+ax.set_xticks([logXf, -DKL, ElogXlivemin, -ndead/nlive, 0], [r"$\log X_\mathrm{f}$", r'$-\mathcal{D}_\mathrm{KL}$', r'$\langle\log X_\mathrm{min}^\mathrm{live}\rangle$', r'$\langle\log X_i\rangle$', '$0$'], fontsize=8)
 ax.set_yticks([])
 ax.set_ylim(0, 1.05)
 ax.margins(x=0)

### 2023-11-09 paper/paper.tex
name change
Minor updates

@@
 
-\title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
-\author[Z. Hu et al.]{Zixiao Hu, Artyom Baryshnikov, Will Handley}
+\title[\texttt{aeons}]{\texttt{aeons}: approximating the end of nested sampling}
+\author[Z. Hu et al.]{Zixiao Hu\thanks{E-mail: zixiao.hu.09@gmail.com},$^{1,2}$  Artem Baryshnikov,$^{1,2}$  Will Handley\thanks{E-mail: wh260@cam.ac.uk},$^{1,2}$
+\\
+% List of institutions
+$^{1}$Astrophysics Group, Cavendish Laboratory, J. J. Thomson Avenue, Cambridge, CB3 0HE, UK\\
+$^{2}$Kavli Institute for Cosmology, Madingley Road, Cambridge, CB3 0HA, UK
+}
 
@@
 \begin{abstract}
-This paper presents new analytic results on the anatomy of nested sampling, from which a technique is developed to estimate the run-time of the algorithm that works for any implementation.
-At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be calculated.
+This paper presents analytic results on the anatomy of nested sampling, from which a technique is developed to estimate the run-time of the algorithm that works for any nested sampling implementation.
+At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be estimated.
 The method finds the right order of magnitude at all times, obtaining the true endpoint within standard error generally around the halfway point, and converges to the true value by the end of the run.
@@
 Currently, existing implementations of nested sampling (e.g.
-\citealt{multinest, polychord, ultranest}) either do not give an indication of run-time, or only provide crude measures of progress that do not directly correspond to the run-time.
+\citealt{multinest, polychord, dynesty, ultranest}) either do not give an indication of run-time, or only provide crude measures of progress that do not directly correspond to the run-time.
 \par
@@
 \begin{figure}
-\begin{Verbatim}[frame=single, commandchars=\\\{\}]
-\textcolor{red}{Predicted endpoint: 25054 +/- 242}
-\textcolor{red}{Progress: [=================>########] 72%}
-___________________
-lives      |   500 |
-phantoms   | 24310 |
-posteriors | 18018 |
-equals     |   245 |
--------------------
-ncluster   =  1/1
-ndead      =  18018
-nposterior =  18018
-nequals    =  249
-nlike      =  4159049
-<nlike>    =  491.04 (9.82 per slice)
-log(Z)     =  -12.55 +/- 0.27
-\end{Verbatim}
+%\begin{Verbatim}[frame=single, commandchars=\\\{\}]
+%\textcolor{red}{Predicted endpoint: 25054 +/- 242}
+%\textcolor{red}{Progress: [=================>########] 72%}
+%___________________
+%lives      |   500 |
+%phantoms   | 24310 |
+%posteriors | 18018 |
+%equals     |   245 |
+%-------------------
+%ncluster   =  1/1
+%ndead      =  18018
+%nposterior =  18018
+%nequals    =  249
+%nlike      =  4159049
+%<nlike>    =  491.04 (9.82 per slice)
+%log(Z)     =  -12.55 +/- 0.27
+%\end{Verbatim}
 \caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}

### 2023-11-09 paper/paper.tex
Commented verbatim

--- 2023-11-09 paper/paper.tex
@@
 \title[Approximating the end of nested sampling]{Approximating the end of nested sampling}
-\author[Z. Hu et al.]{Zixiao Hu, Artyom Baryshnikov, Will Handley}
+\author[Z. Hu et al.]{Zixiao Hu, Artem Baryshnikov, Will Handley}
 

### 2023-11-09 paper/paper.tex
Added some more references

--- 2023-11-09 paper/paper.tex
@@
 \title[\texttt{aeons}]{\texttt{aeons}: approximating the end of nested sampling}
-\author[Z. Hu et al.]{Zixiao Hu\thanks{E-mail: zixiao.hu.09@gmail.com},$^{1,2}$  Artem Baryshnikov,$^{1,2}$  Will Handley\thanks{E-mail: wh260@cam.ac.uk},$^{1,2}$
+\author[Z. Hu et al.]{Zixiao Hu\thanks{E-mail: zixiao.hu.09@gmail.com},$^{1,2}$  Artem Baryshnikov,$^{1,2}$  Will Handley\thanks{E-mail: wh260@cam.ac.uk}$^{1,2}$
 \\
@@
 At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be estimated.
+We test these methods on both toy models and true cosmological nested sampling runs.
 The method finds the right order of magnitude at all times, obtaining the true endpoint within standard error generally around the halfway point, and converges to the true value by the end of the run.
@@
 It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}.
-The core algorithm is unique in its estimation of volumes by \textit{counting}, which makes high-dimensional integration feasible.
-It also avoids problems faced by traditional Bayesian algorithms, such as multi-modality.
+The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested monte carlo integrals which makes high-dimensional integration feasible and robust.
+It also avoids problems that challenge traditional Bayesian samplers, such as posterior multi-modality and phase transitions.
 \par
@@
 Currently, existing implementations of nested sampling (e.g.
-\citealt{multinest, polychord, dynesty, ultranest}) either do not give an indication of run-time, or only provide crude measures of progress that do not directly correspond to the run-time.
+\citealt{multinest, polychord, dnest, dynesty, ultranest, nessai,proxnest}) either do not give an indication of remaining run-time, or only provide crude measures of progress that do not directly correspond to the the true endpoint.
 \par
@@
 \section{Background}\label{sec:background}
-Let us begin with a brief description of the nested sampling algorithm to establish the necessary notation.
-For a more comprehensive treatment, we recommend Skilling's paper and textbook \citet{sivia}, as well as the excellent reviews \citet{Buchner_2023} and \citet{physical_scientists}.
+We begin with a brief description of the nested sampling algorithm to establish the necessary notation.
+For a more comprehensive treatment, we recommend the original \citep{skilling} paper, the \citet{sivia} textbook, as well as the excellent technical review by \citet{Buchner_2023} and Nature review by \citet{physical_scientists}.
 \par

### 2023-11-10 paper/paper.tex
Further updates to text

@@
 The algorithm operates by maintaining a set of $\nlive$ \textit{live points} sampled from the prior, which can vary in number throughout the run \citep{dynamic_ns}.
-At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points}.
-A new point is then drawn from the prior, subject to the constraint that it must have a higher likelihood than the latest dead point.
-Repeating the procedure leads to the live points shrinking around peaks in the likelihood.
+At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points} (illustrated in \cref{fig:dead_measure}).
+New points are then (optionally) drawn from the prior, subject to the constraint that they must have a higher likelihood than the latest dead point~$\mathcal{L}_i$.
+Repeating the procedure leads to the live points compressing around peaks in the likelihood.
+If there are more births than deaths, the number of live points $n_i$ increases, whilst choosing to not generate new points reduces the number of live points.
+The exact creation schedule depends on the dynamic nested sampling strategy.
 \par
@@
 \begin{equation}
-	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \frac{1}{2}\sum_{i=1} \mathcal{L}(X_{i-1}-X_{i+1}),
+	\mathcal{Z} = \int_0^1 \mathcal{L}(X)\ \mathrm{d}X \approx \sum_{i=1} \mathcal{L}_i\ \tfrac{1}{2}(X_{i-1}-X_{i+1}),
 \end{equation}
 where $X(\mathcal{L})$ is the fraction of the prior with a likelihood greater than $\mathcal{L}$.
-The prior volumes $X_i$ are unknown, but can be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
+Whilst the likelihood contour $\mathcal{L}_i$ at each iteration is known, the prior volumes $X_i$ must be statistically estimated as follows: one can define a \textit{shrinkage factor} $t_i$ at each iteration $X_{i} = t_i X_{i-1}$, such that
 \begin{equation}\label{eq:X_dist}
@@
 \section{The anatomy of a nested sampling run}\label{sec:anatomy}
-The following section aims to make an inventory of the information available to us at an intermediate iteration $i$, which we shall eventually use to make endpoint predictions.
-We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression, the log-likelihood increase, the inferred temperature, and the dimensionality of the samples.
+The following sections act as an inventory of the information available to us at an intermediate iteration $i$, which we shall use to make endpoint predictions in \cref{sec:endpoint}.
+We present an anatomy of the progression of a nested sampling run in terms of the prior volume compression (\cref{sec:prior_volume}), the log-likelihood increase (\cref{sec:logL}), the inferred temperature (\cref{sec:temperature}), and the dimensionality of the samples (\cref{sec:dimensionality}).
 
-\subsection{Prior volume}
+\subsection{Prior volume}\label{sec:prior_volume}
 The key feature of nested sampling is that the sampling is controlled by prior volume compression.
@@
 \end{equation}
-The bulk of the posterior lies within a prior volume $X = e^{-\DKL}$, which is the target compression.
-One gets there by iteratively taking steps of size $\Delta \log X_i = -1/n_i$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
+The bulk of the posterior lies within a prior volume ${X = e^{-\DKL}}$, which is the target compression.
+From \cref{eq:t_moments} one gets there by iteratively taking steps of size ${\Delta \log X_i = -1/n_i}$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
 \begin{equation}
@@
 \end{equation}
-A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points (as shown in \cref{fig:dead_measure}), which is exactly needed to overcome the curse of dimensionality.
+A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points , which is exactly needed to overcome the curse of dimensionality.
 \begin{figure*}
@@
 \caption{The dead points of a nested sampling run, recursively zoomed in.
-Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} V/V$, and hence scale invariant.
-The live points (larger dots) are uniform across the prior, plotted for comparison.}
+Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} X/X$, and hence scale invariant until the final set of live points is reached.
+The live points at a given iteration $i$ (larger dots) are uniform across the prior, plotted for comparison.}
 \label{fig:dead_measure}
@@
 \par
-The same is not true for the live points, which are uniformly distributed in prior volume. As a result, the maximum live point is found at 
+The same is not true for the live points, which are uniformly distributed in prior volume, and by the comments in the penultimate paragraph of \cref{sec:background} have ${n_{i+k} = n_i-k}$.
+As a result, the maximum live point is found at
 \begin{equation}\label{eq:Xmin}
-	\langle\log X_\mathrm{min}^{\mathrm{live}}\rangle = \langle\log X_i\rangle - \sum_{k=1}^{\nlive} \frac{1}{n_k} \approx -\frac{i}{\nlive} - \log \nlive - \gamma,
+    \langle\log X_\mathrm{min}^{\mathrm{live}}\rangle = \langle\log X_i\rangle - \sum_{k=1}^{\nlive} \frac{1}{n_{i+k}} \approx -\frac{i}{\nlive} - \log \nlive - \gamma,
 \end{equation}
@@
 \begin{equation}
-	\mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_i) + \sum_{k=1}^{\nlive} \frac{1}{n_k^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6}, 
+    \mathrm{Var}(\log X_\mathrm{min}^{\mathrm{live}}) = \mathrm{Var}(\log X_i) + \sum_{k=1}^{\nlive} \frac{1}{n_{i+k}^2} \approx \frac{i}{\nlive^2} + \frac{\pi^2}{6}, 
 \end{equation}
@@
 \par
-The live points therefore only get us a factor of $\log \nlive$ closer to the posterior bulk.
-In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples look anything like the posterior.
+The live points therefore only get us a factor of $\log \nlive$ closer in volume to the posterior bulk.
+In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples begin populating the posterior typical set.
 One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
@@
 Intuitively, it is because for a sharply peaked likelihood the live points are too diffuse to land there with any significant probability for most of the run.
-\par
 These results are summarised in \cref{fig:logX_distribution}
@@
 \subsection{Log-likelihood}\label{sec:logL}
-It is also useful to observe the distribution of the live and dead points in log-likelihood. We choose to examine a representative case of the $d$-dimensional multivariate Gaussian:
+We now consider the distribution of the live and dead points in log-likelihood. We will examine the representative case of the $d$-dimensional multivariate Gaussian:
 \begin{equation}\label{eq:gaussian_logL}
@@
 with maximum point $\logLmax$ and length scale $\sigma$ to get an insight into the analytics.
-The likelihood monotonically increases towards $\logLmax$, with the posterior samples eventually concentrated around the bulk at
+The likelihood monotonically increases towards $\logLmax$, with \cref{sec:logLmax} showing the posterior samples eventually concentrated around the bulk at
 \begin{equation}
@@
 
-\subsection{Temperature}
+\subsection{Temperature}\label{sec:temperature}
 \subsubsection*{Motivations}
@@
 
-\subsubsection{Microcanonical temperature}
+\subsubsection{Microcanonical temperature}\label{sec:microcanonical_temperature}
 The obvious candidate is the microcanonical temperature $\partial S/\partial E$, where the volume entropy is $\log X$ and the energy is as usual $-\log \Like$.
@@
 Indeed, material science applications \citep{Baldock_2017} use this estimator to monitor the `cooling' progress of nested sampling, with a window size of 1000 iterations.
-\par
-The ensemble bulk can be placed at the current point in several other plausible ways, which are discussed below.
 
-\subsubsection{Canonical temperature}
-Another temperature considered by Habeck is that at which the current energy (i.e. $-\log \Like_i$) is the average energy of the entire ensemble. One can obtain it by inverting 
+\subsubsection{Canonical temperature}\label{sec:canonical_temperature}
+Another temperature considered by \citet{demons} is that at which the current energy (i.e. $-\log \Like_i$) is the average energy of the entire ensemble. One can obtain it by inverting 
 \begin{equation}
@@
 
-\subsubsection{Bayesian temperature}
+\subsubsection{Bayesian temperature}\label{sec:bayesian_temperature}
 We furthermore propose a temperature $\beta_\mathrm{B}$ that is obtained via Bayesian inference, which returns a distribution rather than a point estimate.
@@
 
-\subsection{Dimensionality}
+\subsection{Dimensionality}\label{sec:dimensionality}
 We can immediately use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible.
@@
 \begin{equation}
-    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),
+    \log X_\mathrm{f} = \frac{d}{2}\log 2\sigma^2	+ \log \Gamma^{-1}_{d/2} \left(\epsilon \gamma_i+ \frac{\epsilon\mathcal{Z}_\mathrm{dead}}{ \left( 2\sigma^2 \right)^{d/2}\Like_\mathrm{max}}\right),\label{eq:xf}
 \end{equation}
@@
 \end{equation}
-The distribution of the true posterior in $\log\mathcal{L}$ is
+Providing that $\sigma^d \ll 1$, the posterior can be approximated as
+\begin{equation}
+    P(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}.
+    \label{eqn:PX}
+\end{equation}
+The distribution of the true posterior in $\log\mathcal{L}$ is therefore
 \begin{equation}
@@
 
+WHCOMMENT: How sure are we of this statement?
+I agree nested sampling is slower than gradient descent at maximising, but it still converges exponentially.
+What is the approximation of \cref{eq:xf} in this instance?
+
+WHCOMMENT: Is it worth hiding all these big results in the appendix -- it's not that long, and I think readers must read it to pace themselves through the maximum arguments?
+
+WHCOMMENT: $P(\log\mathcal{L})$ for the dead points?
+
+WHCOMMENT: maxmimum $(n)$ iteration notation?
+
+WHCOMMENT: \cref{fig:last_live_point} move legend.
+
 \bibliographystyle{mnras}

### 2023-11-27 paper/paper.tex
Further updates

@@
     This paper presents analytic results on the anatomy of nested sampling, from which a technique is developed to estimate the run-time of the algorithm that works for any nested sampling implementation.
-At each iteration, the dimensionality of the current set of samples is inferred, then used to extrapolate the known likelihood profile using a Gaussian model function, which allows the remaining evidence and prior volume at termination to be estimated.
 We test these methods on both toy models and true cosmological nested sampling runs.
-The method finds the right order of magnitude at all times, obtaining the true endpoint within standard error generally around the halfway point, and converges to the true value by the end of the run.
+The method gives an order-of-magnitude prediction of the end point at all times, forecasting the true endpoint within standard error generally around the halfway point.
 \end{abstract}
@@
 One should keep in mind that despite the above theoretical reasoning, our introduction of the likelihood transformation was ultimately motivated by our wish to utilise the extra degree of freedom it provides.
-As we will see below, we recommend choosing the exact definition depending on what is useful for the transformation to do.
+As we will see below, we recommend choosing the exact definition depending on what is useful for the problem at hand.
 
@@
 \begin{equation}
-    T \propto \langle \nlive \rangle \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
+    T \propto \langle \tfrac{1}{\nlive} \rangle^{-1} \times \langle \mathcal{T}\{ \Like(\theta) \} \rangle \times \langle f_\mathrm{sampler} \rangle \times \DKL.
 \end{equation}
-The second term is the relatively constant time per likelihood evaluation.
+The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n)$, The second term is the average time per likelihood evaluation.
 The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude.
-The primary unknown during a run, and therefore our primary interest, is the final term: the compression required to get from prior to posterior.
+The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior.
+This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principle interest in this section.
 
 \subsection{The termination prior volume}
-To be exact, one wishes to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}).
-The problem is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is quite far from the posterior bulk.
+Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}).
+The difficulty is that at an intermediate iteration we only know the posterior up to the maximum log-likelihood live point, which until just before the end is far from the posterior bulk.
 \par
@@
 We do this by \textit{extrapolating} the known likelihood profile; that is, the trajectory of $\Like(X)$ traced out by the live and dead points.
+One would never use this predicted posterior to perform inference, since more accuracy can always be achieved by simply finishing the run.
+However, we will demonstrate it is sufficient for making a run-time prediction for $\log X_\mathrm{f}$.
 \par
-One would never use this predicted posterior to do inference, since more accuracy can always be achieved by simply finishing the run.
-However, it is more than sufficient for making a prediction for $\log X_\mathrm{f}$.
 Quantitatively, this proceeds as follows: fit a function $f(X, \phi)$  with some parameters $\phi$ to the known likelihood profile, which allows us to express the prior volume we need to compress to as
@@
 \end{equation}
+or equivalently
 \begin{equation}\label{endpoint}
@@
 Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule.
-The conversion is easiest in the constant $\nlive$ case; at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ will be
+For example in the constant $\nlive$ case at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ would be
 \begin{equation}
@@
     \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
-    \item Take the live point profile and do a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
+    \item Take the live point profile and perform a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
     \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
@@
 \par
-One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown it to be far less stable.
+One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown this approach to be far less stable.
 
 \subsection{Alternative approaches}
-More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could certainly do better than what we have just presented.
-However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more practical approach we have adopted.
+More comprehensive Bayesian approaches, perhaps including \textit{a priori} information about the likelihood or greater flexibility in the fitting function, could likely perform better than what we have just presented.
+However, such methods would not befit \textit{run-time} prediction which has a much more limited computational budget, hence the more pragmatic approach we have adopted.
 Here, we discuss as a benchmark alternative approaches to endpoint estimation that have a comparable computational complexity.
 \subsubsection{Integral progress}
-One alternative approach, used in \textsc{Ultranest} \citep{ultranest}, derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
+An alternative approach used in \textsc{Ultranest} \citep{ultranest} derives a progress bar based on the fraction of the accumulated integral compared to the remaining integral, approximated as 
 \begin{equation}\label{eq:integral_progress}
@@
 \end{equation}
-There are a few problems here.
-Firstly, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width.
-Secondly, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is simply be zero for most of the run.
-Furthermore, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
-\par
-This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.
+This has a several shortcomings.
+First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width.
+Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run.
+Finally, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
+This approach can however be useful in the low dimensions appropriate for \textsc{Ultranest} when the live points are always near the maximum, but in general is less reliable.
 
 \subsubsection{Extrapolating evidence increments}
-Seasoned users of nested sampling might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges.
+Seasoned watchers of nested sampling runs might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges.
 We do this for a spherical Gaussian and compare it to our method.
@@
 The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile.
-The increments are also not quite exponential, since the exponential fit leads to a large over-prediction.
+The increments are also not exactly exponential, since the exponential fit leads to a large over-prediction.
 The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result.
@@
 \caption{Endpoint predictions for a spherical Gaussian, with $1-2\sigma$ uncertainties shaded.
-Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which do considerably worse than the method of extrapolating the likelihood.}
+Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which perform considerably worse than the method of extrapolating the likelihood.}
 \label{fig:inc_predictions}
@@
 \section{Results}\label{sec:results}
-Now that we have a method for predicting the endpoint, it can be tested on a range of distributions.
+We now test the approach established in the previous section on a range of distributions.
 We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
 \subsection{Toy examples}
+Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\cite{anesthetic}.
 \subsubsection{Gaussians}
@@
 The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable.
-\begin{figure*}
+\begin{figure}
 \begin{center}
@@
 \label{fig:elongated_logXfs}
-\end{figure*}
+\end{figure}
 
@@
 choosing $d = 10$ and allowing $\gamma$ to vary.
-The correct estimate is obtained to within standard error by about halfway, but before that is quite off.
-The key limitation is that the estimate is wrong early on not because the compression is anisotropic, or that there is a phase transition; rather, it is a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
+The correct estimate is obtained to within standard error by about halfway, but before that is inaccurate.
+The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or that there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
 \par
-Nonetheless, the right order of magnitude is obtained at all times, so this is not a major concern.
-The Cauchy is also a pathological case, and the same problem does not rear its head in more realistic cases, as we shall see next.
+Nevertheless, the right order of magnitude is obtained at all times, so this remains sufficient for most use-cases.
+The Cauchy is also a pathological case, and the same problem does not in practice appear for more realistic cases, as we shall see next.
 \begin{figure*}
@@
 \par
-The method in general converges to the correct endpoint by about the halfway point, and gets the correct order of magnitude throughout.
+The method in general converges on a correct prediction of endpoint by about halfway, and gets the correct order of magnitude throughout.
 Consistent predictions are obtained for both toy and cosmological examples.
@@
 
-\section{Code availability}
+\section*{Code availability}
 A package implementing the endpoint prediction mechanism is in development for easy plug-in to existing implementations of nested sampling.
 The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.
+
+\section*{Acknowledgements}
+WJH is supported by the Royal Society as a Royal Society University Research Fellow at the University of Cambridge.
+AB was supported by a Cambridge Mathematical Placement and the Royal Society summer studentship.
+ZH was supported by a Royal Society summer studentship.
+We thank Mike Hobson for helpful discussion during ZH Part III viva.
 
@@
 \par
-This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_i,\log\mathcal{L}_\mathrm{max}]$.
+This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they steadily squeeze the interval $[\log\mathcal{L}_i,\log\mathcal{L}_\mathrm{max}]$.
 In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.

### 2023-11-27 paper/paper.tex
incorporating appendix into main body

--- 2023-11-10 paper/paper.tex
@@
 \subsection{Log-likelihood}\label{sec:logL}
-We now consider the distribution of the live and dead points in log-likelihood. We will examine the representative case of the $d$-dimensional multivariate Gaussian:
+We now consider the distribution of the samples in log-likelihood.
+To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
 \begin{equation}\label{eq:gaussian_logL}
-    \log \Like = \logLmax - X^{2/d}/2\sigma^2
+    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
 \end{equation}
-with maximum point $\logLmax$ and length scale $\sigma$ to get an insight into the analytics.
-The likelihood monotonically increases towards $\logLmax$, with \cref{sec:logLmax} showing the posterior samples eventually concentrated around the bulk at
+Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
 \begin{equation}
-    \langle\log\Like\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
+    \mathcal{P}(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}, 
+    \label{eqn:PX}
 \end{equation}
-We can again find the size of each step in log-likelihood, as well as the expected location of the maximum live point. Let us define a likelihood normalised by the distance to the maximum:
+which in terms of $\log\mathcal{L}$ becomes
 \begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i};
-    \label{eq:normalised_likelihood}
+    \mathcal{P}(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
 \end{equation}
-$y=0$ corresponds to the current point $\log\mathcal{L}_i$ and $y=1$ to the maximum $\log\mathcal{L}_\mathrm{max}$. At each iteration, $y$ increases by roughly
+i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$, with mean and variance
 \begin{equation}
-    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}  
+    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
 \end{equation}
-in the large $\nlive$ limit. We can again sum the harmonic series to see that the maximum live point is expected to be at 
+A full picture of the distribution is visualised in grey in \cref{fig:logL_distribution}. 
+\par
+Suppose we are now at iteration $i$, with the latest dead point having likelihood $\log \Like_i$.
+As with the prior volumes, we shall attempt to quantify how much extra information about the likelihood is contained in the live points.
+For convenience, we will normalise the live point likelihoods as
+\begin{equation}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
+    \label{eq:y}
+\end{equation}
+so that $y=0$ corresponds to the latest dead point and $y=1$ to the maximum. At each step, $y$ increases by  
+\begin{equation}
+    \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}, 
+\end{equation}
+which is constant.
+The log-likelihood therefore proceeds towards the maximum geometrically, with a `time constant' of $dn_i/2$.
+We can again sum the harmonic series to find that the maximum live point will be at
 \begin{equation}
@@
 It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.
-
+During the course of the run, the live points are always uniformly distributed over the constrained prior, and hence have probability distribution
+\begin{multline}
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i)^{\frac{d}{2}}} \\
+    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
+    \label{eq:PL}
+\end{multline}
+It is helpful at this stage to parameterise the live point log-likelihoods as
+\begin{equation}
+    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
+    \label{eq:y}
+\end{equation}
+as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_i$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
+\begin{equation}
+    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
+    \label{eq:Py}
+\end{equation}
+We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$.
+Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
+\begin{multline}
+    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}\\ 
+    [0<y_\mathrm{max}^\mathrm{live}<1],
+    \label{eq:Pyhat}
+\end{multline}
+which may be roughly summarised as
+\begin{multline}
+    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
+     \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
+    \label{eq:ymax}
+\end{multline}
+or in the large $d$, $n$ limit
+\begin{align}
+    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
+    \label{eq:ymaxd}\\
+    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
+    \label{eq:ymaxdn}
+\end{align}
+where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
 \subsubsection*{Aside: nested sampling as a maximiser}
@@
 
-\appendix
-\section{The maximum live log-likelihood}\label{sec:logLmax}
-Assume a Gaussian likelihood
-\begin{equation}\label{eq:logL}
-	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - X^{2/d}/2\sigma^2.
-\end{equation}
-Providing that $\sigma^d \ll 1$, the posterior can be approximated as
-\begin{equation}
-    P(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}.
-    \label{eqn:PX}
-\end{equation}
-The distribution of the true posterior in $\log\mathcal{L}$ is therefore
-\begin{equation}
-    P(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
-\end{equation}
-i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$. The posterior average and variance of $\log \mathcal{L}$ are given by
-\begin{equation}
-    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
-\end{equation}
-Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
-\begin{multline}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i)^{\frac{d}{2}}} \\
-    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
-    \label{eq:PL}
-\end{multline}
-It is helpful at this stage to define a parameter
-\begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
-    \label{eq:y}
-\end{equation}
-as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_i$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
-\begin{equation}
-    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
-    \label{eq:Py}
-\end{equation}
-We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$.
-Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
-\begin{multline}
-    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}\\ 
-    [0<y_\mathrm{max}^\mathrm{live}<1],
-    \label{eq:Pyhat}
-\end{multline}
-which may be roughly summarised as
-\begin{multline}
-    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
-     \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
-    \label{eq:ymax}
-\end{multline}
-or in the large $d$, $n$ limit
-\begin{align}
-    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
-    \label{eq:ymaxd}\\
-    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
-    \label{eq:ymaxdn}
-\end{align}
-where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
-\par
-This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they do steadily squeeze the interval $[\log\mathcal{L}_i,\log\mathcal{L}_\mathrm{max}]$.
-In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
-
-WHCOMMENT: How sure are we of this statement?
-I agree nested sampling is slower than gradient descent at maximising, but it still converges exponentially.
-What is the approximation of \cref{eq:xf} in this instance?
-
-WHCOMMENT: Is it worth hiding all these big results in the appendix -- it's not that long, and I think readers must read it to pace themselves through the maximum arguments?
-
-WHCOMMENT: $P(\log\mathcal{L})$ for the dead points?
-
-WHCOMMENT: maxmimum $(n)$ iteration notation?
-
-WHCOMMENT: \cref{fig:last_live_point} move legend.
-
 \bibliographystyle{mnras}

### 2023-11-27 paper/figures/dead_measure.py

@@ -10,17 +10,20 @@ from aeons.utils import *
 
 nlive = 100
 samples = get_samples('planck_gaussian_100', reduced=False)[1]
+ndeads = range(nlive*20, nlive*100, nlive*20)
+prev = np.array([20, 40, 40, 40]) * nlive
+fig, axes = plt.subplots(1,4, figsize=(7,2))
 
 x = 'omegabh2'
 y = 'omegach2'
 
-fig, axes = plt.subplots(1,4, figsize=(7,2))
-
-for ax in axes:
+for i, ax in enumerate(axes):
     ax.set_xticks([])
     ax.set_yticks([])
     ax.set_box_aspect(1)
-    ax.scatter(samples[x], samples[y], s=0.1)
+    points = samples.iloc[ndeads[i] - prev[i] : ndeads[i]]
+    ax.scatter(points[x], points[y], s=0.1)
+    # ax.scatter(samples[x], samples[y], s=0.1)
 
 axes[0].get_children()[-1].get_zorder()
 axes[0].set_zorder(10)
@@ -92,8 +95,8 @@ fig.canvas.draw()
 fig.canvas.flush_events()
 plot_live_points = True
 
-for k, (i, ax0, ax1) in enumerate(zip(range(nlive*20, nlive*100, nlive*20), axes[:-1], axes[1:])):
-    if plot_live_points:
+for k, (i, ax0, ax1) in enumerate(zip(ndeads, axes[:-1], axes[1:])):
+    if plot_live_points:    
         live = samples.live_points(i)
         ax0.scatter(live[x], live[y], s=2, color=f'C{k+1}')
         ax1.scatter(live[x], live[y], s=2, color=f'C{k+1}')
@@ -108,4 +111,3 @@ else:
     filename = 'dead_measure.pdf'
 
 fig.savefig(filename, transparent=False, bbox_inches='tight')
-

### 2023-11-28 paper/paper.tex
name change

@@
 \begin{figure}
-%\begin{Verbatim}[frame=single, commandchars=\\\{\}]
-%\textcolor{red}{Predicted endpoint: 25054 +/- 242}
-%\textcolor{red}{Progress: [=================>########] 72%}
-%___________________
-%lives      |   500 |
-%phantoms   | 24310 |
-%posteriors | 18018 |
-%equals     |   245 |
-%-------------------
-%ncluster   =  1/1
-%ndead      =  18018
-%nposterior =  18018
-%nequals    =  249
-%nlike      =  4159049
-%<nlike>    =  491.04 (9.82 per slice)
-%log(Z)     =  -12.55 +/- 0.27
-%\end{Verbatim}
+\begin{Verbatim}[frame=single, commandchars=\\\{\}]
+\textcolor{red}{Predicted endpoint: 25054 +/- 242}
+\textcolor{red}{Progress: [=================>########] 72%}
+___________________
+lives      |   500 |
+phantoms   | 24310 |
+posteriors | 18018 |
+equals     |   245 |
+-------------------
+ncluster   =  1/1
+ndead      =  18018
+nposterior =  18018
+nequals    =  249
+nlike      =  4159049
+<nlike>    =  491.04 (9.82 per slice)
+log(Z)     =  -12.55 +/- 0.27
+\end{Verbatim}
 \caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}
@@
 \end{equation}
-A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points , which is exactly needed to overcome the curse of dimensionality.
+A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points, which is exactly needed to overcome the curse of dimensionality.
 \begin{figure*}
@@
 \end{equation}
-A full picture of the distribution is visualised in grey in \cref{fig:logL_distribution}. 
+One can also derive the individual distributions for the live and dead points in log-likelihood.
+Note that this is merely the distribution of the \textit{values} of the points, without their nested sampling weights.
+The dead distribution is
+\begin{equation}
+    P(\log\Like) = \frac{\mathrm{d} \log X}{\mathrm{d} \log\Like} P(\log X) \propto \frac{dn}{2(\logLmax - \log\Like)},
+\end{equation}
+since the log-densities of the dead point prior volumes are proportional to the live point number at any given iteration.
+Meanwhile, the live points are uniformly distributed, and so have the distribution
+\begin{multline}
+	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i)^{\frac{d}{2}}} \\
+    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}].
+    \label{eq:PL}
+\end{multline}
+\cref{fig:logL_distribution} shows the distinction between the distributions of the live and dead point values, as well as the posterior. 
 \par
-Suppose we are now at iteration $i$, with the latest dead point having likelihood $\log \Like_i$.
-As with the prior volumes, we shall attempt to quantify how much extra information about the likelihood is contained in the live points.
-For convenience, we will normalise the live point likelihoods as
+How much further do the live points penetrate in log-likelihood?
+We seek the distribution for the maximum likelihood of the live points, $\log\Like_{(n_i)}^{\mathrm{live}}$, where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points.
+For convenience, we will normalise the likelihoods as
 \begin{equation}
     y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
-    \label{eq:y}
+    \label{eq:normalised_likelihood}
 \end{equation}
-so that $y=0$ corresponds to the latest dead point and $y=1$ to the maximum. At each step, $y$ increases by  
+so that $y=0$ corresponds to the latest dead point at $\log\Like_i$ and $y=1$ to the maximum. The live point distribution simplifies to
 \begin{equation}
+    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
+    \label{eq:Py}
+\end{equation}
+Using the result that the maximum of $n_i$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^{n_i})$, we obtain
+\begin{multline}
+    P(y_{(n_i)}^\mathrm{live}) = \frac{n_i d}{2}(1-y_{(n_i)}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_{(n_i)}^\mathrm{live})^{\frac{d}{2}}\right)^{n_i-1}\\ 
+    [0<y_{(n_i)}^\mathrm{live}<1],
+    \label{eq:Pyhat}
+\end{multline}
+which may be roughly summarised as
+\begin{multline}
+    y_{(n_i)}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n_i)}{\Gamma(1+\frac{2}{d}+n_i)} \\
+     \pm \left( \frac{\Gamma(1+n_i)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n_i)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n_i)^2}{\Gamma(1+\frac{2}{d}+n_i)^2}\right)^{\frac{1}{2}},
+    \label{eq:ymax}
+\end{multline}
+or in the large $d$, $n$ limit
+\begin{align}
+    \lim_{d\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
+    \label{eq:ymaxd}\\
+    \lim_{d,n\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
+    \label{eq:ymaxdn}
+\end{align}
+where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
+This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk.
+\par
+Alternatively, the same result arrives intuitively from the fact that at each step, $y$ increases by  
+\begin{equation}\label{eq:dy}
     \lim_{\nlive \to \infty}\Delta y \approx \frac{\mathrm{d} y}{\mathrm{d}\log X} \Delta \log X = \frac{2}{d \nlive}, 
 \end{equation}
-which is constant.
-The log-likelihood therefore proceeds towards the maximum geometrically, with a `time constant' of $dn_i/2$.
-We can again sum the harmonic series to find that the maximum live point will be at
+so that by again summing the harmonic series, we get
 \begin{equation}
-    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d},
+    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d}.
 \end{equation}
-where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points.
-This is a very small fraction in high dimensions, showing again that until the end the live points are far from the posterior bulk, and certainly nowhere near the maximum.
-\par
-The above also implies that the normalised distance between the highest and second highest live point is roughly 
+\cref{eq:dy} also implies that the normalised distance between the highest and second highest live point is roughly 
 \begin{equation}
@@
 It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.
-During the course of the run, the live points are always uniformly distributed over the constrained prior, and hence have probability distribution
-\begin{multline}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i)^{\frac{d}{2}}} \\
-    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
-    \label{eq:PL}
-\end{multline}
-It is helpful at this stage to parameterise the live point log-likelihoods as
-\begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
-    \label{eq:y}
-\end{equation}
-as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_i$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
-\begin{equation}
-    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
-    \label{eq:Py}
-\end{equation}
-We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$.
-Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
-\begin{multline}
-    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}\\ 
-    [0<y_\mathrm{max}^\mathrm{live}<1],
-    \label{eq:Pyhat}
-\end{multline}
-which may be roughly summarised as
-\begin{multline}
-    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
-     \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
-    \label{eq:ymax}
-\end{multline}
-or in the large $d$, $n$ limit
-\begin{align}
-    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
-    \label{eq:ymaxd}\\
-    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
-    \label{eq:ymaxdn}
-\end{align}
-where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
 \subsubsection*{Aside: nested sampling as a maximiser}
@@
 \par
-A more rigorous derivation (\cref{sec:logLmax}) shows that the maximum live point has mean and variance
-\begin{equation}\label{eq:ylivemax}
-	\lim_{d,\nlive\to\infty} y_\mathrm{max}^\mathrm{live} \sim \frac{2\log \nlive}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d}.
-\end{equation}
-Now let us take the dead point to be the termination point with likelihood $\log\Like_\mathrm{f}$ and prior volume $X_\mathrm{f}$, so that
+Let us take the current iteration to be the termination point with likelihood $\log\Like_\mathrm{f}$ and prior volume $X_\mathrm{f}$, so that
 \begin{equation}
@@
 \par
-Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$, and combining this with \eqref{eq:ylivemax} we find
+Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$ and combining this with \eqref{eq:ymaxdn}, we find
 \begin{equation}
@@
 \par
-A plot of the log-likelihood distribution at the end of a run is shown in \cref{fig:logL_distribution}
+Making the termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence.
+However, nested sampling still shrinks around the maximum exponentially, so provided a good termination criterion is chosen, it will get there in reasonable time.
+What we are really interested in is the number of iterations required to get from the posterior bulk to the true maximum, which we will now calculate.
+At the posterior, we have roughly speaking
+\begin{equation}
+    (\log X, \log\Like) = (-\DKL, \logLmax - d/2).
+\end{equation}
+The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum can then be found by inverting \cref{eq:gaussian_logL}:
+\begin{equation}
+    \log X_\delta = -\DKL - \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right),
+\end{equation}
+which corresponds to an additional
+\begin{equation}
+    \Delta i = \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right)
+\end{equation}
+iterations after the bulk is reached.
+One would therefore expect a general distribution to take $\mathcal{O}(d \log d)$ iterations to get from the usual nested sampling stopping point to within an e-fold of the maximum, which can serve as a sanity check for the maximisation procedure.
+\par
+A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.
+
 \begin{figure*}
@@
 Once $X_\mathrm{f}$ is known, the corresponding iteration count depends on the live point schedule.
-For example in the constant $\nlive$ case at each iteration $\log X$ decreases by $1/n$, so the total number of iterations $N_\mathrm{f}$ would be
+For example, in the constant $\nlive$ case $\log X$ decreases by $1/n$ at each iteration, so the total number of iterations $N_\mathrm{f}$ would be
 \begin{equation}
@@
 \par
+Uncertainties in the final estimate are obtained by drawing many samples from the distribution of $d_\mathrm{G}$ defined by the Bayesian temperature, and repeating step two for each.
 One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown this approach to be far less stable.
@@
 \subsection{Toy examples}
-Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\cite{anesthetic}.
+Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\citep{anesthetic}.
 \subsubsection{Gaussians}
@@
 The correct estimate is obtained to within standard error by about halfway, but before that is inaccurate.
-The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or that there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
+The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or because there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
 \par
@@
 Finally, we evaluate the method on real cosmological chains.
-\cref{fig:lcdm_logXfs} presents the endpoints (calculated after the fact) for nested sampling runs for curvature quantification on several common cosmological data sets (details in \citet{curvature_tension}).
+\cref{fig:lcdm_logXfs} presents the endpoints (calculated after the fact) for nested sampling runs for curvature quantification on several common cosmological data sets (details in \citealt{curvature_tension}).
 \par
@@
 \section*{Code availability}
-A package implementing the endpoint prediction mechanism is in development for easy plug-in to existing implementations of nested sampling.
+A package is in development to implement the endpoint prediction mechanism, for easy plug-in to existing implementations of nested sampling.
 The latest updates can be found at \href{https://github.com/zixiao-h/aeons}{github.com/zixiao-h/aeons}.
@@
 
-\appendix
-\section{The maximum live log-likelihood}\label{sec:logLmax}
-Assume a Gaussian likelihood
-\begin{equation}\label{eq:logL}
-	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - X^{2/d}/2\sigma^2.
-\end{equation}
-Providing that $\sigma^d \ll 1$, the posterior can be approximated as
-\begin{equation}
-    P(X) = \tfrac{1}{(2\pi\sigma^2)^{d/2}} e^{-X^{2/d}/2\sigma^2}.
-    \label{eqn:PX}
-\end{equation}
-The distribution of the true posterior in $\log\mathcal{L}$ is therefore
-\begin{equation}
-    P(\log\mathcal{L}) = \frac{1}{\Gamma\left(\frac{d}{2}\right)}e^{\log\mathcal{L}-\log\mathcal{L}_\mathrm{max}} (\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}
-\end{equation}
-i.e. $2(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}) \sim \chi^2_{d}$. The posterior average and variance of $\log \mathcal{L}$ are given by
-\begin{equation}
-    \langle\log\mathcal{L}\rangle_\mathcal{P} = \log\mathcal{L}_\mathrm{max} - \frac{d}{2},  \quad \mathrm{Var}(\log\mathcal{L})_\mathcal{P} = \frac{d}{2}.
-\end{equation}
-Meanwhile, the live points are uniformly distributed over the constrained prior and hence have probability distribution
-\begin{multline}
-	P(\log\mathcal{L}) = \frac{d}{2}\frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L})^{\frac{d}{2}-1}}{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i)^{\frac{d}{2}}} \\
-    [\log\mathcal{L}_i < \log\mathcal{L} <\log\mathcal{L}_\mathrm{max}],
-    \label{eq:PL}
-\end{multline}
-It is helpful at this stage to define a parameter
-\begin{equation}
-    y = \frac{\log\mathcal{L}-\log\mathcal{L}_i}{\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_i}
-    \label{eq:y}
-\end{equation}
-as a normalised measure of how far a point is between the latest dead point and the maximum log-likelihood, with $y=0$ corresponding to $\mathcal{L}_i$ and $y=1$ to $\mathcal{L}_\mathrm{max}$, so that
-\begin{equation}
-    P(y) = \frac{d}{2}(1-y)^{\frac{d}{2}-1} \quad [0<y<1].
-    \label{eq:Py}
-\end{equation}
-We now seek the distribution for the maximum likelihood of the live points, $\log\mathcal{L}_\mathrm{max}^{\mathrm{live}}$.
-Using the result that the maximum of $n$ variables with cumulative distribution $F(y)$ follows $\frac{d}{dy}( 1- (1-F(y))^n)$, we obtain
-\begin{multline}
-    P(y_\mathrm{max}^\mathrm{live}) = \frac{nd}{2}(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}-1}\left(1-(1-y_\mathrm{max}^\mathrm{live})^{\frac{d}{2}}\right)^{n-1}\\ 
-    [0<y_\mathrm{max}^\mathrm{live}<1],
-    \label{eq:Pyhat}
-\end{multline}
-which may be roughly summarised as
-\begin{multline}
-    y_\mathrm{max}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n)}{\Gamma(1+\frac{2}{d}+n)} \\
-     \pm \left( \frac{\Gamma(1+n)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n)^2}{\Gamma(1+\frac{2}{d}+n)^2}\right)^{\frac{1}{2}},
-    \label{eq:ymax}
-\end{multline}
-or in the large $d$, $n$ limit
-\begin{align}
-    \lim_{d\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
-    \label{eq:ymaxd}\\
-    \lim_{d,n\to\infty} y_\mathrm{max}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
-    \label{eq:ymaxdn}
-\end{align}
-where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
-\par
-This shows that in general the live points are nowhere near the maximum log-likelihood at any iteration, though they steadily squeeze the interval $[\log\mathcal{L}_i,\log\mathcal{L}_\mathrm{max}]$.
-In particular, in high dimensions $n$ only gets us harmonically/logarithmically closer, whilst $d$ pushes us linearly further away.
-
-WHCOMMENT: How sure are we of this statement?
-I agree nested sampling is slower than gradient descent at maximising, but it still converges exponentially.
-What is the approximation of \cref{eq:xf} in this instance?
-
-WHCOMMENT: Is it worth hiding all these big results in the appendix -- it's not that long, and I think readers must read it to pace themselves through the maximum arguments?
-
-WHCOMMENT: $P(\log\mathcal{L})$ for the dead points?
-
-WHCOMMENT: maxmimum $(n)$ iteration notation?
-
-WHCOMMENT: \cref{fig:last_live_point} move legend.
-
 \bibliographystyle{mnras}

### 2023-11-28 paper/figures/last_live_point.py

@@ -36,7 +36,10 @@ for ax in axs:
     ax.set_xticks([0, -points.logX().iloc[ndead]], [0, r'$-\log X_i$'])
 axs[0].set_title(r'$\mathcal{L}^\beta$', fontsize=8)
 axs[1].set_title('$\\mathcal{P}(\\log X) = \\mathcal{L}^\\beta X$', fontsize=8)
-axs[0].legend(loc='upper left', fontsize=5)
+
+from matplotlib.patches import Patch
+lines = [Patch(color=color, label=label) for color, label in zip(colors, labels)]
+axs[0].legend(handles=lines, ncol=3, bbox_to_anchor=(1.8, -0.3), fontsize=8)
 # axs[1].margins(x=0)
 
 fig.subplots_adjust(left=.01, right=.99, wspace=.05)

### 2023-11-29 paper/paper.tex
Temporary updates

@@
 We test these methods on both toy models and true cosmological nested sampling runs.
-The method gives an order-of-magnitude prediction of the end point at all times, forecasting the true endpoint within standard error generally around the halfway point.
+The method gives an order-of-magnitude prediction of the end point at all times, forecasting the true endpoint within standard error around the halfway point.
 \end{abstract}
@@
 \end{equation}
+Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
+\begin{equation}
+\mathcal{D}_\mathrm{KL} = \frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
+\end{equation}
+
 One can also derive the individual distributions for the live and dead points in log-likelihood.
@@
 \end{equation}
-In the event that we keep $\epsilon$ in, we replace $\frac{d}{2e}\to \frac{d}{2e}\epsilon^{2/d}$, so we can of course battle the $\frac{d}{2e}$ term, but this becomes exponentially difficult in high dimensions.
+In the event that we retain $\epsilon$, we replace $\frac{d}{2e}\to \frac{d}{2e}\epsilon^{2/d}$, allowing one to battle the $\frac{d}{2e}$ term exponentially as dimensions increase.
 \par
@@
 However, nested sampling still shrinks around the maximum exponentially, so provided a good termination criterion is chosen, it will get there in reasonable time.
-What we are really interested in is the number of iterations required to get from the posterior bulk to the true maximum, which we will now calculate.
-At the posterior, we have roughly speaking
+
+To quantify this statement, consider the number of iterations required to get from the posterior bulk to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
 \begin{equation}
@@
 \begin{equation}
-    \Delta i = \frac{d}{2} \left(\log \frac{d}{2} - \log \delta\right)
+    \Delta i = \frac{nd}{2} \left(\log \frac{d}{2} - \log \delta\right)
 \end{equation}
 iterations after the bulk is reached.
-One would therefore expect a general distribution to take $\mathcal{O}(d \log d)$ iterations to get from the usual nested sampling stopping point to within an e-fold of the maximum, which can serve as a sanity check for the maximisation procedure.
+One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum, which can serve as a sanity check for the maximisation procedure.
 \par

### 2023-11-29 paper/paper.tex
Adjustments made

@@
 \end{equation}
-A constant step size in $\log X$ corresponds to a geometrically constant measure for the dead points, which is exactly needed to overcome the curse of dimensionality.
+A steady step size in $\log X$ corresponds to a geometrically constant measure for the dead points, which is exactly needed to overcome the curse of dimensionality.
 \begin{figure*}
@@
 \begin{equation}
-\mathcal{D}_\mathrm{KL} = \frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
+\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
 \end{equation}
@@
 \begin{equation}
-    P(\log\Like) = \frac{\mathrm{d} \log X}{\mathrm{d} \log\Like} P(\log X) \propto \frac{dn}{2(\logLmax - \log\Like)},
+    P(\log\Like) = \frac{\mathrm{d} \log X}{\mathrm{d} \log\Like} P(\log X) \propto \frac{d\:n_i}{2(\logLmax - \log\Like)},
 \end{equation}
@@
 How much further do the live points penetrate in log-likelihood?
-We seek the distribution for the maximum likelihood of the live points, $\log\Like_{(n_i)}^{\mathrm{live}}$, where we have used the notation of order statistics to denote $x_{(n)}$ as the maximum of $n$ points.
+We seek the distribution for the maximum likelihood of the live points, $\log\Like_{(n_i)}^{\mathrm{live}}$, where we have used the notation of order statistics to denote $x_{(k)}$ as the maximum of $k$ points.
 For convenience, we will normalise the likelihoods as
@@
 \end{multline}
-or in the large $d$, $n$ limit
+or in the large $d$, $n_i$ limit
 \begin{align}
-    \lim_{d\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2H_n}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n))}{3d^2}\right)^{\frac{1}{2}},
+    \lim_{d\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2H_{n_i}}{d} \pm \left(\frac{2(\pi^2 - 6\Psi^{(1)}(1+n_i))}{3d^2}\right)^{\frac{1}{2}},
     \label{eq:ymaxd}\\
-    \lim_{d,n\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2\log n}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
+    \lim_{d,n_i\to\infty} y_{(n_i)}^\mathrm{live} &\sim \frac{2\log n_i}{d} \pm \sqrt{\frac{2}{3}}\frac{\pi}{d},
     \label{eq:ymaxdn}
 \end{align}
-where $\psi^{(1)}$ is the trigamma function and $H_n$ is the $n$th harmonic number.
+where $\psi^{(1)}$ is the trigamma function and $H_{n_i}$ is the $n_i$th harmonic number.
 This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk.
@@
     \boxed{
-        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n}{e} \pm \frac{\pi}{\sqrt{6}e}
+        \log{\mathcal{L}}_\mathrm{max}^\mathrm{live} \approx \log\mathcal{L}_\mathrm{max} - \frac{d}{2e} + \frac{\log n_i}{e} \pm \frac{\pi}{\sqrt{6}e}
     },
@@
 showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood.
-The final set of $n$ live points gets you $\log n/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
+The final set of $n_i$ live points gets you $\log n_i/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
 \par
-Making the termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence.
-However, nested sampling still shrinks around the maximum exponentially, so provided a good termination criterion is chosen, it will get there in reasonable time.
+Making the traditional termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence.
+However, nested sampling still shrinks around the maximum exponentially, so provided a good alternative termination criterion is chosen, it will get there in reasonable time.
 
-To quantify this statement, consider the number of iterations required to get from the posterior bulk to the true maximum, which we will now calculate. At the posterior, we have roughly speaking
+To quantify this statement, consider the number of iterations $\Delta i$ required to get from the posterior bulk to the true maximum, which we will now calculate.
+At the posterior, we have roughly speaking
 \begin{equation}
@@
 \end{equation}
-iterations after the bulk is reached.
-One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum, which can serve as a sanity check for the maximisation procedure.
+iterations after the bulk is reached, where $n$ is the harmonic mean of the number of live points.
+One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum.
+A rule-of-thumb termination criterion could therefore be to run for at least $\tfrac{nd}{2}\log\tfrac{d}{2}$ iterations after the posterior is reached.
 \par
@@
 \end{equation}
-The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n)$, The second term is the average time per likelihood evaluation.
+The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n_i)$, The second term is the average time per likelihood evaluation.
 The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude.

### 2023-11-30 paper/figures/logL_distribution.py
correcting logLlivemax position

@@ -7,7 +7,7 @@ figsettings()
 fig, ax = plt.subplots()
 
 logLmax = 0
-d = 30
+d = 70
 n = 1000
 avlogL = logLmax-d/2
 
@@ -45,7 +45,7 @@ ax.axvline(logLmax, color='k', linestyle=':', lw=1)
 ax.plot([avlogL, logLmax], [l1,l1], 'k-', lw=.5)
 ax.annotate(r'$\frac{d}{2}$', ((avlogL+logLmax)/2,l1+pad_text), ha='center', va='bottom', fontsize=8)
 
-l2 = 0.7
+l2 = 0.68
 ax.plot([avlogL-np.sqrt(d/2), avlogL+np.sqrt(d/2)], [l2, l2], 'k-', lw=.5)
 ax.annotate(r'$\pm\sqrt{\frac{d}{2}}$', (avlogL+0.3,l2+pad_text), ha='left', va='bottom', fontsize=8)
 
@@ -61,24 +61,29 @@ l5 = 0.75
 ax.plot([logLs, logLmaxlive], [l5, l5], 'k-', lw=.5)
 ax.annotate(r'$\frac{\log n}{e}$', ((logLs+logLmaxlive)/2,l5+pad_text), ha='center', va='bottom', fontsize=8)
 
-ax.annotate("Iterations", xytext=(-28, 0.5), xy=(-23, 0.52), arrowprops=dict(arrowstyle="->"), fontsize=8)
+ax.annotate("Iterations", xytext=(-55, 0.5), xy=(-48, 0.52), arrowprops=dict(arrowstyle="->"), fontsize=8)
+
 
 ax.set_xticks([avlogL, 
                logLs, 
                logLmaxlive,
                logLmax])
 ax.set_xticklabels([r'$\langle\log\mathcal{L}\rangle_\mathcal{P}$',
-                    r'$\log\mathcal{L}_i$',
+                    r'$\log\mathcal{L}_f$',
                     r'$\log\mathcal{L}_\mathrm{max}^\mathrm{live}$',
                     r'$\log\mathcal{L}_\mathrm{max}$'],
                     fontsize=8)
+ticks = ax.get_xticklabels()
+ticks[1].set_rotation(25)
+ticks[2].set_rotation(25)
 
 ax.set_yticks([])
 ax.legend(loc='upper left', fontsize=8)
-ax.set_ylim([0,1.4])
+ax.set_ylim([0, 1.4])
[…]

### 2023-11-30 paper/figures/logX_distribution.py

@@ -41,7 +41,7 @@ ax.axvline(-ndead/nlive, color='deepskyblue', lw=.5, ls='--')
 # Arrows showing log n + gamma
 l = 0.32
 pad_text = 0.05
-ax.annotate(r'$\log n+\gamma$', ((-ndead/nlive+ ElogXlivemin)/2 - 3,l+pad_text), ha='center', va='bottom', fontsize=8)
+ax.annotate(r'$-\log n_i-\gamma$', ((-ndead/nlive+ ElogXlivemin)/2 - 3,l+pad_text), ha='center', va='bottom', fontsize=8)
 arrow_start = (-ndead/nlive, l)
 arrow_end = (ElogXlivemin, l)
 arrow = patches.FancyArrowPatch(arrow_start, arrow_end, arrowstyle='->', mutation_scale=5)

### 2023-12-28 paper/paper.tex
correcting logLlivemax position
changes based on johannes' comments

@@
 \begin{equation}
-\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2
+\mathcal{D}_\mathrm{KL} = -\frac{d}{2}\log 2 \Gamma(1+\tfrac{d}{2})^{2/d} \sigma^2.
 \end{equation}
@@
 
-\subsection{The termination prior volume}
+\subsection{The termination prior volume}\label{sec:termination}
 Making the above discussion more precise, we wish to find the compression factor $\log X_ \mathrm{f}$ at which the termination criterion is met, which is larger in magnitude than $\DKL$ (\cref{fig:logX_distribution}).
@@
 Finally, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
-This approach can however be useful in the low dimensions appropriate for \textsc{Ultranest} when the live points are always near the maximum, but in general is less reliable.
+This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.
+\par
+We also note that for implementations whose sampling efficiency $f_\mathrm{sampler}$ varies, the progress indicator of \cref{sec:termination} does not directly correspond to time, but will still be useful if the variation is known \textit{a priori}.
+For example, ellipsoidal samplers such as \textsc{MultiNest} and the MLFriends algorithm in \textsc{UltraNest} reduce in efficiency near the posterior bulk, so an impatient user who does not wish to expend the extra exponential effort for the final few percent can confidently halt earlier if they know they are already near the termination point.
 

### 2024-02-22 paper/paper.tex
new branch for reviewer changes

@@
 
-\section*{Code availability}
+\section*{Data availability}
 A package is in development to implement the endpoint prediction mechanism, for easy plug-in to existing implementations of nested sampling.

### 2024-07-17 paper/paper.tex
moderate revisions for journal submission

@@
 \usepackage{xcolor}
+\usepackage{gensymb}
 \usepackage[T1]{fontenc}
@@
 \newcommand{\logLmax}{\log \Like_\mathrm{max}}
+\newcommand{\dG}{d_\mathrm{G}}
 
@@
 It was immediately adopted for cosmology, and is now used in a wide range of physical sciences including particle physics \citep{Trotta_2008}, materials science \citep{materials} and machine learning \citep{sparse_reconstruction}.
-The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested monte carlo integrals which makes high-dimensional integration feasible and robust.
+The core algorithm is unique in its estimation of volumes by \textit{counting}, coupling together nested Monte Carlo integrals which makes high-dimensional integration feasible and robust.
 It also avoids problems that challenge traditional Bayesian samplers, such as posterior multi-modality and phase transitions.
@@
 At each iteration $i$, the point with the lowest likelihood is removed and added to a list of \textit{dead points} (illustrated in \cref{fig:dead_measure}).
-New points are then (optionally) drawn from the prior, subject to the constraint that they must have a higher likelihood than the latest dead point~$\mathcal{L}_i$.
+New points are then (optionally) drawn from the prior, subject to the constraint that they must have a higher likelihood than the current dead point~$\mathcal{L}_i$.
 Repeating the procedure leads to the live points compressing around peaks in the likelihood.
@@
 \par
-The integral in \eqref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
+The integral in \cref{eq:evidence} is then evaluated by transformation to a one-dimensional integral over the \textit{prior volume} $X$
 \begin{equation}
@@
 The bulk of the posterior lies within a prior volume ${X = e^{-\DKL}}$, which is the target compression.
-From \cref{eq:t_moments} one gets there by iteratively taking steps of size ${\Delta \log X_i = -1/n_i}$, so that when we add up the contribution of each step in \eqref{eq:t_moments} we get
+From \cref{eq:t_moments} one gets there by iteratively taking steps of size ${\Delta \log X_i = -1/n_i}$, so that when we add up the contribution of each step in \cref{eq:t_moments} we get
 \begin{equation}
@@
 \end{center}
-\caption{The dead points of a nested sampling run, recursively zoomed in.
-Their density is constant in $\log X$, which is a geometrically constant measure $\propto \mathrm{d} X/X$, and hence scale invariant until the final set of live points is reached.
-The live points at a given iteration $i$ (larger dots) are uniform across the prior, plotted for comparison.}
+\caption{The dead (small dots) and live points (larger dots) at several stages in a nested sampling run, recursively zoomed in.
+Note that the prior is taken to be uniform, which is reasonable because most nested sampling implementations require the prior to be provided as a transformation to the unit hypercube.
+The live points are uniform across the prior, while the dead points have a geometrically constant density.}
 \label{fig:dead_measure}
@@
 \par
-The same is not true for the live points, which are uniformly distributed in prior volume, and by the comments in the penultimate paragraph of \cref{sec:background} have ${n_{i+k} = n_i-k}$.
+The same is not true for the live points, whose enclosed prior volume values are uniformly distributed, and by the comments in the penultimate paragraph of \cref{sec:background} have ${n_{i+k} = n_i-k}$.
 As a result, the maximum live point is found at
@@
 In other words, it is not until we are around $\log \nlive$ away from $\log X = -\DKL$ that the samples begin populating the posterior typical set.
-One can see from \eqref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
+One can see from \cref{eq:DKL} that the divergence increases linearly with dimension, so for large dimensionalities and typical live point numbers $\lesssim 1000$, this does not happen until near the end of the run.
 \par
@@
 We now consider the distribution of the samples in log-likelihood.
-To get an insight into the analytics, we will examine the representative case of the $d$-dimensional multivariate Gaussian with maximum point $\logLmax$ and length scale $\sigma$:
+To get an insight into the analytics, we will examine the case of the $d$-dimensional multivariate Gaussian likelihood with maximum point $\logLmax$ and length scale $\sigma$, situated in a uniform unit prior:
+\begin{equation}\label{eq:logL_x}
+    \log\Like = \logLmax - \frac{|\bm{x}|^{2}}{2\sigma^{2}}, \quad \bm{x} \in \mathbb{R}^{d}.
+\end{equation}
+This case is common and useful because many likelihoods are near-Gaussian, and most nested sampling implementations require the prior to be specified in terms of a transformation into the unit hypercube i.e.
+in prior probability units.
+Providing that $\sigma^d \ll 1$, the likelihood and posterior can be approximated as
 \begin{equation}\label{eq:gaussian_logL}
-    \log \Like = \logLmax - X^{2/d}/2\sigma^2.
+    \log \Like = \logLmax - X^{2/d}/2\sigma^2,
 \end{equation}
-Providing that $\sigma^d \ll 1$, the posterior can be approximated as 
 \begin{equation}
@@
 \end{equation}
-Given the prior is uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
+Given the prior volumes are uniform $X\sim U(0,1)$, the KL divergence in the $\sigma^d \ll 1$ limit is:
 \begin{equation}
@@
 One can also derive the individual distributions for the live and dead points in log-likelihood.
-Note that this is merely the distribution of the \textit{values} of the points, without their nested sampling weights.
+Note that this is merely the distribution of the $\log \Like$ \textit{values} of the points, without considering the associated prior volumes.
 The dead distribution is
@@
 since the log-densities of the dead point prior volumes are proportional to the live point number at any given iteration.
-Meanwhile, the live points are uniformly distributed, and so have the distribution
+Meanwhile, the prior volumes enclosed by the live points (being independent samples) are always uniformly distributed, and so have the distribution
 \begin{multline}
@@
 \end{equation}
-so that $y=0$ corresponds to the latest dead point at $\log\Like_i$ and $y=1$ to the maximum. The live point distribution simplifies to
+so that $y=0$ corresponds to the most recent dead point at $\log\Like_i$ and $y=1$ to the maximum. The live point distribution simplifies to
 \begin{equation}
@@
 \end{multline}
-which may be roughly summarised as
+which when analytically integrated to find its first and second moments give the mean and variance to be 
 \begin{multline}
     y_{(n_i)}^\mathrm{live} \sim 1-\frac{\Gamma(1+\frac{2}{d})\Gamma(1+n_i)}{\Gamma(1+\frac{2}{d}+n_i)} \\
-     \pm \left( \frac{\Gamma(1+n_i)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n_i)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n_i)^2}{\Gamma(1+\frac{2}{d}+n_i)^2}\right)^{\frac{1}{2}},
+     \pm \left( \frac{\Gamma(1+n_i)\Gamma(1+\frac{4}{d})}{\Gamma(1+\frac{4}{d}+n_i)} - \frac{\Gamma(1+\frac{2}{d})^2 \Gamma(1+n_i)^2}{\Gamma(1+\frac{2}{d}+n_i)^2}\right)^{\frac{1}{2}}.
     \label{eq:ymax}
 \end{multline}
-or in the large $d$, $n_i$ limit
+In the large $d$, $n_i$ limit, taking the first and second order terms in the Taylor expansion simplifies the above expressions to 
 \begin{align}
@@
 where $\psi^{(1)}$ is the trigamma function and $H_{n_i}$ is the $n_i$th harmonic number.
-This is a very small fraction in high dimensions, showing that until the end the live points are far from the posterior bulk.
+The final term is a very small fraction in high dimensions, showing that for Gaussian and near-Gaussian cases the live points are far from the posterior bulk until the very end.
 \par
@@
 \begin{equation}
-    y_{(\nlive)}^{\mathrm{live}} = \frac{2\log \nlive}{d}.
+    y_{(\nlive)}^{\mathrm{live}} \approx \frac{2\log \nlive}{d}.
 \end{equation}
@@
 \end{equation}
-In other words, the highest likelihood point is always at least an order of magnitude greater than the second highest.
+In other words, for a Gaussian the highest likelihood point is always at least an order of magnitude greater than the second highest.
 It is therefore typically the case that nearly all of the posterior mass is concentrated in a single point, the maximum live point, until the very end of the run when the prior volumes have shrunk enough to compensate.
+\par
+The above results assume an exact Gaussian likelihood, which as discussed provides a reasonable first estimate for many physical likelihoods that are near-Gaussian.
+Further work can be done to generalise the analytics to non-Gaussians, though as we see in \cref{sec:results} results obtained via Gaussian assumptions can give the right order of magnitude.
 \subsubsection*{Aside: nested sampling as a maximiser}
@@
 In particular, the latter authors emphasised that posterior samplers such as nested sampling find the bulk of the \textit{mass}, not the maximum of the distribution, but that this can be remedied by tightening the termination criterion.
-We now use the machinery we have developed to put this statement on a more quantitative footing.
+We now use the machinery we have developed to put this statement on a more quantitative footing, again examining the Gaussian case.
 \par
@@
 Note that we have assumed that prior effects are negligible (so $1\approx \infty$), and that $\epsilon \ll 1$ so that the denominator is approximately the accumulated evidence.
-Computing this for \eqref{eq:gaussian_logL}, we find the answer in terms of lower incomplete gamma functions
+Using \cref{eq:gaussian_logL}, we find the Gaussian termination condition in terms of lower incomplete gamma functions to be
 \begin{equation}
@@
 Taking the $X_\mathrm{f}\ll (\sqrt{2}\sigma)^d$ limit (almost certainly valid at termination) we find
-\begin{equation}
+\begin{equation}\label{eq:Xf_approx}
     \lim_{X_\mathrm{f}\ll (\sqrt{2}\sigma)^d} \epsilon \approx \frac{X_\mathrm{f}}{(\sqrt{2}\sigma)^d \ \Gamma\left(1+\frac{d}{2}\right)} = \frac{(\log\mathcal{L}_\mathrm{max}-\log\mathcal{L}_\mathrm{f})^{\frac{d}{2}}}{\Gamma\left(1+\frac{d}{2}\right)}.
@@
 \par
-Putting this together, taking $\mathcal{L}_i$ in \eqref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$ and combining this with \eqref{eq:ymaxdn}, we find
+Putting this together, taking $\mathcal{L}_i$ in \cref{eq:normalised_likelihood} to be $\mathcal{L}_\mathrm{f}$ and combining this with \cref{eq:ymaxdn}, we find
 \begin{equation}
@@
 \end{equation}
-showing that in general nested sampling will finish at a contour $d/2e$ away from the maximum log-likelihood.
-The final set of $n_i$ live points gets you $\log n_i/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still by statistical fluctuation.
+showing that Gaussian nested sampling chains will finish at a contour $d/2e$ away from the maximum log-likelihood.
+The final set of $n_i$ live points gets you $\log n_i/e$ closer, with a chance of getting $\sim\pi/\sqrt{6}e=0.472$ closer still at the $1\sigma$ interval.
 \par
-Making the traditional termination criterion stricter therefore has limited returns in high-dimensions, if it is ultimately still based on the remaining evidence.
+Making the traditional termination criterion stricter therefore has limited returns in high dimensions, if it is ultimately still based on the remaining evidence.
 However, nested sampling still shrinks around the maximum exponentially, so provided a good alternative termination criterion is chosen, it will get there in reasonable time.
-
-To quantify this statement, consider the number of iterations $\Delta i$ required to get from the posterior bulk to the true maximum, which we will now calculate.
+\par
+To quantify this statement, consider the number of iterations $\Delta i$ required to get from the posterior mean to the true maximum, which we will now calculate.
 At the posterior, we have roughly speaking
@@
 \end{equation}
-The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum can then be found by inverting \cref{eq:gaussian_logL}:
+The prior volume $\log X_\delta$ at which the likelihood is within $\delta$ of the maximum of the Gaussian can then be found by inverting \cref{eq:gaussian_logL}:
 \begin{equation}
@@
 which corresponds to an additional
-\begin{equation}
+\begin{equation}\label{eq:delta_i_pd}
     \Delta i = \frac{nd}{2} \left(\log \frac{d}{2} - \log \delta\right)
@@
 iterations after the bulk is reached, where $n$ is the harmonic mean of the number of live points.
-One would therefore expect a general distribution to take $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum.
+One would therefore expect a general distribution to take a similar order of magnitude of steps, that is $\mathcal{O}(\tfrac{n d }{2}\log \tfrac{d}{2})$ iterations to get from the usual nested sampling stopping point to within an $e$-fold of the maximum.
 A rule-of-thumb termination criterion could therefore be to run for at least $\tfrac{nd}{2}\log\tfrac{d}{2}$ iterations after the posterior is reached.
 \par
+We can also put compare this endpoint with the termination condition based on $\epsilon$. From \cref{eq:Xf_approx}, for any $d$, $n$ we have
+\begin{equation}
+    \log X_\mathrm{f} \approx \frac{d}{2}\log 2\sigma^2 + \log\epsilon + \log\Gamma\left(1+\frac{d}{2}\right),
+\end{equation}
+and using \cref{eq:gaussian_logL} to invert $\log\Like = \logLmax-\delta$ gives
+\begin{equation}
+    \log X_\delta=\frac{d}{2}\log 2\sigma^2+\frac{d}{2}\log\delta.
+\end{equation}
+Putting these together means that the difference in $\log X$ between the termination criterion based on evidence fraction $\epsilon$, and a point $\delta$ away from the maximum log-likelihood is
+\begin{equation}
+    \log X_\delta - \log X_\mathrm{f} \approx \frac{d}{2}\log\delta - \log\epsilon - \log\Gamma\left(1+\frac{d}{2}\right).
+\end{equation}
+Getting to within $\mathcal{O}(1)$ of the maximum after the termination criterion therefore takes an additional
+ \begin{equation}
+     \Delta i \approx n \left(\log\Gamma\left(1+\frac{d}{2}\right) - \log\frac{1}{\epsilon}\right)
+\end{equation}
+iterations.
+A larger dimension pushes the maximum away from the termination point, while a smaller $\epsilon$ brings it closer.
+Note the similarity of this equation to \cref{eq:delta_i_pd} via the Stirling approximation, and setting $\epsilon = 1/2$, approximately true at the posterior mean.
+\par
+It automatically follows that the $\epsilon$ one should set to end the run $\mathcal{O}(1)$ away from the maximum is simply
+\begin{equation}
+    \boxed{\epsilon \approx \frac{1}{\Gamma\left(1 + d/2\right)}}.
+\end{equation}
 A summary of the distances between the notable points at the end of a run is shown in \cref{fig:logL_distribution}.
@@
 \subsubsection*{Motivations}
-As shown in the previous section, midway through the run nearly all of the posterior mass is concentrated at a single point.
-However, this does not capture the \textit{structure} of the posterior that has been explored and all of the information it provides.
+In addition to $\log \Like_i$ and $\log X_i$, another useful piece of information at each iteration is the current \textit{temperature}, which we will now define.
 \par
-We have the potential to fix this because nested sampling is invariant to monotonic transformations, so we can transform the likelihood as $\Like \to \Like^{\beta}$ without loss of information by trivially re-weighting the samples.
-Increasing $\beta$ worsens the situation, while $\beta \to 0$ simply gives back the prior.
-There is, on the other hand, a significant intermediate range which makes the samples look like a posterior centred at the present contour, which will allow us to recover the structure of the samples.
-A schematic of the procedure is shown in \cref{fig:last_live_point}.
+Bayesian inference is closely related to thermodynamics, for if one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble at inverse temperature $\beta = 1/T$:
+\begin{equation}
+    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\Like^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}.
+\end{equation}
+Nested sampling therefore calculates the partition function $Z(\beta)$, with the feature that it is agnostic to the temperature; because a nested sampling chain is invariant under monotonic transformations of the likelihood, one can get the posterior at any temperature by setting $\Like \to \Like^\beta$ without losing information.
 \par
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/last_live_point.pdf}
-\end{center}
-\caption{The likelihood and posterior as a function of $\log X$ in the middle of a nested sampling run.
-Almost all of the posterior mass is concentrated at a single point with the highest compression, because it is orders of magnitude higher in likelihood.
-Reducing $\beta$ re-weights to shift the posterior mass; taking $\beta \to 0$ goes too far and gives back the prior, but there is some intermediate $\beta = \beta_i$ which is significant because it centres the posterior at the current contour.}
-\label{fig:last_live_point}
-\end{figure}
+Transforming the likelihood with $\beta$ changes the termination point of the run, because the posterior mass is redistributed.
+In the middle of a run, there is therefore a temperature one can set which makes the current iteration the termination point.
+This `current temperature of termination' provides a measure of progress, marching from $0 \to 1$ as the run converges, though it is uncalibrated in the sense that knowledge of the current temperature gives no indication of when the final temperature $\beta = 1$ will be reached.
+It is of course dependent on the threshold $\epsilon$ chosen for the termination point, and we will label this $\beta$ as $\beta_\epsilon$ to distinguish it from other definitions of the temperature we will later discuss.
 \par
-At this point it is relevant to note the correspondence between Bayesian inference and statistical mechanics, from which the above transform is derived.
-If one equates the parameters to microstates $i$, the negative log-likelihood to the microstate energy $E_i$, and the prior to the density of states $g_i$, then the posterior as given by the generalised Bayes' rule is the canonical ensemble
-\begin{equation}
-    p(E_i) = \frac{g_i e^{-\beta E_i}}{Z(\beta)} \quad \leftrightarrow \quad \mathcal{P}_\beta(\theta) = \frac{\mathcal{L}^{\beta}(\theta)\pi(\theta)}{\mathcal{Z}(\beta)}
-\end{equation}
-at the inverse temperature $\beta = 1/T$.
-As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, but nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours.
+Changing the termination point is perhaps the most obvious consequence of changing $\beta$, but transforming the posterior mass can be useful in other ways.
+$\beta$ is ultimately just an additional degree of freedom in a nested sampling run, to be set at any value one sees fit.
+The primary purpose for introducing it in this paper is that by redistributing the posterior mass like so, one can regularize the Bayesian model dimensionality of an intermediate set of samples, which provides crucial information about the likelihood and hence the termination point.
+This will be elaborated on in \cref{sec:dimensionality}; for now, we discuss several other `temperatures' relevant to nested sampling, including those introduced in previous literature.
+
+\subsubsection{Microcanonical temperature}
+As noted by \citet{demons}, thermal algorithms such as thermodynamic integration \citep{path_sampling} get the evidence by evolving through a series of canonical ensembles via some temperature schedule, or annealing schedule.
+Nested sampling instead maintains \textit{microcanonical} ensembles, which are the iso-likelihood contours.
 Instead of using temperature \citep{simulated_annealing, parallel_tempering} or energy \citep{wang_landau} as a control parameter, nested sampling chooses a series of ensembles with constant relative volume entropy $\Delta \log X$, which allows the algorithm to handle phase transitions \citep{baldock}.
 \par
-Because the temperature of a microcanonical ensemble is a derived property rather than a parameter, there is some freedom in its definition.
-Returning to our original motivation, we make the connection that the temperature is the re-weighting $\Like \to \Like^{\beta}$ which centres the ensemble around the current energy.
-We now present several temperatures that achieve this aim, each of which one can plausibly consider to be the current temperature of a nested sampling run.
-
-\subsubsection{Microcanonical temperature}\label{sec:microcanonical_temperature}
-The obvious candidate is the microcanonical temperature $\partial S/\partial E$, where the volume entropy is $\log X$ and the energy is as usual $-\log \Like$.
+An obvious candidate for the current temperature is therefore the microcanonical temperature $\beta_\mathrm{M} = \partial S/\partial E$, where the volume entropy is $\log X$ and the energy is as usual $-\log \Like$.
 This gives the density of states; as discussed in Skilling's original paper,
 \begin{equation}
-    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like_i}.
+    \beta_\mathrm{M}  = - \frac{\mathrm{d} \log X}{\mathrm{d} \log \Like} \Bigg\vert_{\log \Like_i}
 \end{equation}
-is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X_i$, if we assume differentiability, which is exactly the intuition we were aiming for to put the ensemble bulk at the current contour. 
+is the $\beta$ at which $\Like^{\beta} X$ peaks at $\log X_i$, if differentiability is assumed.
+If the posterior is symmetric as a function of $\log X_i$ about its peak, then this corresponds to $\beta_\epsilon = 1/2$.
 \par
@@
 \subsubsection{Canonical temperature}\label{sec:canonical_temperature}
-Another temperature considered by \citet{demons} is that at which the current energy (i.e. $-\log \Like_i$) is the average energy of the entire ensemble. One can obtain it by inverting 
+Another temperature, considered by \citet{demons}, is the $\beta$ at which the energy of the current dead point (i.e.
+$-\log \Like_i$) is equal to the average energy of all the points, live and dead.
+One can obtain it by inverting
 \begin{equation}
@@
 \end{equation}
-to get the `canonical' temperature $\beta_\mathrm{C}$.
+to find the `canonical' temperature $\beta_\mathrm{C}$ for which the above equation is true.
 While $\beta_\mathrm{M}$ is derived from (the gradient of) a single contour, this temperature uses the entire ensemble.
@@
 The obtained temperatures are consistent with the previous two choices, which may seem oddly coincidental.
-In fact, closer inspection reveals that large values of $P(\beta \mid \log X_i)$ are the temperatures with a large value of the posterior at the present contour, normalised by the corresponding evidence.
-Thus the Bayesian temperature uses the same idea as the microcanonical one, except it accounts for the spread in the result.
+In fact, closer inspection reveals that large values of $P(\beta \mid \log X_i)$ are the temperatures with the largest values of the posterior at the present $\log X_i$, normalised by the corresponding evidence.
+Thus the Bayesian temperature uses the same idea as the microcanonical one (which maximises the posterior), except here the spread in result is accounted for.
 
 \subsubsection*{Comparisons}
-\cref{fig:beta_comparison} shows the three temperatures as a function of compression for two cases, one containing a phase transition and one without.
-They are consistent in both cases when there is a single dominant phase, but differ during a phase transition.
-The canonical temperature is the only one that rises monotonically with compression.
+\cref{fig:beta_posteriors} shows the posterior distribution as a function of $\log X$ for a 32-d spherical Gaussian when set to the above four temperatures, as known at a given iteration roughly 30\% through a run.
+Lower prior volumes are to the right, and 25 samples are taken from the prior volume distribution to capture the nested sampling uncertainties.
+When $\beta = 1$, the untransformed posterior is concentrated at the single point with the highest likelihood.
+This is no longer the case when $\beta$ is set to any of the other values, causing the posterior to have non-vanishing variance.
+The termination temperature can be seen to transform the posterior to have exactly $\epsilon$ of the total evidence to the right hand side of the current iteration (dotted black line).
+$\beta_\mathrm{M}$ makes the peak of the posterior with respect to $\log X$ coincide with the current iteration, and $\beta_\mathrm{C}$ produces a very similar result from its own criteria.
+25 samples are taken from the distribution of the Bayesian temperature and the resultant posterior plotted for each, displaying a much broader range of uncertainty compared to the other plots.
+\par
+\cref{fig:beta_plots} then shows how these four temperatures evolve during the run (i.e.
+as a function of $\log X$) for two cases: one containing a classic slab-spike phase transition and one without.
+The behaviour of the former case is especially useful, because it is well-known to trouble temperature-based samplers.
+\par
+In the case of a single phase all temperatures are consistent, with the exception that this only holds for $\epsilon \approx 0.5$ for the termination temperature.
+This is expected, since we can see from \cref{fig:beta_posteriors} that this choice of epsilon places the current iteration roughly at the peak of the posterior just as $\beta_M$ does.
+Other choices of $\epsilon$ shift the profile up and down.
+This user-dependence can offer more flexibility, but also be a disadvantage as an extra parameter to be tuned.
+The temperatures diverge, however, during a phase transition.
+Notably, with the exception of the canonical temperature they all \textit{decrease} then increase again, in contrast to a typically monotonic annealing schedule.
+It is also evident from the plots that $\beta_\mathrm{B}$ is the only temperature that incorporates a measure of uncertainty.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/beta_posteriors.pdf}
+\end{center}
+\caption{The posterior distribution at an intermediate point 30\% of the way through a nested sampling run for a 32$d$ spherical Gaussian, with the likelihood transformed from  $\Like \to \Like^{\beta}$ for several different choices of $\beta$.
+The untransformed posterior has all its weight at a single point.
+The other $\beta$ produce relatively similar transformations with finite posterior variance.
+Notably, the termination temperature is dependent on a manual choice of $\epsilon$, and the Bayesian temperature includes uncertainty in $\beta$ itself.}
+\label{fig:beta_posteriors}
+\end{figure*}
+
 \begin{figure}
 \begin{center}
-    \includegraphics{figures/beta_comparison.pdf}
+    \includegraphics{figures/beta_plots.pdf}
 \end{center}
-\caption{Inferred temperatures using the microcanonical, canonical and Bayesian definitions.
-The shaded regions show the $1-2\sigma$ uncertainties.
-All are consistent for a single phase, but differ during a phase transition.}
-\label{fig:beta_comparison}
+\caption{Inferred temperatures using the termination ($\epsilon = 10^{-3}$), microcanonical, canonical and Bayesian definitions.
+The shaded region shows the $1-2\sigma$ uncertainties for the Bayesian temperature.
+$\beta_\epsilon$ is offset from the others, but can be shifted up and down depending on the chosen $\epsilon$.
+The other $\beta$ are all are consistent in a single phase, but differ during a phase transition.}
+\label{fig:beta_plots}
+\end{figure}
+Despite the above theoretical justifications, one should remember the original motivation for introducing $\beta$ was to utilise the extra degree of freedom it provides, to transform the likelihood without losing information.
+As we will see in \cref{sec:choosing_the_best_temperature}, one can choose the exact definition depending on what is useful for the problem at hand.
+
+\subsection{Dimensionality}\label{sec:dimensionality}
+We can now use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible.
+\citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality $\dG$, defined as the posterior variance of the information content:
+\begin{equation}\label{eq:d_G}
+    \frac{\dG}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
+    = \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
+\end{equation}
+As is discussed in \citet{Handley_2019}, $\dG$ holds valuable information about the posterior, measuring the number of spatial degrees of freedom in the posterior in the manner of principal component analysis, but without introducing arbitrary cut-off eigenvalues.
+It is exact in the case of a multivariate Gaussian likelihood that is small in length scale relative to the prior, so for non-Gaussian likelihoods $\dG$ can be considered the equivalent dimension if it were actually a Gaussian.
+Furthermore, it can be written as the posterior variance of the log-likelihood:
+\begin{equation}
+    \frac{\dG}{2} = \mathrm{Var}(\log\Like)_\mathcal{P}.
+\end{equation}
+Unfortunately, calculating this quantity in the middle of a run tends to return zero.
+That naturally follows from the results of \cref{sec:logL}, which shows that an intermediate set of samples typically have all their weight concentrated at a single point, the maximum live point, and a single point has vanishing variance.
+This behaviour continues practically until the run is close to termination.
+Re-weighting the likelihood by re-setting the temperature, however, leads this to no longer be the case.
+As shown in \cref{fig:beta_posteriors}, the posterior variance (and hence dimensionality) is zero at $\beta = 1$ but nonzero for the various other temperatures presented.
+\par
+We find that the $\dG$ calculated from these transformed posteriors are exactly the Bayesian model dimensionalities of the \textit{true} posterior.
+\cref{fig:d_G_gaussians} demonstrates this using the microcanonical and Bayesian temperatures, on several spherical Gaussians for which the true value of $\dG$ is known; except for the earliest iterations, the right value is found to within standard error for all the likelhoods.
+The Bayesian temperature has slightly larger error bounds, as a more cautious estimate, but otherwise the estimates are consistent.
+\par
+This provides a crucial piece of information, allowing us to access the Gaussian dimensionality of the true posterior before the run is complete.
+Before we discuss how this is can be used to aid endpoint estimation, we consider how the prior can affect the value of $\dG$ mid-run.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/d_G_gaussians.pdf}
+\end{center}
+\caption{Bayesian model dimensionalities estimated throughout a run for spherical Gaussians of varying dimensions, which were not previously accessible before the end of a run because the posterior samples have vanishing variance.
+Calculating the value from the transformed likelihood $\Like \to \Like^{\beta}$, however, recovers the dimensionality of the final posterior before it is reached, which is shown in the above plots.
+Shading represents the $1-2\sigma$ uncertainties on the estimates $\hat{d}_\mathrm{G}$ of the true posterior dimensionality.
+Only the microcanonical and Bayesian temperatures are shown for clarity of comparison.}
+\label{fig:d_G_gaussians}
+\end{figure*}
+
+\subsubsection*{Aside: Prior effects}\label{sec:prior_effects}
+We now highlight the manner in which the prior limits our knowledge of the dimensionality of the posterior during a nested sampling run, and thus our estimate of the endpoint.
+A simple, but illustrative example is an elongated two-dimensional Gaussian, with parameters $(\theta_1, \theta_2)$, $\bm{\mu} = (0, 0)$, $\Sigma = \mathrm{diag}(0.8, 2.5\times10^{-9})$, prior $\theta_1 \sim U(0, 2)$, $\theta_2\sim U(-10^{-3}, 10^{-3})$.
+After transformation to the unit hypercube prior, the covariance matrix becomes
+\begin{equation}
+    \Sigma = \mathrm{diag}(0.2, 0.01)
+\end{equation}
+Transformation to the hypercube makes it manifest that compared to the posterior, the prior is narrow in the $\theta_1$ direction, and broad in the $\theta_2$ direction.
+The width of the prior \textit{relative} to the posterior plays a crucial part in how the samples evolve over a nested sampling run, which we will now see.
+\par
+Initially, the live point cloud is spread across the whole prior (\cref{fig:d_G_elongated_2D}, panel A).
+When it is time to find a point with a higher log-likelihood, it is much more likely to be in the $\theta_1$ direction because it is less well-constrained than $\theta_2$.
+The samples only compress in the  $\theta_1$ direction, and thus in this phase the live and dead points lie in a single-dimensional subspace of the full two-dimensional posterior, with $\dG = 1$ (\cref{fig:d_G_elongated_2D}, panel B).
+Eventually, the live point cloud converges around $\theta_1$ sufficiently so that it is just as likely to find a higher $\log\Like$ in either parameter direction.
+Only at this point does the point cloud shrink in both directions, lying in the full two-dimensional space of the posterior, with $d_G = 2$ (\cref{fig:d_G_elongated_2D}, panel C).
+\par
+The key point is that in panel B, the live point cloud is actually one-dimensional.
+When comapred to a truly one-dimensional posterior (\cref{fig:d_G_elongated_1D}), at point B the samples look the same in all aspects.
+The consequence is that at point B, we fundamentally cannot know if there are more parameters `waiting' to be constrained, or if we have already found all the relevant directions.
+There are no clever inference tricks to get around it, in the same way that we cannot infer a `spike' from a slab and spike posterior when all the samples have yet to find the `spike'.
+Knowing the total number of parameters of the prior and likelihood to get the `parameter dimensionality' also isn't helpful, because the posterior dimensionality could be significantly different due to parameter degeneracies or non-Gaussian likelihoods.
+\par
+Therefore, unless the prior happens to be chosen to perfectly constrain each principal component of the posterior to the same degree, the prior choice will cause the nested samples to initially only occupy a subspace of the true posterior with a lower dimensionality, only eventually evolving to occupy the full space.
+\begin{figure*}
+\begin{center}
+    \subfloat[a][]{\includegraphics{figures/d_G_elongated_2D_simple.pdf}\label{fig:d_G_elongated_2D}}
+    \subfloat[b][]{\includegraphics{figures/d_G_elongated_1D_simple.pdf}\label{fig:d_G_elongated_1D}}
+\end{center}
+\caption{(a) Dimensionality estimates for a 2$d$ Gaussian that is elongated in one of its dimensions, with $1-2\sigma$ uncertainties shaded.
+The locations of the live points in the prior are shown at three stages, indicated by the connecting lines.
+As can be seen from the live point distribution, the samples only compress in a single direction until much later in the run, and early on it appears as if the other direction is completely unconstrained.
+(b) The same figure repeated for a 1$d$ Gaussian that is actually unconstrained in the other direction.
+At early iterations panel A and B, the live point distribution is identical to the $2d$ case, showing that early on the full posterior dimensionality is unknown and cannot be inferred.
+The Bayesian temperature was used here to transform the likelihood, but the other temperatures give similar results.}
+\label{fig:d_G_elongated}
+\end{figure*}
+\par
+If the widths of the posterior in different directions (in the unit hypercube) are well-separated, like the above case, we see step change in the Bayesian model dimensionality.
+If the length scales are close together, however, the change is more gradual.
+This can be seen when comparing the inferred dimensionality of two $6d$ Gaussians with similar $\DKL$, one with
+\begin{equation}\label{eq:elongated_separated}
+    \Sigma_\mathrm{separated} = \mathrm{diag}(10^{-3}, 10^{-3}, 10^{-3}, 10^{-6}, 10^{-6}, 10^{-6}),
+\end{equation}
+and the other with 
+\begin{equation}\label{eq:elongated_close}
+    \Sigma_\mathrm{close} = \mathrm{diag}(10^{-2}, 10^{-3}, 10^{-4}, 10^{-5}, 10^{-6}, 10^{-7}),
+\end{equation}
+shown in \cref{fig:d_G_elongated_comparison}.
+Given arbitrary prior choices in real nested sampling runs, the length scales of high-dimensional posteriors in the unit hypercube are not necessarily well-separated, so a typical chain will have a dimensionality profile similar to the right-hand side of \cref{fig:d_G_elongated_comparison}.
+\begin{figure}
+\begin{center}
+    \includegraphics{figures/d_G_elongated_comparison.pdf}
+\end{center}
+\caption{Estimated posterior dimensionality throughout a run for two 6$d$ covariant Gaussians.
+The left hand side has a steep separation of length scales between half of the parameters and the other half, while the right hand side has parameters that gradually shrink in length scale.
+These characteristics are reflected in the respective step-like and gradual evolutions of the sample dimensionalities.}
+\label{fig:d_G_elongated_comparison}
 \end{figure}
 \par
-One should keep in mind that despite the above theoretical reasoning, our introduction of the likelihood transformation was ultimately motivated by our wish to utilise the extra degree of freedom it provides.
-As we will see below, we recommend choosing the exact definition depending on what is useful for the problem at hand.
-
-\subsection{Dimensionality}\label{sec:dimensionality}
-We can immediately use the inferred temperature to track how the effective dimensionality of the posterior changes throughout the run, which was previously inaccessible.
-\citet{Handley_2019} demonstrated that at the end of a run, a measure of the number of constrained parameters is given by the Bayesian model dimensionality (BMD), defined as the posterior variance of the information content:
-\begin{equation}
-\frac{d_G}{2} = \int \mathcal{P}(\theta) \left(\log \frac{\mathcal{P}(\theta)}{\pi(\theta)} - \DKL\right)^2 \: \mathrm{d}\theta
-= \langle \mathcal{I}^2 \rangle_\mathcal{P} - \langle \mathcal{I} \rangle^2_\mathcal{P}.
-\end{equation}
-\par
-Calculating the quantity using intermediate set of weighted samples (which is concentrated at a single point) leads to vanishing variance, hence also dimensionality.
-However, we can recover the structure of the posterior together with the true dimensionality by adjusting the temperature.
-Dimensionality estimates are plotted in \cref{fig:d_G_spherical} for a spherical 32-d Gaussian, for which the true dimensionality is known.
-\par
-The different choices of temperature are again consistent, but for the rest of this paper we choose the Bayesian $\beta$, because it provides a better reflection of the uncertainty in the estimate; the others, while fluctuating around the true value, are often many standard errors away from the true value at each single point.
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/d_G_spherical.pdf}
-\end{center}
-\caption{Dimensionality estimates using the different temperatures for a spherical 32-d Gaussian.
-Again, all are consistent, but the Bayesian definition has a uncertainty which includes the true value far more consistently than for the other definitions.}
-\label{fig:d_G_spherical}
-\end{figure}
-
-\subsubsection*{Anisotropic compression}\label{sec:elongated}
-Plots of samples dimensionality against compression also draw attention to the \textit{directions} in which the samples are constrained throughout the run.
-As a concrete example, consider an elongated Gaussian in a unit hypercube prior with $\mu = \bm{0}$ and $\Sigma = \mathrm{diag}\left(10^{-3}, 10^{-3}, 10^{-3}, 10^{-6}, 10^{-6}, 10^{-6}\right)$, for which the dimensionality estimates are plotted in \cref{fig:d_G_elongated}.
-Alongside is a view of the distribution of live points across the prior for two directions with different scales, which shows the level to which those parameters have been constrained at different times.
-\par
-A feature of nested sampling made apparent here is that parameters with high variance are initially `hidden'.
-Compression occurs in the direction which is most likely to have a sample of higher likelihood, and initially it is much easier to find a better point along the direction of a parameter that is poorly constrained.
-Lower variance parameters are constrained much later, and before that happens it appears as though those parameters have a uniform distribution.
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/d_G_elongated.pdf}
-\end{center}
-\caption{Dimensionality estimates for a Gaussian that is elongated in half of its dimensions, with $1-2\sigma$ uncertainties shaded.
-The locations of the live points in the prior are shown at three stages, indicated by the connecting lines.
-As can be seen from the live point distribution, the prior does not compress in the higher variance direction until much later in the run, and early on it appears as if those directions are completely unconstrained.}
-\label{fig:d_G_elongated}
-\end{figure}
-\par
-It is important to appreciate that at lower compression the samples truly lie in a lower-dimensional space, rather than some artefact of the way we view them.
-Anticipating the full dimensionality of the space is therefore just as impossible as that associated with a slab-spike geometry, so in this sense such geometries contain a `compressive phase transition'.
+If we sensibly assume that the existence of more parameters to constrain leads to a longer runtime, the fact that we under-estimate the dimensionality also means that early on, any endpoint algorithm tends to under-estimate the endpoint.
+This is something we will have to live with.
 
@@
 \end{equation}
-The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n_i)$, The second term is the average time per likelihood evaluation.
-The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation and usually does not vary in orders of magnitude.
+The first term is the harmonic mean of the number of live points $\sim\mathcal{O}(n_i)$, which depends on the live point schedule chosen by the user.
+The second term is the average time per likelihood evaluation.
+The third is the average number of evaluations required to replace a dead point with a live point at higher likelihood, which is given by the implementation.
 The final term is the Kullback-Liebler divergence~\cref{eq:DKL}, the compression factor required to get from the prior to the posterior.
-This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principle interest in this section.
+This term is generally outside of user control, in most cases \textit{a priori} unknown, and of principal interest in this section.
+\par
+Our aim is therefore first to quantify the amount of compression required to reach the endpoint at approximately $\log X = -\DKL$, which provides an estimate of the how far the sampler still needs to penetrate into the posterior.
+Once the required compression $\log X_\mathrm{f}$ is known, multiplying by the harmonic mean of the number of live points through the run gives the iteration count.
+Converting the number of nested sampling iterations to a run-time requires knowledge of how the sampling efficiency changes throughout the run, which we will not attempt to do because the variation is highly implementation-specific.
+\par
+On the final point: the iteration progress is roughly proportional to the runtime progress when the sampling efficiency does not vary in order of magnitude.
+When that is not the case, the estimate will still be useful if the variation is known \textit{a priori}.
+For example, ellipsoidal samplers such as \textsc{MultiNest} and the MLFriends algorithm in \textsc{UltraNest} reduce in efficiency near the posterior bulk, so an impatient user who does not wish to expend the extra exponential effort for the final few percent can confidently halt earlier if they know they are already near the termination point.
 
@@
 For example, in the constant $\nlive$ case $\log X$ decreases by $1/n$ at each iteration, so the total number of iterations $N_\mathrm{f}$ would be
-\begin{equation}
-	N_\mathrm{f} = - n\log X_\mathrm{f} .
+\begin{equation}\label{eq:Nf}
+	N_\mathrm{f} = - n\log X_\mathrm{f}.
 \end{equation}
 
-\subsection{How to extrapolate?}
+\subsection{How to extrapolate?}\label{sec:how_to_extrapolate}
+The critical ingredient here is the function $\phi$ used to fit the likelihood profile $\Like(X)$, which should be simple enough to make fitting possible mid-run, but ideally also flexible enough to avoid strong biases.
+Previous work, such as the fitting procedure in the original Nelder-Mead simplex algorithm to estimate the Hessian \citep{nelder_mead}, and the BAMBI nested sampling paper \citep{Graff_2012}, suggest fast Gaussian approximations can be a good candidate for (online) modelling of likelihoods.
+The following method uses a similar approach, but unlike the previous schemes do not fit directly in parameter space.
+\par
 A key observation is that the Bayesian model dimensionality is the equivalent dimension of the posterior if it were actually Gaussian.
 Fitting a Gaussian of this dimension to the likelihood profile therefore makes a reasonable approximation to the true distribution, without explicitly assuming the form of the likelihood function.
+\par
 The parameterisation of the Gaussian that we fit is the same as that given in \cref{sec:logL}, which we shall repeat here for clarity;
@@
 \begin{enumerate}[leftmargin=*]
-    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior at the Bayesian temperature
-    \item Take the live point profile and perform a least squares fit to \eqref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
-    \item Use the likelihood predicted by these parameters to solve \eqref{endpoint} for $X_\mathrm{f}$
+    \item Find the current dimensionality ${d}^{(i)}_G$ of the posterior
+    \item Take the live point profile and perform a least squares fit to \cref{gaussian}, stipulating that $d = {d}^{(i)}_G$ to infer $\logLmax$ and $\sigma$ 
+    \item Use the likelihood predicted by these parameters to solve \cref{endpoint} for $X_\mathrm{f}$
 \end{enumerate}
@@
 \end{equation}
-and $N_\mathrm{f}$ is of course just $-n$ multiplied by this.
+and $N_\mathrm{f}$ is then computed with \cref{eq:Nf}.
 Intuitively, the above procedure can be thought of as inferring the number of constrained parameters, then extrapolating them up to find the point at which they will be fully constrained.
 \par
-Uncertainties in the final estimate are obtained by drawing many samples from the distribution of $d_\mathrm{G}$ defined by the Bayesian temperature, and repeating step two for each.
+Uncertainties in the final estimate are obtained by drawing many samples from the distribution of $\dG$ at the current iteration, and repeating step two for each.
 One might wonder why we do not obtain $d$ via least squares regression together with the other parameters; extensive testing has shown this approach to be far less stable.
+\par
+Thus far, we have not yet specified which temperature we should use to regularize $d_G^{(i)}$.
+We will now empirically assess the performance between the different temperatures across several test likelihoods.
+
+\subsection{Choosing the best temperature}\label{sec:choosing_the_best_temperature}
+We estimate the endpoint using the above procedure for three previously discussed test likelihoods, each in a unit hypercube prior:
+\begin{enumerate}[leftmargin=*]
+    \item Spherical Gaussian; $\sigma = 10^{-3}$
+        \begin{equation}
+        \log \Like = -\frac{|\bm{x}|^2}{2\sigma^2}, \quad \bm{x} \in \mathbb{R}^{32}
+        \end{equation}
+    \item Elongated Gaussian, two well-separated length scales 
+        \begin{equation}
+            \log\Like = -\frac{1}{2} \bm{x}^\intercal \Sigma^{-1}\bm{x}, \quad \bm{x} \in \mathbb{R}^6
+        \end{equation}
+        \begin{equation*}
+            \Sigma_\mathrm{separated} = \mathrm{diag}(10^{-3}, 10^{-3}, 10^{-3}, 10^{-6}, 10^{-6}, 10^{-6}),
+        \end{equation*}
+    \item Elongated Gaussian, gradually changing length scales
+        \begin{equation}
+            \Sigma_\mathrm{close} = \mathrm{diag}(10^{-2}, 10^{-3}, 10^{-4}, 10^{-5}, 10^{-6}, 10^{-7}).
+        \end{equation}
+\end{enumerate}
+Results are shown in \cref{fig:beta_endpoint_comparison}.
+Broadly speaking, the performance is similar, which is expected since the temperatures themselves are relatively consistent with each other (\cref{fig:beta_plots}).
+\par
+No estimates can be calculated for the termination temperature very early on in the run, because there is no choice of $\beta$ that will make the current iteration the termination point.
+For example, this is certainly true when $X_i > \epsilon$.
+\par
+We choose to use the Bayesian temperature in our implementation from this point onwards, because it incorporates the extra degree of uncertainty in the temperature itself, and this broader account of uncertainty is helpful when we are estimating endpoints using limited information.
+However, the choice is to some degree one of personal taste, and it remains reasonable to make estimates with the others.
+\begin{figure*}
+\subfloat{\includegraphics{figures/beta_endpoint_comparison_gauss_32.pdf}\label{gauss_32}}
+\newline
+\subfloat{\includegraphics{figures/beta_endpoint_comparison_correlated_6d.pdf}\label{elongated_gaussian_separated}}
+\newline
+\subfloat{\includegraphics{figures/beta_endpoint_comparison_correlated_6d_close.pdf}\label{elongated_gaussian_close}}
+\caption{Endpoint estimates obtained from the procedure outlined in \cref{sec:how_to_extrapolate}, using different $\beta$ to regularize the $\dG$.
+The true endpoints are shown in dotted red, while the estimates with their $1-2\sigma$ uncertainties are shaded in blue.
+Progress is measured in terms of the compression $-\log X$ along the x-axis, while the estimates on the y-axis are in terms of the predicted final compression $-\log \hat{X}_\mathrm{f}$.
+The estimates are broadly consistent, but notable differences include the failure of the termination temperature to obtain estimates at early iterations, and the larger uncertainty bounds of the Bayesian temperature.}
+\label{fig:beta_endpoint_comparison}
+\end{figure*}
 
@@
 Second, because of the point-like nature of the posterior mid-run, the remaining integral approximated as such holds nearly all of the evidence, so the relative fraction of the accumulated and remaining evidence is almost zero for most of the run.
-Finally, approximation \eqref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
+Finally, approximation \cref{eq:integral_progress} is always an underestimate, because as previously found the maximum live point is generally nowhere near the true maximum.
 This approach can be useful in low dimensions when the live points are always near the maximum, but in general is less reliable.
-\par
-We also note that for implementations whose sampling efficiency $f_\mathrm{sampler}$ varies, the progress indicator of \cref{sec:termination} does not directly correspond to time, but will still be useful if the variation is known \textit{a priori}.
-For example, ellipsoidal samplers such as \textsc{MultiNest} and the MLFriends algorithm in \textsc{UltraNest} reduce in efficiency near the posterior bulk, so an impatient user who does not wish to expend the extra exponential effort for the final few percent can confidently halt earlier if they know they are already near the termination point.
 
@@
 Seasoned watchers of nested sampling runs might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges.
-We do this for a spherical Gaussian and compare it to our method.
-At an intermediate stage of the run, the most recent outputs might look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate}.
-Extrapolating those data to a linear and exponential profile yields endpoint estimates plotted in the graph to the right.
+At an intermediate stage of the run, the most recent outputs look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate} (which in this case was a real run for a spherical Gaussian with $d = 16$, $\sigma = 0.01$).
+The right hand side of \cref{fig:inc_extrapolate} plots the gain in log-evidence at every 500 iterations, then extrapolates them with a linear and exponential profile.
+This is then used to calculate the point at which the gain is so small that the termination condition is met, shown separately for the linear and exponential extrapolations.
 \par
-The linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile.
+Compared to the true endpoint (found by completing the run), the linear extrapolation is clearly an underestimate, since it fails to account for the long tail of the nonlinear profile.
 The increments are also not exactly exponential, since the exponential fit leads to a large over-prediction.
-The predicted endpoint over the course of a run for $d = 16$, $\sigma = 0.01$, as shown in \cref{fig:inc_predictions}, shows the same result.
+The predicted endpoint over the course of a run, as shown in \cref{fig:inc_predictions}, shows the same result.
 One might expect an average to be more accurate, but this tends to be biased towards the exponential prediction, and there is no obvious choice of weighting that would fix this.
@@
 \begin{figure}
-\begin{adjustbox}{valign=t, scale=0.72}
+\begin{adjustbox}{valign=t, scale=0.7}
 \subfloat{\begin{tabular}{|c|c|c|}
@@
 \quad
-\begin{adjustbox}{valign=t}
+\begin{adjustbox}{valign=t, scale=0.9}
     \subfloat{\includegraphics{figures/inc_extrapolate.pdf}}
 \end{adjustbox}
-\caption{Extrapolating the increments of evidence. The left column shows the output of a nested sampling run, and the right column shows the extrapolation.}
+\caption{Extrapolating the increases in evidence $\Delta \log \mathcal{Z}$ to predict the endpoint.
+The left column shows the most recent ten outputs from a nested sampling run currently at $n_\mathrm{dead} = 10,000$, together with the difference every 500 iterations, which appear to converge at a steady rate.
+The right hand side plots these past 10 increments (blue) as a function of the compression $-\log X_\mathrm{f}$, and attempts to extrapolate them via a linear and exponential profile (pink).
+This is used to predict when the increment will be small enough for the termination condition to be met (dashed pink).
+The true progress of the increments, obtained after completing the run, is plotted in black.}
 \label{fig:inc_extrapolate}
@@
 \section{Results}\label{sec:results}
-We now test the approach established in the previous section on a range of distributions.
-We begin by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
-\subsection{Toy examples}
+We now test the approach established in the previous section on a range of distributions, beginning by considering a series of toy examples to explore the capabilities and limitations of the method, before presenting results for real cosmological chains.
+\subsection{Toy examples}\label{sec:logXfs_toy}
 Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\citep{anesthetic}.
-\subsubsection{Gaussians}
-Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:gauss_predictions} as a benchmark for when fitting a Gaussian distribution is exact.
+\subsubsection{Spherical Gaussians}
+Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:logXfs_gaussian} as a benchmark for when fitting a Gaussian distribution is exact.
 In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties.
@@
 \par
-We note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, which can be seen from the middle two plots.
+One can see from the $\log\Like-\log X$ plots in \cref{fig:logL_logX_plots} that the extrapolated likelihood is a good fit to the true likelihood, as expected from a spherical Gaussian.
+We also note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, as indicated by the middle two plots of \cref{fig:logXfs_gaussian}.
 \begin{figure*}
 \begin{center}
-    \includegraphics{figures/gauss_predictions.pdf}
+    \includegraphics{figures/logXfs_gaussian.pdf}
 \end{center}
 \caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left).
+The true endpoints are shown in dotted red, while the estimates with their $1-2\sigma$ uncertainties are shaded in blue.
+Progress is measured in terms of the compression $-\log X$ along the x-axis, while the estimates on the y-axis are in terms of the predicted final compression $-\log \hat{X}_\mathrm{f}$.
 The correct endpoint is obtained for all but the earliest iterations, and the uncertainty is controlled by the number of live points, which can be seen from the two $d = 16$ plots.}
-\label{fig:gauss_predictions}
+\label{fig:logXfs_gaussian}
 \end{figure*}
-We also observe the effect of elongating the Gaussian, using the same example as \cref{sec:elongated}.
-\cref{fig:elongated_logXfs} shows a step-like trend similar to the inferred dimensionalities, reflecting the fact that the full dimensionality is undetectable at lower compression factors.
-The endpoint for a likelihood whose remaining three directions are completely unconstrained coincides with our predictions at early iterations, showing that the two cases are indistinguishable.
-\begin{figure}
-\begin{center}
-    \includegraphics{figures/elongated_logXfs}
-\end{center}
-\caption{Endpoint prediction for an elongated Gaussian.
-At early stages, the full dimensionality is undetectable, and the endpoint is predicted to be the same as for a likelihood with three unconstrained directions.
-Only once the prior has been compressed enough to constrain the other three directions does the prediction converge to the true value.}
-\label{fig:elongated_logXfs}
-\end{figure}
 
-\subsubsection{Cauchy}
-One case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian.
-\cref{fig:cauchy_predictions} shows the predictions for a likelihood of the form
+\subsubsection{Correlated Gaussians}\label{sec:logXfs_correlated}
+We also observe the effect of introducing different length scales and correlations to the Gaussian, generically 
 \begin{equation}
-	\log\mathcal{L} = \log\mathcal{L}_\mathrm{max} - \frac{1 + d}{2} \log \left(1 + \frac{X^2}{\gamma^2}\right),
+    \log\Like = - \frac{1}{2} \bm{x}^{\intercal} \Sigma^{-1} \bm{x}, \quad \bm{x} \in \mathbb{R}^{d}.
 \end{equation}
-choosing $d = 10$ and allowing $\gamma$ to vary.
-The correct estimate is obtained to within standard error by about halfway, but before that is inaccurate.
-The key limitation is that the estimate is wrong early on, not because the compression is anisotropic, or because there is a phase transition; but rather as a limitation of the reducing the likelihood to a Gaussian via the BMD, which is itself less stable for a Cauchy.
+Again, all are run with a constant number of 500 live points.
+Starting from the left of \cref{fig:logXfs_correlated}, the first chain has the covariance matrix given in \cref{eq:elongated_separated}, featuring three parameters with $\sigma^2 = 10^{-3}$ and three with $\sigma^2=10^{-6}$.
+The endpoint predictions are step-like; as discussed in \cref{sec:prior_effects}, it only jumps to the correct endpoint when the prior is compressed enough to sample the $\sigma^2=10^{-3}$ parameters, which are initially better constrained by the prior.
+The second run from the left has the same $6d$ covariance matrix, except the three $\sigma^2=10^{-3}$ parameters are made completely unconstrained, so that the posterior only has three constrained parameters.
+The corresponding endpoint matches that obtained at early stages in the full $6d$ posterior, meaning that at those early stages the two cases are impossible to distinguish, and underprediction cannot be avoided.
 \par
-Nevertheless, the right order of magnitude is obtained at all times, so this remains sufficient for most use-cases.
-The Cauchy is also a pathological case, and the same problem does not in practice appear for more realistic cases, as we shall see next.
+Equivalently, one can consider the inaccuracy as a breakdown of the approximation that takes the Gaussian likelihood \cref{eq:logL_x} to the fitting function \cref{eq:gaussian_logL}, which only holds when the prior is wide compared to the posterior.
+Since the live points resemble the prior in the beginning of the run, and some parameters are constrained before others, the approximation may only hold for some of the parameters, leading to the $\log\Like-\log X$ curve having a lower dimensionality at the start.
+\par
+The third chain from the left ($\Sigma$ given in \cref{eq:elongated_close}) then shows the effect of `smoothing' the parameter variances, so that they step down from $10^{-2}$ to $10^{-7}$ one power of ten at a time.
+Mirroring the behaviour shown in \cref{fig:d_G_elongated_comparison}, the predictions gradually converge to the truth as those parameters with narrower priors are discovered in gradual succession, instead of making a sudden jump.
+Finally, the fourth chain from the left rotates the covariance matrix in \cref{eq:elongated_separated} by $30\degree$, demonstrating that the above behaviour persists when the `principal components' of the posterior are no longer parallel to the parameter axes.
+\par
+All of these chains nevertheless surmise the correct endpoint by about halfway, and always give the right order of magnitude.
+The $\log\Like-\log X$ plot at 30\% of the way through for the elongated Gaussian (second from the left in \cref{fig:logL_logX_plots}) shows that the fitted profiles are of the same shape as the true likelihood, since both are Gaussian.
+However, the extrapolation undershoots because as discussed, prior effects initially cause the sample dimensionality to be lower than the true dimensionality of the posterior.
 \begin{figure*}
 \begin{center}
-    \includegraphics{figures/cauchy_predictions.pdf}
+    \includegraphics{figures/logXfs_correlated.pdf}
 \end{center}
-\caption{Predictions for the Cauchy distribution for various widths $\gamma$ and $\nlive = 500$.
-The endpoint is underestimated for the first half of the run, but this is a limitation of the Gaussian approximation rather than a lack of information mid-run.}
-
-\label{fig:cauchy_predictions}
+\caption{Endpoint predictions for four $6d$ Gaussians with different covariance matrices.
+From the left, the first chain has parameters with two distinct length scales, so the predictions show a step-like trend due to prior effects.
+The second chain has three unconstrained parameters in place of three parameters in the first chain, so the predictions are identical to the early stages of the first chain.
+The third plot shows that if the parameters have a gradual spectrum of length scales, the prior effects fade away more gradually, reflected in the gradual convergence of the estimates.
+The fourth and final plot shows that the prior effects, most evident in the chains with two distinct length scales, persist when the posterior components are not parallel to the parameter axes.}
+\label{fig:logXfs_correlated}
 \end{figure*}
 
-\subsection{Cosmological examples}
-Finally, we evaluate the method on real cosmological chains.
-\cref{fig:lcdm_logXfs} presents the endpoints (calculated after the fact) for nested sampling runs for curvature quantification on several common cosmological data sets (details in \citealt{curvature_tension}).
+\subsubsection{Non-Gaussian likelihoods}
+We now look at how well the extrapolation procedure works when the likelihoods are significantly non-Gaussian, since one might expect fitting a Gaussian to fail. 
+First, we consider the 10-dimensional Rosenbrock function with $a = 1$,  $b = 100$:
+\begin{equation}
+    \log \Like = - \sum_{i = 1}^{9} \left(a - x_i\right)^2 + b\left(x_{i+1} - x_i^2\right)^2, \quad \bm{x} \in \mathbb{R}^{10},
+\end{equation}
+which is important because it has strong correlations between the parameters which do not factorize, such that the concept of dimensionality is less clear.
+Another case that might be expected to cause problems is the pathological Cauchy distribution, which is far from a Gaussian:
+\begin{equation}
+    \log\Like = - \frac{1+d}{2} \log \left( 1 + \frac{|\bm{x}|^2}{\gamma^2}\right), \quad \bm{x} \in \mathbb{R}^8
+\end{equation}
+which we test with $\gamma = 10^{-4}$ and $d = 8$.
+Results are shown in \cref{fig:logXfs_non_gaussian}.
+The Rosenbrock initially underestimates the endpoint, getting the right answer with around 5,000 iterations of 15,529 remaining, though the prediction at worst hovers at around 60-70\% of the actual run time.
+Prior effects contribute to the under-estimation in the same manner as all of the previous non-spherical likelihoods, but \cref{fig:logL_logX_plots} reveals that the relatively poor fit (but still reasonable) of a Gaussian to the Rosenbrock function also plays a role.
 \par
-The \textsc{SH0ES}, \textsc{BAO} and \textsc{lensing} chains are `easy' low $\DKL$ inferences, so it is expected that the correct endpoint is inferred practically from the start.
-The \textsc{Planck} endpoints, on the other hand, are not correct until at least midway through.
-However, this is expected from the covariance of the \textsc{Planck} likelihood, which consists of principal components of many scales and therefore elongated in many dimensions.
-It is therefore of the same class as the elongated Gaussian presented in \cref{sec:elongated}; the samples exist in a lower dimensional subspace mid-run, which slowly increases to the full dimensionality only at the end of the run.
+The Cauchy also under-predicts at first, obtaining the correct endpoint within standard error by about halfway through its 39,236 iterations.
+The posterior here is spherically symmetric, so compression is isotropic and the prior affects all parameters equally.
+Because there are no associated under-estimates of the posterior dimensionality, the inaccuracies are solely due to the poor fit of a Gaussian to a Cauchy, which can be seen in the rightmost plot of \cref{fig:logL_logX_plots} as much more drastic than the other likelihoods; the Bayesian model dimensionality is itself less stable for Cauchy likelihoods.
+Nevertheless, the right order of magnitude is obtained at all times, so this remains sufficient for most use-cases.
+The Cauchy is also a pathological case, and the same problem does not in practice appear for more realistic examples, as we shall see in the next section.
 \begin{figure*}
 \begin{center}
-    \includegraphics{figures/lcdm_logXfs.pdf}
+    \includegraphics{figures/logXfs_non_gaussian.pdf}
+\end{center}
+\caption{Endpoint predictions throughout for two non-Gaussian likelihoods, a Rosenbrock and a Cauchy, with the red dotted line showing the true endpoint and the blue shading indicating the $1-2\sigma$ uncertainties.
+The endpoint is initially underestimated for the Rosenbrock, due to a combination of prior effects and the breaking down of the Gaussian model.
+The Cauchy is also underestimated for the first half of the run, but this is solely a limitation of the Gaussian approximation because the likelihood is spherically symmetric.}
+\label{fig:logXfs_non_gaussian}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics{figures/logL_logX_plots.pdf}
+\caption{Fitted extrapolations compared to the true likelihood for four different test likelihoods.
+The live points at 30\% through the run for each example are fitted a to Gaussian, extrapolated to the higher likelihoods that have not yet explored, and compared to the ground truth.
+It is clear that truly Gaussian likelihoods are modelled well, while the model deteriorates with increasing non-Gaussianity (seen in the Rosenbrock and Cauchy likelihoods).}
+\label{fig:logL_logX_plots}
+\end{figure*}
+
+\subsection{Cosmological examples}\label{sec:cosmological_examples}
+Finally, we evaluate the method on real cosmological chains.
+\cref{fig:logXfs_lcdm} presents endpoint estimates (calculated after the fact) for nested sampling runs on curvature quantification for several common cosmological data sets (details in \citealt{curvature_tension}).
+\par
+The \textsc{SH0ES}, \textsc{BAO} and \textsc{lensing} chains are `easy' low $\DKL$ inferences, so it is somewhat expected that the correct endpoint is inferred practically from the start.
+The \textsc{Planck} endpoints, on the other hand, are not correct until at least midway through.
+However, this is expected from the fact that the \textsc{Planck} likelihood is elongated in many dimensions with a gradual spectrum of length scales.
+It is therefore subject to the same prior effects as the elongated Gaussians in \cref{sec:logXfs_toy}; the samples exist in a lower dimensional subspace mid-run, which slowly increases to the full dimensionality only at the end of the run.
+\begin{figure*}
+\begin{center}
+    \includegraphics{figures/logXfs_lcdm.pdf}
 \end{center}
 \caption{Endpoint predictions for cosmological likelihoods.
-The first three low $\DKL$ inferences get the correct endpoint from the start, while the \textsc{Planck} chain takes longer to converge because the likelihood is highly covariant and thus subject to anisotropic compression, which makes the samples lie in a lower dimensional subspace mid-run.}
-\label{fig:lcdm_logXfs}
+The first three low $\DKL$ inferences are close to the correct endpoint from the start, while the \textsc{Planck} chain takes longer to converge because the likelihood is elongated in many dimensions, and thus subject to prior effects which make the samples lie in a lower dimensional subspace mid-run.}
+\label{fig:logXfs_lcdm}
 \end{figure*}
@@
 We have derived new analytic results to make an anatomy of nested sampling, understanding the progression of a run via the compression of prior volume, the increase in log-likelihood, the inferred temperature schedule, and the convergence of the sample dimensionality.
-From these analyses, we developed a method for predicting the endpoint of a nested sampling run, by using the inferred Bayesian model dimensionality mid-run to extrapolate the known likelihood profile.
+From these analyses, we developed a method for predicting the endpoint of a nested sampling run by using the inferred Bayesian model dimensionality mid-run to extrapolate the known likelihood profile.
 \par
@@
 Consistent predictions are obtained for both toy and cosmological examples.
-The accuracy is typically limited by the information available mid-run, either because of a phase transition or because the anisotropy of the nested sampling compression.
+The accuracy is typically limited by the information available mid-run, due to different prior widths relative to the posterior.
 Pathological distributions, such as a Cauchy, lead to less stable inferences of the dimensionality and expose the limitations of a Gaussian approximation, though the order of magnitude is still correct.
 \par
-Further work can be done to experiment with more flexible basis functions for regression of the likelihood profile, so that it is less dependent on the Gaussian approximation.
+Further work can be done to generalise the results derived in \cref{sec:anatomy} to non-Gaussians, as well as experiment with more flexible basis functions for regression of the likelihood profile so that it is less dependent on the Gaussian approximation.
 

### 2024-07-17 review_response/first_submission.tex
moderate revisions for journal submission

--- 2023-12-28 paper/paper.tex
@@
 log(Z)     =  -12.55 +/- 0.27
-\end{Verbatim}
+end{Verbatim}
 \caption{Output from \textsc{PolyChord} for a typical nested sampling run. The predicted endpoint, shown in red, is calculated using the method described in this paper.}

### 2024-07-17 paper/figures/beta_comparison.py

@@ -1,30 +1,0 @@
-import matplotlib.pyplot as plt
-from util import read_from_txt, figsettings, plot_std
-figsettings()
-
-fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
-
-iterations, logbetas_grad, logbetas_logL, logbetas_mean, logbetas_std = read_from_txt('data/logbetas/gauss_32.txt')
-logXs = read_from_txt('data/logXs/gauss_32.txt')[0][iterations.astype(int)]
-plot_std(-logXs, logbetas_mean, logbetas_std, label=f'Bayesian', ax=axs[0], color='red')
-axs[0].plot(-logXs, logbetas_logL, color='orange', label='Canonical')
-axs[0].plot(-logXs, logbetas_grad, color='black', lw=2, label='Microcanonical')
-
-iterations, logbetas_grad, logbetas_logL, logbetas_mean, logbetas_std = read_from_txt(f'data/logbetas/slab_spike.txt')
-logXs = read_from_txt('data/logXs/slab_spike.txt')[0][iterations.astype(int)]
-plot_std(-logXs, logbetas_mean, logbetas_std, label=f'Bayesian', ax=axs[1], color='red')
-axs[1].plot(-logXs, logbetas_logL, color='orange', label='Canonical')
-axs[1].plot(-logXs, logbetas_grad, color='black', lw=2, label='Microcanonical')
-
-for ax in axs:
-    ax.set_xlabel(r"$-\log X$")
-
-axs[0].set_ylabel(r'$\log \beta$')    
-axs[1].set_ylim(-8, 8)
-
-handles, labels = axs[0].get_legend_handles_labels()
-axs[0].legend(handles[::-1], labels[::-1], ncol=3, bbox_to_anchor=(2.2, -0.4), fontsize=7)
-axs[0].set_title('Spherical Gaussian', fontsize=7)
-axs[1].set_title('Slab-spike Gaussian', fontsize=7)
-fig.suptitle('Temperature vs compression for different distributions', y=1.15, fontsize=8)
-fig.savefig('beta_comparison.pdf', bbox_inches='tight', pad_inches=0)

### 2024-07-17 paper/figures/beta_endpoint_comparison.py

@@ -1,0 +1,44 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+
+num_chain = 1
+ylim = (0, 1.5)
+
+chains = ['gauss_32', 'correlated_6d', 'correlated_6d_close']
+chains_titles = ['Gaussian', 'Elongated Gaussian, two well-separated length scales', 'Elongated Gaussian, gradually changing length scales']
+chain, chain_title = chains[num_chain], chains_titles[num_chain]
+methods = ['term_3', 'grad', 'logL', 'post']
+beta_types = ['Termination, $\\epsilon=10^{-3}$', 'Microcanonical', 'Canonical', 'Bayesian']
+
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.8))
+name, samples = get_samples(chain)
+model = EndModel(samples)
+endpoint_true = model.true_endpoint()
+for i, ax in enumerate(axs):
+    logXf_true = samples.logX().iloc[endpoint_true]
+    iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/{methods[i]}/{name}.txt')
+    logXfs = np.array(logXfs)
+    logXs = samples.logX().iloc[iterations]
+    plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=ylim, color='deepskyblue', ax=ax)
+    ax.set_xlim(0, -logXs.iloc[-1] * 1.1)
+    if num_chain == 0:
+        ax.set_title(beta_types[i], fontsize=6)
+
+for j in range(1, 4):
+    axs[j].set_yticks([])
+
+fig.supxlabel('$-\\log X$', y=0.05)
+fig.supylabel('$-\\log \\hat{X}_f$')
+fig.suptitle(f'{chain_title} ({endpoint_true:,} total iterations)', y=0.95)
+fig.tight_layout()
+fig.savefig(f'{current_dir}/beta_endpoint_comparison_{chain}.pdf', bbox_inches='tight')

### 2024-07-17 paper/figures/beta_plots.py

@@ -1,0 +1,60 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+
+fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
+
+for chain, ax in zip(['gauss_8', 'slab_spike'], axs):
+    name, samples = get_samples(chain)
+    model = EndModel(samples)
+    endpoint_true = model.true_endpoint()
+    logXf_true = samples.logX().iloc[endpoint_true]
+    samples['beta_logL'] = get_betas_logL(samples)
+
+    Npoints = 25
+    iterations = make_iterations(endpoint_true, Npoints)
+    logbetas_term = np.zeros(Npoints)
+    logbetas_term_half = np.zeros(Npoints)
+    logbetas_grad = np.zeros(Npoints)
+    logbetas_post_mean = np.zeros(Npoints)
+    logbetas_post_std = np.zeros(Npoints)
+    logbetas_logL = np.log(samples.beta_logL.iloc[iterations])
+
+    for i, ndead in enumerate(iterations):
+        points = points_at_iteration(samples, ndead)
+        logbetas_term[i] = np.log(get_beta_end(points, ndead, epsilon=1e-3))
+        logbetas_term_half[i] = np.log(get_beta_end(points, ndead, epsilon=0.5))
+        logbetas_grad[i] = get_logbeta_grad(points, ndead)
+        logbetas_post_mean[i], logbetas_post_std[i] = get_logbeta_post(points, ndead)
+        print('\r', f'Iteration {ndead}/{iterations[-1]}', end='')
+        
+    logXs = samples.logX().iloc[iterations]
+    ax.plot(-logXs, logbetas_term, color='C0', label='Termination ($\\epsilon=10^{{-3}}$)')
+    ax.plot(-logXs, logbetas_term_half, ls='--', color='C0', label='Termination ($\\epsilon=0.5$)', zorder=100)
+    ax.plot(-logXs, logbetas_grad, color='black', label='Microcanonical')
+    ax.plot(-logXs, logbetas_logL, color='orange', label='Canonical')
[…]

### 2024-07-17 paper/figures/beta_posteriors.py

@@ -1,0 +1,62 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+name, samples = get_samples('gauss_32')
+samples['betas_logL'] = get_betas_logL(samples)
+
+ndead = 15000
+points = points_at_iteration(samples, ndead)
+logbeta_term = np.log(get_beta_end(points, ndead))
+logbeta_term_half = np.log(get_beta_end(points, ndead, epsilon=0.5))
+logbeta_grad = get_logbeta_grad(points, ndead, interval=250)
+logbeta_logL = np.log(samples['betas_logL'].iloc[ndead])
+logbeta_post = get_logbeta_post(points, ndead)
+logbetas = [1, logbeta_term, logbeta_grad, logbeta_logL, logbeta_post]
+beta_titles = ['$\\beta=1$', 'Termination $\\beta$', 'Microcanonical $\\beta$', 'Canonical $\\beta$', 'Bayesian $\\beta$']
+
+import matplotlib.gridspec as gridspec
+fig = plt.figure(figsize=(7, 3))
+gs = gridspec.GridSpec(2, 6)
+ax0 = fig.add_subplot(gs[0, :2])
+ax1 = fig.add_subplot(gs[0, 2:4])
+ax2 = fig.add_subplot(gs[0, 4:])
+ax3 = fig.add_subplot(gs[1, 1:3])
+ax4 = fig.add_subplot(gs[1, 3:5])
+axs = [ax0, ax1, ax2, ax3, ax4]
+
+for i, ax in enumerate(axs):
+    if i == 4:
+        for j in range(25):
+            logbeta_post_samp = np.random.normal(*logbetas[i])
+            logLX = np.exp(logbeta_post_samp) * points.logL + points.logX()
+            ax.plot(-points.logX(10), np.exp(logLX - logLX.max()), lw=.5, color='coral', alpha=.05)
+    else:
+        logLX = np.exp(logbetas[i]) * points.logL + points.logX()
+        ax.plot(-points.logX(25), np.exp(logLX - logLX.max()), lw=.5, color='coral', alpha=.25)
+    ax.set_xlim(0, 40)
[…]

### 2024-07-17 paper/figures/cauchy_predictions.py

@@ -1,24 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from util import read_from_txt, figsettings, plot_quantiles, endpoints, DKLs
-figsettings()
-
-fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
-for i, exp in enumerate([2, 3, 4, 5]):
-    ax = axs[i]
-    name = f'cauchy_8{exp}'
-    true_endpoint = endpoints[name]
-    iterations, *logXfs = read_from_txt(f'data/logXfs/{name}.txt')
-    logXfs = np.array(logXfs)
-    logX_all = read_from_txt(f'data/logXs/{name}.txt')[0]
-    logXs = logX_all[iterations.astype(int)]
-    logXf_true = logX_all[true_endpoint]
-    plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 2), color='deepskyblue', ax=ax)
-    ax.set_title(f'$\\gamma=10^{{{-exp}}}$, $\\mathcal{{D}}_\\mathrm{{KL}} = {DKLs[name]:.1f}$', fontsize=8)
-    ax.set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
-    ax.margins(x=0)
-
-fig.supxlabel('$-\\log X$', y=0.02)
-fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.01)
-fig.tight_layout()
-fig.savefig('cauchy_predictions.pdf', pad_inches=0, bbox_inches='tight')

### 2024-07-17 paper/figures/d_G_elongated.py

@@ -1,56 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from matplotlib.gridspec import GridSpec
-from aeons.utils import points_at_iteration
-from util import plot_quantiles, read_from_txt, figsettings
-figsettings()
-
-iterations, *d_Gs = read_from_txt(f'data/d_Gs/correlated_6d.txt')
-logX_all = read_from_txt(f'data/logXs/correlated_6d.txt')[0]
-logXs = logX_all[iterations.astype(int)]
-d_Gs = np.array(d_Gs)
-
-fig = plt.figure(figsize=(3.5, 3))
-gs = GridSpec(2, 3, wspace=0.1, hspace=0.1)
-ax1 = fig.add_subplot(gs[0, 0], zorder=1)
-ax2 = fig.add_subplot(gs[0, 1], zorder=1)
-ax3 = fig.add_subplot(gs[0, 2], zorder=1)
-ax4 = fig.add_subplot(gs[1, :], zorder=0)
-axs = [ax1, ax2, ax3, ax4]
-
-logXstars = [-0.5, -8, -30]
-# Find index of logX closest to logXstar
-ndeads = [np.argmin(np.abs(logX_all - logXstar)) for logXstar in logXstars]
-logL, logLbirth = read_from_txt(f'data/logLs/correlated_6d.txt')
-
-for i, ndead in enumerate(ndeads):
-    ax = axs[i]
-    live_points = points_at_iteration(samples, ndead).iloc[ndead:]
-    live_points.drop_weights().plot.scatter(0, 4, ax=ax, s=5, color='lightseagreen')
-    ax.set_aspect('equal')
-    ax.set_xlim(0, 1)
-    ax.set_ylim(0, 1)
-    ax.set_xticks([])
-    ax.set_yticks([])
-    ax.set_xlabel('')
-    ax.set_ylabel('')
-    # ax4.axvline(-logXstars[i], lw=1)
-    
-ax2.set_xlabel(r"$\sigma = 10^{-3}$", fontsize=8)
-ax2.xaxis.set_label_position('top') 
-ax1.set_ylabel(r"$\sigma = 10^{-6}$", fontsize=8)
-plot_quantiles(-logXs, d_Gs, 6, (0, 1.2), color='lightseagreen', ax=ax4)
-ax4.set_xlabel(r'$-\log X$', fontsize=8)
-ax4.set_ylabel(r'$\hat{d}_G$', fontsize=8, rotation=0, labelpad=10)
[…]

### 2024-07-17 paper/figures/d_G_elongated_comparison.py

@@ -1,0 +1,27 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+chains = ['correlated_6d', 'correlated_6d_close']
+titles = ['$\\Sigma_\\mathrm{{separated}}$', '$\\Sigma_\\mathrm{{close}}$']
+fig, axs = plt.subplots(1, 2, figsize=(3.3, 1.5))
+
+for chain, title, ax in zip(chains, titles, axs):
+    name, samples = get_samples(chain, reduced=False)
+    iterations, *d_Gs = read_from_txt(f'{data_dir}/d_Gs/post/{name}.txt')
+    logXs = samples.logX().iloc[iterations]
+    plot_quantiles(-logXs, d_Gs, 6, (0, 1.3), ax=ax, color='lightseagreen')
+    ax.set_title(title, fontsize=8)
+    ax.set_xlabel('$-\\log X$')
+    ax.set_yticks([0, 3, 6], [0, 3, 6])
+
+axs[0].set_ylabel('$d_\\mathrm{G}$')
+fig.tight_layout()
+fig.savefig(f'{current_dir}/d_G_elongated_comparison.pdf', bbox_inches='tight')

### 2024-07-17 paper/figures/d_G_elongated_simple.py

@@ -1,0 +1,81 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+chain = 'elongated_1D_simple'
+name, samples = get_samples(chain, reduced=False)
+iterations, *d_Gs = read_from_txt(f'{data_dir}/d_Gs/post/{name}.txt')
+logXs = samples.logX().iloc[iterations]
+if chain == 'elongated_2D_simple':
+    dims = [0.5, 1.02, 2.05]
+    d_G_true = 2
+    suptitle = '2D posterior'
+    ylabel = '$\\hat{d}_G$'
+else:
+    dims = [0.5, 1, 1]
+    d_G_true = 1
+    suptitle = '1D posterior'
+    ylabel = ''
+
+
+import matplotlib.gridspec as gridspec
+from anesthetic import make_2d_axes
+
+fig = plt.figure(figsize=(3.5, 3))
+gs = gridspec.GridSpec(2, 3, height_ratios=[1, 1])
+fig, ax0 = make_2d_axes([0, 1], diagonal=False, upper=False, subplot_spec=gs[0, 0], fig=fig, zorder=1)
+fig, ax1 = make_2d_axes([0, 1], diagonal=False, upper=False, subplot_spec=gs[0, 1], fig=fig, zorder=1)
+fig, ax2 = make_2d_axes([0, 1], diagonal=False, upper=False, subplot_spec=gs[0, 2], fig=fig, zorder=1)
+
+for ax in [ax0, ax1, ax2]:
+    samples.plot_2d(ax)
+    ax = ax.iloc[0, 0]
+    ax.set_aspect('equal')
+    ax.set_xticks([])
+    ax.set_yticks([])
+    ax.set_xlabel('')
+    ax.set_ylabel('')
+    ax.set_xlim(-0.5, 0.5)
+    ax.set_ylim(-0.5, 0.5)
+
+ax0.iloc[0, 0].set_ylabel('$\\theta_1$', fontsize=8, rotation=0, labelpad=10)
+ax1.iloc[0, 0].set_title('$\\theta_2$', fontsize=8)
+ax0.iloc[0, 0].text(-0.4, 0.3, 'A')
[…]

### 2024-07-17 paper/figures/d_G_gaussians.py

@@ -1,0 +1,42 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.5))
+chains = ['gauss_4', 'gauss_8', 'gauss_16', 'gauss_32']
+titles = ['4', '8', '16', '32']
+
+for ax, chain, title in zip(axs, chains, titles):
+    name, samples = get_samples(chain)
+    model = EndModel(samples)
+    true_endpoint = model.true_endpoint()
+    logXf_true = samples.logX().iloc[true_endpoint]
+
+    iterations, *d_Gs = read_from_txt(f'{data_dir}/d_Gs/post/{chain}.txt')
+    logXs = samples.logX().iloc[iterations]
+    plot_quantiles(-logXs, d_Gs, int(title), ylim=(0, 1.5), ax=ax, color='red')
+    
+    iterations, *d_Gs = read_from_txt(f'{data_dir}/d_Gs/grad/{chain}.txt')
+    logXs = samples.logX().iloc[iterations]
+    plot_quantiles(-logXs, d_Gs, int(title), ylim=(0, 1.5), ax=ax, color='black')
+    
+    ax.set_title(f'$d = {title}$', fontsize=8)
+    ax.set_yticks([0, int(title)], [0, int(title)])
+
+from matplotlib.patches import Patch
+custom_patches = [Patch(color='black', label='Microcanonical'), Patch(color='red', label='Bayesian')]
+fig.legend(handles=custom_patches, bbox_to_anchor=(0.95, 0.19), ncol=2, fontsize=8)
+
+fig.supxlabel(r'$-\log X$', y=0.05)
+fig.supylabel(r'$\hat{d}_\mathrm{G}$', x=0.02, rotation=0)
+fig.tight_layout()
+
+fig.savefig(f'{current_dir}/d_G_gaussians.pdf', pad_inches=0, bbox_inches='tight')

### 2024-07-17 paper/figures/d_G_spherical.py

@@ -1,28 +1,0 @@
-import numpy as np
-import matplotlib.pyplot as plt
-from aeons.utils import *
-from aeons.regress import *
-from aeons.endpoint import *
-from aeons.plotting import *
-from aeons.beta import *
-figsettings()
-
-name, samples = get_samples('gauss_32')
-iterations_grad, *d_Gs_grad = read_from_txt(f'{data_dir}/d_Gs/grad/{name}.txt')
-iterations_logL, *d_Gs_logL = read_from_txt(f'{data_dir}/d_Gs/logL/{name}.txt')
-iterations_post, *d_Gs_post = read_from_txt(f'{data_dir}/d_Gs/post/{name}.txt')
-d_Gs_grad, d_Gs_logL, d_Gs_post = np.array(d_Gs_grad), np.array(d_Gs_logL), np.array(d_Gs_post)
-logXs_grad = samples.logX().iloc[iterations_grad]
-logXs_logL = samples.logX().iloc[iterations_logL]
-logXs_post = samples.logX().iloc[iterations_post]
-
-fig, ax = plt.subplots(figsize=(3.5, 1.3))
-plot_quantiles(-logXs_post, d_Gs_post, samples.d_G(), ylim=(0, 1.5), ax=ax, color='red', label='Bayesian')
-plot_quantiles(-logXs_logL, d_Gs_logL, samples.d_G(), ylim=(0, 1.5), ax=ax, color='orange', label='Canonical')
-plot_quantiles(-logXs_grad, d_Gs_grad, samples.d_G(), ylim=(0, 1.5), ax=ax, color='black', label='Microcanonical')
-handles, labels = ax.get_legend_handles_labels()
-ax.legend(handles[::-1], labels[::-1], ncol=3, bbox_to_anchor=(1.05, -0.4))
-ax.set_xlabel(r'$-\log X$')
-ax.set_ylabel(r'$d_G$', rotation=0, labelpad=10)
-ax.set_title('Intermediate dimension for spherical 32-d Gaussian')
-fig.savefig('d_G_spherical.pdf', bbox_inches='tight', pad_inches=0)

### 2024-07-17 paper/figures/elongated_logXfs.py

@@ -4,6 +4,8 @@ from aeons.utils import *
 from aeons.plotting import *
 figsettings()
 
+import os
+dir_path = os.path.dirname( os.path.realpath( __file__ ) )
 
 name, samples = get_samples('correlated_6d')
 iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
@@ -16,11 +18,15 @@ logXf_3d = get_samples('correlated_3d')[1].logX().iloc[endpoints['correlated_3d'
 fig, ax = plt.subplots(figsize=(3, 1.6))
 plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 1.2), color='deepskyblue', ax=ax)
 
-ax.axhline(-logXf_3d, color='orange', linestyle='--', linewidth=.5)
-ax.set_title('Elongated Gaussian', fontsize=8)
-ax.set_yticks([0, -logXf_3d, -logXf_true], [0, r"$-\log X_\mathrm{f}^\mathrm{3d}$", r"$-\log X_\mathrm{f}^\mathrm{true}$"], rotation=0)
-ax.set_xlabel('$-\\log X$')
-ax.margins(x=0)
+ax.axhline(-logXf_true, color='red', linestyle='--', linewidth=.5, label='$d=6$ endpoint (truth)')
+ax.axhline(-logXf_3d, color='orange', linestyle='--', linewidth=.5, label='$d=3$ endpoint')
 
-fig.savefig('elongated_logXfs.pdf', pad_inches=0, bbox_inches='tight')
+ax.set_yticks([0, -logXf_3d, -logXf_true], [0, f"{-logXf_3d:.1f}", f"{-logXf_true:.1f}"], rotation=0)
+ax.set_title('Elongated Gaussian', fontsize=8)
+ax.set_xlabel('$-\\log X$')
+ax.set_ylabel('$-\\log \\hat{X}_\\mathrm{f}$')
+ax.margins(x=0)
+ax.legend()
+
+fig.savefig(dir_path + '/' + 'elongated_logXfs.pdf', pad_inches=0, bbox_inches='tight')
 

### 2024-07-17 paper/figures/last_live_point.py

@@ -1,46 +1,0 @@
-import matplotlib.pyplot as plt
-from aeons.utils import *
-from aeons.beta import *
-import matplotlib.patches as patches
-figsettings()
-
-name, samples = get_samples('planck_gaussian')
-ndead = 8000
-points = points_at_iteration(samples, ndead)
-beta_grad = np.exp(get_logbeta_grad(points, ndead))
-betas = [0, beta_grad, 1]
-labels = ['$\\beta=1$', '$\\beta = \\beta_i$', '$\\beta = 0$']
-colors = ['darkgray', 'orange', 'black']
-
-fig, axs = plt.subplots(1, 2, figsize=(3.5, 1.3))
-for i, beta in enumerate(betas):
-    axs[0].plot(-points.logX(), np.exp(beta*(points.logL - points.logL.max())), 'x', ms=1, color=colors[i], label=labels[i])
-    axs[0].plot(-points.logX(), np.exp(beta*(points.logL - points.logL.max())), lw=.5, alpha=.2, color=colors[i])
-    logw = beta * points.logL + points.logX()
-    logw -= logw.max()
-    axs[1].fill_between(-points.logX(), 0, np.exp(logw), lw=.8, color=colors[i])
-axs[0].plot(-points.logX().iloc[-1], np.exp(points.logL.iloc[-1] - points.logL.max()), 'x', ms=2, color='k')
-
-l = .2
-arrow_start = (6, l)
-arrow_end = (2, l)
-arrow = patches.FancyArrowPatch(arrow_start, arrow_end, arrowstyle='->', mutation_scale=5)
-axs[0].add_patch(arrow)
-axs[0].annotate(r'Lower $\beta$', xy=(4, l), xytext=(4, l+.1), ha='center', va='bottom', fontsize=5)
-axs[1].axvline(x=-points.logX().iloc[ndead], ls='--', color='k', lw=.5)
-
-for ax in axs:
-    ax.margins(x=0.02)
-    ax.set_ylim(-0.01, 1.1)
-    ax.set_yticks([])
-    ax.set_xticks([0, -points.logX().iloc[ndead]], [0, r'$-\log X_i$'])
-axs[0].set_title(r'$\mathcal{L}^\beta$', fontsize=8)
-axs[1].set_title('$\\mathcal{P}(\\log X) = \\mathcal{L}^\\beta X$', fontsize=8)
-
-from matplotlib.patches import Patch
[…]

### 2024-07-17 paper/figures/logL_logX_plots.py

@@ -1,0 +1,48 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+from aeons.likelihoods import full
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+chains = ['gauss_8', 'correlated_6d_close', 'rosenbrock_10', 'cauchy_82']
+titles = ['Spherical Gaussian', 'Elongated Gaussian', 'Rosenbrock', 'Cauchy']
+fig, axs = plt.subplots(1, 4, figsize=(7, 2))
+
+for chain, title, ax in zip(chains, titles, axs):
+    name, samples = get_samples(chain)
+    model = EndModel(samples)
+    endpoint_true = model.true_endpoint()
+    logXf_true = samples.logX().iloc[endpoint_true]
+    samples = samples.iloc[:endpoint_true]
+    
+    ndead = int(endpoint_true * 0.3)
+    points = points_at_iteration(samples, ndead)
+    logL, X_mean, nk, logZdead = data(points)
+    d_G, _ = get_d_G_post(points, ndead)
+    
+    for i in range(25):
+        X = generate_Xs(nk)
+        d = np.random.choice(d_G)
+        theta = params_from_d(logL[ndead:], X[ndead:], d)
+        logL_model = full.func(np.exp(samples.logX()), theta)
+        ax.plot(-samples.logX().iloc[ndead:], logL_model.iloc[ndead:], lw=.5, color='deepskyblue')
+    ax.plot(-samples.logX().iloc[ndead:], samples.logL.iloc[ndead:], lw=2, color='black')
+    ax.set_xlabel('$-\\log X$')
+    ax.set_title(f'{title}\n ({len(samples):,} total iterations)', fontsize=8)
+    ax.set_yticks([])
+axs[0].set_ylabel('$\\log \\mathcal{L}$')
+
+from matplotlib.lines import Line2D
+black_patch = Line2D([0], [0], color='black', lw=2)
+green_patch = Line2D([0], [0], color='lightseagreen', lw=.5)
+axs[0].legend([black_patch, green_patch], ['True likelihood', 'Fitted likelihoods'], loc='lower right')
+fig.tight_layout()
+
+fig.savefig(f'{current_dir}/logL_logX_plots.pdf', bbox_inches='tight')

### 2024-07-17 paper/figures/logXfs_correlated.py

@@ -1,0 +1,39 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
+chains = ['correlated_6d', 'correlated_3d', 'correlated_6d_close', 'correlated_6d_rotated']
+titles = ['$\\Sigma_\\mathrm{{separated}}$', '$\\Sigma_{{3d}}$', '$\\Sigma_\\mathrm{{close}}$', '$\\Sigma_\\mathrm{{separated, rotated}}$']
+
+
+for chain, title, ax in zip(chains, titles, axs):
+    name, samples = get_samples(chain)
+    model = EndModel(samples)
+    true_endpoint = model.true_endpoint()
+    logXf_true = samples.logX().iloc[true_endpoint]
+    
+    iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
+    logXfs = np.array(logXfs)
+    logXs = samples.logX().iloc[iterations]
+    plot_quantiles(-logXs, -logXfs, -logXf_true, (0, 1.3), ax=ax)
+    ax.set_title(f'{title}\n{true_endpoint:,} total iterations', fontsize=7)    
+    ax.set_ylim(0, 45)
+    ax.margins(x=0.01)
+
+for ax in axs.flat[1:]:
+    ax.set_yticks([])
+    
+fig.supxlabel('$-\\log X$', y=0.05)
+fig.supylabel('$-\\log \\hat{{X}}_\\mathrm{{f}}$', x=0)
+fig.tight_layout()
+
+fig.savefig(f'{current_dir}/logXfs_correlated.pdf', pad_inches=0, bbox_inches='tight')

### 2024-07-17 paper/figures/logXfs_gaussian.py

@@ -1,25 +1,33 @@
+import numpy as np
 import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
 from aeons.utils import *
-from aeons.plotting import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
 figsettings()
 
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
 d = [8, 16, '16_2000', 32]
+titles = [r'$d = 8$', r'$d = 16$', r'$d = 16, n_\mathrm{live} = 2000$', r'$d = 32$']
 fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
 axs = axs.flatten()
 for i, nDims in enumerate(d):
     name, samples = get_samples(f'gauss_{nDims}')
-    true_endpoint = endpoints[name]
+    model = EndModel(samples)
+    true_endpoint = model.true_endpoint()
     logXf_true = samples.logX().iloc[true_endpoint]
+    
     iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
     logXfs = np.array(logXfs)
     logXs = samples.logX().iloc[iterations]
     plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 1.5), color='deepskyblue', ax=axs[i])
-    axs[i].set_title(rf'$d$ = {nDims}, $\mathcal{{D}}_\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
+    axs[i].set_title(f'{titles[i]}\n{true_endpoint:,} total iterations', fontsize=7)
     axs[i].set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
     axs[i].margins(x=0)
 
-axs[2].set_title(r"$d$ = 16, $n_\mathrm{live} = 2000$", fontsize=8)
 fig.supxlabel('$-\\log X$', y=0.02, fontsize=8)
-fig.supylabel('$-\\log \hat{X}_\mathrm{f}$', fontsize=8, x=0.01)
+fig.supylabel('$-\\log \\hat{X}_\\mathrm{f}$', fontsize=8, x=0.01)
 fig.tight_layout()
-fig.savefig('gauss_predictions.pdf', pad_inches=0, bbox_inches='tight')
+fig.savefig(f'{current_dir}/logXfs_gaussian.pdf', bbox_inches='tight')

### 2024-07-17 paper/figures/logXfs_lcdm.py

@@ -1,28 +1,35 @@
 import numpy as np
 import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
 from aeons.utils import *
-from aeons.regress import *
-from aeons.endpoint import *
-from aeons.plotting import *
 from aeons.beta import *
+from aeons.plotting import plot_quantiles
+from aeons.likelihoods import full
 figsettings()
 
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
 chains = ['SH0ES', 'BAO', 'lensing', 'planck']
+titles = ['SH0ES', 'BAO', 'LENSING', 'PLANCK']
 fig, axs = plt.subplots(1, 4, figsize=(7, 1.6))
-for i, chain in enumerate(chains):
-    ax = axs[i]
+for ax, chain, title in zip(axs, chains, titles):
     name, samples = get_samples(chain)
-    true_endpoint = endpoints[name]
+    model = EndModel(samples)
+    true_endpoint = model.true_endpoint()
+    logXf_true = samples.logX().iloc[true_endpoint]
+    
     iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
     logXfs = np.array(logXfs)
     logXs = samples.logX().iloc[iterations]
     logXf_true = samples.logX().iloc[true_endpoint]
     plot_quantiles(-logXs, -logXfs, -logXf_true, ylim=(0, 1.5), color='deepskyblue', ax=ax)
-    ax.set_title(f'{name.upper()}, $\\mathcal{{D}}_\\mathrm{{KL}} = {samples.D_KL():.1f}$', fontsize=8)
+    ax.set_title(f'{title}\n{true_endpoint:,} total iterations', fontsize=7)
     ax.set_yticks([0, -logXf_true], [0, f"{-logXf_true:.1f}"])
     ax.margins(x=0)
 
 fig.supxlabel('$-\\log X$', y=0.02)
 fig.supylabel('$-\\log \\hat{X_\\mathrm{f}}$', x=0.01)
 fig.tight_layout()
-fig.savefig('lcdm_logXfs.pdf', pad_inches=0, bbox_inches='tight')
+
+fig.savefig(f'{current_dir}/logXfs_lcdm.pdf', pad_inches=0, bbox_inches='tight')

### 2024-07-17 paper/figures/logXfs_non_gaussian.py

@@ -1,0 +1,36 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from aeons.endpoint import EndModel
+from aeons.utils import *
+from aeons.beta import *
+from aeons.plotting import plot_quantiles
+from aeons.likelihoods import full
+figsettings()
+
+import os
+current_dir = os.path.dirname(os.path.abspath(__file__))
+
+
+fig, axs = plt.subplots(1, 2, figsize=(4, 1.5))
+chains = ['rosenbrock_10', 'cauchy_84']
+titles = ['Rosenbrock', 'Cauchy']
+
+for ax, chain, title in zip(axs, chains, titles):
+    name, samples = get_samples(chain)
+    model = EndModel(samples)
+    true_endpoint = model.true_endpoint()
+    logXf_true = samples.logX().iloc[true_endpoint]
+
+    iterations, *logXfs = read_from_txt(f'{data_dir}/logXfs/post/{name}.txt')
+    logXfs = np.array(logXfs)
+    logXs = samples.logX().iloc[iterations]
+    plot_quantiles(-logXs, -logXfs, -logXf_true, (0, 1.5), ax=ax)
+    ax.set_title(f'{title}\n{true_endpoint:,} total iterations', fontsize=7)
+    ax.margins(x=0.01)
+
+axs[1].set_ylim(0, 150)
+fig.supxlabel('$-\\log X$', y=0.05)
+fig.supylabel('$-\\log \\hat{{X}}_\\mathrm{{f}}$', x=0)
+fig.tight_layout()
+
+fig.savefig(f'{current_dir}/logXfs_non_gaussian.pdf', pad_inches=0, bbox_inches='tight')

### 2024-07-17 paper/figures/phase_transition.py

@@ -1,20 +1,0 @@
-import matplotlib.pyplot as plt
-from aeons.utils import *
-from aeons.plotting import *
-figsettings()
-
-samples = pickle_in(f'{aeons_dir}/data/predictions/gauss/two.pkl')
-iterations, logXfs, logXfs_std, true_endpoint = read_from_txt(f'{aeons_dir}/data/predictions/gauss/two_logXfs.txt')
-iterations, d_Gs, d_Gs_std = read_from_txt(f'{aeons_dir}/data/predictions/gauss/two_d_Gs.txt')
-
-logXf_true = samples.logX().iloc[int(true_endpoint[0])]
-logXs = samples.logX().iloc[iterations]
-endpoints, endpoints_std = calc_endpoints(iterations, logXs, logXfs, logXfs_std, nlive=705)
-
-fig, axs = plt.subplots(1, 2, figsize=(4, 1.2))
-plot_std(-logXs, -logXfs, logXfs_std, -logXf_true, ylim=(0.6, 1.1), color='deepskyblue', ax=axs[0])
-plot_std(-logXs, d_Gs, d_Gs_std, d_Gs[-1], (0, 1.3), color='orange', ax=axs[1])
-axs[0].set_title('$-\\log \hat{X_\mathrm{f}}$')
-axs[1].set_title(f'$\\tilde{{d}}$')
-fig.supxlabel('$-\\log X$', y=-0.2)
-fig.savefig('phase_transition.pdf', pad_inches=0, bbox_inches='tight')

### 2024-07-17 aeons/beta.py
cleanup and package updates

@@ -46,6 +46,36 @@ def get_betas_logL(samples):
     f = interp1d(logL, beta)
     return f(samples.logL)
 
+def get_beta_end(points, ndead, epsilon=1e-3):
+    logX = points.logX()
+    logL = points.logL
+    # Check if already terminated
+    logw = logL + logX
+    logw = logw - logw.max()
+    def func(beta):
+        logw = beta * logL + logX
+        w = np.exp(logw - logw.max())
+        Zdead = np.sum(w[:ndead])
+        Zlive = np.sum(w[ndead:])
+        if (Zdead == 0):
+            return np.inf
+        return Zlive/Zdead - epsilon
+    
+    from scipy import optimize
+    from scipy.special import logsumexp
+    logZdead = logsumexp(logw[:ndead])
+    logZlive = logsumexp(logw[ndead:])
+    try:
+        if (logZlive - logZdead < np.log(epsilon)):
+            res = optimize.root_scalar(func, bracket=[1, 1e3])
+            return res.root
+        else:
+            res = optimize.root_scalar(func, bracket=[0, 1])
+            return res.root
+    except:
+        print("No beta found")
+        return 0
+
 ### d_G
 def get_d_G_grad(points, ndead, Nset=25):
     logbeta = get_logbeta_grad(points, ndead)
@@ -63,6 +93,11 @@ def get_d_G_post(points, ndead, Nset=25):
     betas_post = np.exp(np.random.normal(logbeta_mean, logbeta_std, Nset))
     d_G_post = points.d_G(beta=betas_post)
     return d_G_post.values, f"{len(betas_post)} samples"
+
+def get_d_G_epsilon(points, ndead, Nset=25):
+    beta = get_beta_end(points, ndead)
+    d_G = points.d_G(Nset, beta=beta)
+    return d_G.values, ""
 
 def get_d_G_prop(points, ndead):
     betas = np.logspace(-10, 0, 1000)

### 2024-07-17 aeons/endpoint.py

@@ -80,11 +80,14 @@ class EndModel:
             points = self.points(ndead)
             d_G = d_G_method(points, ndead, Nset)[0]
             from scipy.stats import gaussian_kde
-            kde = gaussian_kde(d_G)
+            if d_G.std() < 1e-16:
+                ds = [d_G[0]]
+                print('Warning: d_G is constant')
+            else:
+                ds = gaussian_kde(d_G).resample()
+                ds = ds[ds > 0]
             logL, X_mean, nk, logZdead = data(points)
             logXf_set = np.zeros(Nset)
-            ds = kde.resample()
-            ds = ds[ds > 0]
             for j in range(Nset):
                 X = generate_Xs(nk)
                 d = np.random.choice(ds)

### 2024-07-17 aeons/utils.py

@@ -263,15 +263,11 @@ def get_beta(points, ndead):
 def get_dGs(get_beta, samples, iterations, Nset=10, **kwargs):
     """Get the dG of the samples at each iteration for a given function get_beta."""
     iterations = iterations.astype(int)
-    d_Gs = np.zeros(len(iterations))
-    d_Gs_std = np.zeros(len(iterations))
+    d_Gs = np.zeros((len(iterations), Nset))
     for i, iteration in enumerate(iterations):
         points = points_at_iteration(samples, iteration)
         beta = get_beta(points, iteration, **kwargs)
         points = points.set_beta(beta)
-        d_Gs_i = points.d_G(Nset)
-        d_Gs[i] = d_Gs_i.mean()
-        d_Gs_std[i] = d_Gs_i.std()
-        # Print progress bar
-        print('\r', f'Iteration {iteration} of {iterations[-1]}, {d_Gs[i]}', end='')
-    return d_Gs, d_Gs_std
+        d_Gs[i] = points.d_G(Nset)
+        print('\r', f'Iteration {iteration} of {iterations[-1]}', end='')
+    return d_Gs

### 2024-07-17 review_response/paper.tex
second minor revision for journal submission

--- 2024-07-17 paper/paper.tex
@@
 We now consider the distribution of the samples in log-likelihood.
-To get an insight into the analytics, we will examine the case of the $d$-dimensional multivariate Gaussian likelihood with maximum point $\logLmax$ and length scale $\sigma$, situated in a uniform unit prior:
+To get an insight into the analytics, we will examine the case of the isotropic, $d$-dimensional multivariate Gaussian likelihood with maximum point $\logLmax$ and length scale $\sigma$, situated in a uniform unit prior:
 \begin{equation}\label{eq:logL_x}
@@
 \par
-We can also put compare this endpoint with the termination condition based on $\epsilon$. From \cref{eq:Xf_approx}, for any $d$, $n$ we have
+We can also compare this endpoint with the termination condition based on $\epsilon$. From \cref{eq:Xf_approx}, for any $d$, $n$ we have
 \begin{equation}
@@
 \subsubsection*{Comparisons}
-\cref{fig:beta_posteriors} shows the posterior distribution as a function of $\log X$ for a 32-d spherical Gaussian when set to the above four temperatures, as known at a given iteration roughly 30\% through a run.
+\cref{fig:beta_posteriors} shows the posterior distribution as a function of $\log X$ for a 32-d isotropic Gaussian when set to the above four temperatures, as known at a given iteration roughly 30\% through a run.
 Lower prior volumes are to the right, and 25 samples are taken from the prior volume distribution to capture the nested sampling uncertainties.
@@
 \end{center}
-\caption{The posterior distribution at an intermediate point 30\% of the way through a nested sampling run for a 32$d$ spherical Gaussian, with the likelihood transformed from  $\Like \to \Like^{\beta}$ for several different choices of $\beta$.
+\caption{The posterior distribution at an intermediate point 30\% of the way through a nested sampling run for a 32$d$ isotropic Gaussian, with the likelihood transformed from  $\Like \to \Like^{\beta}$ for several different choices of $\beta$.
 The untransformed posterior has all its weight at a single point.
@@
 $\beta_\epsilon$ is offset from the others, but can be shifted up and down depending on the chosen $\epsilon$.
-The other $\beta$ are all are consistent in a single phase, but differ during a phase transition.}
+The other $\beta$ are all consistent in a single phase, but differ during a phase transition.}
 \label{fig:beta_plots}
@@
 We find that the $\dG$ calculated from these transformed posteriors are exactly the Bayesian model dimensionalities of the \textit{true} posterior.
-\cref{fig:d_G_gaussians} demonstrates this using the microcanonical and Bayesian temperatures, on several spherical Gaussians for which the true value of $\dG$ is known; except for the earliest iterations, the right value is found to within standard error for all the likelhoods.
+\cref{fig:d_G_gaussians} demonstrates this using the microcanonical and Bayesian temperatures, on several isotropic Gaussians for which the true value of $\dG$ is known; except for the earliest iterations, the right value is found to within standard error for all the likelihoods.
 The Bayesian temperature has slightly larger error bounds, as a more cautious estimate, but otherwise the estimates are consistent.
@@
 This provides a crucial piece of information, allowing us to access the Gaussian dimensionality of the true posterior before the run is complete.
-Before we discuss how this is can be used to aid endpoint estimation, we consider how the prior can affect the value of $\dG$ mid-run.
+Before we discuss how this can be used to aid endpoint estimation, we consider how the prior can affect the value of $\dG$ mid-run.
 \begin{figure*}
@@
 \end{center}
-\caption{Bayesian model dimensionalities estimated throughout a run for spherical Gaussians of varying dimensions, which were not previously accessible before the end of a run because the posterior samples have vanishing variance.
+\caption{Bayesian model dimensionalities estimated throughout a run for isotropic Gaussians of varying dimensions, which were not previously accessible before the end of a run because the posterior samples have vanishing variance.
 Calculating the value from the transformed likelihood $\Like \to \Like^{\beta}$, however, recovers the dimensionality of the final posterior before it is reached, which is shown in the above plots.
@@
 \par
-The key point is that in panel B, the live point cloud is actually one-dimensional.
-When comapred to a truly one-dimensional posterior (\cref{fig:d_G_elongated_1D}), at point B the samples look the same in all aspects.
+The key point is that in panel B, the live point cloud really only varies in one dimension, despite the true posterior having two constrained parameters.
+When compared to another $2d$ posterior which truly only has one constrained parameter, with the other being flat (\cref{fig:d_G_elongated_1D}), we find that at point B the samples look the same in all aspects.
 The consequence is that at point B, we fundamentally cannot know if there are more parameters `waiting' to be constrained, or if we have already found all the relevant directions.
@@
 \par
-Our aim is therefore first to quantify the amount of compression required to reach the endpoint at approximately $\log X = -\DKL$, which provides an estimate of the how far the sampler still needs to penetrate into the posterior.
+Our aim is therefore first to quantify the amount of compression required to reach the endpoint at approximately $\log X = -\DKL$, which provides an estimate of how far the sampler still needs to penetrate into the posterior.
 Once the required compression $\log X_\mathrm{f}$ is known, multiplying by the harmonic mean of the number of live points through the run gives the iteration count.
@@
 The critical ingredient here is the function $\phi$ used to fit the likelihood profile $\Like(X)$, which should be simple enough to make fitting possible mid-run, but ideally also flexible enough to avoid strong biases.
-Previous work, such as the fitting procedure in the original Nelder-Mead simplex algorithm to estimate the Hessian \citep{nelder_mead}, and the BAMBI nested sampling paper \citep{Graff_2012}, suggest fast Gaussian approximations can be a good candidate for (online) modelling of likelihoods.
-The following method uses a similar approach, but unlike the previous schemes do not fit directly in parameter space.
+Previous work, such as the fitting procedure in the original Nelder-Mead simplex algorithm to estimate the Hessian \citep{nelder_mead}, and the BAMBI nested sampling paper \citep{Graff_2012}, suggest fast Gaussian approximations can be a good candidate for (online) modeling of likelihoods.
+The following method uses a similar approach, but unlike the previous schemes does not fit directly in parameter space.
 \par
@@
 \begin{enumerate}[leftmargin=*]
-    \item Spherical Gaussian; $\sigma = 10^{-3}$
+    \item Isotropic Gaussian; $\sigma = 10^{-3}$
         \begin{equation}
@@
 \end{equation}
-This has a several shortcomings.
+This has several shortcomings.
 First, run-time is proportional to compression rather than accumulation of the integral, since it takes just as long to traverse the width of the bulk as it does any other width.
@@
 Seasoned watchers of nested sampling runs might be curious how the method compares to simply extrapolating the increments of evidence to roughly estimate when the evidence converges.
-At an intermediate stage of the run, the most recent outputs look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate} (which in this case was a real run for a spherical Gaussian with $d = 16$, $\sigma = 0.01$).
+At an intermediate stage of the run, the most recent outputs look something like that shown in the first two columns of the table in \cref{fig:inc_extrapolate} (which in this case was a real run for a isotropic Gaussian with $d = 16$, $\sigma = 0.01$).
 The right hand side of \cref{fig:inc_extrapolate} plots the gain in log-evidence at every 500 iterations, then extrapolates them with a linear and exponential profile.
@@
 \includegraphics{figures/inc_predictions.pdf}
-\caption{Endpoint predictions for a spherical Gaussian, with $1-2\sigma$ uncertainties shaded.
+\caption{Endpoint predictions for a isotropic Gaussian, with $1-2\sigma$ uncertainties shaded.
 Extrapolating the evidence increments in a linear/exponential manner under/over-predicts the endpoint, both of which perform considerably worse than the method of extrapolating the likelihood.}
@@
 \subsection{Toy examples}\label{sec:logXfs_toy}
-Throughout the toy examples we make use the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\citep{anesthetic}.
-\subsubsection{Spherical Gaussians}
-Predictions for spherical Gaussians of various dimensions are shown in \cref{fig:logXfs_gaussian} as a benchmark for when fitting a Gaussian distribution is exact.
+Throughout the toy examples we make use of the perfect nested sampling~\citep{Keeton_2011,2018BayAn..13..873H} framework implemented in \textsc{anesthetic}~\citep{anesthetic}.
+\subsubsection{Isotropic Gaussians}
+Predictions for isotropic Gaussians of various dimensions are shown in \cref{fig:logXfs_gaussian} as a benchmark for when fitting a Gaussian distribution is exact.
 In all endpoint prediction plots, the shading indicates the $1-2\sigma$ uncertainties.
-All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$.
-The correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained.
+All were run with $\nlive = 500$ except for one $\nlive=2000$ for comparison, with each Gaussian having a width of $\sigma = 0.01$..
 \par
-One can see from the $\log\Like-\log X$ plots in \cref{fig:logL_logX_plots} that the extrapolated likelihood is a good fit to the true likelihood, as expected from a spherical Gaussian.
+This case serves as a useful benchmark for the case where there are no correlations or length scale differences between the parameters, and it is reassuring that the correct endpoint is recovered to within standard error at all points except the very beginning, when the parameters have hardly been constrained.
+One can see from the $\log\Like-\log X$ plots in \cref{fig:logL_logX_plots} that the extrapolated likelihood is a good fit to the true likelihood, as expected from a isotropic Gaussian.
 We also note that as with nested sampling in general, increasing $\nlive$ improves the resolution and reliability of the inferences, as indicated by the middle two plots of \cref{fig:logXfs_gaussian}.
@@
 \end{center}
-\caption{Endpoint predictions for a spherical Gaussian run with $\nlive=500$ (except for the third plot from left).
+\caption{Endpoint predictions for a isotropic Gaussian run with $\nlive=500$ (except for the third plot from left).
 The true endpoints are shown in dotted red, while the estimates with their $1-2\sigma$ uncertainties are shaded in blue.
@@
 The Cauchy also under-predicts at first, obtaining the correct endpoint within standard error by about halfway through its 39,236 iterations.
-The posterior here is spherically symmetric, so compression is isotropic and the prior affects all parameters equally.
+The posterior here is spherical, so compression is isotropic and the prior affects all parameters equally.
 Because there are no associated under-estimates of the posterior dimensionality, the inaccuracies are solely due to the poor fit of a Gaussian to a Cauchy, which can be seen in the rightmost plot of \cref{fig:logL_logX_plots} as much more drastic than the other likelihoods; the Bayesian model dimensionality is itself less stable for Cauchy likelihoods.
@@
 \caption{Fitted extrapolations compared to the true likelihood for four different test likelihoods.
-The live points at 30\% through the run for each example are fitted a to Gaussian, extrapolated to the higher likelihoods that have not yet explored, and compared to the ground truth.
-It is clear that truly Gaussian likelihoods are modelled well, while the model deteriorates with increasing non-Gaussianity (seen in the Rosenbrock and Cauchy likelihoods).}
+The live points at 30\% through the run for each example are fitted to a Gaussian, extrapolated to the higher likelihoods that have not yet been explored, and compared to the ground truth.
+It is clear that truly Gaussian likelihoods are modeled well, while the model deteriorates with increasing non-Gaussianity (seen in the Rosenbrock and Cauchy likelihoods).}
 \label{fig:logL_logX_plots}

### 2024-07-17 review_response/previous_submission.tex
second minor revision for journal submission

--- 2024-07-17 paper/paper.tex
@@
 nlike      =  4159049
-<nlike>    =  491.04 (9.82 per slice)
+<nlike>    =  41.04 (9.82 per slice)
 log(Z)     =  -12.55 +/- 0.27
